```{r sn1_ancombc, message=FALSE, eval=FALSE}
set.seed(1234) # set seed for reproducibility
sn1_ancombc_G121e <- ancombc2(
  data = sn1_physeq_G121e,
  assay_name = "counts",
  tax_level = "genus", # change to agglomerate analysis to a higher taxonomic range
  fix_formula = "dataset",
  p_adj_method = "holm",
  pseudo_sens = TRUE,
  prv_cut = 0,
  lib_cut = 0,
  s0_perc = 0.05,
  group = "dataset",
  struc_zero = TRUE,
  neg_lb = FALSE,
  alpha = 0.05,
  n_cl = 2,
  verbose = TRUE,
  global = FALSE,
  pairwise = FALSE,
  dunnet = FALSE,
  trend = FALSE,
  iter_control = list(tol = 1e-5, max_iter = 20, verbose = FALSE),
  em_control = list(tol = 1e-5, max_iter = 100),
  mdfdr_control = list(fwer_ctrl_method = "holm", B = 100),
  trend_control = NULL
)

any(sn1_ancombc_G121e$res$`passed_ss_datasetMicro-scale` == TRUE &
  sn1_ancombc_G121e$res$`q_datasetMicro-scale` < 0.05)

sn1_ancombc <- sn1_ancombc_G121e$res %>%
  select(
    genus = taxon,
    lfc = `lfc_datasetMicro-scale`,
    qval = `q_datasetMicro-scale`,
    passed_ss = `passed_ss_datasetMicro-scale`,
    diff = `diff_datasetMicro-scale`
  ) %>%
  filter(qval < 0.05) %>%
  arrange(qval, lfc) %>%
  mutate(
    hjust_label = ifelse(lfc > 0, -0.05, 1.05),
    q_label = paste0("q=", signif(round(qval, 2), 2)),
    q_label = ifelse(q_label == "q=0", "q<0.01", q_label)
  ) %>%
  ggplot(aes(
    x = lfc,
    y = fct_reorder(genus, lfc),
    fill = passed_ss
  )) +
  geom_col() +
  geom_text(
    aes(label = q_label, hjust = hjust_label),
    size = 3.2,
    color = "black"
  ) +
  scale_fill_manual(values = c("darkgrey", "#e35d51")) +
  geom_hline(yintercept = 0) +
  custom_ggplot_theme +
  labs(
    # title = "c",
    x = "log2FoldChange",
    y = "Genus"
  ) +
  guides(fill = guide_legend(title = "Eligible for Differential \nAbundance Testing")) +
  theme(legend.position = "bottom") +
  coord_cartesian(xlim = c(-1.2, 1.2))
sn1_ancombc
```


##### Mapping Pangenome

###### Select Unique Genes
```{r mmseq_unique_genes, eval=FALSE}
gene_pa <- read_tsv("../data/data_preparation/gene_presence_absence.Rtab", show_col_types = FALSE) %>%
  rename(gene = Gene)

gene_persistent <- gene_pa %>%
  mutate(n_genome = rowSums(across(where(is.numeric)))) %>%
  arrange(n_genome) %>%
  filter(n_genome == 8) %>%
  pull(gene)
gene_unique <- gene_pa %>%
  mutate(n_genome = rowSums(across(where(is.numeric)))) %>%
  arrange(n_genome) %>%
  filter(n_genome == 1) %>%
  pull(gene)

gene_ani <- read_csv("ppang/MY_GENES_FAMILIES_CP/all_alignments_with_identity.csv", show_col_types = FALSE)
# gene_matrix <- read_csv("ppang/matrix.csv", show_col_types = FALSE)

# Filter duplicated (query, target) pairs
diff_identities <- gene_ani %>%
  filter(
    query != target,
    query %in% gene_unique,
    target %in% gene_unique
  ) %>%
  group_by(query, target) %>%
  filter(n() > 1) %>%
  summarise(
    values = list(identity),
    min_val = min(identity),
    max_val = max(identity),
    diff = max(identity) - min(identity),
    n_unique = length(unique(identity)),
    .groups = "drop"
  ) %>%
  filter(diff > 0) # only keep actual differences

gene_ani_filtered <- gene_ani %>%
  filter(
    query != target,
    query %in% gene_unique,
    target %in% gene_unique
  ) %>%
  select(query, target, identity) %>%
  mutate(identity = unlist(identity)) %>%
  pivot_wider(
    names_from = target,
    values_from = identity,
    values_fill = NA,
    values_fn = max # use max to handle duplicates
  )

selected_genes <- gene_ani %>%
  filter(
    query != target,
    query %in% gene_unique,
    target %in% gene_unique
  ) %>%
  left_join(gene_pa %>%
    mutate(n_genome = rowSums(across(where(is.numeric)))) %>%
    arrange(n_genome) %>%
    filter(n_genome == 1) %>%
    select(-n_genome) %>%
    pivot_longer(!gene, names_to = "genome", values_to = "bin_count"), by = join_by(query == gene)) %>%
  left_join(gene_pa %>%
    mutate(n_genome = rowSums(across(where(is.numeric)))) %>%
    arrange(n_genome) %>%
    filter(n_genome == 1) %>%
    select(-n_genome) %>%
    pivot_longer(!gene, names_to = "genome", values_to = "bin_count"), by = join_by(target == gene)) %>%
  mutate(test = ifelse(genome.x == genome.y, "keep", "exclude")) %>%
  group_by(genome.x) %>%
  mutate(test1 = sum(test))

filter(test %in% c("keep")) %>%
  distinct(target, .keep_all = TRUE) %>%
  count(genome.x)

group_by(query) %>%
  summarise(
    n = sum(!is.na(identity)),
    mean = mean(identity)
  ) %>%
  filter(
    n < 4,
    # mean < 1
  )
```

###### Load Mapping Data

```{r load_pg, message=FALSE, eval=FALSE, warning=FALSE, comments=""}
# Read count
read_counts_pg <- read_tsv("../data/data_preparation/coverm_pangenome.tsv", show_col_types = FALSE) %>%
  rename(gene = 1) %>%
  pivot_longer(!gene, names_to = "data", values_to = "counts") %>%
  mutate(sample = str_extract(data, "M\\d+")) %>%
  group_by(gene, sample) %>%
  summarise(counts = sum(counts), .groups = "drop") %>%
  pivot_wider(names_from = "sample", values_from = "counts")
# Covered Bases
basehits_pg <- read_tsv("../data/data_preparation/contigs.covered_bases.tsv", show_col_types = FALSE) %>%
  rename(gene = 1) %>%
  pivot_longer(!gene, names_to = "data", values_to = "counts") %>%
  mutate(sample = str_extract(data, "M\\d+")) %>%
  group_by(gene, sample) %>%
  summarise(counts = sum(counts), .groups = "drop") %>%
  pivot_wider(names_from = "sample", values_from = "counts")
# Gene Length
length_pg <- read_tsv("../data/data_preparation/GPB.tsv", show_col_types = FALSE) %>%
  mutate(length = stop - start) %>%
  select(gene, length, nb_copy_in_genome, partition) %>%
  arrange(match(gene, read_counts_pg$gene)) %>%
  filter(as.character(gene) %in% as.character(basehits_pg$gene))
```


###### Filter and normalise data
```{r filter_pg, eval=FALSE}
gene_coverage_pg <- basehits_pg %>%
  mutate(across(where(is.numeric), ~ . / length_pg$length))

min_coverage_pg <- 0.3
read_counts_filt_pg <- gene_coverage_pg %>%
  mutate(across(where(is.numeric), ~ ifelse(. > min_coverage_pg, 1, 0))) %>%
  mutate(across(-1, ~ . * read_counts_pg[[cur_column()]]))
readlength_pg <- 150 # change if sequencing read length is different
gene_counts_filt_pg <- read_counts_filt_pg %>%
  mutate(across(where(is.numeric), ~ . / (length_pg$length / readlength_pg)))
gene_counts_pg <- read_counts_pg %>%
  mutate(across(where(is.numeric), ~ . / (length_pg$length / readlength_pg)))
```


###### Retain gene of Interest
```{r filter_pg, eval=FALSE}
genome_info <- gene_pa %>%
  mutate(n_genome = rowSums(across(where(is.numeric)))) %>%
  arrange(n_genome) %>%
  filter(n_genome == 1) %>%
  select(-n_genome) %>%
  #filter(gene %in% c(selected_genes$query)) %>%
  pivot_longer(!gene, names_to = "genome", values_to = "bin_count") %>%
  filter(bin_count > 0)

bin_count <- genome_info %>%
  group_by(genome) %>%
  summarise(sum_bin = sum(bin_count))



count_pg <- left_join(gene_counts_filt_pg %>%
  filter(gene %in% c(genome_info$gene)) %>%
  select(gene, where(~ !all(. == 0))) %>%
  pivot_longer(!gene, names_to = "microsample", values_to = "count") %>%
  filter(count > 0), genome_info, by = "gene") %>%
  left_join(gene_coverage_pg %>%
    pivot_longer(!gene, names_to = "microsample", values_to = "gene_coverage"), by = c("microsample", "gene")) %>%
  group_by(microsample, genome) %>%
  summarise(
    genome_count = sum(count),
    genome_coverage = mean(gene_coverage),
    n_gene = sum(bin_count)
  ) %>%
  left_join(bin_count, by = "genome") %>%
  left_join(macroquant_metrics_filt_30 %>% select(microsample, animal, cryosection, batch), by = "microsample") %>%
  mutate(gene_perc = n_gene / sum_bin) %>% filter(gene_perc > 0.2)

ggplot(
  count_pg %>% arrange(microsample) %>%
  filter(
    cryosection %in% c("G121eI103A", "G103bI301A"),
    !microsample %in% c("M300840", "M301068", "M301085", "M301084") # , # wrong coordinates
    # circul %in% c("Y")
  ) ,
  aes(x = genome_count, y = microsample, fill = genome)
) + # Group by multiple taxonomic levels
  geom_bar(position = "fill", stat = "identity") +
  facet_wrap(~ interaction(animal, cryosection, batch), scales = "free_y") +
  labs(title = "Pangenomic mapping")

pa_bin_000118 <- count_pg %>%
  filter(genome %in% c("GPB:bin_000118")) %>%
  mutate(pa_bin_000118 = ifelse(genome_count > 0, "P", "A")) %>%
  select(microsample, pa_bin_000118)
```


#### Spatial Pattern
##### Data selection
```{r data_spatial, message=FALSE, eval=FALSE}
metadata <- macroquant_alpha_div_filt_30 %>%
  filter(
    type_binomial %in% c("positive"),
    cov_filtering %in% c("Retained By Filtering"),
    !section %in% "Ileum",
    !is.na(Xcoord),
    !animal %in% "G103b"
  )

comm_data <- macroquant_genome_counts_filt_30 %>%
  select(genome, metadata$microsample) %>%
  t() %>%
  data.frame() %>%
  row_to_names(row_number = 1) %>%
  mutate(across(everything(), as.numeric))

table(rowSums(comm_data) > 0)
all(metadata$microsample == rownames(comm_data))
table(is.na(comm_data))
table(metadata$section, metadata$cryosection)

comm_ca <- comm_data[metadata$section == "Caecum right", ]
comm_ca <- comm_ca[, colSums(comm_ca) > 0]
comm_co <- comm_data[metadata$section == "Colon", ]
comm_co <- comm_co[, colSums(comm_co) > 0]
metadata_ca <- metadata[metadata$section == "Caecum right", ]
metadata_co <- metadata[metadata$section == "Colon", ]

table(metadata$section, metadata$cryosection)
table(metadata$section, metadata$animal)
```

##### Transform the data for posterior multivariate analyses

```{r data_spatial_clr, comment="", message=FALSE, warning=FALSE}
comm_ca_zeroRepl <- cmultRepl(comm_ca, method = "GBM", output = "prop", z.warning = 0.95)
comm_co_zeroRepl <- cmultRepl(comm_co, method = "GBM", output = "prop", z.warning = 0.95)

metadata_ca_clr <- metadata_ca[metadata_ca$microsample %in% rownames(comm_ca_zeroRepl), ]
metadata_co_clr <- metadata_co[metadata_co$microsample %in% rownames(comm_co_zeroRepl), ]

clr_transform <- function(x) {
  log(x) - mean(log(x), na.rm = TRUE)
}
comm_ca_clr <- data.frame(t(apply(comm_ca_zeroRepl, 1, clr_transform)))
comm_co_clr <- data.frame(t(apply(comm_co_zeroRepl, 1, clr_transform)))
```

##### Diversity partitioning

```{r diversity_partitioning, comment="", message=FALSE, warning=FALSE}
## Diversity partitioning in caecum and colon

ca_adipart <- adipart(as.matrix(comm_ca > 0) * 1, as.matrix(metadata_ca[, c("microsample", "cryosection")]),
  index = "richness", weights = "unif", relative = FALSE, nsimul = 999, method = "r2dtable"
)

co_adipart <- adipart(as.matrix(comm_co > 0) * 1, as.matrix(metadata_co[, c("microsample", "cryosection")]),
  index = "richness", weights = "unif", relative = FALSE, nsimul = 999, method = "r2dtable"
)

ca <- cbind(
  c(ca_adipart$oecosimu$mean[1], ca_adipart$statistic[1]),
  c(ca_adipart$oecosimu$mean[3], ca_adipart$statistic[3]),
  c(ca_adipart$oecosimu$mean[2], ca_adipart$statistic[2])
)
rownames(ca) <- c("observed", "expected")
colnames(ca) <- c("alpha", "beta", "gamma")

co <- cbind(
  c(co_adipart$oecosimu$mean[1], co_adipart$statistic[1]),
  c(co_adipart$oecosimu$mean[3], co_adipart$statistic[3]),
  c(co_adipart$oecosimu$mean[2], co_adipart$statistic[2])
)
rownames(co) <- c("observed", "expected")
colnames(co) <- c("alpha", "beta", "gamma")


par(mfrow = c(1, 2))
bp <- barplot(ca,
  beside = TRUE,
  col = c("black", "white"),
  ylim = c(0, 110)
)
text(x = apply(bp, 2, mean), y = apply(ca, 2, max) + 7, labels = "***")
title(main = "Caecum", font.main = 4)
bp <- barplot(co,
  beside = TRUE,
  col = c("black", "white"),
  ylim = c(0, 110)
)
text(x = apply(bp, 2, mean), y = apply(co, 2, max) + 7, labels = "***")
title(main = "Colon", font.main = 4)
par(mfrow = c(1, 1))
```

##### Explore Spatial Pattern
**Note:** The spatial_cryosections() function needs to be loaded.
```{r spatial_cryosections_fun, message=FALSE, eval=FALSE}
spatial_cryosections <- function(cryosection_list, metadata_df, comm_clr) {
  cryosection_dfs <- list()
  mantel_results <- list()
  mantelcor_results <- list()
  decay_dfs <- list()
  distance_decay_plots <- list()
  structure_results <- list()

  for (cryosection in cryosection_list) {
    # Filter metadata for this section
    metadata_data <- metadata_df %>%
      filter(cryosection == !!cryosection, !is.na(.data$Xcoord), !is.na(.data$Ycoord))

    # Filter community data
    comm_data <- comm_clr %>%
      data.frame() %>%
      rownames_to_column(var = "microsample") %>%
      filter(microsample %in% metadata_data$microsample) %>%
      column_to_rownames(var = "microsample")

    cryosection_dfs[[cryosection]] <- list(
      comm_clr = comm_data,
      metadata = metadata_data
    )

    # Mantel correlogram
    mantel <- vegan::mantel(
      dist(comm_data),
      dist(metadata_data[, c("Xcoord", "Ycoord")]),
      permutations = 999
    )
    mantel_results[[cryosection]] <- mantel

    # Mantel correlogram
    correlog <- vegan::mantel.correlog(
      D.eco = dist(comm_data),
      D.geo = dist(metadata_data[, c("Xcoord", "Ycoord")]),
      nperm = 999
    )
    mantelcor_results[[cryosection]] <- correlog

    # Distance decay
    toplot <- data.frame(
      spat_dist = as.numeric(dist(metadata_data[, c("Xcoord", "Ycoord")])),
      comm_dist = as.numeric(dist(comm_data))
    )
    decay_dfs[[cryosection]] <- toplot

    # Plot
    p <- ggplot(toplot, aes(x = spat_dist, y = comm_dist)) +
      # geom_point() +
      geom_smooth() +
      xlab("Spatial distance (μm)") +
      ylab("Aitchison \ndistance") +
      custom_ggplot_theme +
      # ggtitle(paste("Distance Decay -", cryosection))
      ggtitle(paste(cryosection))
    distance_decay_plots[[cryosection]] <- p

    # Print summary
    # print(paste("Summary for", cryosection))
    # print(summary(lm(comm_dist ~ spat_dist, data = toplot)))

    #### Complex spatial structures
    # set.seed(111)
    # spat_SWNs<-listw.candidates(data.frame(metadata_data[,c("Xcoord","Ycoord")]),
    # nb = c("gab"),
    # weights = c("fdown", "fup"),
    # y_fdown = c(2,5), y_fup = c(0.1,0.5))
    # W_sel <- listw.select(comm_data, spat_SWNs, MEM.autocor = "positive",
    # p.adjust = T, method="FWD")
    # structure_results[[cryosection]] <- W_sel
  }
  return(list(
    cryosection_dfs = cryosection_dfs,
    mantel_results = mantel_results,
    mantelcor_results = mantelcor_results,
    decay_dfs = decay_dfs,
    distance_decay_plots = distance_decay_plots,
    structure_results = structure_results
  ))
}
```

###### Ceacum Cryosection
```{r cryosection_ca, message=FALSE, eval=FALSE}
# Define your cryosection list
cryosection_ca <- metadata %>%
  filter(
    section == "Caecum right",
    !is.na(Xcoord)
  ) %>%
  count(cryosection) %>%
  filter(n > 20) %>%
  pull(cryosection)

# Run analysis
results_ca <- spatial_cryosections(
  cryosection_list = cryosection_ca,
  metadata_df = metadata_ca_clr,
  comm_clr = comm_ca_clr
)
```

####### Mantel test
```{r caecum_spatial_mantel, comment="", message=FALSE, warning=FALSE}
mantel_pvals_df <- do.call(rbind, lapply(names(results_ca$mantel_results), function(cryosection) {
  data.frame(
    cryosection = cryosection,
    r2 = results_ca$mantel_results[[cryosection]]$statistic,
    p_value = results_ca$mantel_results[[cryosection]]$signif
  )
}))
mantel_pvals_df
# p < 0.05 → Statistically significant correlation between two distance matrices
# p > 0.05 → No significant correlation between two distance matrices
```


####### Mantel Correlogram Stat
```{r caecum_spatial_mantelcor, comment="", message=FALSE, warning=FALSE}
for (cryosection in names(results_ca$mantelcor_results)) {
  print(cryosection)
  correlog_df <- data.frame(results_ca$mantelcor_results[[cryosection]]$mantel.res)
  # sig_classes <- correlog_df[correlog_df$`Pr.Mantel.`< 0.05, ]
  print(correlog_df)
}
```

####### Mantel Correlogram Plots
```{r caecum_spatial_mantelcor_plot, comment="", message=FALSE, warning=FALSE}
# Set up a multi-panel layout (adjust cols/rows to fit your data)
n <- length(cryosection_ca)
cols <- ceiling(sqrt(n))
rows <- ceiling(n / cols)

par(mfrow = c(rows, cols), mar = c(4, 4, 2, 1)) # Set margins and layout
# Loop over and plot each Mantel correlogram
for (cryosection in names(results_ca$mantelcor_results)) {
  plot(results_ca$mantelcor_results[[cryosection]])
  title(main = cryosection, line = 1)
}
```

#######  Distance Decay Stat
```{r caecum_spatial_decay, comment="", message=FALSE, warning=FALSE}
decay_pvals_df <- do.call(rbind, lapply(names(results_ca$decay_dfs), function(cryosection) {
  anova_result <- summary(lmp(comm_dist ~ spat_dist, data = results_ca$decay_dfs[[cryosection]]))
  data.frame(
    cryosection = cryosection,
    `Pr(Prob)` = anova_result$coefficients["spat_dist", "Pr(Prob)"]
  )
}))
decay_pvals_df
# Performs an ANOVA to check if spat_dist significantly explains variation in comm_dist.
# p < 0.05 → Significant correlation: Spatial distance influences microbial composition.
# p > 0.05 → No significant relationship: Microbial communities do not change with distance.
```

#######  Distance Decay plots

```{r caecum_spatial_decay, comment="", message=FALSE, warning=FALSE}
combined_plot <- cowplot::plot_grid(plotlist = results_ca$distance_decay_plots, ncol = 2)
print(combined_plot)
```

####### Complex spatial structures
```{r caecum_spatial_complex, comment="", message=FALSE, warning=FALSE}
names(results_ca$cryosection_dfs)
# problem with G121eI103A
cryosection_list <- c("G121eI101B", "G121eI102A", "G121eI103B", "G121eI104A", "G121eI104B", "G121eI104C")

# Initialize the list once, outside the loop
candidat_list <- list()
# for (cryosection in names(results$cryosection_dfs)) {
for (cryosection in cryosection_list) {
  set.seed(111)

  metadata_data <- results_ca$cryosection_dfs[[cryosection]]$metadata
  comm_data <- results_ca$cryosection_dfs[[cryosection]]$comm_clr

  spat_SWNs <- listw.candidates(
    data.frame(metadata_data[, c("Xcoord", "Ycoord")]),
    nb = c("gab"),
    weights = c("fdown", "fup"),
    y_fdown = c(2, 5),
    y_fup = c(0.1, 0.5)
  )

  W_sel <- listw.select(
    comm_data,
    spat_SWNs,
    MEM.autocor = "positive",
    p.adjust = TRUE,
    method = "FWD"
  )

  candidat_list[[cryosection]] <- W_sel
}

# Combine all candidate data frames into one with cryosection column
candidat_df <- do.call(rbind, lapply(names(candidat_list), function(cryosection) {
  data.frame(
    cryosection = cryosection,
    candidat_list[[cryosection]]$candidates
  )
}))

candidat_df %>%
  filter(Pvalue < 0.05)
```


###### Colon Cryosection
```{r cryosection_c0, message=FALSE, eval=FALSE}
# Define your cryosection list
cryosection_co <- metadata %>%
  filter(
    section == "Colon",
    !is.na(Xcoord)
  ) %>%
  count(cryosection) %>%
  filter(n > 20) %>%
  pull(cryosection)

# Run analysis
results_co <- spatial_cryosections(
  cryosection_list = cryosection_co,
  metadata_df = metadata_co_clr,
  comm_clr = comm_co_clr
)
```

####### Mantel test
```{r colon_spatial_mantel, comment="", message=FALSE, warning=FALSE}
mantel_pvals_df <- do.call(rbind, lapply(names(results_co$mantel_results), function(cryosection) {
  data.frame(
    cryosection = cryosection,
    r2 = results_co$mantel_results[[cryosection]]$statistic,
    p_value = results_co$mantel_results[[cryosection]]$signif
  )
}))
mantel_pvals_df
# p < 0.05 → Statistically significant correlation between two distance matrices
# p > 0.05 → No significant correlation between two distance matrices
```


####### Mantel Correlogram Stat
```{r colon_spatial_mantelcor, comment="", message=FALSE, warning=FALSE}
for (cryosection in names(results_co$mantelcor_results)) {
  print(cryosection)
  correlog_df <- data.frame(results_co$mantelcor_results[[cryosection]]$mantel.res)
  # sig_classes <- correlog_df[correlog_df$`Pr.Mantel.`< 0.05, ]
  print(correlog_df)
}
```

####### Mantel Correlogram Plots
```{r colon_spatial_mantelcor_plot, comment="", message=FALSE, warning=FALSE}
# Set up a multi-panel layout (adjust cols/rows to fit your data)
n <- length(cryosection_co)
cols <- ceiling(sqrt(n))
rows <- ceiling(n / cols)

par(mfrow = c(rows, cols), mar = c(4, 4, 2, 1)) # Set margins and layout
# Loop over and plot each Mantel correlogram
for (cryosection in names(results_co$mantelcor_results)) {
  plot(results_co$mantelcor_results[[cryosection]])
  title(main = cryosection, line = 1)
}
```

#######  Distance Decay Stat
```{r colon_spatial_decay, comment="", message=FALSE, warning=FALSE}
decay_pvals_df <- do.call(rbind, lapply(names(results_co$decay_dfs), function(cryosection) {
  anova_result <- summary(lmp(comm_dist ~ spat_dist, data = results_co$decay_dfs[[cryosection]]))
  data.frame(
    cryosection = cryosection,
    `Pr(Prob)` = anova_result$coefficients["spat_dist", "Pr(Prob)"]
  )
}))
summary(lmp(comm_dist ~ spat_dist, data = results_co$decay_dfs$G121eO301A))

# Performs an ANOVA to check if spat_dist significantly explains variation in comm_dist.
# p < 0.05 → Significant correlation: Spatial distance influences microbial composition.
# p > 0.05 → No significant relationship: Microbial communities do not change with distance.
```

#######  Distance Decay plots

```{r colon_spatial_decay, comment="", message=FALSE, warning=FALSE}
combined_plot <- cowplot::plot_grid(plotlist = results_co$distance_decay_plots, ncol = 2)
print(combined_plot)
```

```{r cryosection_c0, message=FALSE, eval=FALSE}
# G121eI104C and G121eO301A
summary(lmp(comm_dist ~ spat_dist, data = results_ca$decay_dfs$G121eI104C))
summary(lmp(comm_dist ~ spat_dist, data = results_co$decay_dfs$G121eO301A))



mantel_co_grob <- ggplot(
  data.frame(results_co$mantelcor_results$G121eO301A$mantel.res) %>%
    mutate(sign_label = ifelse(`Pr.Mantel.` <= 0.05, "Significant", "Not significant")) %>%
    filter(!is.na(Mantel.cor)),
  aes(x = class.index, y = Mantel.cor)
) +
  geom_point(aes(fill = sign_label), shape = 23, size = 3) +
  geom_line() +
  scale_fill_manual(values = c("Significant" = "black", "Not significant" = "white")) +
  labs(
    title = "G121eO301A",
    x = "Distance Class Index",
    y = "Mantel Correlation",
    color = "Significance"
  ) +
  geom_hline(yintercept = 0, color = "red") +
  custom_ggplot_theme +
  theme(legend.position = "none")

mantel_ca_grob <- ggplot(
  data.frame(results_ca$mantelcor_results$G121eI104C$mantel.res) %>%
    mutate(sign_label = ifelse(`Pr.Mantel.` <= 0.05, "Significant", "Not significant")) %>%
    filter(!is.na(Mantel.cor)),
  aes(x = class.index, y = Mantel.cor)
) +
  geom_point(aes(fill = sign_label), shape = 23, size = 3) +
  geom_line() +
  scale_fill_manual(values = c("Significant" = "black", "Not significant" = "white")) +
  labs(
    title = "G121eI104C",
    x = "Distance Class Index",
    y = "Mantel Correlation",
    color = "Significance"
  ) +
  geom_hline(yintercept = 0, color = "red") +
  custom_ggplot_theme +
  theme(legend.position = "none")


combined_plot <- cowplot::plot_grid(results_co$distance_decay_plots$G121eO301A,
  mantel_co_grob,
  results_ca$distance_decay_plots$G121eI104C,
  mantel_ca_grob,
  ncol = 2,
  labels = c("a", "b", "c", "d"),
  label_size = 16
)
ggsave("mssm_figures/FigS4_2.png", combined_plot, width = 10, height = 9, dpi = 300)


plot(results_ca$mantelcor_results$G121eI104C, main = "G121eI104C")
plot(results_co$mantelcor_results$G121eO301A, main = "G121eO301A")
```


####### Complex spatial structures
```{r colon_spatial_complex, comment="", message=FALSE, warning=FALSE}
names(results_co$cryosection_dfs)
# Initialize the list once, outside the loop
candidat_list <- list()
for (cryosection in names(results_co$cryosection_dfs)) {
  # for (cryosection in cryosection_list) {
  set.seed(111)

  metadata_data <- results_co$cryosection_dfs[[cryosection]]$metadata
  comm_data <- results_co$cryosection_dfs[[cryosection]]$comm_clr

  spat_SWNs <- listw.candidates(
    data.frame(metadata_data[, c("Xcoord", "Ycoord")]),
    nb = c("gab"),
    weights = c("fdown", "fup"),
    y_fdown = c(2, 5),
    y_fup = c(0.1, 0.5)
  )

  W_sel <- listw.select(
    comm_data,
    spat_SWNs,
    MEM.autocor = "positive",
    p.adjust = TRUE,
    method = "FWD"
  )

  candidat_list[[cryosection]] <- W_sel
}

# Combine all candidate data frames into one with cryosection column
candidat_df <- do.call(rbind, lapply(names(candidat_list), function(cryosection) {
  data.frame(
    cryosection = cryosection,
    candidat_list[[cryosection]]$candidates
  )
}))

candidat_df %>%
  filter(Pvalue < 0.05)
```




####### Colon G121eO301A  

```{r G121eO301A_mem, comment="", message=FALSE, warning=FALSE}
# Set your target cryosection
cryosection <- "G121eO301A"
# Extract MEMs and metadata
mem_data <- candidat_list[[cryosection]]$best$MEM.select
metadata_coords <- results_co$cryosection_dfs[[cryosection]]$metadata[, c("Xcoord", "Ycoord")]
mem_names <- candidat_list[[cryosection]]$best$summary$variables
# Determine layout based on number of MEM axes
n_mem <- ncol(mem_data)
rows <- ceiling(sqrt(n_mem))
cols <- ceiling(n_mem / rows)
# Set up plot layout
par(mfrow = c(rows, cols), mar = c(2, 2, 2, 1))
# Plot all MEM axes
for (i in 1:n_mem) {
  s.value(metadata_coords,
    mem_data[, i],
    sub = paste(cryosection, mem_names[i]),
    csub = 2,
    include.origin = FALSE,
    csize = 0.8
  )
}
# Reset layout
par(mfrow = c(1, 1))
```


```{r G121eO301A_mem_stat, comment="", message=FALSE, warning=FALSE}
s.value(metadata_coords,
  rda(
    results_co$cryosection_dfs[[cryosection]]$comm_clr,
    mem_data
  )$CCA$u[, 1],
  sub = c("Spatial patterns of Colon"),
  csub = 2,
  include.origin = F,
  csize = 0.8
)

RsquareAdj(rda(
  results_co$cryosection_dfs[[cryosection]]$comm_clr,
  mem_data
))
```

###### Fix: rlqESLTP of the Colon cryosection

```{r load_data_rlqESLTP, comment="", message=FALSE, warning=FALSE}
source("data/JEC_1743_sm_apps5.txt")
```

```{r prepare_data_rlqESLTP, comment="", message=FALSE, warning=FALSE}
# comm_co_G121eO301A <- results_co$cryosection_dfs[[cryosection]]$comm_clr
# metadata_co_G121eO301A <- results_co$cryosection_dfs[[cryosection]]$metadata
comm_co_G121eO301A <- comm_co %>%
  as.data.frame() %>%
  rownames_to_column(var = "microsample") %>%
  bind_cols(metadata_co) %>%
  filter(cryosection == "G121eO301A") %>%
  select(contains("bin_"))
metadata_co_G121eO301A <- metadata_co %>%
  filter(cryosection == "G121eO301A")

comp <- decostand(comm_co_G121eO301A, MARGIN = 1, method = "hellinger") # Standardize composition by rows
colnames(comp) <- gsub("\\.", ":", names(comp))
env <- data.frame(
  log_seq_counts = log(metadata_co_G121eO301A$total_after_filtering.total_reads),
  div = rowSums(comp > 0)
) # Environmental matrix

genome_gifts <- macromag_gifts[rownames(macromag_gifts) %in% colnames(comp), ]
genome_funct <- genome_gifts %>%
  data.frame() %>%
  rownames_to_column(var = "id") %>% # Preserve row names as a column
  pivot_longer(-id, names_to = "column", values_to = "value") %>% # Reshape to long format
  mutate(group = substr(column, 1, 3)) %>% # Extract the first 3 characters of column names
  group_by(id, group) %>% # Group by row (id) and prefix
  summarise(mean_value = mean(value, na.rm = TRUE), .groups = "drop") %>% # Calculate rowMeans for each group
  pivot_wider(names_from = group, values_from = mean_value) %>% # Reshape back to wide format
  column_to_rownames(var = "id")
genome_funct <- genome_funct[, -c(19:21)]

phy <- macromag_tree # Phylogenetic tree constructed with Phylomatic V3 + Phylocom

phy <- drop.tip(phy, setdiff(phy$tip.label, rownames(genome_funct)))
spa <- metadata_co_G121eO301A[, c("Xcoord", "Ycoord")] # Matrix of X Y spatial coordinates
comp <- comp[, match(phy$tip.label, colnames(comp))]
genome_funct <- genome_funct[match(phy$tip.label, rownames(genome_funct)), ]

mean(phy$tip.label == colnames(comp))

# Change the phylogenetic tree (an object of the class "phylo") into an object of the class "phylog", used by ade4
phylog <- newick2phylog(write.tree(phy)) # => "phylog" class
colnames(comp) <- gsub(":", "_", colnames(comp))
rownames(genome_funct) <- gsub(":", "_", rownames(genome_funct))
```

###### Fix: Correspondence analysis of the composition matrix 

```{r correspondence_rlqESLTP, comment="", message=FALSE, warning=FALSE}
coacomp <- dudi.coa(comp, scan = FALSE, nf = 108)
summary(coacomp)
```

###### Fix: Spatial analysis 

###### Fix: Gabriel graph

```{r gabriel_graph, comment="", message=FALSE, warning=FALSE}
# Create the Gabriel graph
nb1 <- graph2nb(gabrielneigh(as.matrix(spa)), sym = T)
nb1

# Spatial autocorrelation in diversity and sequence count distribution
sp.correlogram(nb1, log(env$div), order = 8, method = "I")
sp.correlogram(nb1, env$log_seq_counts, order = 8, method = "I")

lw1 <- nb2listw(nb1) # gives a neighbours list with spatial weights (i.e. matrix W)
lw1
```

```{r gabriel_graph_plot, comment="", message=FALSE, warning=FALSE}
par(mfrow = c(1, 1))
plot(nb1, spa, pch = 21, bg = "red")
title(main = "Gabriel Graph")
class(nb1)

# [1] "nb"
# This is the Gabriel graph -the graph that in this case defines which points
# are connected

# The matrix of spatial variables is obtained as the eigenvectors of a
# neighbour matrix. This matrix is analysed by principal component analysis.
nb1.neigh <- nb2neig(nb1)
vecspa <- scores.neig(nb1.neigh)
pcaspa <- dudi.pca(vecspa, coacomp$lw, scan = FALSE, nf = ncol(vecspa))
summary(pcaspa)
```

###### Fix: PCA analysis of the env matrix
```{r env_pca, comment="", message=FALSE, warning=FALSE}
pcaenv <- dudi.pca(env, row.w = coacomp$lw, scannf = FALSE, nf = 2)
summary(pcaenv)
```

###### Fix: The distances between species based on their biological traits, analyzed by PCoA.

```{r distances_traits, comment="", message=FALSE, warning=FALSE}
# Distance matrices for traits separately
listdis <- ldist.ktab(ktab.list.df(list(genome_funct)), c("Q"), scan = FALSE)
# choose 1

# Distance matrix for traits together
disT <- dist.ktab(ktab.list.df(list(genome_funct)), c("Q"), scan = FALSE)
# choose 1
pcotraits <- dudi.pco(disT, coacomp$cw, full = TRUE)
summary(pcotraits)
```

###### Fix: The distances between species based on their phylogenetic relatedness, analysed by PCoA.

```{r distances_phylogeny, comment="", message=FALSE, warning=FALSE}
pcophy <- dudi.pco(as.dist(as.matrix(phylog$Wdist)[names(comp), names(comp)]), coacomp$cw, full = TRUE)
summary(pcophy)

## 11. Tests for phylogenetic signals in traits
## *******************************************
# Function rtest.decdiv() in appendix S5, Pavoine et al., 2011. [1]
# All traits together
phystot <- rtest.decdiv(phylog, rep(1, 108), as.dist(as.matrix(disT)[names(phylog$leaves), names(phylog$leaves)]),
  nrep = 99, vranking = "droot", optiontest = "less", ties.method = "average", option = 3
)
phystot

# All traits independently
rtest.decdiv(phylog, rep(1, ncol(comp)), as.dist(as.matrix(listdis$D01)[
  names(phylog$leaves),
  names(phylog$leaves)
]),
nrep = 99, vranking = "droot", optiontest = "less",
ties.method = "average", option = 3
)
rtest.decdiv(phylog, rep(1, ncol(comp)), as.dist(as.matrix(listdis$D02)[
  names(phylog$leaves),
  names(phylog$leaves)
]),
nrep = 99, vranking = "droot", optiontest = "less",
ties.method = "average", option = 3
)
rtest.decdiv(phylog, rep(1, ncol(comp)), as.dist(as.matrix(listdis$D03)[
  names(phylog$leaves),
  names(phylog$leaves)
]),
nrep = 99, vranking = "droot", optiontest = "less",
ties.method = "average", option = 3
)
rtest.decdiv(phylog, rep(1, ncol(comp)), as.dist(as.matrix(listdis$D05)[
  names(phylog$leaves),
  names(phylog$leaves)
]),
nrep = 99, vranking = "droot", optiontest = "less",
ties.method = "average", option = 3
)
rtest.decdiv(phylog, rep(1, ncol(comp)), as.dist(as.matrix(listdis$D06)[
  names(phylog$leaves),
  names(phylog$leaves)
]),
nrep = 99, vranking = "droot", optiontest = "less",
ties.method = "average", option = 3
)
rtest.decdiv(phylog, rep(1, ncol(comp)), as.dist(as.matrix(listdis$D07)[
  names(phylog$leaves),
  names(phylog$leaves)
]),
nrep = 99, vranking = "droot", optiontest = "less",
ties.method = "average", option = 3
)
rtest.decdiv(phylog, rep(1, ncol(comp)), as.dist(as.matrix(listdis$D08)[
  names(phylog$leaves),
  names(phylog$leaves)
]),
nrep = 99, vranking = "droot", optiontest = "less",
ties.method = "average", option = 3
)
rtest.decdiv(phylog, rep(1, ncol(comp)), as.dist(as.matrix(listdis$D09)[
  names(phylog$leaves),
  names(phylog$leaves)
]),
nrep = 99, vranking = "droot", optiontest = "less",
ties.method = "average", option = 3
)
rtest.decdiv(phylog, rep(1, ncol(comp)), as.dist(as.matrix(listdis$B01)[
  names(phylog$leaves),
  names(phylog$leaves)
]),
nrep = 99, vranking = "droot", optiontest = "less",
ties.method = "average", option = 3
)
rtest.decdiv(phylog, rep(1, ncol(comp)), as.dist(as.matrix(listdis$B02)[
  names(phylog$leaves),
  names(phylog$leaves)
]),
nrep = 99, vranking = "droot", optiontest = "less",
ties.method = "average", option = 3
)
rtest.decdiv(phylog, rep(1, ncol(comp)), as.dist(as.matrix(listdis$B03)[
  names(phylog$leaves),
  names(phylog$leaves)
]),
nrep = 99, vranking = "droot", optiontest = "less",
ties.method = "average", option = 3
)
rtest.decdiv(phylog, rep(1, ncol(comp)), as.dist(as.matrix(listdis$B06)[
  names(phylog$leaves),
  names(phylog$leaves)
]),
nrep = 99, vranking = "droot", optiontest = "less",
ties.method = "average", option = 3
)
rtest.decdiv(phylog, rep(1, ncol(comp)), as.dist(as.matrix(listdis$B07)[
  names(phylog$leaves),
  names(phylog$leaves)
]),
nrep = 99, vranking = "droot", optiontest = "less",
ties.method = "average", option = 3
)
rtest.decdiv(phylog, rep(1, ncol(comp)), as.dist(as.matrix(listdis$B08)[
  names(phylog$leaves),
  names(phylog$leaves)
]),
nrep = 99, vranking = "droot", optiontest = "less",
ties.method = "average", option = 3
)
rtest.decdiv(phylog, rep(1, ncol(comp)), as.dist(as.matrix(listdis$B09)[
  names(phylog$leaves),
  names(phylog$leaves)
]),
nrep = 99, vranking = "droot", optiontest = "less",
ties.method = "average", option = 3
)
rtest.decdiv(phylog, rep(1, ncol(comp)), as.dist(as.matrix(listdis$B10)[
  names(phylog$leaves),
  names(phylog$leaves)
]),
nrep = 99, vranking = "droot", optiontest = "less",
ties.method = "average", option = 3
)

## Update the PCOA of traits to remove traits without phylogenetic signal
# Distance matrix for traits together
disT <- dist.ktab(ktab.list.df(list(genome_funct[, -c(4, 8, 13)])), c("Q"), scan = FALSE)
# choose 1
pcotraits <- dudi.pco(disT, coacomp$cw, full = TRUE)
summary(pcotraits)
GIFT_db
```

###### Fix: Extended RLQ analysis

```{r extended_rlq, comment="", message=FALSE, warning=FALSE}
rlqmix <- rlqESLTP(pcaenv, pcaspa, coacomp, pcotraits, pcophy, scan = F, nf = 2)
barplot(rlqmix$eig)
rlqmix$eig[1] / sum(rlqmix$eig)
# [1] 0.73
rlqmix$eig[2] / sum(rlqmix$eig)
# [1] 0.12
```

**1st axis**
```{r extended_rlq_axis1, comment="", message=FALSE, warning=FALSE}
plot(rlqmix, xy = spa, ax = 1, wh = "S")
plot(rlqmix, phy = phylog, ax = 1, wh = "P")
plot(rlqmix, traits = genome_funct[, -c(4, 8, 13)], ax = 1, type = "Q", wh = "T")
plot(rlqmix, env = pcaenv$tab, ax = 1, type = "Q", wh = "E")

s.value(metadata_co_G121eO301A[, c("Xcoord", "Ycoord")],
  rlqmix$lR[, 1],
  sub = "Patterns in S and E",
  csub = 1.5,
  include.origin = F,
  csize = 0.8
)

plot(rlqmix, env = pcaenv$tab, ax = 1, type = "Q", wh = "E")

dotchart.phylog(phylog, rlqmix$lQ[names(phylog$leaves), 1],
  cleav = 0, cdot = 1,
  scaling = F, yjoi = 0, cex.axis = 1.5, sub = "patterns in T and P", csub = 0
)


# Create data frame with coordinates and values
df <- data.frame(
  X = metadata_co_G121eO301A$Xcoord,
  Y = metadata_co_G121eO301A$Ycoord,
  Value = rlqmix$lR[, 1]
)
ggplot(df, aes(x = X, y = Y, color = Value, size = abs(Value))) +
  geom_point(alpha = 0.8) +
  scale_color_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  guides(size = FALSE) + # optional: hide size legend
  labs(subtitle = "Patterns in S and E", color = "Value") +
  theme_minimal() +
  theme(
    plot.subtitle = element_text(size = 14, face = "italic"),
    axis.title = element_blank()
  ) +
  coord_fixed() # THIS fixes the aspect ratio so distances on x and y are equal


# 1. Species scores (Q)
df_species <- data.frame(
  Species = rownames(rlqmix$lQ),
  Axis1 = rlqmix$lQ[, 1],
  Axis2 = rlqmix$lQ[, 2]
)

# 2. Environmental variables loadings (from pcaenv)
df_env <- data.frame(
  EnvVar = rownames(pcaenv$co),
  Axis1 = pcaenv$co[, 1],
  Axis2 = pcaenv$co[, 2]
)

# 3. Scale environmental variables loadings for arrows (adjust multiplier as needed)
arrow_mult <- 3
df_env$Axis1 <- df_env$Axis1 * arrow_mult
df_env$Axis2 <- df_env$Axis2 * arrow_mult

# 4. Plot
ggplot() +
  geom_point(data = df_species, aes(x = Axis1, y = Axis2), color = "blue") +
  geom_segment(
    data = df_env, aes(x = 0, y = 0, xend = Axis1, yend = Axis2),
    arrow = arrow(length = unit(0.3, "cm")), color = "red"
  ) +
  geom_text_repel(
    data = df_env, aes(x = Axis1, y = Axis2, label = EnvVar),
    color = "red"
  ) +
  labs(x = "RLQ Axis 1", y = "RLQ Axis 2", title = "RLQ plot of species and environmental variables") +
  theme_minimal() +
  coord_equal()

df <- data.frame(
  leaf = factor(names(phylog$leaves), levels = rev(names(phylog$leaves))), # reverse order so top = first leaf
  score = rlqmix$lQ[names(phylog$leaves), 1]
)

ggplot(df, aes(x = score, y = leaf)) +
  geom_point(size = 3, color = "blue") +
  labs(x = "RLQ Axis 1", y = "Phylogeny leaves", subtitle = "Patterns in T and P") +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 8))

p_tree <- ggtree(phylog$tre) + theme_tree2()
p_dot <- ggplot(df, aes(x = score, y = leaf)) +
  geom_point(size = 3, color = "blue") +
  labs(x = "RLQ Axis 1", y = NULL, subtitle = "Patterns in T and P") +
  theme_minimal() +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank())

p_tree + p_dot + plot_layout(widths = c(1, 1))

png("mssm_figures/FigS43.png", width = 800, height = 1200, res = 150) # adjust resolution & size

png("mssm_figures/plot_a.png", width = 800, height = 400, res = 150)
s.value(
  metadata_co_G121eO301A[, c("Xcoord", "Ycoord")],
  rlqmix$lR[, 1],
  sub = "Patterns in S and E",
  csub = 1.5,
  include.origin = FALSE,
  csize = 0.8
)
mtext("a", side = 3, line = 0.5, at = par("usr")[1], font = 2, cex = 1.5)
dev.off()

png("mssm_figures/plot_b.png", width = 800, height = 400, res = 150)
plot(rlqmix, env = pcaenv$tab, ax = 1, type = "Q", wh = "E")
# mtext("b", side = 3, line = 0.5, at = par("usr")[1], font = 2, cex = 1.5)
dev.off()

png("mssm_figures/plot_c.png", width = 800, height = 600, res = 150)
dotchart.phylog(
  phylog,
  rlqmix$lQ[names(phylog$leaves), 1],
  cleav = 0,
  cdot = 1,
  scaling = FALSE,
  yjoi = 0,
  cex.axis = 1.5,
  sub = "patterns in T and P",
  csub = 0
)
# mtext("c", side = 3, line = 0.5, at = par("usr")[1], font = 2, cex = 1.5)
dev.off()
install.packages("magick")
library(magick)

img_a <- image_read("mssm_figures/plot_a.png")
img_b <- image_read("mssm_figures/plot_b.png")
img_c <- image_read("mssm_figures/plot_c.png")

combined <- image_append(c(img_b, img_c), stack = TRUE)
image_write(combined, "mssm_figures/combined_plots.png")
```


**2nd axis**

```{r extended_rlq_axis2, comment="", message=FALSE, warning=FALSE}
plot(rlqmix, xy = spa, ax = 2, wh = "S")
plot(rlqmix, phy = phylog, ax = 2, wh = "P")
plot(rlqmix, traits = genome_funct, ax = 2, type = "Q", wh = "T")
plot(rlqmix, env = pcaenv$tab, ax = 2, type = "Q", wh = "E")
```

