# Loading and Tidying Data
**Note:** This step requires access to the raw data files located at xxxx. If you prefer to skip this step, you can proceed to the next section, making sure to load the R object that contains the necessary files for the subsequent analysis in location.

## Import Sample metadata
```{r sample_metadata_mssm, eval=FALSE}
mssm_sample_metadata <- airtable("4-MSE-Info", "appKakM1bnKSekwuW") %>%  #get base ID from Airtable browser URL
  read_airtable(., fields = c(
    "ID", "LabBatch_text", "IntestinalSection", "SampleType",
    "Xcoord", "Ycoord", "SizeApprox", "cryosection_text", "buffer_text", "Collection_method_text",
    "Collection_Success", "Collection_attempts", "UsedCycles", "animal_temp", "Protocol_text",
    "adapter_type_text", "nmole_L_100_200_bp", "nmole_L_200_1000_bp"
  ), id_to_col = TRUE) %>% 
  filter(
    LabBatch_text %in% c(
      #"MSEB0001", "MSEB0002", "MSEB0003", "MSEB0004", "MSEB0005",
      "MSEB0006", "MSEB0009", "MSEB0010", "MSEB0011", "MSEB0012",
      "MSEB0014", "MSEB0015"
    ),
    !SampleType %in% c("ControlHuman")
  ) %>%
  rename(
    batch = LabBatch_text, microsample = ID, section = IntestinalSection, 
    type = SampleType, cryosection = cryosection_text, 
    buffer = buffer_text, collection_method = Collection_method_text,
    collection = Collection_Success, collection_attempts = Collection_attempts,
    cycles = UsedCycles, animal = animal_temp, size = SizeApprox, 
    protocol = Protocol_text, adapter = adapter_type_text
  ) %>%
  select(
    microsample, section, type, batch, cryosection, buffer, Xcoord, Ycoord, 
    size, collection_method, collection, collection_attempts, cycles, animal, 
    protocol, adapter, nmole_L_100_200_bp, nmole_L_200_1000_bp
  ) %>%
  mutate(
    type_binomial = ifelse(type == "Positive", "positive", "negative"),
    buffer = case_when(
      buffer == "B01" ~ "L01",
      buffer == "B02" ~ "L02",
      buffer == "B03" ~ "L03",
      buffer == "B04" ~ "L04",
      buffer == "B09" ~ "L05",
      buffer == "B11" ~ "L06",
      buffer == "B12" ~ "L07",
      buffer == "B13" ~ "L08",
      TRUE ~ buffer
    )
  ) %>%
  unnest(c(section, Xcoord, Ycoord, size, collection, cycles, nmole_L_100_200_bp, nmole_L_200_1000_bp)) %>%
  arrange(microsample)
#write.table(mssm_sample_metadata, "data/data_preparation/mssm_sample_metadata.tsv")
#mssm_sample_metadata <- read.table("data/data_preparation/mssm_sample_metadata.tsv")
```

## Sequencing Data Quality 
### Read Quality: Pre- vs Post-Filtering (fastp)
**Note:** The parse_fastp() function in index needs to be loaded.
```{r read_quality, eval=FALSE}
mssm_json_data <- fromJSON("data/data_preparation/multiqc_data.json")  #Download file from ERDA
mssm_fastp <- parse_fastp(mssm_json_data) %>%
  rename_with(~ gsub("^(summary|filtering_result|adapter_cutting|polyx_trimming)\\.", "", .)) %>%
  select(
    sample_id,
    before_filtering.total_reads,
    after_filtering.total_reads,
    before_filtering.total_bases,
    after_filtering.total_bases,
    before_filtering.gc_content,
    after_filtering.gc_content,
    low_quality_reads,
    adapter_trimmed_reads,
    adapter_trimmed_bases,
    duplication.rate
  ) %>%
  mutate(microsample = str_remove(sample_id, "\\.lib\\d+$")) %>%
  filter(grepl("^M", microsample)) %>%
  group_by(microsample) %>%
  summarise(
    across(contains("reads"), sum, .names = "total_{.col}"),
    across(contains("bases"), sum, .names = "total_{.col}"),
    across(ends_with("content"), mean, .names = "percent_{.col}"),
    across(ends_with("rate"), mean, .names = "percent_{.col}")
  )
```


### Contamination Assessment: Host, Human, and Environmental Reads
```{r mapping_eukaryots_references, eval=FALSE}
mssm_euk_mapping <- read_tsv("https://sid.erda.dk/share_redirect/aTWbNfAf1o/results/preprocess_bowtie2_multiqc_report_data/multiqc_samtools_stats.txt") %>%
  mutate(
    reference = case_when(
      grepl("GRCh38", Sample, ignore.case = TRUE) ~ "human",
      grepl("GRCg7b", Sample, ignore.case = TRUE) ~ "chicken",
      grepl("Sscrofa11.1", Sample, ignore.case = TRUE) ~ "pig",
      TRUE ~ NA_character_
    ),
    microsample = str_extract(Sample, "M\\d+")
  ) %>%
  filter(
    !is.na(reference), # Remove rows where reference is NA
    grepl("^M", microsample)
  ) %>%
  select(Sample, microsample, reference, raw_total_sequences, reads_mapped, reads_unmapped) %>%
  group_by(microsample, reference) %>%
  summarise(
    raw_total_sequences = sum(raw_total_sequences, na.rm = TRUE),
    reads_mapped = sum(reads_mapped, na.rm = TRUE),
    reads_unmapped = sum(reads_unmapped, na.rm = TRUE)
  ) %>%
  pivot_wider(names_from = reference, values_from = c(raw_total_sequences, reads_mapped, reads_unmapped))
```

### Genome-Centric Mapping Statistics: Macroscale MAG Catalogue
```{r mapping_macro_catalogue, eval=FALSE}
# download.file("https://sid.erda.dk/share_redirect/aTWbNfAf1o/results/quantify.zip", "data/data_preparation/quantify.zip", mode = "wb")
mssm_mag_mapping <- read.table(unz("data/data_preparation/quantify.zip", "multiqc_samtools_stats.txt"), header = T, sep = "\t", quote = "\"") %>%
  mutate(microsample = str_extract(Sample, "M\\d+")) %>%
  filter(grepl("^M", microsample)) %>%
  select(Sample, microsample, raw_total_sequences, reads_mapped, reads_unmapped) %>%
  group_by(microsample) %>%
  summarise(
    raw_total_sequences_mag = sum(raw_total_sequences, na.rm = TRUE),
    reads_mapped_mag = sum(reads_mapped, na.rm = TRUE),
    reads_unmapped_mag = sum(reads_unmapped, na.rm = TRUE)
  )
```

### Read-Centric Mapping Statistics: Kraken (genomes) and SingleM (single-copy marker genes)
```{r mapping_singlem, eval=FALSE}
mssm_singlem <- read_tsv("https://sid.erda.dk/share_redirect/aTWbNfAf1o/results/preprocess/singlem/microbial_fraction.tsv") %>%
  mutate(microsample = str_extract(sample, "M\\d+")) %>%
  filter(grepl("^M", microsample)) %>%
  select(-c(warning, sample)) %>%
  group_by(microsample) %>%
  summarise(
    read_fraction_singlem = (mean(read_fraction, na.rm = TRUE)) / 100 #,
    # bacterial_archaeal_bases_singlem = sum(bacterial_archaeal_bases, na.rm = TRUE),
    # average_bacterial_archaeal_genome_size_singlem = mean(average_bacterial_archaeal_genome_size, na.rm = TRUE)
  )
```

```{r mapping_kraken, eval=FALSE}
mssm_kraken <- read_tsv("https://sid.erda.dk/share_redirect/aTWbNfAf1o/results/preprocess_kraken2_multiqc_report_data/multiqc_kraken.txt") %>%
  mutate(microsample = str_extract(Sample, "M\\d+")) %>%
  filter(grepl("^M", microsample)) %>%
  select(Sample, microsample, D) %>%
  separate(D, into = c("Eukaryota", "Bacteria", "Archaea", "Viruses"), sep = ", ", remove = TRUE) %>%
  mutate(across(!all_of("microsample"), ~ parse_number(.))) %>%
  group_by(microsample) %>%
  summarise(
    kr_Eukaryota = sum(Eukaryota, na.rm = TRUE),
    kr_Bacteria = sum(Bacteria, na.rm = TRUE),
    kr_Archaea = sum(Archaea, na.rm = TRUE),
    kr_Viruses = sum(Viruses, na.rm = TRUE),
  ) %>%
  mutate(
    total_kraken = kr_Eukaryota + kr_Bacteria + kr_Archaea + kr_Viruses,
    bacteria_pct_kraken = kr_Bacteria / total_kraken,
    eukaryota_pct_kraken = kr_Eukaryota / total_kraken
  )
```

### Redundancy & Coverage Estimation: Nonpareil
```{r estimates_nonpareil, eval=FALSE}
mssm_nonpareil <- read_tsv("https://sid.erda.dk/share_redirect/aTWbNfAf1o/results/preprocess_nonpareil_multiqc_report_data/nonpareil-table.txt") %>%
  mutate(microsample = str_extract(Sample, "M\\d+")) %>%
  filter(grepl("^M", microsample)) %>%
  select(microsample, Redundancy, Coverage, Diversity) %>%
  group_by(microsample) %>%
  summarise(
    redundancy_nonpareil = (mean(Redundancy, na.rm = TRUE)) / 100,
    coverage_nonpareil = (mean(Coverage, na.rm = TRUE)) / 100,
    diversity_nonpareil = mean(Diversity, na.rm = TRUE)
  )
```

### Combined Quality Statistics
```{r data_quality_df, eval=FALSE}
mssm_quality <- reduce(
  list(
    mssm_sample_metadata,
    mssm_fastp,
    mssm_euk_mapping,
    mssm_mag_mapping,
    mssm_singlem,
    mssm_kraken,
    mssm_nonpareil
  ),
  full_join,
  by = "microsample"
)  %>%
  mutate(
    lowquality_reads_pct = ((total_before_filtering.total_reads - total_after_filtering.total_reads) / total_before_filtering.total_reads),
    adapter_contamination_pct = (total_adapter_trimmed_bases / total_before_filtering.total_bases),
    host_pct = (reads_mapped_chicken / total_before_filtering.total_reads),
    human_pct = (reads_mapped_human / total_before_filtering.total_reads),
    pig_pct = (reads_mapped_pig / total_before_filtering.total_reads),
    prokreads_pct = (raw_total_sequences_mag / total_before_filtering.total_reads),
    mapmag_rawreads_pct = (reads_mapped_mag / total_before_filtering.total_reads),
    mapmag_filteredreads_pct = (reads_mapped_mag / total_after_filtering.total_reads),
    mapmag_prokreads_pct = (reads_mapped_mag / raw_total_sequences_mag)
  ) %>%
  filter(microsample %in% c(mssm_sample_metadata$microsample))


```

## Input Preparation For Metagenomic Analysis

### Abundance Estimation Using Macro-scale MAGs Catalogue 95 ANI

#### Import Genome Counts

```{r mag_readcount_macroquant, eval=FALSE}
macroquant_read_counts <- read_tsv("https://sid.erda.dk/share_redirect/aTWbNfAf1o/results/quantify/coverm/genome.count.REF0015-g_mg_hybrid-0.95.tsv.gz", show_col_types = FALSE) %>%
  rename(genome = 1) %>%
  pivot_longer(!genome, names_to = "data", values_to = "counts") %>%
  mutate(sample = substr(data, 1, 7)) %>%
  group_by(genome, sample) %>%
  summarise(counts = sum(counts), .groups = "drop") %>%
  pivot_wider(names_from = "sample", values_from = "counts") 
```

```{r mag_coveredbases_macroquant, eval=FALSE}
macroquant_covered_bases <- read_tsv("https://sid.erda.dk/share_redirect/aTWbNfAf1o/results/quantify/coverm/genome.covered_bases.REF0015-g_mg_hybrid-0.95.tsv.gz", show_col_types = FALSE) %>%
  rename(genome = 1) %>%
  pivot_longer(!genome, names_to = "data", values_to = "counts") %>%
  mutate(sample = substr(data, 1, 7)) %>%
  group_by(genome, sample) %>%
  summarise(counts = sum(counts), .groups = "drop") %>%
  pivot_wider(names_from = "sample", values_from = "counts")
```

#### Macro-scale MAGs Catalogue
##### Taxonomy
This is the raw taxonomy table generated by GTDBtk, which is simplified for downstream analyses.
```{r mag_taxonomy_macro, eval=FALSE}
macromag_taxonomy <- read_tsv("https://sid.erda.dk/share_redirect/BTwCSjaWz2/REF0015-g_mg_hybrid/gtdbtk.summary.tsv.gz", show_col_types = FALSE) %>%
  rename(genome = user_genome) %>%
  mutate(genome = str_replace_all(genome, "\\.fa", "")) %>%
  separate(classification, c("domain", "phylum", "class", "order", "family", "genus", "species"), sep = ";") %>%
  select(genome, domain, phylum, class, order, family, genus, species) %>%
  mutate(
    domain = sub("^d__", "", domain), # Remove prefix
    phylum = sub("^p__", "", phylum),
    class = sub("^c__", "", class),
    order = sub("^o__", "", order),
    family = sub("^f__", "", family),
    genus = sub("^g__", "", genus),
    species = sub("^s__", "", species)
  ) %>%
  arrange(match(genome, macroquant_read_counts$genome)) %>%
  filter(genome %in% macroquant_read_counts$genome) %>%
  mutate(
    species = if_else(
      is.na(species) | species == "",
      paste0("unclassified_", genus),
      species
    )
  ) %>%
  mutate(
    species = if_else(
      is.na(species) | species == "unclassified_",
      paste0("unclassified_", family),
      species
    )
  ) %>%
  mutate(
    genus = if_else(
      is.na(genus) | genus == "",
      paste0("unclassified_", family),
      genus
    )
  )
```

##### Quality
```{r mag_quality_macro, eval=FALSE}
macromag_quality <- read_tsv("https://sid.erda.dk/share_redirect/BTwCSjaWz2/REF0015-g_mg_hybrid/checkm2.quality_report.tsv.gz", show_col_types = FALSE) %>%
  # rename first column
  rename(genome = 1) %>%
  # remove .fa from the end of each genome name
  mutate(genome = str_replace_all(genome, "\\.fa", "")) %>%
  # select relevant columns and rename
  select(genome, Completeness, Contamination, Coding_Density, Genome_Size, Contig_N50) %>%
  rename(
    completeness = Completeness,
    contamination = Contamination,
    coding_density = Coding_Density,
    length = Genome_Size,
    N50_length = Contig_N50
  ) %>%
  # estimate genome circularity (N50 should be equal to genome length)
  mutate(circularity = N50_length / length) %>%
  mutate(circularity = ifelse(circularity >= 1, "1", "0")) %>%
  # update completeness and contamination for genomes with circularity of 1
  mutate(
    completeness = ifelse(circularity == "1", 100, completeness),
    contamination = ifelse(circularity == "1", 0, contamination)
  ) %>%
  # estimate ratio of: genome length / median bacterial genome length
  mutate(length_ratio = length / median(length, na.rm = TRUE))

macromag_genomemetadata <- macromag_taxonomy %>%
  left_join(macromag_quality, by = join_by(genome == genome)) %>% # join quality
  arrange(match(genome, macroquant_read_counts$genome)) %>%
  filter(genome %in% macroquant_read_counts$genome)

```

##### Functional Annotation

```{r mag_annotation_macro, eval=FALSE}
# macromag_annotations <- read_tsv("https://sid.erda.dk/share_redirect/BTwCSjaWz2/REF0015-g_mg_hybrid/dram.annotations.tsv.gz", show_col_types = FALSE) %>%
macromag_annotations <- read_tsv("data/data_preparation/annotations.tsv", show_col_types = FALSE) %>%
  rename(gene = 1, genome = 2) %>%
  filter(genome %in% macroquant_read_counts$genome)
macromag_gifts <- distill(macromag_annotations, GIFT_db, genomecol = 2, annotcol = c(9, 10, 19))
GIFT_db_df <- GIFT_db
```

##### Phylogenetic Tree

```{r mag_tree_macro, eval=FALSE}
macromag_tree <- read.tree("https://sid.erda.dk/share_redirect/BTwCSjaWz2/REF0015-g_mg_hybrid/gtdbtk.backbone.bac120.classify.tree")
macromag_tree$tip.label <- str_replace_all(macromag_tree$tip.label, "'", "") # remove single quotes in MAG names
macromag_tree <- keep.tip(macromag_tree, tip = macroquant_read_counts$genome) # keep only MAG tips
```

#### Coverage Filtering & Genome Count Normalization

##### Coverage filtering
**Note:** Genomes that have less than 30% of their length covered by reads are turned into zeros to account for the random allocation of reads across genomes due to mapping heuristics. 
```{r coverage_fitering_macroquant, eval=FALSE}
macroquant_coverage <- macroquant_covered_bases %>% # calculate coverage values by dividing the number of covered bases by the length of each genome
  mutate(across(where(is.numeric), ~ . / macromag_genomemetadata$length))

macroquant_read_counts_filtered <- list()
macroquant_summary_filtered <- list()
# Define thresholds
coverage_thresholds <- c(0.4, 0.35, 0.3, 0.25, 0.2, 0.15, 0.1, 0.05)
# Loop over thresholds
for (threshold in coverage_thresholds) {
  df <- macroquant_coverage %>%
    mutate(across(where(is.numeric), ~ ifelse(. > threshold, 1, 0))) %>%
    mutate(across(-1, ~ . * macroquant_read_counts[[cur_column()]]))
  summary <- df %>%
    summarise(across(where(is.numeric), \(x) sum(x, na.rm = TRUE))) %>%
    pivot_longer(
      cols = where(is.numeric), # columns to pivot (all numeric here)
      names_to = "microsample", # new column for former column names
      values_to = paste0("read_count_", threshold * 100) # new column for values
    )
  name <- paste0("filt_", threshold * 100)
  macroquant_read_counts_filtered[[name]] <- df
  macroquant_summary_filtered[[name]] <- summary
}
macroquant_merged_summary <- reduce(macroquant_summary_filtered, left_join, by = "microsample")

min_coverage <- 0.3
macroquant_read_counts_filt_30 <- macroquant_coverage %>%
  mutate(across(where(is.numeric), ~ ifelse(. > min_coverage, 1, 0))) %>%
  mutate(across(-1, ~ . * macroquant_read_counts[[cur_column()]]))

all(macroquant_read_counts_filtered$filt_30 == macroquant_read_counts_filt_30)

```

##### Generate genome count table
After filtering the low-coverage reads, read counts are transformed into genome counts using genome-length and read-length information.
```{r genome_counts_macroquant, eval=FALSE}
readlength <- 150 # change if sequencing read length is different
macroquant_genome_counts_filt_30 <- macroquant_read_counts_filt_30 %>%
  mutate(across(where(is.numeric), ~ . / (macromag_genomemetadata$length / readlength))) 
macroquant_genome_counts  <- macroquant_read_counts %>%
  mutate(across(where(is.numeric), ~ . / (macromag_genomemetadata$length / readlength))) 
```

#### Alpha Diversity
**Note:** The calculate_alpha_diversity() function needs to be loaded.
```{r alpha_diversities_macroquant, eval=FALSE}
# Calculate alpha diversity for filtered genome counts data
macroquant_metrics_filt_30 <- calculate_alpha_diversity(
  input_data = macroquant_genome_counts_filt_30,
  metadata_name = mssm_quality,
  dataset_name = "filt_30",
  tree_name = macromag_tree
) %>% filter(!microsample %in% c("Undeter"))
``` 

```{r nMOL, eval=FALSE}
# Calculate alpha diversity for filtered genome counts data
macroquant_metrics_filt_30 %>% 
  group_by(type_binomial, cov_filtering, size) %>% 
  summarise(nmole_median=median(nmole_L_200_1000_bp,na.rm = TRUE),
            nmole_IQR1=quantile(nmole_L_200_1000_bp, probs = c(0.25), na.rm = TRUE),
            nmole_IQR2=quantile(nmole_L_200_1000_bp, probs = c(0.75), na.rm = TRUE),
            nmole_min=min(nmole_L_200_1000_bp, na.rm = TRUE),
            nmole_max=max(nmole_L_200_1000_bp, na.rm = TRUE))

```


### Abundance Estimation Using Micro-scale MAGs Catalogue

#### Import Genome Counts

```{r mag_readcount_microquant, eval=FALSE}
microquant_read_counts <- read_tsv("https://sid.erda.dk/share_redirect/dn35KSggCd/genome.count.REF0025-MSEB0011-0.95.tsv.gz", show_col_types = FALSE) %>%
  rename(genome = 1) %>%
  pivot_longer(!genome, names_to = "data", values_to = "counts") %>%
  mutate(sample = str_extract(data, "M\\d+")) %>%  # Extract the sample name
  group_by(genome, sample) %>%
  summarise(counts = sum(counts), .groups = "drop") %>%
  pivot_wider(names_from = "sample", values_from = "counts")

```

```{r mag_coveredbases_microquant, eval=FALSE}
microquant_covered_bases <- read_tsv("https://sid.erda.dk/share_redirect/dn35KSggCd/genome.covered_bases.REF0025-MSEB0011-0.95.tsv.gz", show_col_types = FALSE) %>%
  rename(genome = 1) %>%
  pivot_longer(!genome, names_to = "data", values_to = "counts") %>%
  mutate(sample = str_extract(data, "M\\d+")) %>%
  group_by(genome, sample) %>%
  summarise(counts = sum(counts), .groups = "drop") %>%
  pivot_wider(names_from = "sample", values_from = "counts")
```

#### Micro-scale MAGs Catalogue
##### Taxonomy
This is the raw taxonomy table generated by GTDBtk, which is simplified for downstream analyses.
```{r mag_taxonomy_micro, eval=FALSE}
micromag_taxonomy <- read_tsv("https://sid.erda.dk/share_redirect/BTwCSjaWz2/REF0025-MSEB0011/gtdbtk.summary.tsv", show_col_types = FALSE) %>%
  rename(genome = user_genome) %>%
  mutate(genome = str_replace_all(genome, "\\.fa", "")) %>%
  separate(classification, c("domain", "phylum", "class", "order", "family", "genus", "species"), sep = ";") %>%
  select(genome, domain, phylum, class, order, family, genus, species) %>%
    mutate(
      domain = sub("^d__", "", domain), # Remove prefix
      phylum = sub("^p__", "", phylum), 
      class = sub("^c__", "", class), 
      order = sub("^o__", "", order), 
      family = sub("^f__", "", family), 
      genus = sub("^g__", "", genus), 
      species = sub("^s__", "", species)
    )  %>% 
  arrange(match(genome, microquant_read_counts$genome)) %>% 
  filter(genome %in% microquant_read_counts$genome) %>%
  mutate(
    species = if_else(
      is.na(species) | species == "",
      paste0("unclassified_", genus),
      species
    )
  ) %>%
  mutate(
    species = if_else(
      is.na(species) | species == "unclassified_",
      paste0("unclassified_", family),
      species
    )
  ) %>%
  mutate(
    genus = if_else(
      is.na(genus) | genus == "",
      paste0("unclassified_", family),
      genus
    )
  ) 


```

##### Quality
```{r mag_quality_micro, eval=FALSE}
micromag_quality <- read_tsv("https://sid.erda.dk/share_redirect/BTwCSjaWz2/REF0025-MSEB0011/checkm2.quality_report.tsv", show_col_types = FALSE) %>%
  # rename first column
  rename(genome = 1) %>% 
  # remove .fa from the end of each genome name
  mutate(genome = str_replace_all(genome, "\\.fa", "")) %>%
  # select relevant columns and rename
  select(genome, Completeness, Contamination, Coding_Density, Genome_Size, Contig_N50) %>% 
  rename(completeness = Completeness, 
         contamination = Contamination, 
         coding_density = Coding_Density, 
         length = Genome_Size,
         N50_length = Contig_N50) %>%
  # estimate genome circularity (N50 should be equal to genome length)
  mutate(circularity = N50_length/length) %>%
  mutate(circularity = ifelse(circularity >= 1, "1", "0")) %>%
  # update completeness and contamination for genomes with circularity of 1
  mutate(completeness = ifelse(circularity == "1", 100, completeness),
         contamination = ifelse(circularity == "1", 0, contamination)) %>%   
  # estimate ratio of: genome length / median bacterial genome length
  mutate(length_ratio = length/median(length, na.rm = TRUE))

micromag_genomemetadata <- micromag_taxonomy %>%
  left_join(micromag_quality, by = join_by(genome == genome)) %>% # join quality
  arrange(match(genome, microquant_read_counts$genome)) %>% 
  filter(genome %in% microquant_read_counts$genome)


```

##### Functional Annotation

```{r mag_annotation_micro, eval=FALSE}
#micromag_annotations <- read_tsv("https://sid.erda.dk/share_redirect/BTwCSjaWz2/REF0025-MSEB0011/dram.annotations.tsv.gz", show_col_types = FALSE) %>%
#micromag_annotations <- read_tsv("data/data_preparation/micro_annotations.tsv.xz", show_col_types = FALSE) %>%
#  rename(gene = 1, genome = 2)  %>%
#  filter(genome %in% microquant_read_counts$genome)
#micromag_gifts <- distill(micromag_annotations, GIFT_db, genomecol = 2, annotcol = c(9, 10, 19))
```

##### Phylogenetic Tree

```{r mag_tree_micro, eval=FALSE}
micromag_tree <- read.tree("https://sid.erda.dk/share_redirect/BTwCSjaWz2/REF0025-MSEB0011/gtdbtk.backbone.bac120.classify.tree")
micromag_tree$tip.label <- str_replace_all(micromag_tree$tip.label, "'", "") # remove single quotes in MAG names
micromag_tree <- keep.tip(micromag_tree, tip = microquant_read_counts$genome) # keep only MAG tips
```

#### Coverage Filtering & Genome Count Normalization

##### Coverage filtering
**Note:** Genomes that have less than 30% of their length covered by reads are turned into zeros to account for the random allocation of reads across genomes due to mapping heuristics. 
```{r coverage_fitering_microquant, eval=FALSE}
microquant_coverage<- microquant_covered_bases %>% # calculate coverage values by dividing the number of covered bases by the length of each genome
  mutate(across(where(is.numeric), ~ . / micromag_genomemetadata$length))

microquant_read_counts_filtered <- list()
microquant_summary_filtered <- list()
# Define thresholds
coverage_thresholds <- c(0.4, 0.35, 0.3, 0.25, 0.2, 0.15, 0.1, 0.05)
# Loop over thresholds
for (threshold in coverage_thresholds) {
  df <- microquant_coverage%>%
    mutate(across(where(is.numeric), ~ ifelse(. > threshold, 1, 0))) %>%
    mutate(across(-1, ~ . * microquant_read_counts[[cur_column()]]))
  summary <- df %>%
    summarise(across(where(is.numeric), \(x) sum(x, na.rm = TRUE)))%>%
    pivot_longer(
      cols = where(is.numeric), # columns to pivot (all numeric here)
      names_to = "microsample", # new column for former column names
      values_to = paste0("read_count_", threshold * 100) # new column for values
    )
  name <- paste0("filt_", threshold * 100)
  microquant_read_counts_filtered[[name]] <- df
  microquant_summary_filtered[[name]] <- summary
}
microquant_merged_summary <- reduce(microquant_summary_filtered, left_join, by = "microsample")

min_coverage <- 0.3
microquant_read_counts_filt_30 <- microquant_coverage %>%
  mutate(across(where(is.numeric), ~ ifelse(. > min_coverage, 1, 0))) %>%
  mutate(across(-1, ~ . * microquant_read_counts[[cur_column()]]))

all(microquant_read_counts_filtered$filt_30 == microquant_read_counts_filt_30)

```

##### Generate genome count table
After filtering the low-coverage reads, read counts are transformed into genome counts using genome-length and read-length information.
```{r genome_counts_microquant, eval=FALSE}
readlength <- 150 # change if sequencing read length is different
microquant_genome_counts_filt_30 <- microquant_read_counts_filt_30 %>%
  mutate(across(where(is.numeric), ~ . / (micromag_genomemetadata$length / readlength)))
microquant_genome_counts <- microquant_read_counts %>%
  mutate(across(where(is.numeric), ~ . / (micromag_genomemetadata$length / readlength)))
```

#### Alpha Diversity
**Note:** The `calculate_alpha_diversity()` function needs to be loaded.

```{r alpha_diversities_microquant, eval=FALSE}
# Calculate alpha diversity for filtered genome counts data
microquant_metrics_filt_30 <- calculate_alpha_diversity(
  input_data = microquant_read_counts_filt_30,
  metadata_name = mssm_quality,
  dataset_name = "filt_30",
  tree_name = micromag_tree
) %>% filter(!microsample %in% c("Undeter"))

``` 

### Abundance Estimation Using Macro-scale MAGs Catalogue 98 ANI
##### Mapping Data
```{r load_98ANI, eval=FALSE}
# Read count
macroquant_read_counts_98 <- read_tsv("data/data_preparation/genome.count.REF0015-g_mg_hybrid-0.98.tsv", show_col_types = FALSE) %>%
  rename(genome = 1) %>%
  pivot_longer(!genome, names_to = "data", values_to = "counts") %>%
  mutate(sample = str_extract(data, "M\\d+")) %>%
  group_by(genome, sample) %>%
  summarise(counts = sum(counts), .groups = "drop") %>%
  pivot_wider(names_from = "sample", values_from = "counts")
# Covered bases
macroquant_covered_bases_98 <- read_tsv("data/data_preparation/genome.covered_bases.REF0015-g_mg_hybrid-0.98.tsv", show_col_types = FALSE) %>%
  rename(genome = 1) %>%
  pivot_longer(!genome, names_to = "data", values_to = "counts") %>%
  mutate(sample = str_extract(data, "M\\d+")) %>%
  group_by(genome, sample) %>%
  summarise(counts = sum(counts), .groups = "drop") %>%
  pivot_wider(names_from = "sample", values_from = "counts")
# Taxonomy
macromag_taxonomy_98 <- read_tsv("https://sid.erda.dk/share_redirect/BTwCSjaWz2/REF0015-g_mg_hybrid/gtdbtk.summary.tsv.gz", show_col_types = FALSE) %>%
  rename(genome = user_genome) %>%
  mutate(genome = str_replace_all(genome, "\\.fa", "")) %>%
  separate(classification, c("domain", "phylum", "class", "order", "family", "genus", "species"), sep = ";") %>%
  select(genome, domain, phylum, class, order, family, genus, species) %>%
  arrange(match(genome, macroquant_read_counts_98$genome)) %>%
  filter(genome %in% macroquant_covered_bases_98$genome) %>%
  select(genome, domain, phylum, class, order, family, genus, species) %>%
  mutate(
    domain = sub("^d__", "", domain), # Remove prefix
    phylum = sub("^p__", "", phylum),
    class = sub("^c__", "", class),
    order = sub("^o__", "", order),
    family = sub("^f__", "", family),
    genus = sub("^g__", "", genus),
    species = sub("^s__", "", species)
  ) %>%
  mutate(
    species = if_else(
      is.na(species) | species == "",
      paste0("unclassified_", genus),
      species
    )
  ) %>%
  mutate(
    species = if_else(
      is.na(species) | species == "unclassified_",
      paste0("unclassified_", family),
      species
    )
  ) %>%
  mutate(
    genus = if_else(
      is.na(genus) | genus == "",
      paste0("unclassified_", family),
      genus
    )
  )

# Genome quality
macromag_quality_98 <- read_tsv("https://sid.erda.dk/share_redirect/BTwCSjaWz2/REF0015-g_mg_hybrid/checkm2.quality_report.tsv.gz", show_col_types = FALSE) %>%
  rename(genome = 1) %>%
  mutate(genome = str_replace_all(genome, "\\.fa", "")) %>%
  arrange(match(genome, macroquant_read_counts_98$genome)) %>%
  select(genome, Completeness, Contamination, Coding_Density, Genome_Size, Contig_N50) %>%
  mutate(circul = ifelse(Genome_Size == Contig_N50, "Y", "N")) %>%
  rename(completeness = Completeness, contamination = Contamination, coding_density = Coding_Density, length = Genome_Size) %>%
  filter(genome %in% macroquant_covered_bases_98$genome)

# Genome Metadata
macromag_genomemetadata_98 <- macromag_taxonomy_98 %>%
  left_join(macromag_quality_98, by = join_by(genome == genome)) # join quality
```

##### Filter and Normalise
```{r filter_98ANI, eval=FALSE}
macroquant_coverage_98 <- macroquant_covered_bases_98 %>%
  mutate(across(where(is.numeric), ~ . / macromag_genomemetadata_98$length))
min_coverage_98 <- 0.3
macroquant_read_counts_filt_30_98 <- macroquant_coverage_98 %>%
  mutate(across(where(is.numeric), ~ ifelse(. > min_coverage, 1, 0))) %>%
  mutate(across(-1, ~ . * macroquant_read_counts_98[[cur_column()]]))
readlength_98 <- 150 # change if sequencing read length is different
macroquant_counts_filt_30_98 <- macroquant_read_counts_filt_30_98 %>%
  mutate(across(where(is.numeric), ~ . / (macromag_genomemetadata_98$length / readlength)))
macroquant_genome_counts_98 <- macroquant_read_counts_98 %>%
  mutate(across(where(is.numeric), ~ . / (macromag_genomemetadata_98$length / readlength)))
```


## Macro-scale Metagenomics (MACRO)
### Merge Genome Count to MSSM
```{r genome_count_combined, eval=FALSE}
macro_genome_count_macro_filt_30 <- read_tsv("data/data_preparation/conventional_REF0015-g_mg_hybrid.0.95_genome_count_filt.tsv") %>%
  filter(rowSums(select(., -genome)) != 0)

if (all(macro_genome_count_macro_filt_30$genome %in% macromag_genomemetadata$genome)) {
  mssmmacroquant_vs_macro_genome_count_30 <- full_join(macroquant_genome_counts_filt_30 %>% select(genome, mssm_sample_metadata$microsample), macro_genome_count_macro_filt_30, by = "genome") %>%
    select(where(~ !all(. == 0))) %>%
    filter(if_any(where(is.numeric), ~ . != 0)) %>%
    column_to_rownames("genome") %>%
    replace(is.na(.), 0)
} else {
  message("Not all genomes in MACRO are present in MSSM")
}

```

### Merge Metadata to MSSM
```{r sample_metadata_combined, eval=FALSE}
mssmmacroquant_vs_macro_sample_metadata <- macroquant_metrics_filt_30 %>%
  select(microsample, section, batch, cryosection, buffer, animal, cycles, type_binomial, 
         total_before_filtering.total_reads, total_after_filtering.total_reads, richness) %>%
  filter(microsample %in% names(mssmmacroquant_vs_macro_genome_count_30)) %>%
  bind_rows(
    tibble(
      microsample = "D300487",
      animal = "G103b",
      total_before_filtering.total_reads = 21962564*2,
      total_after_filtering.total_reads = 21962564*2,
      section = "right caecum digesta",
      type_binomial = "positive",
      batch = NA, cryosection = NA, buffer = NA  # Fill these with appropriate values or NA
    ),
    tibble(
      microsample = "D300499",
      animal = "G121e",
      total_before_filtering.total_reads = 24855916*2,
      total_after_filtering.total_reads = 24855916*2,
      section = "right caecum digesta",
      type_binomial = "positive",
      batch = NA, cryosection = NA, buffer = NA  # Same here
    )
  ) %>%
  mutate(group = ifelse(section == "right caecum digesta", "Macro-scale Metagenomics", "Micro-scale Spatial Metagenomics")) %>% 
  column_to_rownames("microsample")
```

### Combined Phyloseq Object
```{r phyloseq, eval=FALSE}
mssmmacroquant_vs_macro_taxonomy <- macromag_taxonomy %>% filter(genome %in% c(rownames(mssmmacroquant_vs_macro_genome_count_30))) %>% column_to_rownames("genome")

physeq_all <- phyloseq(
  mssmmacroquant_vs_macro_genome_count_30 %>% otu_table(., taxa_are_rows = TRUE),
  mssmmacroquant_vs_macro_sample_metadata %>% sample_data(),
  mssmmacroquant_vs_macro_taxonomy %>% as.matrix() %>% tax_table()
)
physeq_all  <- prune_taxa(taxa_sums(physeq_all) > 0, physeq_all)

physeq_G121e <- subset_samples(physeq_all, animal %in% c("G121e"))
physeq_G121e  <- prune_taxa(taxa_sums(physeq_G121e) > 0, physeq_G121e)
physeq_G103b <- subset_samples(physeq_all, animal %in% c("G103b"))
physeq_G103b  <- prune_taxa(taxa_sums(physeq_G103b) > 0, physeq_G103b)
physeq_genus <- tax_glom(physeq_all, "genus")
physeq_genus_relabund <- transform_sample_counts(physeq_genus, function(x) x / sum(x))


```


### Amalia: Mock Community (BAC)
### Amalia: Fluorescence In Situ hybridization (FISH)

### Confocal microscopy of DAPI-stained microsections

```{r DAP_quantification_df, eval=FALSE}
dapi_df <- read_csv("data/data_preparation/data_DAPI_bact_quant_selected.csv", show_col_types = FALSE) %>%
  separate(Organ_source, into = c("Age_1", "Section"), sep = " ") %>% 
  group_by(Cryosection) %>% # Group by the column(s) containing duplicates
  mutate(Sequence = row_number()) %>%
  ungroup()  %>%
  mutate(
    Age = ifelse(Age_1 == "old", "35-day-old Animal", "14-day-old Animal"),
    Animal = str_sub(Cryosection, 1, 6),
    merge_variable=paste(Animal, Sequence, sep = "_"),
    #Distance_from_one_side=ifelse(Distance_from_one_side == 0, 1, Distance_from_one_side)
  ) %>%
  filter(
    case_when(
      Age == "14-day-old Animal" ~ Distance_from_one_side < 2500,  
      Age == "35-day-old Animal" ~ Distance_from_one_side < 4000,  
      TRUE ~ TRUE))  %>% 
  mutate(Section = recode(Section,
      "caecum" = "Caecum",
      "colon" = "Colon",
      "ileum" = "Ileum"
    ))
```

## Fragment Analyzer

```{r frag_anal_df, eval=FALSE}
sn3_5_dreamprep_df <- read_csv("data/data_preparation/CSV_2vs1plate_2FAplates.csv",
  col_types = cols_only(
    "Sample ID" = col_character(),
    "Range" = col_character(),
    "ng/uL" = col_double(),
    "% Total" = col_double(),
    "nmole/L" = col_double(),
    "Avg. Size" = col_double(),
    "Plate" = col_character(),
    "Type" = col_character(),
    "Concentration" = col_character()
  )
) %>%
  rename(
    sample = "Sample ID",
    plate = "Plate", type = "Type", concentration = "Concentration",
    ng_uL = "ng/uL", nmole_L = "nmole/L",
    percentage_total = "% Total",
    size = "Avg. Size"
  ) %>%
  mutate(Range = factor(Range, levels = c("35 bp to 100 bp", "100 bp to 200 bp", "200 bp to 1000 bp")))

```

## SNP level microdiversity (Lorikeet)

```{r lorikeet_df, eval=FALSE}
ani_files <- list.files("data/data_preparation/", pattern = "_population_ani.tsv$", full.names = TRUE)
# Return a named list of cleaned ANI matrices
ani_list <- purrr::map(ani_files, function(file_path) {
  # Extract bin ID
  bin_id <- stringr::str_extract(basename(file_path), "GPB_bin_\\d+")

  # Extract sample labels
  labels <- tibble(raw = readLines(file_path)) %>%
    filter(str_starts(raw, "##sample")) %>%
    mutate(
      ID = str_extract(raw, "(?<=ID=)\\d+") %>% as.integer(),
      SAMPLE = str_extract(raw, "M30[0-9]{4}")
    ) %>%
    select(ID, SAMPLE)

  # Read and clean ANI matrix
  ani <- readr::read_tsv(file_path, comment = "#", show_col_types = FALSE) %>%
    left_join(labels, by = c("SampleID" = "ID")) %>%
    select(-SampleID) %>%
    column_to_rownames(var = "SAMPLE") %>%
    set_names(labels$SAMPLE) %>%
    select(where(~ !all(is.na(.)))) %>%
    select(where(~ !all(is.nan(.)))) %>% 
    filter(!if_all(everything(), is.nan))

  return(ani)
})
# Set bin IDs as list names
names(ani_list) <- stringr::str_extract(basename(ani_files), "GPB_bin_\\d+")

```


## Wrap working objects
In the last step, the objects that are needed for downstream analyses are stored in an R object.

```{r wrap_objects, eval=FALSE}
#save.image(file = "data/data_preparation/25_06_27data.Rdata")
#save.image(file = "data/data_preparation/25_07_01data.Rdata")
save.image(file = "data/data_preparation/25_07_11_data.Rdata")
save(
  mssm_sample_metadata,
  macromag_genomemetadata,
  micromag_genomemetadata,
  macromag_taxonomy,
  micromag_taxonomy,
  macromag_gifts,
  macromag_tree,
  micromag_tree,
  macroquant_metrics_filt_30,
  microquant_metrics_filt_30,
  macroquant_genome_counts_filt_30,
  microquant_genome_counts_filt_30,
  macroquant_read_counts_filt_30,
  microquant_read_counts_filt_30,
  physeq_all,
  physeq_genus,
  physeq_genus_relabund,
  dapi_df, 
  sn3_5_dreamprep_df,
  macroquant_counts_filt_30_98,
  macromag_genomemetadata_98,
  ani_list,
  GIFT_db_df,
  file = "data/data_preparation/25_07_11_selected_data.Rdata"
)
```

