# Loading and Tidying Data
**Note:** This step requires the raw data files available on Zenodo (see links below) and requires setting eval = TRUE in the code chunks below.
If you prefer to skip this step, you can continue to the next sections by ensuring that the R object containing the necessary data is loaded from resources/working_data_object.Rdata.

## Load data from Zenodo
```{r data_zenodo, warning = FALSE, comments = "", message = FALSE, eval = FALSE}
# 3D'omics data source - Micro-scale spatial metagenomics
dir.create("resources/zenodo", recursive = TRUE, showWarnings = FALSE)
url <- "https://zenodo.org/api/records/17091770/files-archive"
download.file(url, "resources/zenodo.zip", mode = "wb")
unzip("resources/zenodo.zip", exdir = "resources/zenodo", junkpaths = TRUE)
file.remove("resources/zenodo.zip")

# Poultry salmonella experiment (micro-scale)
dir.create("resources/zenodo/microscale_catalogue", recursive = TRUE, showWarnings = FALSE)
urls <- c(
  "https://zenodo.org/records/17091749/files/checkm2.quality_report.tsv.gz?download=1",
  "https://zenodo.org/records/17091749/files/gtdbtk.backbone.bac120.classify.tree?download=1",
  "https://zenodo.org/records/17091749/files/dram.annotations.tsv.gz?download=1",
  "https://zenodo.org/records/17091749/files/gtdbtk.summary.tsv.gz?download=1"
)
filenames <- basename(sub("\\?.*", "", urls))
destfiles <- file.path("resources/zenodo/microscale_catalogue", filenames)
mapply(download.file, urls, destfiles, MoreArgs = list(mode = "wb"), method = "curl") 

# 3D'omics reference genome catalogue - Poultry salmonella experiment
dir.create("resources/zenodo/macroscale_catalogue", recursive = TRUE, showWarnings = FALSE)
urls <- c(
  "https://zenodo.org/records/17091731/files/checkm2.quality_report.tsv.gz?download=1",
  "https://zenodo.org/records/17091731/files/dram.annotations.tsv.gz?download=1",
  "https://zenodo.org/records/17091731/files/gtdbtk.backbone.bac120.classify.tree?download=1",
  "https://zenodo.org/records/17091731/files/gtdbtk.summary.tsv.gz?download=1"
)
filenames <- basename(sub("\\?.*", "", urls))
destfiles <- file.path("resources/zenodo/macroscale_catalogue", filenames)
mapply(download.file, urls, destfiles, MoreArgs = list(mode = "wb"), method = "curl") 
```

## Import Sample metadata
```{r sample_metadata_mssm, warning = FALSE, comments = "", message = FALSE, eval = FALSE}
mssm_sample_metadata <- read_tsv("resources/zenodo/mssm_sample_metadata.tsv", show_col_types = FALSE)
```

## Sequencing Data Quality 
### Read Quality: Pre- vs Post-Filtering (fastp)
**Note:** The parse_fastp() function in index needs to be loaded.
```{r read_quality, warning = FALSE, comments = "", message = FALSE, eval = FALSE}
mssm_fastp <- read_tsv("resources/zenodo/mssm_fastp.tsv", show_col_types = FALSE)
```


### Contamination Assessment: Host, Human, and Environmental Reads
```{r mapping_eukaryots_references, warning = FALSE, comments = "", message = FALSE, eval = FALSE}
mssm_euk_mapping <- read_tsv("resources/zenodo/mssm_euk_mapping.tsv", show_col_types = FALSE)  
```

### Genome-Centric Mapping Statistics: Macroscale MAG Catalogue
```{r mapping_macro_catalogue, warning = FALSE, comments = "", message = FALSE, eval = FALSE}
mssm_mag_mapping<- read_tsv("resources/zenodo/mssm_mag_mapping.tsv", show_col_types = FALSE)  
```

### Read-Centric Mapping Statistics: Kraken (genomes) and SingleM (single-copy marker genes)
```{r mapping_singlem, warning = FALSE, comments = "", message = FALSE, eval = FALSE}
mssm_singlem <- read_tsv("resources/zenodo/mssm_singlem.tsv", show_col_types = FALSE)  
```

```{r mapping_kraken, warning = FALSE, comments = "", message = FALSE, eval = FALSE}
mssm_kraken <- read_tsv("resources/zenodo/mssm_kraken.tsv", show_col_types = FALSE)  
```

### Redundancy & Coverage Estimation: Nonpareil
```{r estimates_nonpareil, warning = FALSE, comments = "", message = FALSE, eval = FALSE}
mssm_nonpareil <- read_tsv("resources/zenodo/mssm_nonpareil.tsv", show_col_types = FALSE)  
```

### Combined Quality Statistics
```{r data_quality_df, warning = FALSE, comments = "", message = FALSE, eval = FALSE}
mssm_quality <- reduce(
  list(
    mssm_sample_metadata,
    mssm_fastp,
    mssm_euk_mapping,
    mssm_mag_mapping,
    mssm_singlem,
    mssm_kraken,
    mssm_nonpareil
  ),
  full_join,
  by = "microsample"
)  %>%
  mutate(
    lowquality_reads_pct = ((total_before_filtering.total_reads - total_after_filtering.total_reads) / total_before_filtering.total_reads),
    adapter_contamination_pct = (total_adapter_trimmed_bases / total_before_filtering.total_bases),
    host_pct = (reads_mapped_chicken / total_before_filtering.total_reads),
    human_pct = (reads_mapped_human / total_before_filtering.total_reads),
    pig_pct = (reads_mapped_pig / total_before_filtering.total_reads),
    prokreads_pct = (raw_total_sequences_mag / total_before_filtering.total_reads),
    mapmag_rawreads_pct = (reads_mapped_mag / total_before_filtering.total_reads),
    mapmag_filteredreads_pct = (reads_mapped_mag / total_after_filtering.total_reads),
    mapmag_prokreads_pct = (reads_mapped_mag / raw_total_sequences_mag)
  ) 
```

## Input Preparation For Metagenomic Analysis

### Abundance Estimation Using Macro-scale MAGs Catalogue 95 ANI

#### Import Genome Counts & Coverage

```{r mag_readcount_macroquant, warning = FALSE, comments = "", message = FALSE, eval = FALSE}
macroquant_read_counts <- read_tsv("resources/zenodo/macroquant_read_counts.tsv", show_col_types = FALSE)  
```

```{r mag_coveredbases_macroquant, warning = FALSE, comments = "", message = FALSE, eval = FALSE}
macroquant_covered_bases <- read_tsv("resources/zenodo/macroquant_covered_bases.tsv", show_col_types = FALSE)  
```

#### Macro-scale MAGs Catalogue

##### Taxonomy {-}
This is the raw taxonomy table generated by GTDBtk, which is simplified for downstream analyses.
```{r mag_taxonomy_macro, warning = FALSE, comments = "", message = FALSE, eval = FALSE}
macroscale_catalogue_taxonomy <- read_tsv("resources/zenodo/macroscale_catalogue/gtdbtk.summary.tsv.gz", show_col_types = FALSE) %>%
  rename(genome = user_genome) %>%
  mutate(genome = str_replace_all(genome, "\\.fa", "")) %>%
  separate(classification, c("domain", "phylum", "class", "order", "family", "genus", "species"), sep = ";") %>%
  select(genome, domain, phylum, class, order, family, genus, species) %>%
  mutate(
    domain = sub("^d__", "", domain), # Remove prefix
    phylum = sub("^p__", "", phylum),
    class = sub("^c__", "", class),
    order = sub("^o__", "", order),
    family = sub("^f__", "", family),
    genus = sub("^g__", "", genus),
    species = sub("^s__", "", species)
  ) %>% 
  mutate(
    species = if_else(
      is.na(species) | species == "",
      paste0("unclassified_", genus),
      species
    )
  ) %>%
  mutate(
    species = if_else(
      is.na(species) | species == "unclassified_",
      paste0("unclassified_", family),
      species
    )
  ) %>%
  mutate(
    genus = if_else(
      is.na(genus) | genus == "",
      paste0("unclassified_", family),
      genus
    )
  )

macromag_taxonomy <- macroscale_catalogue_taxonomy %>% 
  arrange(match(genome, macroquant_read_counts$genome)) %>%
  filter(genome %in% macroquant_read_counts$genome) 
```

##### Quality {-}
```{r mag_quality_macro, warning = FALSE, comments = "", message = FALSE, eval = FALSE}
macroscale_catalogue_quality <- read_tsv("resources/zenodo/macroscale_catalogue/checkm2.quality_report.tsv.gz", show_col_types = FALSE) %>%
  # rename first column
  rename(genome = 1) %>%
  # remove .fa from the end of each genome name
  mutate(genome = str_replace_all(genome, "\\.fa", "")) %>%
  # select relevant columns and rename
  select(genome, Completeness, Contamination, Coding_Density, Genome_Size, Contig_N50) %>%
  rename(
    completeness = Completeness,
    contamination = Contamination,
    coding_density = Coding_Density,
    length = Genome_Size,
    N50_length = Contig_N50
  ) %>%
  # estimate genome circularity (N50 should be equal to genome length)
  mutate(circularity = N50_length / length) %>%
  mutate(circularity = ifelse(circularity >= 1, "1", "0")) %>%
  # update completeness and contamination for genomes with circularity of 1
  mutate(
    completeness = ifelse(circularity == "1", 100, completeness),
    contamination = ifelse(circularity == "1", 0, contamination)
  ) %>%
  # estimate ratio of: genome length / median bacterial genome length
  mutate(length_ratio = length / median(length, na.rm = TRUE))

macromag_genomemetadata <- macromag_taxonomy %>%
  left_join(macroscale_catalogue_quality, by = join_by(genome == genome)) %>% # join quality
  arrange(match(genome, macroquant_read_counts$genome)) %>%
  filter(genome %in% macroquant_read_counts$genome)

```

##### Functional Annotation {-}

```{r mag_annotation_macro, warning = FALSE, comments = "", message = FALSE, eval = FALSE}
macromag_annotations <- read_tsv("resources/zenodo/macroscale_catalogue/dram.annotations.tsv.gz", show_col_types = FALSE) %>%
  rename(gene = 1, genome = 2) %>%
  filter(genome %in% macroquant_read_counts$genome)
macromag_gifts <- distill(macromag_annotations, GIFT_db, genomecol = 2, annotcol = c(9, 10, 19))
GIFT_db_df <- GIFT_db
```

##### Phylogenetic Tree {-}

```{r mag_tree_macro, warning = FALSE, comments = "", message = FALSE, eval = FALSE}
# upload from zenodo
macromag_tree <- read.tree("resources/zenodo/macroscale_catalogue/gtdbtk.backbone.bac120.classify.tree")
macromag_tree$tip.label <- str_replace_all(macromag_tree$tip.label, "'", "") # remove single quotes in MAG names
macromag_tree <- keep.tip(macromag_tree, tip = macroquant_read_counts$genome) # keep only MAG tips
```

#### Coverage Filtering & Count Normalization

##### Coverage filtering {-}
**Note:** Genomes that have less than 30% of their length covered by reads are turned into zeros to account for the random allocation of reads across genomes due to mapping heuristics. 
```{r coverage_fitering_macroquant, warning = FALSE, comments = "", message = FALSE, eval = FALSE}
macroquant_coverage <- macroquant_covered_bases %>% # calculate coverage values by dividing the number of covered bases by the length of each genome
  mutate(across(where(is.numeric), ~ . / macromag_genomemetadata$length))
min_coverage <- 0.3
macroquant_read_counts_filt_30 <- macroquant_coverage %>%
  mutate(across(where(is.numeric), ~ ifelse(. > min_coverage, 1, 0))) %>%
  mutate(across(-1, ~ . * macroquant_read_counts[[cur_column()]]))
```

##### Generate genome count table {-}
After filtering the low-coverage reads, read counts are transformed into genome counts using genome-length and read-length information.
```{r genome_counts_macroquant, warning = FALSE, comments = "", message = FALSE, eval = FALSE}
readlength <- 150 # change if sequencing read length is different
macroquant_genome_counts_filt_30 <- macroquant_read_counts_filt_30 %>%
  mutate(across(where(is.numeric), ~ . / (macromag_genomemetadata$length / readlength))) 
macroquant_genome_counts  <- macroquant_read_counts %>%
  mutate(across(where(is.numeric), ~ . / (macromag_genomemetadata$length / readlength))) 
```

#### Alpha Diversity Metrics
**Note:** The calculate_alpha_diversity() function needs to be loaded.
```{r alpha_diversities_macroquant, warning = FALSE, comments = "", message = FALSE, eval = FALSE}
# Calculate alpha diversity for filtered genome counts data
macroquant_metrics_filt_30 <- calculate_alpha_diversity(
  input_data = macroquant_genome_counts_filt_30,
  metadata_name = mssm_quality,
  dataset_name = "filt_30",
  tree_name = macromag_tree
) 
``` 


### Abundance Estimation Using Micro-scale MAGs Catalogue

#### Import Genome Counts & Coverage

```{r mag_readcount_microquant, warning = FALSE, comments = "", message = FALSE, eval = FALSE}
microquant_read_counts <- read_tsv("resources/zenodo/microquant_read_counts.tsv", show_col_types = FALSE)  
```

```{r mag_coveredbases_microquant, warning = FALSE, comments = "", message = FALSE, eval = FALSE}
microquant_covered_bases <- read_tsv("resources/zenodo/microquant_covered_bases.tsv", show_col_types = FALSE)  
```

#### Micro-scale MAGs Catalogue
##### Taxonomy {-}
This is the raw taxonomy table generated by GTDBtk, which is simplified for downstream analyses.
```{r mag_taxonomy_micro, warning = FALSE, comments = "", message = FALSE, eval = FALSE}
micromag_taxonomy <- read_tsv("resources/zenodo/microscale_catalogue/gtdbtk.summary.tsv.gz", show_col_types = FALSE) %>%
  rename(genome = user_genome) %>%
  mutate(genome = str_replace_all(genome, "\\.fa", "")) %>%
  separate(classification, c("domain", "phylum", "class", "order", "family", "genus", "species"), sep = ";") %>%
  select(genome, domain, phylum, class, order, family, genus, species) %>%
    mutate(
      domain = sub("^d__", "", domain), # Remove prefix
      phylum = sub("^p__", "", phylum), 
      class = sub("^c__", "", class), 
      order = sub("^o__", "", order), 
      family = sub("^f__", "", family), 
      genus = sub("^g__", "", genus), 
      species = sub("^s__", "", species)
    )  %>% 
  arrange(match(genome, microquant_read_counts$genome)) %>% 
  filter(genome %in% microquant_read_counts$genome) %>%
  mutate(
    species = if_else(
      is.na(species) | species == "",
      paste0("unclassified_", genus),
      species
    )
  ) %>%
  mutate(
    species = if_else(
      is.na(species) | species == "unclassified_",
      paste0("unclassified_", family),
      species
    )
  ) %>%
  mutate(
    genus = if_else(
      is.na(genus) | genus == "",
      paste0("unclassified_", family),
      genus
    )
  ) 
```

##### Quality {-}
```{r mag_quality_micro, warning = FALSE, comments = "", message = FALSE, eval = FALSE}
micromag_quality <- read_tsv("resources/zenodo/microscale_catalogue/checkm2.quality_report.tsv.gz", show_col_types = FALSE) %>%
  # rename first column
  rename(genome = 1) %>% 
  # remove .fa from the end of each genome name
  mutate(genome = str_replace_all(genome, "\\.fa", "")) %>%
  # select relevant columns and rename
  select(genome, Completeness, Contamination, Coding_Density, Genome_Size, Contig_N50) %>% 
  rename(completeness = Completeness, 
         contamination = Contamination, 
         coding_density = Coding_Density, 
         length = Genome_Size,
         N50_length = Contig_N50) %>%
  # estimate genome circularity (N50 should be equal to genome length)
  mutate(circularity = N50_length/length) %>%
  mutate(circularity = ifelse(circularity >= 1, "1", "0")) %>%
  # update completeness and contamination for genomes with circularity of 1
  mutate(completeness = ifelse(circularity == "1", 100, completeness),
         contamination = ifelse(circularity == "1", 0, contamination)) %>%   
  # estimate ratio of: genome length / median bacterial genome length
  mutate(length_ratio = length/median(length, na.rm = TRUE))

micromag_genomemetadata <- micromag_taxonomy %>%
  left_join(micromag_quality, by = join_by(genome == genome)) %>% # join quality
  arrange(match(genome, microquant_read_counts$genome)) %>% 
  filter(genome %in% microquant_read_counts$genome)
```

##### Phylogenetic Tree {-}

```{r mag_tree_micro, warning = FALSE, comments = "", message = FALSE, eval = FALSE}
# upload from zenodo
micromag_tree <- read.tree("resources/zenodo/microscale_catalogue/gtdbtk.backbone.bac120.classify.tree")
micromag_tree$tip.label <- str_replace_all(micromag_tree$tip.label, "'", "") # remove single quotes in MAG names
micromag_tree <- keep.tip(micromag_tree, tip = microquant_read_counts$genome) # keep only MAG tips
```

#### Coverage Filtering & Count Normalization

##### Coverage filtering {-}
**Note:** Genomes that have less than 30% of their length covered by reads are turned into zeros to account for the random allocation of reads across genomes due to mapping heuristics. 
```{r coverage_fitering_microquant, warning = FALSE, comments = "", message = FALSE, eval = FALSE}
microquant_coverage<- microquant_covered_bases %>% # calculate coverage values by dividing the number of covered bases by the length of each genome
  mutate(across(where(is.numeric), ~ . / micromag_genomemetadata$length))
min_coverage <- 0.3
microquant_read_counts_filt_30 <- microquant_coverage %>%
  mutate(across(where(is.numeric), ~ ifelse(. > min_coverage, 1, 0))) %>%
  mutate(across(-1, ~ . * microquant_read_counts[[cur_column()]]))
```

##### Generate genome count table {-}
After filtering the low-coverage reads, read counts are transformed into genome counts using genome-length and read-length information.
```{r genome_counts_microquant, warning = FALSE, comments = "", message = FALSE, eval = FALSE}
readlength <- 150 # change if sequencing read length is different
microquant_genome_counts_filt_30 <- microquant_read_counts_filt_30 %>%
  mutate(across(where(is.numeric), ~ . / (micromag_genomemetadata$length / readlength)))
microquant_genome_counts <- microquant_read_counts %>%
  mutate(across(where(is.numeric), ~ . / (micromag_genomemetadata$length / readlength)))
```

#### Alpha Diversity Metrics
**Note:** The `calculate_alpha_diversity()` function needs to be loaded.

```{r alpha_diversities_microquant, warning = FALSE, comments = "", message = FALSE, eval = FALSE}
# Calculate alpha diversity for filtered genome counts data
microquant_metrics_filt_30 <- calculate_alpha_diversity(
  input_data = microquant_read_counts_filt_30,
  metadata_name = mssm_quality,
  dataset_name = "filt_30",
  tree_name = micromag_tree
) 
``` 

### Abundance Estimation Using Macro-scale MAGs Catalogue 98 ANI

##### Mapping and Catalogue Data
```{r load_98ANI, warning = FALSE, comments = "", message = FALSE, eval = FALSE}
# Read count
macroquant_read_counts_98 <- read_tsv("resources/zenodo/macroquant_read_counts_98.tsv", show_col_types = FALSE)  
# Covered bases
macroquant_covered_bases_98 <- read_tsv("resources/zenodo/macroquant_covered_bases_98.tsv", show_col_types = FALSE)  
# Taxonomy
macromag_taxonomy_98 <- macroscale_catalogue_taxonomy %>%
  arrange(match(genome, macroquant_read_counts_98$genome)) %>%
  filter(genome %in% macroquant_covered_bases_98$genome) 
# Genome Metadata
macromag_genomemetadata_98 <- macromag_taxonomy_98 %>%
  left_join(macroscale_catalogue_quality, by = join_by(genome == genome)) %>% # join quality
  arrange(match(genome, macroquant_read_counts_98$genome)) %>% 
  filter(genome %in% macroquant_read_counts_98$genome)
```

##### Filter and Normalise
```{r filter_98ANI, warning = FALSE, comments = "", message = FALSE, eval = FALSE}
macroquant_coverage_98 <- macroquant_covered_bases_98 %>%
  mutate(across(where(is.numeric), ~ . / macromag_genomemetadata_98$length))
min_coverage_98 <- 0.3
macroquant_read_counts_filt_30_98 <- macroquant_coverage_98 %>%
  mutate(across(where(is.numeric), ~ ifelse(. > min_coverage, 1, 0))) %>%
  mutate(across(-1, ~ . * macroquant_read_counts_98[[cur_column()]]))
readlength_98 <- 150 # change if sequencing read length is different
macroquant_counts_filt_30_98 <- macroquant_read_counts_filt_30_98 %>%
  mutate(across(where(is.numeric), ~ . / (macromag_genomemetadata_98$length / readlength)))
macroquant_genome_counts_98 <- macroquant_read_counts_98 %>%
  mutate(across(where(is.numeric), ~ . / (macromag_genomemetadata_98$length / readlength)))
```


## Macro-scale Metagenomics (MACRO)
### Merge Genome Count to MSSM
```{r genome_count_combined, warning = FALSE, comments = "", message = FALSE, eval = FALSE}
macro_genome_count_macro_filt_30 <- read_tsv("resources/zenodo/macro_genome_count_macro_filt_30.tsv", show_col_types = FALSE)  
if (all(macro_genome_count_macro_filt_30$genome %in% macromag_genomemetadata$genome)) {
  mssmmacroquant_vs_macro_genome_count_30 <- full_join(macroquant_genome_counts_filt_30 %>% select(genome, mssm_sample_metadata$microsample), macro_genome_count_macro_filt_30, by = "genome") %>%
    select(where(~ !all(. == 0))) %>%
    filter(if_any(where(is.numeric), ~ . != 0)) %>%
    column_to_rownames("genome") %>%
    replace(is.na(.), 0)
} else {
  message("Not all genomes in MACRO are present in MSSM")
}
```

### Merge Metadata to MSSM
```{r sample_metadata_combined, warning = FALSE, comments = "", message = FALSE, eval = FALSE}
mssmmacroquant_vs_macro_sample_metadata <- macroquant_metrics_filt_30 %>%
  select(microsample, section, batch, cryosection, buffer, animal, cycles, type, 
         total_before_filtering.total_reads, total_after_filtering.total_reads, richness) %>%
  filter(microsample %in% names(mssmmacroquant_vs_macro_genome_count_30)) %>%
  bind_rows(
    tibble(
      microsample = "D300487",
      animal = "G103b",
      total_before_filtering.total_reads = 21962564*2,
      total_after_filtering.total_reads = 21962564*2,
      section = "right caecum digesta",
      type_binomial = "positive",
      batch = NA, cryosection = NA, buffer = NA  # Fill these with appropriate values or NA
    ),
    tibble(
      microsample = "D300499",
      animal = "G121e",
      total_before_filtering.total_reads = 24855916*2,
      total_after_filtering.total_reads = 24855916*2,
      section = "right caecum digesta",
      type_binomial = "positive",
      batch = NA, cryosection = NA, buffer = NA  # Same here
    )
  ) %>%
  mutate(group = ifelse(section == "right caecum digesta", "Macro-scale Metagenomics", "Micro-scale Spatial Metagenomics")) %>%
  column_to_rownames("microsample")
```

### Combined Phyloseq Object
```{r phyloseq, warning = FALSE, comments = "", message = FALSE, eval = FALSE}
mssmmacroquant_vs_macro_taxonomy <- macromag_taxonomy %>% filter(genome %in% c(rownames(mssmmacroquant_vs_macro_genome_count_30))) %>% column_to_rownames("genome")

physeq_all <- phyloseq(
  mssmmacroquant_vs_macro_genome_count_30 %>% otu_table(., taxa_are_rows = TRUE),
  mssmmacroquant_vs_macro_sample_metadata %>% sample_data(),
  mssmmacroquant_vs_macro_taxonomy %>% as.matrix() %>% tax_table()
)
physeq_all  <- prune_taxa(taxa_sums(physeq_all) > 0, physeq_all)

physeq_G121e <- subset_samples(physeq_all, animal %in% c("G121e"))
physeq_G121e  <- prune_taxa(taxa_sums(physeq_G121e) > 0, physeq_G121e)
physeq_G103b <- subset_samples(physeq_all, animal %in% c("G103b"))
physeq_G103b  <- prune_taxa(taxa_sums(physeq_G103b) > 0, physeq_G103b)
physeq_genus <- tax_glom(physeq_all, "genus")
physeq_genus_relabund <- transform_sample_counts(physeq_genus, function(x) x / sum(x))


```


## Mock Community (BAC)

### Import Sample Metadata
```{r sample_metadata_mock, warning=FALSE, comments="", message=FALSE, eval = FALSE}
sample_metadata_mock <- read_tsv("resources/zenodo/sample_metadata_mock.tsv", show_col_types = FALSE)
```

### Import Genome Metadata
```{r genome_metadata_mock, warning=FALSE, comments="", message=FALSE, eval = FALSE}
genome_metadata_mock <- tibble(genome=c("SR","XR","MO","PA"),
                       species=c("Stenotrophomonas rhizophila","Xanthomonas retroflexus","Microbacterium oxydans","Paenibacillus amylolyticus"),
                       length=c(4216903,4655568,3989887,7273724),
                       gram=c("negative","negative","positive","positive"))
```

### Import Genome Counts & Coverage

```{r load_count_mock, message=FALSE, warning = FALSE, comments = "", eval = FALSE}
read_counts_mock <- read_tsv("resources/zenodo/read_counts_mock.tsv", show_col_types = FALSE)
```

```{r load_hits_mock, message=FALSE, warning = FALSE, comments = "", eval = FALSE}
genome_covered_bases_mock<- read_tsv("resources/zenodo/genome_covered_bases_mock.tsv", show_col_types = FALSE)
```

## Fluorescence In Situ hybridization (FISH)
#### Confocal microscopy of DAPI-stained microsections
```{r DAP_quantification_df, warning = FALSE, comments = "", message = FALSE, eval = FALSE}
dapi_df <- read_tsv("resources/zenodo/dapi_df.tsv", show_col_types = FALSE)  
```

### Abundance Estimation using FISH


## Fragment Analyzer

```{r frag_anal_df, warning = FALSE, comments = "", message = FALSE, eval = FALSE}
sn3_5_dreamprep_df <- read_tsv("resources/zenodo/sn3_5_dreamprep_df.tsv", show_col_types = FALSE)  
```

## SNP level microdiversity (Lorikeet)

```{r lorikeet_df, warning = FALSE, comments = "", message = FALSE, eval = FALSE}
ani_files <- list.files("resources/zenodo/", pattern = "_population_ani.tsv$", full.names = TRUE)

ani_list <- purrr::map(ani_files, function(file_path) {
  bin_id <- stringr::str_extract(basename(file_path), "GPB_bin_\\d+")

  labels <- tibble(raw = readLines(file_path)) %>%
    filter(str_starts(raw, "##sample")) %>%
    mutate(
      ID = str_extract(raw, "(?<=ID=)\\d+") %>% as.integer(),
      SAMPLE = str_extract(raw, "M30[0-9]{4}")
    ) %>%
    select(ID, SAMPLE)

  ani <- readr::read_tsv(file_path, comment = "#", show_col_types = FALSE) %>%
    left_join(labels, by = c("SampleID" = "ID")) %>%
    select(-SampleID) %>%
    column_to_rownames(var = "SAMPLE") %>%
    set_names(labels$SAMPLE) %>%
    select(where(~ !all(is.na(.)))) %>%
    select(where(~ !all(is.nan(.)))) %>% 
    filter(!if_all(everything(), is.nan))

  return(ani)
})
# Set bin IDs as list names
names(ani_list) <- stringr::str_extract(basename(ani_files), "GPB_bin_\\d+")
```


## Wrap working objects
In the last step, the objects that are needed for downstream analyses are stored in an R object.
```{r wrap_objects, warning = FALSE, comments = "", message = FALSE, eval = FALSE}
save.image(file = "resources/working_data_object.Rdata")
```


