# MSSM Method Development

## Load data
```{r load_data_main_manuscript, warning = FALSE, comments = "", message = FALSE}
load("data/25_07_11_selected_data.Rdata") # load the data from 1-data_preparation
```

## Microbial Genomic Reference Data

### Macro-scale MAGs Catalogue Summary Statistics 
```{r stat_macromag, warning = FALSE, comments = "", message = FALSE}
cat("Completeness median = ", round(median(macromag_genomemetadata$completeness), 2), "\n")
cat(
  "Completeness IQR = ", round(quantile(macromag_genomemetadata$completeness, probs = c(0.25)), 2),
  "-", round(quantile(macromag_genomemetadata$completeness, probs = c(0.75)), 2), "\n"
)
cat("Contamination median = ", round(median(macromag_genomemetadata$contamination), 2), "\n")
cat(
  "Contamination IQR = ", round(quantile(macromag_genomemetadata$contamination, probs = c(0.25)), 2),
  "-", round(quantile(macromag_genomemetadata$contamination, probs = c(0.75)), 2)
)
```

### Micro-scale MAGs Catalogue Summary Statistics 
```{r stat_micromag, warning = FALSE, comments = "", message = FALSE}
cat("Completeness median = ", round(median(micromag_genomemetadata$completeness), 2), "\n")
cat(
  "Completeness IQR = ", round(quantile(micromag_genomemetadata$completeness, probs = c(0.25)), 2),
  "-", round(quantile(micromag_genomemetadata$completeness, probs = c(0.75)), 2), "\n"
)
cat("Contamination median = ", round(median(micromag_genomemetadata$contamination), 2), "\n")
cat(
  "Contamination IQR = ", round(quantile(micromag_genomemetadata$contamination, probs = c(0.25)), 2),
  "-", round(quantile(micromag_genomemetadata$contamination, probs = c(0.75)), 2)
)
```
### Amalia: Macro vs Micro MAGs: Completeness and Abundance (Fig. 2a)

```{r fig_2a, warning = FALSE, comments = "", message = FALSE}
```

### Amalia: Macro MAGs Overview (Fig. 2b)

```{r fig_2b, warning = FALSE, comments = "", message = FALSE}
```

### Amalia: Micro MAGs Overview (Suppl. Fig. 1)

```{r suppl_fig_1, warning = FALSE, comments = "", message = FALSE}
```

### Comparative Analysis of Genome Catalogues (Suppl. Note 1)

```{r suppl_note_1, warning = FALSE, comments = "", message = FALSE}
cat("Number Genomes in Macro-scale MAGs Catalogue =", length(macromag_taxonomy$genome), "\n")
cat("Number Genomes in Micro-scale MAGs Catalogue =", length(micromag_taxonomy$genome), "\n")
```

#### Data selection

```{r suppl_note_1_samples, warning = FALSE, comments = "", message = FALSE}
sn1_sample_id <- microquant_metrics_filt_30 %>% # select microquant dataset
  filter(
    batch %in% c("MSEB0011"),
    type_binomial %in% c("positive")
  ) %>% pull(microsample)

sn1_phylo_count <- full_join(
  microquant_genome_counts_filt_30 %>%
    select(genome, all_of(sn1_sample_id)) %>%
    rename_with(~ paste0(.x, "_micro"), .cols = -genome),
  macroquant_genome_counts_filt_30 %>%
    select(genome, all_of(sn1_sample_id)) %>%
    rename_with(~ paste0(.x, "_macro"), .cols = -genome),
  by = "genome"
) %>%
  mutate(across(
    where(is.numeric),
    ~ replace_na(.x, 0)
  )) %>%
  filter(if_any(where(is.numeric), ~ . != 0)) %>%
  select(
    genome,
    where(~ !is.numeric(.x) || sum(.x, na.rm = TRUE) != 0)
  ) %>%
  column_to_rownames("genome")


sn1_phylo_metadata <- bind_rows(
  microquant_metrics_filt_30 %>%
    mutate(
      sample_id = microsample,
      microsample = paste0(microsample, "_micro"),
      dataset = "Micro-scale"
    ),
  macroquant_metrics_filt_30 %>%
    mutate(
      sample_id = microsample,
      microsample = paste0(microsample, "_macro"),
      dataset = "Macro-scale"
    )
) %>%
  filter(microsample %in% c(names(sn1_phylo_count))) %>%
  column_to_rownames("microsample")

sn1_phylo_taxonomy <- bind_rows(micromag_taxonomy, macromag_taxonomy) %>%
  filter(genome %in% c(rownames(sn1_phylo_count))) %>%
  column_to_rownames("genome")

sn1_physeq <- phyloseq(
  sn1_phylo_count %>% otu_table(., taxa_are_rows = TRUE),
  sn1_phylo_metadata %>% sample_data(),
  sn1_phylo_taxonomy %>% as.matrix() %>% tax_table()
)

sn1_physeq_species <- tax_glom(sn1_physeq, taxrank = "species")
if (nrow(data.frame(sn1_physeq_species@tax_table) %>%
  rownames_to_column("genome") %>%
  count(species) %>%
  filter(n > 1)) > 0) {
  cat("warning: Some species appear more than once in the taxonomy table")
} else {
  cat("All species are unique")
}

sn1_physeq_G121e <- subset_samples(sn1_physeq_species, animal == "G121e")
sn1_physeq_G121e <- prune_taxa(taxa_sums(sn1_physeq_G121e) > 0, sn1_physeq_G121e)
```

#### Read Counts 
```{r SN1_read_count, warning = FALSE, comments = "", message = FALSE}
sn1_read_df <- bind_rows(
  lapply(list(
    micro = microquant_read_counts_filt_30 %>%
      select(genome, all_of(sn1_sample_id)),
    macro = macroquant_read_counts_filt_30 %>%
      select(genome, all_of(sn1_sample_id))
  ), function(df) {
    df %>%
      summarise(across(-1, sum)) %>%
      pivot_longer(cols = everything(), names_to = "microsample", values_to = "number_reads")
  }),
  .id = "source"
) %>% pivot_wider(names_from = source, values_from = number_reads)

# Run Spearman's correlation test
sn1_cor_result <- cor.test(sn1_read_df$micro, sn1_read_df$macro, method = "spearman")
# Extract values
rho <- sn1_cor_result$estimate
pval <- sn1_cor_result$p.value
n <- length(sn1_read_df$micro)
# Report
cat(sprintf("Spearman's ρ = %.2f, p = %.3f, n = %d", rho, pval, n), "\n")
cat("Macro-scale genome catalogue mean read counts =", mean(sn1_read_df$macro) / 1000000, "\n")
cat("Micro-scale genome catalogue mean read counts =", mean(sn1_read_df$micro) / 1000000, "\n")


sn1_read_count <- sn1_read_df %>%
  ggplot(aes(x = macro, y = micro)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  stat_cor(
    method = "spearman", label.x = Inf, label.y = -Inf,
    hjust = 1.1, vjust = -1.1, size = 4
  ) + # adds the correlation
  custom_ggplot_theme +
  labs(
    x = "Mapped Reads \nMacro-scale Catalogue",
    y = "Mapped Reads \nMicro-scale Catalogue"
  )
# sn1_read_count
```

#### Alpha diversity (Fig. S1.1a)

```{r sn1_alpha_df, warning = FALSE, comments = "", message = FALSE}
sn1_alpha_df <- sample_data(sn1_physeq_species) %>%
  data.frame() %>%
  filter(richness > 0) %>%
  select(dataset, sample_id, richness, neutral, phylogenetic) %>%
  rename(`MAGs Catalogue` = dataset, microsample = sample_id)

sn1_alpha_df_wd <- sn1_alpha_df %>%
  pivot_wider(
    names_from = `MAGs Catalogue`,
    names_glue = "{`MAGs Catalogue`}_{.value}",
    values_from = c(richness, neutral, phylogenetic)
  )

sn1_alpha_df_lg <- sn1_alpha_df %>%
  pivot_longer(
    cols = c(richness, neutral, phylogenetic),
    names_to = "metric",
    values_to = "value"
  ) %>%
  mutate(metric = factor(metric,
    levels = c("richness", "neutral", "phylogenetic")
  ))
```


```{r FigS1_1a, warning = FALSE, comments = "", message = FALSE}
figs1_1a <- sn1_alpha_df_lg %>%
  filter(
    !metric %in% c("phylogenetic"),
    (metric == "neutral" & value > 1) |
      (metric == "richness" & value > 0)
  ) %>%
  mutate(metric = fct_recode(metric,
    "Richness" = "richness",
    "Neutral" = "neutral"
  )) %>%
  ggplot(aes(x = metric, y = value, fill = `MAGs Catalogue`)) +
  geom_boxplot() +
  # geom_violin(trim = TRUE, alpha = 0.7, position = position_dodge(0.8), na.rm = TRUE) +  # Violin plot
  geom_jitter(
    position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.8),
    size = 1.5, alpha = 0.3
  ) +
  theme_minimal() +
  scale_fill_manual(values = c("#79b1a3", "#f58262")) +
  custom_ggplot_theme +
  labs(
    # title = "a",
    x = "Alpha Diversity Metric",
    y = ""
  ) +
  theme(legend.position = "bottom")
figs1_1a
# ggsave(filename = "figures/FigS1_1a.pdf", figs1_1a, width = 12, height = 5)
# ggsave(filename = "figures/FigS1_1a.png", figs1_1a, width = 12, height = 5)
```

```{r paired_Wilcoxon_richness, warning = FALSE, comments = "", message = FALSE}
wilcox.test(sn1_alpha_df_wd$`Micro-scale_richness`, sn1_alpha_df_wd$`Macro-scale_richness`, paired = TRUE, na.action = na.omit) %>%
  tidy() # 5 nas?
```

```{r paired_Wilcoxon_neutral, warning = FALSE, comments = "", message = FALSE}
wilcox.test(sn1_alpha_df_wd$`Micro-scale_neutral`, sn1_alpha_df_wd$`Macro-scale_neutral`, paired = TRUE, na.action = na.omit) %>%
  tidy() # 5 nas?
```

#### Detected Taxa
```{r sn1_alpha_taxa, warning = FALSE, comments = "", message = FALSE}
microscale_taxa <- left_join(
  microquant_genome_counts_filt_30 %>%
    select(genome, all_of(sn1_sample_id)) %>%
    filter(rowSums(across(where(is.numeric))) > 0) %>% select(genome) %>% as.data.frame(),
  micromag_taxonomy,
  by = "genome"
)

macroscale_taxa <- left_join(
  macroquant_genome_counts_filt_30 %>%
    select(genome, all_of(sn1_sample_id)) %>%
    filter(rowSums(across(where(is.numeric))) > 0) %>% select(genome) %>% as.data.frame(),
  macromag_taxonomy,
  by = "genome"
)
```


```{r sn1_alpha_unique_species, warning = FALSE, comments = "", message = FALSE}
micro_unique_species <- setdiff(microscale_taxa$species, macroscale_taxa$species)
macro_unique_species <- setdiff(macroscale_taxa$species, microscale_taxa$species)
sn1_unique_species_df <- tibble(
  species = c(micro_unique_species, macro_unique_species),
  source = c(
    rep("Micro-scale", length(micro_unique_species)),
    rep("Macro-scale", length(macro_unique_species))
  )
)
cat("Number Unique Species in Micro-scale MAGs Catalogue =", length(micro_unique_species), "\n")
cat("Number Unique Species in Macro-scale MAGs Catalogue =", length(macro_unique_species), "\n")
```


```{r sn1_alpha_unique_genera,warning = FALSE, comments = "", message = FALSE}
micro_unique_genera <- setdiff(microscale_taxa$genus, macroscale_taxa$genus)
macro_unique_genera <- setdiff(macroscale_taxa$genus, microscale_taxa$genus)
sn1_unique_genera_df <- tibble(
  genera = c(micro_unique_genera, macro_unique_genera),
  source = c(
    rep("Micro-scale", length(micro_unique_genera)),
    rep("Macro-scale", length(macro_unique_genera))
  )
)
cat("Number Unique Genera in Micro-scale MAGs Catalogue =", length(micro_unique_genera), "\n")
cat("Number Unique Genera in Macro-scale MAGs Catalogue =", length(macro_unique_genera), "\n")
```

#### Beta diversity (Fig. S1.1b, Fig. S1.1c)
```{r sn1_beta_dfs, warning = FALSE, comments = "", message = FALSE}
sn1_betacount_df <- data.frame(otu_table(sn1_physeq_species)) %>%
  rownames_to_column("genome") %>%
  melt(id = "genome") %>%
  setNames(c("genome", "microsample", "count")) %>%
  left_join(data.frame(tax_table(sn1_physeq_species)) %>%
    rownames_to_column("genome"), by = "genome") %>%
  select(species, microsample, count) %>%
  filter(count > 0) %>%
  pivot_wider(names_from = species, values_from = count, values_fill = 0) %>% # change here
  select(where(~ any(. != 0))) %>%
  mutate(across(-microsample, ~ . / rowSums(across(where(is.numeric))) * 100)) %>%
  column_to_rownames(var = "microsample")
```

```{r FigS1_1b, warning = FALSE, comments = "", message = FALSE}
sn1_beta_obj <- perform_pca(sn1_betacount_df)
sn1_beta_metadata <- data.frame(sample_data(sn1_physeq_species)) %>%
  rownames_to_column("microsample") %>%
  filter(microsample %in% c(rownames(sn1_beta_obj$df_clr))) %>%
  rename(
    `MAGs Catalogue` = dataset,
    Animal = animal
  )
sn1_scores <- rownames_to_column(as.data.frame(sn1_beta_obj$pca_result$x), var = "microsample")
sn1_scores <- left_join(sn1_scores, sn1_beta_metadata, by = join_by(microsample == microsample))
# Calculate variance explained for each principal component
variance_explained <- (sn1_beta_obj$pca_result$sdev^2) / sum(sn1_beta_obj$pca_result$sdev^2) * 100
# Plot the first two principal components
figs1_1b <- ggplot() +
  geom_point(
    data = sn1_scores,
    aes(x = PC1, y = PC2, group = microsample, color = `MAGs Catalogue`, shape = Animal),
    size = 3, alpha = 0.5
  ) +
  scale_color_manual(values = c("#79b1a3", "#f58262")) +
  # Connect the points by group (sample_id)
  geom_path(
    data = sn1_scores,
    aes(x = PC1, y = PC2, group = sample_id),
    linewidth = 0.7, alpha = 0.8, color = "lightgrey"
  ) +
  labs(
    # title = "b",
    x = paste0("PC1: ", round(variance_explained[1], 2), "% variance explained"),
    y = paste0("PC2: ", round(variance_explained[2], 2), "% variance explained")
  ) +
  scale_x_continuous(limits = c(-max(abs(sn1_scores$PC1)), max(abs(sn1_scores$PC1)))) + # Calculate symmetric limits based on data
  scale_y_continuous(limits = c(-max(abs(sn1_scores$PC2)), max(abs(sn1_scores$PC2)))) + # Calculate symmetric limits based on data
  geom_hline(yintercept = 0, color = "darkgrey") + # Horizontal line at y = 0
  geom_vline(xintercept = 0, color = "darkgrey") + # Vertical line at x = 0
  custom_ggplot_theme +
  theme(legend.position = "bottom")
figs1_1b
# ggsave(filename = "figures/FigS1_1b.pdf", figs1_1b, width = 12, height = 9)
# ggsave(filename = "figures/FigS1_1b.png", figs1_1b, width = 12, height = 9)
data.frame(otu_table(tax_glom(sn1_physeq_G121e, taxrank = "species"))) %>% 
  rownames_to_column("genome")  %>% 
  pivot_longer(cols = -c(genome), names_to = "microsample", values_to = "abundance") %>% 
  mutate(dataset = str_sub(microsample, -5, -1)) %>% 
  group_by(dataset, genome) %>% 
  summarise(sum_abundance=sum(abundance)) %>% 
  filter(sum_abundance==0) %>% 
  left_join(data.frame(tax_table(tax_glom(sn1_physeq_G121e, taxrank = "species"))) %>% 
  rownames_to_column("genome"), by = "genome" ) %>% arrange(species)
```


```{r FigS1_1c_aldex, warning = FALSE, comments = "", message = FALSE}
structural_zero <- data.frame(otu_table(tax_glom(sn1_physeq_G121e, taxrank = "species"))) %>% 
  rownames_to_column("genome")  %>% 
  pivot_longer(cols = -c(genome), names_to = "microsample", values_to = "abundance") %>% 
  mutate(dataset = str_sub(microsample, -5, -1)) %>% 
  group_by(dataset, genome) %>% 
  summarise(sum_abundance=sum(abundance)) %>% 
  filter(sum_abundance==0) %>% pull(genome)
  
  
sn1_aldex_G121e_counts <- data.frame(otu_table(tax_glom(sn1_physeq_G121e, "species"))) %>% 
  rownames_to_column("genome") %>% 
  #filter(!genome %in%c(structural_zero)) %>% 
  column_to_rownames("genome")
sn1_aldex_G121e_pseudo_counts <- round(sn1_aldex_G121e_counts * 1e6)
sn1_aldex_G121e_conditions <- data.frame(sample_data(tax_glom(sn1_physeq_G121e, "species"))) %>% pull(dataset)
# Perform the ALDEx2 differential abundance test
# Use 't' test for two groups; 'glm' for multi-class or covariates
sn1_aldex_G121e_results <- aldex(sn1_aldex_G121e_pseudo_counts,
  sn1_aldex_G121e_conditions,
  test = "t", 
  effect = TRUE, # include effect sizes
  denom = "all", # normalization denominator
  mc.samples = 256, # number of Monte Carlo instances
  verbose = TRUE
)
head(sn1_aldex_G121e_results)
# - we.ep: expected p-value from Welch's t-test
# - wi.ep: expected p-value from Wilcoxon test
# - we.eBH: BH-adjusted p-value (q-value)
# - effect: effect size (difference in clr means / variability)
# Filter
sn1_aldex_G121e_sig <- sn1_aldex_G121e_results[sn1_aldex_G121e_results$we.eBH < 0.05, ]
sn1_aldex_G121e_effect <- sn1_aldex_G121e_sig[abs(sn1_aldex_G121e_sig$effect) > 0.5, ]

sn1_aldex <-
  sn1_aldex_G121e_effect %>%
  rownames_to_column("genome") %>%
  left_join(data.frame(tax_table(tax_glom(sn1_physeq_G121e, "species"))) %>% rownames_to_column("genome"), by = "genome")

# Create horizontal barplot
figs1_1c <- sn1_aldex %>%
  mutate(structural_zero=ifelse(genome %in% c(structural_zero), "Y", "N")) %>% 
  ggplot(., aes(x = diff.btw, y = reorder(species, effect), fill = abs(effect) > 1)) +
  geom_bar(stat = "identity", show.legend = FALSE) +
  coord_flip() +
  labs(
    x = "Median difference in \nCLR values between groups",
    y = "Species"
  ) +
  scale_fill_manual(values = c("TRUE" = "#e35d51", "FALSE" = "darkgrey")) +
  custom_ggplot_theme + facet_wrap(~structural_zero, scale="free") + 
  theme(axis.text.x = element_text(angle = 40, hjust = 1))
figs1_1c
```

#### Fig. S1
```{r FigS1, warning = FALSE, comments = "", message = FALSE}
figs1 <- cowplot::plot_grid(figs1_1a, figs1_1b,
  ncol = 1, labels = c("a", "b"),
  rel_heights = c(1, 1), label_size = 16
)
ggsave("figures/FigS1.png", figs1, width = 10, height = 11, dpi = 300)
```


```{r ancombc2_G121e, message=FALSE, eval=FALSE}
set.seed(1234) # set seed for reproducibility
ancombc_output_sn1_physeq_G121e <- ancombc2(
  data = sn1_physeq_G121e,
  assay_name = "counts",
  tax_level = "species", # change to agglomerate analysis to a higher taxonomic range
  fix_formula = "dataset",
  p_adj_method = "holm",
  pseudo_sens = TRUE,
  prv_cut = 0,
  lib_cut = 0,
  s0_perc = 0.05,
  group = "dataset",
  struc_zero = TRUE,
  neg_lb = FALSE,
  alpha = 0.05,
  n_cl = 2,
  verbose = TRUE,
  global = FALSE,
  pairwise = FALSE,
  dunnet = FALSE,
  trend = FALSE,
  iter_control = list(tol = 1e-5, max_iter = 20, verbose = FALSE),
  em_control = list(tol = 1e-5, max_iter = 100),
  mdfdr_control = list(fwer_ctrl_method = "holm", B = 100),
  trend_control = NULL
)

names(ancombc_output_sn1_physeq_G121e$res)
any(ancombc_output_sn1_physeq_G121e$res$`passed_ss_datasetMicro-scale` == TRUE &
  ancombc_output_sn1_physeq_G121e$res$`q_datasetMicro-scale` < 0.05)

ancom <- ancombc_output_sn1_physeq_G121e$res %>%
  select(
    genus = taxon,
    lfc = `lfc_datasetMicro-scale`,
    qval = `q_datasetMicro-scale`,
    passed_ss = `passed_ss_datasetMicro-scale`,
    diff = `diff_datasetMicro-scale`
  ) %>%
  filter(qval < 0.05) %>%
  arrange(qval, lfc) %>%
  mutate(
    hjust_label = ifelse(lfc > 0, -0.05, 1.05),
    q_label = paste0("q=", signif(round(qval, 2), 2)),
    q_label = ifelse(q_label == "q=0", "q<0.01", q_label)
  ) %>%
  ggplot(aes(
    x = lfc,
    y = fct_reorder(genus, lfc),
    fill = passed_ss
  )) +
  geom_col() +
  geom_text(
    aes(label = q_label, hjust = hjust_label),
    size = 3.2,
    color = "black"
  ) +
  scale_fill_manual(values = c("darkgrey", "#e35d51")) +
  geom_hline(yintercept = 0) +
  custom_ggplot_theme +
  labs(
    #title = "c",
    x = "log2FoldChange",
    y = "Genus"
  ) +
  guides(fill = guide_legend(title = "Eligible for Differential \nAbundance Testing")) +
  theme(legend.position = "bottom") +
  coord_cartesian(xlim = c(-1.2, 1.2))


```

```{r ancom, warning = FALSE, comments = "", message = FALSE}
G121e_df <- data.frame(otu_table(tax_glom(sn1_physeq_G121e, taxrank = "species"))) %>%
  rownames_to_column("genome") %>%
  melt(id = "genome") %>%
  setNames(c("genome", "microsample", "count")) %>%
  left_join(data.frame(tax_table(tax_glom(sn1_physeq_G121e, taxrank = "species"))) %>%
    rownames_to_column("genome"), by = "genome") %>%
  select(species, microsample, count) %>%
  filter(count > 0) %>%
  pivot_wider(names_from = species, values_from = count, values_fill = 0) %>% # change here
  select(where(~ any(. != 0))) %>%
  mutate(across(-microsample, ~ . / rowSums(across(where(is.numeric))) * 100)) %>%
  column_to_rownames(var = "microsample")
G121e_df_obj <- perform_pca(G121e_df)
G121e_df_metadata <- data.frame(sample_data(tax_glom(sn1_physeq_G121e, taxrank = "species"))) %>%
  rownames_to_column("microsample") %>%
  filter(microsample %in% c(rownames(G121e_df_obj$df_clr))) %>%
  rename(
    `MAGs Catalogue` = dataset,
    Animal = animal
  )


ancom_species <- ancombc_output_sn1_physeq_G121e$res %>%
  select(
    genus = taxon,
    lfc = `lfc_datasetMicro-scale`,
    qval = `q_datasetMicro-scale`,
    passed_ss = `passed_ss_datasetMicro-scale`,
    diff = `diff_datasetMicro-scale`
  ) %>%
  filter(qval < 0.05, passed_ss == TRUE) %>% pull(genus)
  
  
G121e_df_obj$df_clr %>% rownames_to_column("microsample") %>% 
  pivot_longer(cols=(-microsample), names_to = "species", values_to = "clr") %>%  
  left_join(data.frame(tax_table(tax_glom(sn1_physeq_G121e, taxrank = "species"))) %>% rownames_to_column("genome"), by = "species") %>% 
  #left_join(data.frame(structural_zero) %>% mutate(genome=structural_zero), by= "genome") %>% 
  mutate(dataset = str_sub(microsample, -5, -1),
         structural_zero=ifelse(genome %in% c(structural_zero), "Y", "N"),
         ancom=ifelse(genome %in% c(ancom_species), "Y", "N")) %>% 
  #filter(structural_zero=="N") %>% 
  arrange(dataset, clr) %>% 
  ggplot(aes(x = species, y = clr, fill=dataset)) +
  geom_boxplot() +
  geom_jitter(
    position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.8),
    size = 1.5, alpha = 0.3
  ) +
  theme_minimal() +
  scale_fill_manual(values = c("#79b1a3", "#f58262")) +
  custom_ggplot_theme +
  labs(
    title = "",
    x = "",
    y = ""
  ) +
  theme(legend.position = "none") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_grid(structural_zero~ancom, scale="free") #+ coord_flip() #+theme(axis.text.y = element_blank())
```

#### Supplementary Table 1
```{r sup_table_1_comparison, warning = FALSE, comments = "", message = FALSE}
total_microsamples <- length(unique(mivsma_phylo_lg %>% filter(
  group %in% c("Micro-scale Spatial Metagenomics"),
  animal %in% c("G121e"),
  !section %in% c("Colon", "Ileum")
) %>% pull(microsample)))

mivsma_phylo_lg %>%
  filter(
    group %in% c("Micro-scale Spatial Metagenomics"),
    animal %in% c("G121e"),
    !section %in% c("Colon", "Ileum"),
    genus %in% mivsma_35d_micro_unique
  ) %>%
  group_by(genus) %>%
  summarise(
    # Prevalence = sum(abundance > 0),
    Prevalence_Total = round(sum(abundance > 0) / total_microsamples, 3),
    mean_abundance = round(mean(abundance), 3)
  )
```

## FISH probes design?

```{r x1, warning = FALSE, comments = "", message = FALSE}
```



### Tissue lysis and sequencing library preparation (Suppl. Note 3)

#### Impact of Lysis Conditions on Microbial Taxonomic Profiling (Suppl. Note 3.2)

##### Amalia: Mock Community (Fig. S3.2.1)

```{r x2, warning = FALSE, comments = "", message = FALSE}
```

##### Microsamples (Fig. S3.2.2)

###### Data selection

```{r sn3_2_df, warning = FALSE, comments = "", message = FALSE}
sn3_2_df <- macroquant_metrics_filt_30 %>%
  filter(
    batch %in% c("MSEB0006", "MSEB0010"), # Caeucum (MSEB0006) and Colon (MSEB0010)
    section %in% c("Caecum right", "Colon"),
    cycles %in% c(13, 14, 15),
    type_binomial %in% c("positive"),
    richness > 0
  ) %>%
  mutate(
    collection_attempts = as.factor(collection_attempts),
    buffer = as.factor(buffer)
  )

sn3_2_df %>%
  group_by(section, buffer) %>%
  summarise(n_samples = n())
sn3_2_df %>%
  group_by(section, buffer, collection_attempts) %>%
  summarise(n_samples = n(), .groups = "keep") %>%
  pivot_wider(names_from = collection_attempts, values_from = n_samples)
```

###### Visual Comparison of Sequencing Performance (Fig. S3.2.2)

```{r FigS3_2_2, warning = FALSE, comments = "", message = FALSE}
ordered_metrics <- c(
  "total_after_filtering.total_reads",
  "mapmag_prokreads_pct",
  "percent_duplication.rate",
  "richness",
  "neutral",
  "phylogenetic" # ,
  # "read_fraction_singlem"
)

figs3_2_2 <- sn3_2_df %>%
  mutate(mapmag_prokreads_pct=mapmag_prokreads_pct*100,
         percent_duplication.rate=percent_duplication.rate*100) %>% 
  pivot_longer(
    cols = -c(names(mssm_sample_metadata), "cov_filtering"),
    names_to = "metric",
    values_to = "value"
  ) %>%
  filter(metric %in% ordered_metrics) %>%
  mutate(
    metric = factor(metric, levels = ordered_metrics),
    metric = fct_recode(metric,
      "a. Sequencing Yield \nPost-Quality Filtering" = "total_after_filtering.total_reads",
      "b. Reads Mapping \nMAG Catalogue (%)" = "mapmag_prokreads_pct",
      "c. Reads \nDuplication Rate (%)" = "percent_duplication.rate",
      "d. Richness" = "richness",
      "e. Neutral" = "neutral",
      "f. Phylogenetic" = "phylogenetic"
    ),
    section = recode(section,
      "Caecum right" = "Caecum",
      "Colon" = "Colon"
    )
  ) %>%
  ggplot(aes(x = section, y = value, fill = buffer)) +
  geom_boxplot() +
  facet_wrap(~metric, scales = "free_y", ncol = 3) +
  labs(
    fill = "Lysis",
    y = "",
    x = "Intestinal section"
  ) +
  # geom_violin(trim = TRUE, alpha = 0.7, position = position_dodge(0.8), na.rm = TRUE) +  # Violin plot
  geom_jitter(
    position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.8),
    size = 1.5, alpha = 0.3
  ) +
  scale_fill_manual(values = c("#79b1a3", "#f58262", "black")) +
  custom_ggplot_theme +
  theme(legend.position = "bottom")

figs3_2_2
# ggsave(filename = "figures/FigS3_2_2.pdf", figs3_2_2, width = 8, height = 7)
# ggsave(filename = "figures/FigS3_2_2.png", figs3_2_2, width = 8, height = 7)
```

###### Statistical Comparison of Sequencing Performance
**Note:** The fit_and_analyze_model() function needs to be loaded.

###### a. Reads Retained Post Quality Filtering
```{r sn3_2_a_caecum, warning = FALSE, comments = "", message = FALSE}
# Counts (integers ≥0): poisson, quasipoisson (accounts overdispersion)
sn3_2_a_ca <- fit_and_analyze_model(
  model = "glm",
  distribution = "quasipoisson", # DHARMa significant deviation
  response_var = "total_after_filtering.total_reads",
  explanatory_var = "collection_attempts + buffer",
  data = sn3_2_df %>% filter(section %in% c("Caecum right"))
)
sn3_2_a_ca$anova
# md <- glm(total_after_filtering.total_reads~collection_attempts + buffer, family = "quasipoisson", data = sn3_2_df #%>%  filter(section %in% c("Caecum right")))
# broom::tidy(Anova(md, test.statistic = "Wald")) # ANOVA with F test
```

```{r sn3_2_a_colon, warning = FALSE, comments = "", message = FALSE}
# Counts (integers ≥0): poisson, quasipoisson (accounts overdispersion)
sn3_2_a_co <- fit_and_analyze_model(
  model = "glm",
  distribution = "quasipoisson", # DHARMa significant deviation
  response_var = "total_after_filtering.total_reads",
  explanatory_var = "collection_attempts + buffer",
  data = sn3_2_df %>% filter(section %in% c("Colon"))
)
sn3_2_a_co$anova
```

###### b. Reads Mapping MAG Catalogue (%)

```{r sn3_2_b_caecum, warning = FALSE, comments = "", message = FALSE}
# Proportions (0–1 continuous):	quasibinomial GLM (accounts overdispersion)
sn3_2_b_ca <- fit_and_analyze_model(
  model = "glm",
  distribution = "quasibinomial", # DHARMa significant deviation
  response_var = "mapmag_prokreads_pct",
  explanatory_var = "collection_attempts + buffer",
  data = sn3_2_df %>% filter(section %in% c("Caecum right"))
)
sn3_2_b_ca$anova
```


```{r sn3_2_b_colon, warning = FALSE, comments = "", message = FALSE}
# Proportions (0–1 continuous):	quasibinomial GLM (accounts overdispersion)
sn3_2_b_co <- fit_and_analyze_model(
  model = "glm",
  distribution = "quasibinomial", # DHARMa significant deviation
  response_var = "mapmag_prokreads_pct",
  explanatory_var = "collection_attempts + buffer",
  data = sn3_2_df %>% filter(section %in% c("Colon"))
)
sn3_2_b_co$anova
```

###### c. Reads Duplication Rate (%)

```{r sn3_2_c_caecum, warning = FALSE, comments = "", message = FALSE}
# Proportions (0–1 continuous):	quasibinomial GLM (accounts overdispersion)
sn3_2_c_ca <- fit_and_analyze_model(
  model = "glm",
  distribution = "quasibinomial", # DHARMa significant deviation
  response_var = "percent_duplication.rate",
  explanatory_var = "collection_attempts + buffer",
  data = sn3_2_df %>% filter(section %in% c("Caecum right"))
)
sn3_2_c_ca$anova
```

```{r sn3_2_c_colon, warning = FALSE, comments = "", message = FALSE}
# Proportions (0–1 continuous):	quasibinomial GLM (accounts overdispersion)
sn3_2_c_co <- fit_and_analyze_model(
  model = "glm",
  distribution = "quasibinomial", # DHARMa significant deviation
  response_var = "percent_duplication.rate",
  explanatory_var = "collection_attempts + buffer",
  data = sn3_2_df %>% filter(section %in% c("Colon"))
)
sn3_2_c_co$anova
```

###### d. Richness
```{r sn3_2_d_caecum, warning = FALSE, comments = "", message = FALSE}
# Counts (integers ≥0): poisson, quasipoisson (accounts overdispersion)
sn3_2_d_ca <- fit_and_analyze_model(
  model = "glm",
  distribution = "quasipoisson", # DHARMa significant deviation
  response_var = "richness",
  explanatory_var = "collection_attempts + buffer",
  data = sn3_2_df %>% filter(section %in% c("Caecum right"))
)
sn3_2_d_ca$anova
# md <- glm(total_after_filtering.total_reads~collection_attempts + buffer, family = "quasipoisson", data = sn3_2_df #%>%  filter(section %in% c("Caecum right")))
# broom::tidy(Anova(md, test.statistic = "Wald")) # ANOVA with F test
```


```{r sn3_2_d_colon, warning = FALSE, comments = "", message = FALSE}
# Counts (integers ≥0): poisson, quasipoisson (accounts overdispersion)
sn3_2_d_co <- fit_and_analyze_model(
  model = "glm",
  distribution = "quasipoisson", # DHARMa significant deviation
  response_var = "richness",
  explanatory_var = "collection_attempts + buffer",
  data = sn3_2_df %>% filter(section %in% c("Colon"))
)
sn3_2_d_co$anova
```


###### e. Neutral" = "neutral",
```{r sn3_2_e_caecum, warning = FALSE, comments = "", message = FALSE}
# # Continuous floats (any real number): lm or gaussian, Gamma (continuous positive data, skewed to the right)
sn3_2_e_ca <- fit_and_analyze_model(
  model = "glm",
  distribution = "gaussian", # DHARMa significant deviation KS
  response_var = "neutral",
  explanatory_var = "collection_attempts + buffer",
  data = sn3_2_df %>% filter(section %in% c("Caecum right"))
)
# plot(sn3_2_e_ca$simResidual)
sn3_2_e_ca$anova
```

```{r sn3_2_e_colon, warning = FALSE, comments = "", message = FALSE}
# Counts (integers ≥0): poisson, quasipoisson (accounts overdispersion)
sn3_2_e_co <- fit_and_analyze_model(
  model = "glm",
  distribution = "gaussian",
  response_var = "neutral",
  explanatory_var = "collection_attempts + buffer",
  data = sn3_2_df %>% filter(section %in% c("Colon"))
)
# plot(sn3_2_e_co$simResidual)
sn3_2_e_co$anova
```

###### f. Phylogenetic" = "phylogenetic"
```{r sn3_2_f_caecum, warning = FALSE, comments = "", message = FALSE}
# # Continuous floats (any real number): lm or gaussian, Gamma (continuous positive data, skewed to the right)
sn3_2_f_ca <- fit_and_analyze_model(
  model = "glm",
  distribution = "gaussian", # DHARMa significant deviation KS
  response_var = "phylogenetic",
  explanatory_var = "collection_attempts + buffer",
  data = sn3_2_df %>% filter(section %in% c("Caecum right"))
)
plot(sn3_2_f_ca$simResidual)
sn3_2_f_ca$anova
```

```{r sn3_2_f_colon, warning = FALSE, comments = "", message = FALSE}
# Counts (integers ≥0): poisson, quasipoisson (accounts overdispersion)
sn3_2_f_co <- fit_and_analyze_model(
  model = "glm",
  distribution = "gaussian",
  response_var = "phylogenetic",
  explanatory_var = "collection_attempts + buffer",
  data = sn3_2_df %>% filter(section %in% c("Colon"))
)
plot(sn3_2_f_co$simResidual)
sn3_2_f_co$anova
```      

###### Detected Taxa

```{r sn3_2_taxa, warning = FALSE, comments = "", message = FALSE}
sn3_2_taxa <- left_join(
  macroquant_genome_counts_filt_30 %>%
    select(genome, all_of(sn3_2_df$microsample)) %>%
    filter(rowSums(across(where(is.numeric))) > 0) %>%
    melt(id = "genome", variable.name = "microsample"),
  macromag_taxonomy,
  by = "genome"
) %>%
  left_join(macroquant_metrics_filt_30, by = "microsample") %>%
  filter(
    value > 0,
    section %in% c("Colon")
  )


sn3_2_unique_L06 <- setdiff(
  sn3_2_taxa %>% filter(buffer %in% c("L06")) %>% distinct(genus),
  sn3_2_taxa %>% filter(buffer %in% c("L07")) %>% distinct(genus)
)
sn3_2_unique_L07 <- setdiff(
  sn3_2_taxa %>% filter(buffer %in% c("L07")) %>% distinct(genus),
  sn3_2_taxa %>% filter(buffer %in% c("L06")) %>% distinct(genus)
)

sn3_2_unique_genus_df <- tibble(
  genus = c(sn3_2_unique_L06$genus, sn3_2_unique_L07$genus),
  source = c(
    rep("L06", length(sn3_2_unique_L06$genus)),
    rep("L07", length(sn3_2_unique_L07$genus))
  )
)

cat("Number Unique Genus in L06 =", length(sn3_2_unique_L06$genus), "\n")
cat("Number Unique Genus in L07 =", length(sn3_2_unique_L07$genus), "\n")
(length(sn3_2_unique_L07$genus) / sn3_2_taxa %>%
  distinct(genus) %>%
  count()) * 100
```      

#### Performance Evaluation Across Intestinal Sections (Suppl. Note 3.3)

###### Data selection 

```{r sn3_3_df, warning = FALSE, comments = "", message = FALSE}
sn3_3_df <- macroquant_metrics_filt_30 %>%
  filter(
    batch %in% c("MSEB0006", "MSEB0010"),
    section %in% c("Caecum right", "Ileum", "Colon"),
    # cycles %in% c(13, 14, 15),
    type_binomial %in% c("positive"),
    richness > 0
  ) %>%
  mutate(
    collection_attempts = as.factor(collection_attempts),
    buffer = as.factor(buffer),
    cycles = as.factor(cycles)
  )

sn3_3_df %>%
  filter(!cycles %in% c("19")) %>%
  group_by(section, buffer) %>%
  summarise(n_samples = n())

sn3_3_df %>%
  filter(!cycles %in% c("19")) %>%
  group_by(section, buffer, collection_attempts) %>%
  summarise(n_samples = n(), .groups = "keep") %>%
  pivot_wider(names_from = collection_attempts, values_from = n_samples)
```
###### Visual Comparison of Sequencing Performance (Fig. S3.3.1)

```{r FigS3_3_1, warning = FALSE, comments = "", message = FALSE}
ordered_metrics <- c(
  "total_after_filtering.total_reads",
  "lowquality_reads_pct",
  # "adapter_contamination_pct",
  "host_pct",
  "mapmag_prokreads_pct",
  "read_fraction_singlem",
  "richness"
)

figs3_3_1 <- sn3_3_df %>%
  filter(!cycles %in% c("19")) %>%
  mutate(lowquality_reads_pct=lowquality_reads_pct*100,
         host_pct= host_pct*100,
         mapmag_prokreads_pct=mapmag_prokreads_pct*100) %>% 
  pivot_longer(
    cols = -c(names(mssm_sample_metadata), "cov_filtering"),
    names_to = "metric",
    values_to = "value"
  ) %>%
  filter(metric %in% ordered_metrics) %>%
  mutate(
    metric = factor(metric, levels = ordered_metrics),
    metric = fct_recode(metric,
      "a. Sequencing Yield \nPost-Quality Filtering" = "total_after_filtering.total_reads",
      "b. Low Quality \nReads (%)" = "lowquality_reads_pct",
      "c. Reads Mapping \nEukaryotic Host (%)" = "host_pct",
      "d. Reads Mapping \nMAG Catalogue (%)" = "mapmag_prokreads_pct",
      "e. Microbial Fraction  \nBased On SCG (%)" = "read_fraction_singlem",
      "f. Richness" = "richness"
    ),
    section = recode(section,
      "Caecum right" = "Caecum",
      "Colon" = "Colon"
    )
  ) %>%
  ggplot(aes(x = section, y = value, fill = section)) +
  geom_boxplot() +
  facet_wrap(~metric, scales = "free_y", ncol = 3) +
  labs(
    fill = "",
    y = "",
    x = "Intestinal section"
  ) +
  # geom_violin(trim = TRUE, alpha = 0.7, position = position_dodge(0.8), na.rm = TRUE) +  # Violin plot
  geom_jitter(
    position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.8),
    size = 1.5, alpha = 0.3
  ) +
  scale_fill_manual(values = c("#79b1a3", "#f58262", "#e3c7a0")) +
  custom_ggplot_theme +
  theme(legend.position = "none")

figs3_3_1
# ggsave(filename = "figures/FigS3_3_1.pdf", figs3_3_1, width = 8, height = 7)
# ggsave(filename = "figures/FigS3_3_1.png", figs3_3_1, width = 8, height = 7)
```

```{r validate_section_conclusion, warning = FALSE, comments = "", message = FALSE}
macroquant_metrics_filt_30 %>% 
  filter(#animal == "G121e", 
         !cycles %in% c(19), section=="Caecum right") %>%
  ggplot(aes(x = section, y = human_pct*100, fill = batch)) +
  geom_boxplot(alpha = 0.6) +
  theme_minimal() +
  labs(
    title = "Low-Quality Reads (%) in Colon (G121e)",
    x = "Section",
    y = "Low-Quality Reads (%)",
    fill = "Batch"
  ) +
  # geom_violin(trim = TRUE, alpha = 0.7, position = position_dodge(0.8), na.rm = TRUE) +  # Violin plot
  geom_jitter(
    position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.8),
    size = 1.5, alpha = 0.3
  ) 

macroquant_metrics_filt_30 %>% 
  filter(#animal == "G121e", 
         !cycles %in% c(19), section=="Caecum right") %>%
  ggplot(aes(x = section, y = lowquality_reads_pct*100, fill = batch)) +
  geom_boxplot(alpha = 0.6) +
  theme_minimal() +
  labs(
    title = "Low-Quality Reads (%) in Colon (G121e)",
    x = "Section",
    y = "Low-Quality Reads (%)",
    fill = "Batch"
  ) +
  # geom_violin(trim = TRUE, alpha = 0.7, position = position_dodge(0.8), na.rm = TRUE) +  # Violin plot
  geom_jitter(
    position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.8),
    size = 1.5, alpha = 0.3
  ) 


macroquant_metrics_filt_30 %>% 
  filter(animal == "G121e", 
         type_binomial == "positive", !cycles %in% c(19)) %>%
  ggplot(aes(x = section, y = mapmag_prokreads_pct, fill = batch)) +
  geom_boxplot(alpha = 0.6) +
  theme_minimal() +
  labs(
    title = "Low-Quality Reads (%) in Colon (G121e)",
    x = "Section",
    y = "Reads Mapping MAG Catalogue (%)",
    fill = "Batch"
  ) +
  # geom_violin(trim = TRUE, alpha = 0.7, position = position_dodge(0.8), na.rm = TRUE) +  # Violin plot
  geom_jitter(
    position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.8),
    size = 1.5, alpha = 0.3
  ) 
```


###### Statistical Comparison of Sequencing Performance

###### a. Reads Retained Post Quality Filtering
```{r sn3_3_a, warning = FALSE, comments = "", message = FALSE}
# Counts (integers ≥0): poisson, quasipoisson (accounts overdispersion)
sn3_3_a <- fit_and_analyze_model(
  model = "glm",
  distribution = "quasipoisson", # DHARMa significant deviation
  response_var = "total_after_filtering.total_reads",
  explanatory_var = "collection_attempts + section",
  data = sn3_3_df %>% filter(!cycles %in% c("19"))
)
# plot(a_sectioncomp$simResidual)
sn3_3_a$anova
```

###### b. Low Quality Reads (%)

```{r sn3_3_b, warning = FALSE, comments = "", message = FALSE}
# Proportions (0–1 continuous):	quasibinomial GLM (accounts overdispersion)
sn3_3_b <- fit_and_analyze_model(
  model = "glm",
  distribution = "quasibinomial",
  response_var = "lowquality_reads_pct",
  explanatory_var = "collection_attempts + section",
  data = sn3_3_df %>% filter(!cycles %in% c("19"))
)
sn3_3_b$anova

sn3_3_b$model_fit %>% plot_model(., type = "pred", terms = c("collection_attempts", "section"))
sn3_3_b$model_fit %>% plot_model(., type = "pred", terms = c("section"))
sn3_3_b$model_fit %>% ggpredict(., terms = c("section"))
```

###### c. Reads Mapping Eukaryotic Host (%)

```{r sn3_3_c, warning = FALSE, comments = "", message = FALSE}
# Proportions (0–1 continuous):	quasibinomial GLM (accounts overdispersion)
sn3_3_c <- fit_and_analyze_model(
  model = "glm",
  distribution = "quasibinomial",
  response_var = "host_pct",
  explanatory_var = "collection_attempts + section",
  data = sn3_3_df %>% filter(!cycles %in% c("19"))
)
sn3_3_c$anova
sn3_3_c$model_fit %>% plot_model(., type = "pred", terms = c("collection_attempts", "section"))
sn3_3_c$model_fit %>% plot_model(., type = "pred", terms = c("section"))
sn3_3_c$model_fit %>% ggpredict(., terms = c("section"))
```


###### d. Reads Mapping MAG Catalogue (%)

```{r d_sectioncomp, warning = FALSE, comments = "", message = FALSE}
# Proportions (0–1 continuous):	quasibinomial GLM (accounts overdispersion)
sn3_3_d <- fit_and_analyze_model(
  model = "glm",
  distribution = "quasibinomial",
  response_var = "mapmag_prokreads_pct",
  explanatory_var = "collection_attempts + section",
  data = sn3_3_df %>% filter(!cycles %in% c("19"))
)
sn3_3_d$anova
sn3_3_d$model_fit %>% plot_model(., type = "pred", terms = c("collection_attempts", "section"))
sn3_3_d$model_fit %>% plot_model(., type = "pred", terms = c("section"))
sn3_3_d$model_fit %>% ggpredict(., terms = c("section"))
```

###### e. Microbial Fraction Based On Single-Copy Gene Abundance (%)

```{r e_sectioncomp, warning = FALSE, comments = "", message = FALSE}
chart.Correlation(sn3_3_df %>%
  filter(section %in% c("Ileum")) %>%
  select(
    "read_fraction_singlem",
    "mapmag_prokreads_pct"
  ), method = "spearman")
# correlation

# Proportions (0–1 continuous):	quasibinomial GLM (accounts overdispersion)
sn3_3_e <- fit_and_analyze_model(
  model = "glm",
  distribution = "quasibinomial",
  response_var = "read_fraction_singlem",
  explanatory_var = "collection_attempts + section",
  data = sn3_3_df %>% filter(!cycles %in% c("19"))
)
sn3_3_e$anova
sn3_3_e$model_fit %>% plot_model(., type = "pred", terms = c("collection_attempts", "section"))
sn3_3_e$model_fit %>% plot_model(., type = "pred", terms = c("section"))
sn3_3_e$model_fit %>% ggpredict(., terms = c("section"))
```

###### Colon: 19 Versus 15 Cycles
```{r Boxplot_19versus15Cycles, warning = FALSE, comments = "", message = FALSE}
ordered_metrics <- c(
  "total_before_filtering.total_reads",
  "total_after_filtering.total_reads",
  "lowquality_reads_pct",
  "percent_duplication.rate"
)

sn_3_3_15vs19 <- sn3_3_df %>%
  filter(section %in% c("Colon")) %>%
  pivot_longer(
    cols = -c(names(mssm_sample_metadata), "cov_filtering"),
    names_to = "metric",
    values_to = "value"
  ) %>%
  filter(metric %in% ordered_metrics) %>%
  mutate(
    metric = factor(metric, levels = ordered_metrics),
    metric = fct_recode(metric,
      "a. Sequencing yield \n(Number Reads)" = "total_before_filtering.total_reads",
      "b. Reads Retained \nPost Quality Filtering" = "total_after_filtering.total_reads",
      "c. Low Quality Reads (%)" = "lowquality_reads_pct",
      "d. Reads \nDuplication Rate (%)" = "percent_duplication.rate"
    )
  ) %>%
  ggplot(aes(x = cycles, y = value, fill = buffer)) +
  geom_boxplot() +
  facet_wrap(~metric, scales = "free_y", ncol = 2) +
  labs(
    fill = "Lysis Treatment",
    y = "",
    x = "Intestinal section"
  ) +
  # geom_violin(trim = TRUE, alpha = 0.7, position = position_dodge(0.8), na.rm = TRUE) +  # Violin plot
  geom_jitter(
    position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.8),
    size = 1.5, alpha = 0.3
  ) +
  scale_fill_manual(values = c("#79b1a3", "#f58262")) +
  custom_ggplot_theme +
  theme(legend.position = "bottom")

sn_3_3_15vs19
```

```{r reads_cyclecomp, warning = FALSE, comments = "", message = FALSE}
# Counts (integers ≥0): poisson, quasipoisson (accounts overdispersion)
readsbf_cyclecomp <- fit_and_analyze_model(
  model = "glm",
  distribution = "quasipoisson", # DHARMa significant deviation
  response_var = "total_before_filtering.total_reads",
  explanatory_var = "collection_attempts + cycles",
  data = sn3_3_df %>% filter(section %in% c("Colon"))
)
# plot(reads_cyclecomp$simResidual)
readsbf_cyclecomp$anova
readsbf_cyclecomp$model_fit %>% ggpredict(., terms = c("cycles"))
```


```{r readsfilt_cyclecomp, warning = FALSE, comments = "", message = FALSE}
# Counts (integers ≥0): poisson, quasipoisson (accounts overdispersion)
readshq_cyclecomp <- fit_and_analyze_model(
  model = "glm",
  distribution = "quasipoisson", # DHARMa significant deviation
  response_var = "total_after_filtering.total_reads",
  explanatory_var = "collection_attempts + cycles",
  data = sn3_3_df %>% filter(section %in% c("Colon"))
)
# plot(reads_cyclecomp$simResidual)
readshq_cyclecomp$anova
readshq_cyclecomp$model_fit %>% ggpredict(., terms = c("cycles"))
```

```{r quality_cyclecomp, warning = FALSE, comments = "", message = FALSE}
# Proportions (0–1 continuous):	quasibinomial GLM (accounts overdispersion)
quality_cyclecomp <- fit_and_analyze_model(
  model = "glm",
  distribution = "quasibinomial",
  response_var = "lowquality_reads_pct",
  explanatory_var = "collection_attempts + cycles",
  data = sn3_3_df %>% filter(section %in% c("Colon"))
)
quality_cyclecomp$anova
quality_cyclecomp$model_fit %>% ggpredict(., terms = c("cycles"))
```

```{r duplication_cyclecomp, warning = FALSE, comments = "", message = FALSE}
# Proportions (0–1 continuous):	quasibinomial GLM (accounts overdispersion)
duplication_cyclecomp <- fit_and_analyze_model(
  model = "glm",
  distribution = "quasibinomial", # DHARMa significant deviation
  response_var = "percent_duplication.rate",
  explanatory_var = "collection_attempts + cycles",
  data = sn3_3_df %>% filter(section %in% c("Colon"))
)
duplication_cyclecomp$anova
```

###### Community Composition of Caecum & Colon (Fig. S3.3.2)

```{r physeq_section, warning = FALSE, comments = "", message = FALSE}
sn3_3_physeq <- subset_samples(
  physeq_all,
  section %in% c("Caecum right", "Colon") &
    !cycles %in% c(19) &
    animal %in% c("G121e") &
    type_binomial == "positive"
)

sn3_3_physeq <- prune_taxa(taxa_sums(sn3_3_physeq) > 0, sn3_3_physeq)
# sn3_3_physeq <- filter_taxa(sn3_3_physeq, function(x) sum(x > 0) > (0.05 * length(x)), TRUE)
sn3_3_aldex_G121e_counts <- data.frame(otu_table(sn3_3_physeq))
sn3_3_aldex_G121e_pseudo_counts <- round(sn3_3_aldex_G121e_counts * 1e6)
sn3_3_aldex_G121e_conditions <- data.frame(sample_data(sn3_3_physeq)) %>% pull(section)
# Perform the ALDEx2 differential abundance test
# Use 't' test for two groups; 'glm' for multi-class or covariates
sn3_3_aldex_G121e_results <- aldex(sn3_3_aldex_G121e_pseudo_counts,
  sn3_3_aldex_G121e_conditions,
  test = "t", # or "wilcox"
  effect = TRUE, # include effect sizes
  denom = "all", # normalization denominator
  mc.samples = 256, # number of Monte Carlo instances
  verbose = TRUE
)
head(sn3_3_aldex_G121e_results)
# - we.ep: expected p-value from Welch's t-test
# - wi.ep: expected p-value from Wilcoxon test
# - we.eBH: BH-adjusted p-value (q-value)
# - effect: effect size (difference in clr means / variability)
# Filter
sn3_3_aldex_G121e_sig <- sn3_3_aldex_G121e_results[sn3_3_aldex_G121e_results$we.eBH < 0.05, ]
sn3_3_aldex_G121e_effect <- sn3_3_aldex_G121e_sig[abs(sn3_3_aldex_G121e_sig$effect) > 0.5, ]

sn3_3_aldex <-
  sn3_3_aldex_G121e_effect %>%
  rownames_to_column("genome") %>%
  left_join(data.frame(tax_table(sn3_3_physeq)) %>% rownames_to_column("genome"), by = "genome")

# Create horizontal barplot
sn3_3_aldex %>%
  ggplot(., aes(x = reorder(genome, effect), y = diff.btw, fill = abs(effect) > 1)) +
  geom_bar(stat = "identity", show.legend = FALSE) +
  coord_flip() +
  labs(
    x = "Genus",
    y = "Median difference in CLR values between groups"
  ) +
  scale_fill_manual(values = c("TRUE" = "#e35d51", "FALSE" = "darkgrey")) +
  theme_minimal(base_size = 14)
```


**Note:** The pivot_phylo() function needs to be loaded.
```{r physeq_section_lg, warning = FALSE, comments = "", message = FALSE}
sn3_3_physeq_lg <- pivot_phylo(
  phyloseq_obj = subset_samples(
    physeq_all,
    batch %in% c("MSEB0006", "MSEB0010") &
      section %in% c("Caecum right", "Colon") &
      !cycles %in% c(19) &
      type_binomial == "positive"
  ),
  glom = TRUE,
  tax_transform = TRUE,
  taxon_level = "order",
  tr_method = "compositional"
)
```

```{r figs3_3_2, warning = FALSE, comments = "", message = FALSE}
figs3_3_2 <- sn3_3_physeq_lg %>%
  mutate(section = ifelse(section == "Caecum right", "Caecum", section)) %>%
  ggplot(
    aes(x = microsample, y = abundance, fill = order)
  ) +
  geom_bar(stat = "identity", colour = "white", linewidth = 0.1) +
  scale_fill_manual(values = order_colors, drop = FALSE) +
  labs(
    x = "Microsamples",
    y = "Relative abundance",
    fill = "Order"
  ) +
  facet_wrap(~section, scales = "free", ncol = 1) +
  guides(fill = guide_legend(ncol = 4)) +
  custom_ggplot_theme +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank()) +
  theme(legend.position = "bottom")

figs3_3_2
# ggsave(filename = "figures/FigS3_3_2.pdf", figs3_3_2, width = 8, height = 7)
# ggsave(filename = "figures/FigS3_3_2.png", figs3_3_2, width = 8, height = 7)
```


###### Confocal microscopy of DAPI-stained microsections 
```{r sn3_3_dapi_df, warning = FALSE, comments = "", message = FALSE}
sn3_3_dapi_df <- dapi_df %>%
  group_by(Animal, Sequence, Age, Section) %>%
  summarise(
    n = n(),
    Distance_from_one_side_mn = mean(Distance_from_one_side, na.rm = TRUE),
    Distance_from_one_side_sd = ifelse(n > 1, sd(Distance_from_one_side), NA_real_),
    mean = mean(Bacterial_count),
    median = median(Bacterial_count),
    sd_na = ifelse(n > 1, sd(Bacterial_count), NA_real_), # If only 1 obs, return NA
    sd = ifelse(n() > 1, sd(Bacterial_count, na.rm = TRUE), 0), # Avoid NA SD
    sem = ifelse(n > 1, sd / sqrt(n), 0), # Set SEM to 0 if only 1 observation
    q1 = quantile(Bacterial_count, 0.25), # First quartile
    q3 = quantile(Bacterial_count, 0.75), # Third quartile
    .groups = "drop"
  )

dapi_df %>%
  group_by(Age, Section) %>%
  summarise(
    n = n(),
    mean = mean(Bacterial_count),
    median = median(Bacterial_count),
    sd_na = ifelse(n > 1, sd(Bacterial_count), NA_real_), # If only 1 obs, return NA
    sd = ifelse(n() > 1, sd(Bacterial_count, na.rm = TRUE), 0), # Avoid NA SD
    q1 = quantile(Bacterial_count, 0.25), # First quartile
    q3 = quantile(Bacterial_count, 0.75), # Third quartile
    .groups = "drop"
  )
```

####### Visual Comparison (Fig. S3.3.3a)

```{r figs3_3_3a, warning = FALSE, comments = "", message = FALSE}
figs3_3_3a <- ggplot(sn3_3_dapi_df, aes(x = Distance_from_one_side_mn, y = median, color = Section)) +
  geom_line(linewidth = 1) + # Mean line
  geom_ribbon(aes(ymin = q1, ymax = q3, fill = Section),
    alpha = 0.2
  ) + # SEM shading
  labs(
    title = "a",
    x = "Distance from Villi (μm)",
    y = "Bacterial Counts (Median)"
  ) +
  facet_wrap(~Age, nrow = 2, scale = "free_x") +
  scale_fill_manual(values = c("#79b1a3", "#f58262", "#e3c7a0"), drop = FALSE) +
  scale_color_manual(values = c("#79b1a3", "#f58262", "#e3c7a0"), drop = FALSE) +
  custom_ggplot_theme +
  theme(legend.position = "bottom")

figs3_3_3a
# ggsave(filename = "figures/FigS3_3_3a.pdf", figs3_3_3a, width = 8, height = 7)
# ggsave(filename = "figures/FigS3_3_3a.png", figs3_3_3a, width = 8, height = 7)
```


### Design considerations: Controls (Suppl. Note 3.4)

###### Visual Comparison of Sequencing Performance (Fig. S3.4.1)
```{r FigS3_4_1, warning = FALSE, comments = "", message = FALSE}
ordered_metrics <- c(
  "total_before_filtering.total_reads",
  "lowquality_reads_pct",
  "adapter_contamination_pct",
  "human_pct",
  "raw_total_sequences_mag",
  "mapmag_prokreads_pct"
)

figs3_4_1 <- macroquant_metrics_filt_30 %>%
  filter(!section %in% c("Ileum")) %>%
  mutate(batch = gsub("MSEB00", "B", batch),
         lowquality_reads_pct=lowquality_reads_pct*100,
         adapter_contamination_pct= adapter_contamination_pct*100,
         human_pct=human_pct*100,
         mapmag_prokreads_pct=mapmag_prokreads_pct*100) %>%
  pivot_longer(
    cols = -c(names(mssm_sample_metadata), "cov_filtering"),
    names_to = "metric",
    values_to = "value"
  ) %>%
  filter(metric %in% ordered_metrics) %>%
  mutate(
    metric = factor(metric, levels = ordered_metrics),
    metric = fct_recode(metric,
      "a. Sequencing Yield Pre-quality Filtering" = "total_before_filtering.total_reads",
      "b. Low Quality Reads (%)" = "lowquality_reads_pct",
      "c. Adapter Contamination (% bases)" = "adapter_contamination_pct",
      "d. Reads Mapping Human (%)" = "human_pct",
      "e. Read Counts for Bacterial Quantification " = "raw_total_sequences_mag",
      "f. Reads Mapping MAG Catalogue (%)" = "mapmag_prokreads_pct"
    ),
    type_binomial = recode(type_binomial,
      "positive" = "Microsample",
      "negative" = "Negative"
    )
  ) %>%
  ggplot(aes(x = batch, y = value, fill = type_binomial)) +
  geom_boxplot() +
  facet_wrap(~metric, scales = "free_y", ncol = 1) +
  labs(
    fill = "",
    y = "",
    x = "Batch"
  ) +
  # geom_violin(trim = TRUE, alpha = 0.7, position = position_dodge(0.8), na.rm = TRUE) +  # Violin plot
  geom_jitter(
    position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.8),
    size = 1, alpha = 0.3
  ) +
  scale_fill_manual(values = c("#f58262", "#79b1a3")) +
  custom_ggplot_theme +
  theme(legend.position = "bottom")

figs3_4_1
# ggsave(filename = "figures/FigS3_4_1.pdf", figs3_4_1, width = 8, height = 11)
# ggsave(filename = "figures/FigS3_4_1.png", figs3_4_1, width = 8, height = 11)
```


#### Amalia: Centered Log-Ratio (CLR) Transformation And Principal Component Analysis (PCA)

###### FigS3_4_2

```{r FigS3_4_2, warning = FALSE, comments = "", message = FALSE}
```

###### FigS3_4_3

```{r FigS3_4_3, warning = FALSE, comments = "", message = FALSE}
```


###### Visual Comparison of Coverage Cutoff Filtering (Fig. S3.4.4)

```{r FigS3_4_4, warning = FALSE, comments = "", message = FALSE}
figs3_4_4 <- macroquant_metrics_filt_30 %>%
  mutate(
    batch = gsub("MSEB00", "B", batch),
    type_binomial = recode(type_binomial,
      "positive" = "Positive",
      "negative" = "Negative"
    )
  ) %>%
  # group_by(type_binomial, cov_filtering) %>%
  group_by(batch, type_binomial, cov_filtering) %>%
  summarise(n_samples = n()) %>%
  ggplot(aes(fill = cov_filtering, x = batch, y = n_samples)) +
  geom_bar(position = "fill", stat = "identity") + # Stacked 100% bar
  geom_text(aes(label = n_samples),
    position = position_fill(vjust = 0.5),
    color = "black", size = 6
  ) +
  facet_wrap(~type_binomial) +
  scale_fill_manual(values = c("#f58262", "#79b1a3")) +
  labs(
    fill = "",
    y = "",
    x = "Batch"
  ) +
  custom_ggplot_theme +
  theme(legend.position = "bottom")

figs3_4_4
# ggsave(filename = "figures/FigS3_4_4.pdf", figs3_4_4, width = 8, height = 6)
# ggsave(filename = "figures/FigS3_4_4.png", figs3_4_4, width = 8, height = 6)
```


```{r sn3_4_physeq_lg, warning = FALSE, comments = "", message = FALSE}
sn3_4_physeq_lg <- pivot_phylo(
  phyloseq_obj = physeq_all,
  glom = TRUE,
  tax_transform = TRUE,
  taxon_level = "genus",
  tr_method = "compositional"
)
sn3_4_physeq_lg %>%
  group_by(type_binomial) %>%
  summarise(n_unique_microsamples = n_distinct(microsample))
# negative	19
# positive	501
sn3_4_physeq_lg %>%
  group_by(genus, type_binomial) %>%
  summarise(
    n_samples = n(),
    mean_ra = mean(abundance)
  ) %>%
  filter(genus %in% c(sn3_4_physeq_lg %>%
    filter(type_binomial %in% c("negative")) %>% pull(genus)))
```


```{r figs3_4_5, warning = FALSE, comments = "", message = FALSE}
figs3_4_5 <- sn3_4_physeq_lg %>%
  filter(type_binomial %in% c("negative")) %>%
  ggplot(
    aes(x = microsample, y = abundance, fill = genus)
  ) +
  geom_bar(stat = "identity", colour = "white", linewidth = 0.1) +
  # scale_fill_manual(values = order_colors, drop = FALSE) +
  labs(
    x = "Negative reactions",
    y = "Relative abundance",
    fill = "Genus"
  ) +
  # facet_wrap(~batch, scale = "free") +
  guides(fill = guide_legend(ncol = 4)) +
  custom_ggplot_theme +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank()) +
  theme(legend.position = "bottom")
figs3_4_5
# ggsave(filename = "figures/FigS3_4_5.pdf", figs3_4_5, width = 8, height = 7)
# ggsave(filename = "figures/FigS3_4_5.png", figs3_4_5, width = 8, height = 7)
```

### Amalia: Design considerations: Microsample Sizes

```{r x2, warning = FALSE, comments = "", message = FALSE}
```

#### Amalia: MSSM spatial resolution evaluation (Fig. 3)
```{r x3, warning = FALSE, comments = "", message = FALSE}
```

### Resource optimisation and throughput (Suppl. Note 3.5)
```{r sn3_5_df, warning = FALSE, comments = "", message = FALSE}
sn3_5_df <- macroquant_metrics_filt_30 %>%
  filter(
    batch %in% c("MSEB0009", "MSEB0014", "MSEB0015"),
    type_binomial %in% c("positive"),
    size %in% c("5000"),
    !between(as.numeric(sub("M", "", microsample)), 300651, 300665)
  ) %>%
  mutate(
    protocol = ifelse(protocol %in% c("ULV2_Full"), "100%", "50%"),
    comparison = ifelse(batch %in% c("MSEB0009"), "Cost", "Throughput"),
    lib_method = ifelse(batch %in% c("MSEB0015"), "Fluent (1 plate set-up)", "Manual"),
    comparison_grid = ifelse(comparison == "Cost", protocol, lib_method)
  ) %>%
  mutate(
    comparison = as.factor(comparison),
    comparison_grid = as.factor(comparison_grid)
  ) # %>% count(comparison, batch)
 table(sn3_5_df$comparison_grid)
```

```{r sn3_5_total_pct, warning = FALSE, comments = "", message = FALSE}
sn3_5_df_t <- sn3_5_df %>%
  group_by(comparison_grid) %>%
  # summarise(mean=mean(total_after_filtering.total_reads))
  summarise(
    total_reads = sum(total_before_filtering.total_reads),
    low_quality = sum(total_before_filtering.total_reads - total_after_filtering.total_reads),
    mapping_host = sum(reads_mapped_chicken),
    mapping_human = sum(reads_mapped_human),
    mapping_pig = sum(reads_mapped_pig),
    mapping_bacteria = sum(reads_mapped_mag),
    unmapped_bacteria = sum(reads_unmapped_mag)
  ) %>%
  mutate(other = total_reads - (low_quality + mapping_host + mapping_human + mapping_pig + mapping_bacteria + unmapped_bacteria)) %>%
  mutate(across(-c(comparison_grid, total_reads),
    ~ round((. / total_reads) * 100, 1),
    .names = "{.col}_pct"
  )) %>%
  select(comparison_grid, ends_with("_pct")) %>%
  t() %>%
  as.data.frame()
colnames(sn3_5_df_t) <- sn3_5_df_t[1, ]
sn3_5_df_t <- sn3_5_df_t[-1, ] # remove the first row
sn3_5_df_t
```

```{r sn3_5_median_pct, warning = FALSE, comments = "", message = FALSE}
sn3_5_df_t <- sn3_5_df %>%
  mutate(other = total_before_filtering.total_reads - ((total_before_filtering.total_reads - total_after_filtering.total_reads)
  + reads_mapped_chicken + reads_mapped_human + reads_mapped_pig + reads_mapped_mag + reads_unmapped_mag)) %>%
  select(microsample, comparison_grid, total_before_filtering.total_reads, total_low_quality_reads, reads_mapped_chicken, reads_mapped_human, reads_mapped_pig, reads_mapped_mag, reads_unmapped_mag, other) %>%
  mutate(across(-c(microsample, comparison_grid, total_before_filtering.total_reads),
    ~ round((. / total_before_filtering.total_reads) * 100, 1),
    .names = "{.col}_pct"
  )) %>%
  select(microsample, comparison_grid, ends_with("_pct")) %>%
  group_by(comparison_grid) %>%
  summarise(
    low_quality_md = median(total_low_quality_reads_pct),
    mapping_host_md = median(reads_mapped_chicken_pct),
    mapping_human_md = median(reads_mapped_human_pct),
    mapping_pig_md = median(reads_mapped_pig_pct),
    mapping_bacteria_md = median(reads_mapped_mag_pct),
    unmapped_bacteria_md = median(reads_unmapped_mag_pct),
    other_md = median(other_pct)
  ) %>%
  select(comparison_grid, ends_with("_md")) %>%
  t() %>%
  as.data.frame()
colnames(sn3_5_df_t) <- sn3_5_df_t[1, ]
sn3_5_df_t <- sn3_5_df_t[-1, ] # remove the first row
sn3_5_df_t
```


```{r FigS3_5_1a, warning = FALSE, comments = "", message = FALSE}
ordered_metrics <- c(
  # "total_before_filtering.total_reads",
  "total_low_quality_reads",
  "reads_mapped_chicken",
  "reads_mapped_human",
  "reads_mapped_pig",
  "reads_mapped_mag",
  "reads_unmapped_mag",
  "reads_other"
)

figs3_5_1a <- sn3_5_df %>%
  mutate(other = total_before_filtering.total_reads - ((total_before_filtering.total_reads - total_after_filtering.total_reads)
  + reads_mapped_chicken + reads_mapped_human + reads_mapped_pig + reads_mapped_mag + reads_unmapped_mag)) %>%
  pivot_longer(
    cols = -c(names(mssm_sample_metadata), comparison, lib_method, comparison_grid, cov_filtering),
    names_to = "metric",
    values_to = "value"
  ) %>%
  filter(metric %in% ordered_metrics) %>%
  mutate(microsample = as.factor(microsample)) %>%
  ggplot(aes(x = microsample, y = value, fill = metric)) +
  geom_bar(position = "fill", stat = "identity") +
  scale_fill_manual(values = c(
    "#346254", "#4a6ab7", "#8c1c47",
    "#9c8464", "#c49d4b", "#462410", "lightblue"
  )) +
  labs(x = "Microsample", y = "", fill = "Read type") +
  facet_nested(. ~ comparison + comparison_grid, scales = "free", space = "free", switch = "y") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  custom_ggplot_theme +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank()) +
  theme(legend.position = "bottom") +
  guides(fill = guide_legend(title = "Proportion \nSequencing \nReads"))

figs3_5_1a
# ggsave(filename = "figures/FigS3_5_1a.pdf", figs3_5_1a, width = 8, height = 6)
# ggsave(filename = "figures/FigS3_5_1a.png", figs3_5_1a, width = 8, height = 6)
```

```{r sn3_5_dreamprep_df , warning = FALSE, comments = "", message = FALSE}
# Prepare data: extract plate and filter
sn3_5_dreamprep_df_f <- sn3_5_dreamprep_df %>% 
  #group_by(Range, type, concentration) %>% 
  #summarise(mean=mean(nmole_L))
  mutate(plate = str_extract(sample, "^Plate \\d+")) %>%
  filter(
    !Range %in% c("35 bp to 100 bp"),
    concentration %in% c("0.1"),
    type %in% c("Positive")
  ) %>% # remove rows with that level
  mutate(
    Range = droplevels(Range),
    plate = case_when(
      grepl("Plate 1", plate, ignore.case = TRUE) ~ "Single-Plate",
      grepl("Plate 2", plate, ignore.case = TRUE) ~ "Dual-Plate A",
      grepl("Plate 3", plate, ignore.case = TRUE) ~ "Dual-Plate B",
      TRUE ~ NA_character_
    ),
    plate = factor(plate, levels = c("Single-Plate", "Dual-Plate A", "Dual-Plate B")),
    concentration = paste0("DNA Input ", concentration, " ng/μL"),
    Range = case_when(
      grepl("100 bp to 200 bp", Range, ignore.case = TRUE) ~ "Non-specific Short Fragments",
      grepl("200 bp to 1000 bp", Range, ignore.case = TRUE) ~ "Library",
      TRUE ~ NA_character_
    )
  )
```


```{r FigS3_5_1b, warning = FALSE, comments = "", message = FALSE}
sn3_5_dreamprep_df_f %>% 
  filter(!nmole_L==0) %>% 
  group_by(Range, type, concentration) %>% 
  summarise(mean=mean(nmole_L))

comparisons <- list(c("Single-Plate", "Dual-Plate A"), c("Dual-Plate A", "Dual-Plate B"), c("Single-Plate", "Dual-Plate B"))

figs3_5_1b <- sn3_5_dreamprep_df_f %>%
  #filter(concentration == "DNA Input 0.01 ng/μL") %>% 
  ggplot(aes(x = plate, y = nmole_L, fill = plate)) +
  geom_boxplot(position = position_dodge(width = 0.75)) +
  geom_boxplot() +
  geom_point(position = position_jitterdodge(0.2)) +
  stat_compare_means(
    comparisons = comparisons,
    method = "wilcox.test",
    label = "p.signif", # <-- show stars, not p =
    p.adjust.method = "BH",
    size = 4,
    vjust = 0.2,
    # step.increase = 0.02,   # Reduces vertical distance between labels
    tip.length = 0.01
  ) +
  # scale_y_continuous(trans="log10") +
  facet_wrap(~Range, scales = "free") +
  labs(
    y = "nmol/μL",
    x = "",
  ) +
  geom_jitter(
    position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.8),
    size = 1.5, alpha = 0.3
  ) +
  scale_fill_manual(values = c("#79b1a3", "#f58262", "#e3c7a0")) +
  custom_ggplot_theme +
  theme(legend.position = "none")

figs3_5_1b
```

```{r FigS3_5_1, warning = FALSE, comments = "", message = FALSE}
figs3_5_1 <- cowplot::plot_grid(figs3_5_1a, figs3_5_1b,
  ncol = 1, labels = c("a", "b"),
  label_size = 16
)
figs3_5_1
ggsave("figures/FigS3_5_1.png", figs3_5_1, width = 10, height = 13, dpi = 300)
# ggsave("figures/FigS3_5_1.pdf", figs3_5_1, width = 10, height = 13)
```

