# Main analysis

```{r load_data_main, echo = TRUE,  warning = FALSE, comments = "", message = FALSE}
load("resources/working_data_object.Rdata") 
```

## Microsample collection (Suppl.Fig. 2)

```{r figs2_plot, echo = TRUE, warning=FALSE, comments="", message=FALSE, fig.height=5, fig.width=8}
collection_success_df <- macroquant_metrics_filt_30 %>%
  filter(type == "positive", collection_attempts > 0) %>%
  mutate(batch = gsub("MSEB00", "B", batch))

batch_levels <- c(sort(unique(collection_success_df$batch)), "All batches")
size_levels <- c(500, 1500, 2500, 5000, 25000, 50000)

all_batches_plot_vertical <- collection_success_df %>%
  bind_rows(collection_success_df %>% mutate(batch = "All batches")) %>%
  mutate(
    batch = factor(batch, levels = batch_levels),
    collection = factor(collection, levels = c("Fail", "Success")),
    collection_attempts = factor(collection_attempts, levels = c(4, 3, 2, 1)),
    batch_group = if_else(batch == "All batches", "All batches", "Individual batches")
  ) %>%
  count(batch, batch_group, collection, collection_attempts, name = "n") %>%
  left_join(
    collection_success_df %>%
      bind_rows(collection_success_df %>% mutate(batch = "All batches")) %>%
      mutate(
        batch = factor(batch, levels = batch_levels),
        batch_group = if_else(batch == "All batches", "All batches", "Individual batches")
      ) %>%
      count(batch, batch_group, name = "n_total"),
    by = c("batch", "batch_group")
  ) %>%
  mutate(frac_signed = if_else(collection == "Fail", -n / n_total, n / n_total)) %>%

  ggplot(aes(x = batch, y = frac_signed, fill = collection_attempts, alpha = batch_group)) +
  geom_col(position = position_stack(reverse = TRUE), width = 0.85) +
  geom_hline(yintercept = 0, color = "grey20", linewidth = 1.1) +
  geom_vline(xintercept = length(batch_levels) - 0.5, linetype = "dashed", color = "grey40") +
  geom_text(
    data = collection_success_df %>%
      bind_rows(collection_success_df %>% mutate(batch = "All batches")) %>%
      mutate(
        batch = factor(batch, levels = batch_levels),
        batch_group = if_else(batch == "All batches", "All batches", "Individual batches")
      ) %>%
      count(batch, batch_group, name = "n_total"),
    aes(
      x = batch,
      y = 1.02,
      label = paste0("n=", n_total),
      fontface = if_else(as.character(batch) == "All batches", "bold", "plain")
    ),
    inherit.aes = FALSE,
    angle = 0,
    vjust = 0,
    size = 2.7,
    show.legend = FALSE
  ) +
  scale_alpha_manual(values = c("Individual batches" = 0.75, "All batches" = 1), guide = "none") +
  scale_fill_manual(
    values = c("4" = "#685d79", "3" = "#ac6d83", "2" = "#d9727f", "1" = "#fcbb6d"),
    breaks = c("4", "3", "2", "1"),
    labels = c("4", "3", "2", "1"),
    drop = FALSE,
    guide = guide_legend(
      nrow = 1,
      byrow = TRUE,
      override.aes = list(alpha = 1, linewidth = 0)
    )
  ) +
  scale_y_continuous(
    limits = c(-0.5, 1.05),
    breaks = c(-0.5, -0.25, 0, 0.25, 0.5, 0.75, 1.0),
    labels = c("50%", "25%", "0%", "25%", "50%", "75%", "100%"),
    expand = c(0, 0)
  ) +
  coord_cartesian(ylim = c(-0.55, 1.08), clip = "off") +
  theme_minimal() +
  custom_ggplot_theme +
  labs(x = "Batch", y = "Fraction of samples (bottom: Fail, top: Success)", fill = "Number of collection attempts") +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
    plot.margin = margin(15, 15, 15, 15)
  )


b14_size_plot_vertical <- collection_success_df %>%
  filter(batch == "B14") %>%
  mutate(
    size = factor(size, levels = size_levels),
    collection = factor(collection, levels = c("Fail", "Success")),
    collection_attempts = factor(collection_attempts, levels = c(4, 3, 2, 1))
  ) %>%
  count(size, collection, collection_attempts, name = "n") %>%
  left_join(
    collection_success_df %>%
      filter(batch == "B14") %>%
      mutate(size = factor(size, levels = size_levels)) %>%
      count(size, name = "n_total"),
    by = "size"
  ) %>%
  mutate(frac_signed = if_else(collection == "Fail", -n / n_total, n / n_total)) %>%
  ggplot(aes(x = size, y = frac_signed, fill = collection_attempts)) +
  # Stacked bars by attempts; reverse order keeps visual direction consistent.
  geom_col(position = position_stack(reverse = TRUE), width = 0.85) +
  # Shared zero line separates fail/success regions.
  geom_hline(yintercept = 0, color = "grey20", linewidth = 1.1) +
  # Per-area n-labels above bars.
  geom_text(
    data = collection_success_df %>%
      filter(batch == "B14") %>%
      mutate(size = factor(size, levels = size_levels)) %>%
      count(size, name = "n_total"),
    aes(x = size, y = 1.02, label = paste0("n=", n_total)),
    inherit.aes = FALSE,
    angle = 0,
    vjust = 0,
    size = 2.7
  ) +
  scale_fill_manual(values = c("4" = "#685d79", "3" = "#ac6d83", "2" = "#d9727f", "1" = "#fcbb6d"), drop = FALSE) +
  guides(fill = "none") +
  # Same diverging scale and labels as left panel for direct interpretability.
  scale_y_continuous(
    limits = c(-0.5, 1.05),
    breaks = c(-0.5, -0.25, 0, 0.25, 0.5, 0.75, 1.0),
    labels = c("50%", "25%", "0%", "25%", "50%", "75%", "100%"),
    expand = c(0, 0)
  ) +
  coord_cartesian(ylim = c(-0.55, 1.08), clip = "off") +
  theme_minimal() +
  custom_ggplot_theme +
  labs(x = "B14: Microsample area", y = NULL, fill = "Number of collection attempts") +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
    plot.margin = margin(15, 15, 15, 15)
  )

figs4_alt_vertical <- all_batches_plot_vertical + b14_size_plot_vertical +
  plot_layout(
    widths = c(length(batch_levels), length(size_levels)),
    guides = "collect"
  ) &
  theme(legend.position = "bottom", legend.box = "vertical")
figs4_alt_vertical
```

## Microsample Area Validation and Lower Detection Limit

```{r Fig2_prep, echo=TRUE,  warning = FALSE, comments = "", message = FALSE}
custom_ggplot_theme_LMD_size <- custom_ggplot_theme +
  theme(
        legend.text = element_text(size = 7),
        legend.title = element_text(size = 7),
        axis.text = element_text(size = 5),
        axis.title = element_text(size = 7, face = "bold"),
        strip.text = element_text(size = 7, face = "bold", color = "black"),
        plot.title = element_text(size = 7, face = "bold", color = "black"),
        plot.margin = unit(c(0, 0, 0, 0), "cm"))

size_colors <- c("500" = '#ffdf9e', "1500" = '#ffc273', "2500" = '#e56969', "5000" = '#c1558b', "25000" = '#8a49a1',"50000" = '#4f5bd5')
prc_mapped_var <-  "mapmag_prokreads_pct" # "mapmag_filteredreads_pct" # "mapmag_rawreads_pct" # 

# Make a-diversity df tidy for plotting
macroquant_metrics_filt_30_tidy_M14_pos_retained <- macroquant_metrics_filt_30 %>%
  pivot_longer(
    cols = c(richness, neutral, phylogenetic), 
    names_to = "metric",                      
    values_to = "value") %>%
  mutate(metric = factor(metric, levels = c("richness", "neutral", "phylogenetic")),
         size = factor(size, levels = c(500, 1500, 2500, 5000, 25000, 50000))) %>%
  filter(
    batch == "MSEB0014",
    cov_filtering == "Retained By Filtering",
    type == "positive"
  ) 

macroquant_genome_counts_filt_30_tidy_M14_pos <- macroquant_genome_counts_filt_30_tidy %>%
  filter(
    batch == "MSEB0014",
    cov_filtering == "Retained By Filtering",
    type == "positive"
  )
```

### Fig. 2b
```{r Fig2b,  echo=TRUE, warning = FALSE, comments = "", message = FALSE, fig.height = 5, fig.width = 8}
# Plot a-diversity metrics for different microsample sizes
var <- "value"
summary_stats_B14_alpha_div <- macroquant_metrics_filt_30_tidy_M14_pos_retained %>%
  group_by(size, metric) %>%
  summarise(
    median = median(.data[[var]], na.rm = TRUE),
    IQR    = IQR(.data[[var]], na.rm = TRUE),
    Q1     = quantile(.data[[var]], 0.25, na.rm = TRUE),
    Q3     = quantile(.data[[var]], 0.75, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    label_txt = paste0(
      "", round(median, 1), "\n",
      " [", round(Q1, 1), "–", round(Q3, 1), "]"
    )
  )

mapmag_summary_by_size <- macroquant_metrics_filt_30_tidy_M14_pos_retained %>%
  group_by(size) %>%
  summarise(
    median_pct = median(.data[[prc_mapped_var]] * 100, na.rm = TRUE),
    Q1 = quantile(.data[[prc_mapped_var]] * 100, 0.25, na.rm = TRUE),
    Q3 = quantile(.data[[prc_mapped_var]] * 100, 0.75, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    size = factor(size, levels = levels(macroquant_metrics_filt_30_tidy_M14_pos_retained$size)),
    metric = factor("Mapped MAG (%)", levels = "Mapped MAG (%)"),
    mapmag_bin = cut(
      median_pct,
      breaks = c(0, 20, 40, 60, 80, 100),
      include.lowest = TRUE,
      labels = c("0-20", "20-40", "40-60", "60-80", "80-100")
    ),
    label_txt = paste0(round(median_pct, 1), "\n[", round(Q1, 1), "-", round(Q3, 1), "]")
  )

mapmag_bin_colors <- c(
  "0-20" = "#E7E3D4",
  "20-40" = "#C9C3A5",
  "40-60" = "#9FAF88",
  "60-80" = "#6F8B6B",
  "80-100" = "#3F5F45"
)

fig2b_top <- macroquant_metrics_filt_30_tidy_M14_pos_retained %>%
  mutate(
    mapmag_pct = .data[[prc_mapped_var]] * 100,
    mapmag_bin = cut(
      mapmag_pct,
      breaks = c(0, 20, 40, 60, 80, 100),
      include.lowest = TRUE,
      labels = c("0-20", "20-40", "40-60", "60-80", "80-100")
    )
  ) %>%
  ggplot(aes(x = size, y = value)) +
    geom_boxplot(outlier.shape = NA, size = 0.3, fill = "white") +
    geom_jitter(
      aes(fill = mapmag_bin),
      shape = 21, color = "black",
      width = 0.2, alpha = 1,
      size = 1.5, stroke = 0.2
    ) +
    scale_fill_manual(values = mapmag_bin_colors, name = "Mapped MAG (%)", drop = FALSE) +
    scale_y_continuous(
      limits = c(0, NA),
      expand = expansion(mult = c(0.05, 0.45))
    ) +
    geom_text(
      data = summary_stats_B14_alpha_div,
      aes(x = size, y = median, label = label_txt),
      inherit.aes = FALSE,
      vjust = -0.6,
      size = 2, alpha = 1, color = "black"
    ) +
    facet_nested(metric ~ size, scales = "free", space = "fixed", switch = "y",
                 strip = strip_themed(
                   background_x = elem_list_rect(
                     fill = size_colors,
                     colour = "white",
                     size = 0.8))) +
    theme_minimal() +
    custom_ggplot_theme_LMD_size +
    labs(x = NULL, y = NULL) +
    theme(
      axis.text.x = element_blank(),
      axis.text.y = element_text(color = "black"),
      axis.ticks.y = element_line(color = "black", linewidth = 0.25),
      axis.ticks.length.y = unit(0.08, "cm"),
      panel.grid = element_blank(),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      panel.grid.major.x = element_blank(),
      panel.grid.major.y = element_blank(),
      panel.grid.minor.x = element_blank(),
      panel.grid.minor.y = element_blank()
    )

fig2b_mapmag_panel <- ggplot(mapmag_summary_by_size, aes(x = size, y = 1, fill = mapmag_bin)) +
  geom_tile(width = 1.2, height = 0.45, color = "black", linewidth = 0.25) +
  geom_text(aes(label = label_txt), color = "white", size = 2.2, lineheight = 0.95) +
  scale_fill_manual(values = mapmag_bin_colors, name = "Mapped MAG (%)", drop = FALSE, guide = "none") +
  facet_nested(metric ~ size, scales = "free_x", space = "fixed", switch = "y",
               strip = strip_themed(
                 background_x = elem_list_rect(
                   fill = size_colors,
                   colour = "white",
                   size = 0.8),
                 background_y = elem_list_rect(
                   fill = "#e9eef4",
                   colour = "white",
                   size = 0.8))) +
  theme_minimal() +
  custom_ggplot_theme_LMD_size +
  labs(x = NULL, y = NULL) +
  theme(
    legend.position = "none",
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.minor.y = element_blank(),
    strip.text.x = element_blank(),
    strip.background.x = element_blank()
  )

fig2b_combined <- fig2b_top / fig2b_mapmag_panel +
  patchwork::plot_layout(heights = c(4, 0.3), guides = "collect") &
  theme(legend.position = "right")
fig2b_combined

```

```{r Fig2_mapped_mag_median, echo=TRUE,  warning = FALSE, comments = "", message = FALSE}
# Mapped MAG median
# Mapped MAG median
macroquant_metrics_filt_30_tidy_M14_pos_retained %>%
  group_by(size) %>%
  summarise(
    median_pct = median(.data[[prc_mapped_var]] * 100, na.rm = TRUE),
    IQR_pct    = IQR(.data[[prc_mapped_var]] * 100, na.rm = TRUE),
    Q1     = quantile(.data[[prc_mapped_var]] * 100, 0.25, na.rm = TRUE),
    Q3     = quantile(.data[[prc_mapped_var]] * 100, 0.75, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    label_txt = paste0(
      "", round(median_pct, 2), "\n",
      " [", round(Q1, 2), "–", round(Q3, 2), "]"
    )
  ) %>%
  knitr::kable(digits = 3)
```

### Fig. 2c
```{r Fig2c, warning = FALSE, echo=TRUE, comments = "", message = FALSE, fig.height = 6, fig.width = 8}
# Select data for PCA and make wide df 
LMD_size_df_wide <- macroquant_genome_counts_filt_30_tidy_M14_pos %>%
  select(genome, microsample, count,) %>% #  size
  pivot_wider(names_from = genome,values_from = count,values_fill = 0) %>%
  select(where(~ any(. != 0)) | where(is.character) | where(is.factor)) %>%  
  mutate(across(-c(microsample), ~ . / rowSums(across(where(is.numeric))) * 100 )) %>%
  column_to_rownames(var = "microsample")

LMD_size_df_wide <- remove_samples_or_taxa(df = LMD_size_df_wide, 
                                           min_samples_per_taxon = 6,
                                           min_taxa_per_sample = 14)

LMD_size_pca <- perform_pca(LMD_size_df_wide, z_delete = TRUE)

fig2c <- plot_pca(LMD_size_pca$pca_result,
         samples_color_metadata = "size", 
         samples_color_value = size_colors, 
         samples_shape_metadata = "animal",
         loadings_color_metadata = "order", 
         loadings_color_value = order_colors, 
         loadings_taxon_level = "species",
         sample_metadata = macroquant_metrics_filt_30, 
         genome_metadata = macromag_genomemetadata, 
         order_colors= order_colors, 
         scaling_factor_value = 1,
         custom_ggplot_theme = custom_ggplot_theme_LMD_size,
         loadings_number = 10,
         group_overlay = "kde",
         overlay_group_metadata = "size",
         overlay_legend = FALSE,
         kde_contour_type = "filled",
         kde_contour_var = "ndensity",
         kde_bins = 4,
         kde_breaks = c(0.20, 0.50, 0.65, 0.92),
         kde_adjust = c(1.5, 1.5),
         kde_alpha = 0.22,
         plot_outliers = FALSE,
         kde_exclude_outliers_from_fit = TRUE,
         percent_outliers = 0.1)
 
fig2c
```

#### PERMANOVA
```{r PERMANOVA_different_areas, echo = TRUE,  warning = FALSE, comments = "", message = FALSE}
LMD_size_dist_matrix <- vegdist(LMD_size_pca$df_clr_dist, method = "euclidean")

LMD_size_pca_metadata_adonis <- macroquant_metrics_filt_30 %>%
  filter(microsample %in% rownames(LMD_size_pca$df_clr_dist)) %>%
  arrange(match(microsample, rownames(LMD_size_pca$df_clr_dist)))
all(LMD_size_pca_metadata_adonis$microsample == rownames(LMD_size_pca$df_clr_dist))

adonis2_result <- adonis2(
  LMD_size_dist_matrix ~ size,
  data = LMD_size_pca_metadata_adonis,
  permutations = 9999,
  by = "terms"
)
adonis2_result %>%
  knitr::kable()

bd_size <- betadisper(LMD_size_dist_matrix, LMD_size_pca_metadata_adonis$size)
permutest(bd_size)

```


### Fig. 2d
```{r Fig2d, echo=TRUE, warning = FALSE, comments = "", message = FALSE, fig.height = 6, fig.width =8}
LMD_size_species_prev_raw <- macroquant_genome_counts_filt_30_tidy_M14_pos %>%
  filter(!is.na(species), species != "") %>%
  group_by(microsample, size, order, species) %>%
  summarise(relabun = sum(count, na.rm = TRUE), .groups = "drop")

# prevalence
size_sample_counts_prev <- macroquant_genome_counts_filt_30_tidy_M14_pos %>%
  distinct(size, microsample) %>%
  count(size, name = "n_samples")

species_prevalence_by_size_prev <- LMD_size_species_prev_raw %>%
  group_by(size, species) %>%
  summarise(n_present = n_distinct(microsample), .groups = "drop") %>%
  right_join(
    tidyr::expand_grid(
      size = size_sample_counts_prev$size,
      species = unique(LMD_size_species_prev_raw$species)
    ),
    by = c("size", "species")
  ) %>%
  mutate(n_present = dplyr::coalesce(n_present, 0L)) %>%
  left_join(size_sample_counts_prev, by = "size") %>%
  mutate(prevalence = n_present / n_samples)

# prevalence criterion
species_keep_prev <- species_prevalence_by_size_prev %>%
  group_by(species) %>%
  summarise(n_sizes_pass = sum(prevalence >= 0.20), .groups = "drop") %>%
  filter(n_sizes_pass >= 2) %>%
  pull(species)

# plotting table
# bars = prevalence, dots = median relative abundance within each size/species
species_prev_plot_df <- LMD_size_species_prev_raw %>%
  filter(species %in% species_keep_prev) %>%
  group_by(size, species, order) %>%
  summarise(
    median_relabun = median(relabun, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  left_join(
    species_prevalence_by_size_prev %>% select(size, species, prevalence),
    by = c("size", "species")
  )

species_prev_order <- species_prev_plot_df %>%
  group_by(species) %>%
  summarise(overall_prevalence = mean(prevalence, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(overall_prevalence)) %>%
  pull(species)

# top PCA-loading species
LMD_size_loading_species_prev <- LMD_size_pca$pca_result$rotation[, 1:2] %>%
  as.data.frame() %>%
  rownames_to_column(var = "genome") %>%
  mutate(
    PC1 = PC1 * 1.5,
    PC2 = PC2 * 1.5,
    abs_loading = sqrt(PC1^2 + PC2^2)
  ) %>%
  left_join(macromag_genomemetadata %>% select(genome, species, order), by = "genome") %>%
  filter(!is.na(species), species != "") %>%
  arrange(desc(abs_loading)) %>%
  group_by(species, order) %>%
  summarise(abs_loading = max(abs_loading, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(abs_loading)) %>%
  slice_head(n = 10)

loading_species_set_prev <- unique(LMD_size_loading_species_prev$species)

species_axis_labels_prev <- tibble(species = species_prev_order) %>%
  mutate(species_label = ifelse(species %in% loading_species_set_prev, paste0("* ", species), species))

bar_width_prev <- 0.85
point_max_size_prev <- 2.2

fig2d_species_prevalence <- species_prev_plot_df %>%
  left_join(species_axis_labels_prev, by = "species") %>%
  mutate(species_label = factor(species_label, levels = rev(species_axis_labels_prev$species_label))) %>%
  ggplot(aes(x = prevalence * 100, y = species_label)) +
    geom_col(aes(fill = order), width = bar_width_prev, alpha = 0.85, color = "black", linewidth = 0.1) +
    geom_point(aes(size = median_relabun), shape = 16, color = "black", alpha = 0.9) +
    facet_nested(. ~ size,
                 scales = "free", space = "free", switch = "y",
                 strip = strip_themed(
                   background_x = elem_list_rect(
                     fill = size_colors,
                     colour = "white",
                     size = 0.8))) +
    scale_fill_manual(values = order_colors, name = "Order") +
    scale_size_continuous(name = "Median rel. abun.", range = c(0.25, point_max_size_prev)) +
    scale_x_continuous(
      limits = c(0, 110),
      breaks = c(0, 25, 50, 75, 100),
      expand = expansion(mult = c(0, 0))
    ) +
    labs(x = "Prevalence (%)", y = NULL) +
    theme_minimal() +
    custom_ggplot_theme_LMD_size +
    theme(panel.spacing = unit(0.2, "lines"))

fig2d_species_prevalence
```


## Assessment of MSSM Performance Across Intestinal Regions 

### Summary Statistic

```{r section_stat,  echo=TRUE, warning = FALSE, comments = "", message = FALSE}
section_df <- macroquant_metrics_filt_30 %>%
  filter(
    batch %in% c("MSEB0006", "MSEB0010"),
    section %in% c("Caecum right", "Ileum", "Colon"),
    type %in% c("positive"),
    richness > 0
  ) %>%
  mutate(
    collection_attempts = as.factor(collection_attempts),
    buffer = as.factor(buffer),
    cycles = as.factor(cycles)
  )

section_stat <- section_df %>% 
  filter(!cycles %in% c("19")) %>%
  group_by(section) %>%
  summarise(
    `a. Low Quality Reads (%)`   = format_median_iqr(lowquality_reads_pct*100),
    `b. Reads Mapping Eukaryotic Host (%)`     = format_median_iqr(host_pct*100),
    `c. Reads Mapping MAG Catalogue (%)`       = format_median_iqr(mapmag_prokreads_pct*100),
    `d. Microbial Fraction  Based On SCG (%)`  = format_median_iqr(read_fraction_singlem*100),
    `e. Richness`  = format_median_iqr(richness),
    .groups = "drop"
  ) %>% 
  as.data.frame() %>% 
  column_to_rownames("section") %>% 
  t() 

knitr::kable(
  section_stat,
  booktabs = TRUE,
  align = "lccc",
  caption = "Median percentage (IQR)"
)
```

```{r boxplot_section_comparison,  echo=TRUE, warning = FALSE, comments = "", message = FALSE, fig.width=8, fig.height=8}
ordered_metrics <- c(
  "lowquality_reads_pct",
  "host_pct",
  "mapmag_prokreads_pct",
  "read_fraction_singlem",
  "richness"
)

boxplot_section_comparison <- section_df %>%
  filter(!cycles %in% c("19")) %>%
  mutate(
    lowquality_reads_pct = lowquality_reads_pct * 100,
    host_pct = host_pct * 100,
    mapmag_prokreads_pct = mapmag_prokreads_pct * 100,
    read_fraction_singlem = read_fraction_singlem * 100
  ) %>%
  pivot_longer(
    cols = -c(names(mssm_sample_metadata), "cov_filtering"),
    names_to = "metric",
    values_to = "value"
  ) %>%
  filter(metric %in% ordered_metrics) %>%
  mutate(
    metric = factor(metric, levels = ordered_metrics),
    metric = fct_recode(metric,
      "a. Low Quality \nReads (%)" = "lowquality_reads_pct",
      "b. Reads Mapping \nEukaryotic Host (%)" = "host_pct",
      "c. Reads Mapping \nMAG Catalogue (%)" = "mapmag_prokreads_pct",
      "d. Microbial Fraction  \nBased On SCG (%)" = "read_fraction_singlem",
      "e. Richness" = "richness"
    ),
    section = recode(section,
      "Caecum right" = "Caecum",
      "Colon" = "Colon"
    )
  ) %>%
  ggplot(aes(x = section, y = value, fill = section)) +
  geom_boxplot(alpha = 0.6) +
  facet_wrap(~metric, scales = "free_y", ncol = 3) +
  labs(
    fill = "",
    y = "",
    x = "Intestinal section"
  ) +
  # geom_violin(trim = TRUE, alpha = 0.7, position = position_dodge(0.8), na.rm = TRUE) +  # Violin plot
  geom_jitter(
    position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.8),
    size = 1.5, alpha = 0.3
  ) +
  scale_fill_manual(values = c("#2A9D8F", "#264653", "#E76F51")) +
  custom_ggplot_theme +
  theme(legend.position = "none")

boxplot_section_comparison
```

Library-building performance, evaluated using sequencing and diversity metrics, was highest in the caecum, lower in the colon, and poorest in the ileum.

### Statistical Comparison Of Sequencing Performance
**Note:** The fit_and_analyze_model(), format_anova() and run_metric() function needs to be loaded.

```{r section_test,  echo=TRUE, warning = FALSE, comments = "", message = FALSE}
metrics <- tibble::tribble(
  ~metric,                         ~response_var,                    ~model, ~distribution,
  "Low quality reads (%)",        "lowquality_reads_pct",          "glm", "quasibinomial",
  "Host mapping (%)",             "host_pct",                      "glm", "quasibinomial",
  "MAG mapping (%)",              "mapmag_prokreads_pct",         "glm", "quasibinomial",
  "Single-copy gene fraction (%)","read_fraction_singlem",        "glm", "quasibinomial",
  "Richness",           "richness",                            "glm",  "quasipoisson"

)

section_test <- metrics %>%
  purrr::pmap_dfr(function(metric, response_var, model, distribution) {
    run_metric(
      metric = metric,
      response_var = response_var,
      model = model,
      distribution = distribution,
      section_name = NULL,
      explanatory_var = "collection_attempts + section",
      term_name = "section",
      data = section_df %>%
        dplyr::filter(!cycles %in% c("19"))
    )
  }) %>%
  dplyr::select(report)

knitr::kable(
  section_test,
  booktabs = TRUE,
  align = "ll"
  )

```


```{r e_sectioncomp,  echo=TRUE, warning = FALSE, comments = "", message = FALSE}
chart.Correlation(section_df %>%
  filter(section %in% c("Ileum")) %>%
  select(
    "read_fraction_singlem",
    "mapmag_prokreads_pct"
  ), method = "spearman")
```
Single-copy gene analysis showed a moderate correlation with the proportion of reads mapping to the reference post-filtering. 

```{r animal_coverage,  echo=TRUE, warning = FALSE, comments = "", message = FALSE}
animal_section_df <- macroquant_metrics_filt_30 %>%
  filter(
    section %in% c("Caecum right", "Colon", "Ileum")
  ) %>% 
  group_by(animal, section, cov_filtering) %>%
  summarise(
    count = n(),
    .groups = "drop"
  )

ggplot(animal_section_df,
       aes(x = animal,
           y = count,
           fill = cov_filtering)) +
  geom_col(position = "stack") +
  facet_wrap(~section) +
  scale_fill_manual(
    values = c(
      "Retained By Filtering" = "#2A9D8F",
      "Excluded By Filtering" = "#E76F51"
    )
  ) +
  labs(
    x = "Animal",
    y = "Number of observations",
    fill = "Filtering status"
  ) +
  pnas_theme +
  theme(legend.position = "bottom")
```
MSSM performance at 5,000 μm² in the caecum of a 35-day-old (G121e) animal matched that of a 14-day-old (G103b) individual. Colon samples from the 14-day-old animal performed poorlydue to insufficient genome coverage.

### Confocal Microscopy Of DAPI-stained Microsections (Fig. S5a)

```{r figs5a,  echo=TRUE, warning = FALSE, comments = "", message = FALSE, fig.width=8, fig.height=9}
dapi <- dapi_df %>%
  group_by(Animal, Sequence, Age, Section) %>%
  summarise(
    n = n(),
    Distance_from_one_side_mn = mean(Distance_from_one_side, na.rm = TRUE),
    Distance_from_one_side_sd = ifelse(n > 1, sd(Distance_from_one_side), NA_real_),
    mean = mean(Bacterial_count),
    median = median(Bacterial_count),
    sd_na = ifelse(n > 1, sd(Bacterial_count), NA_real_), # If only 1 obs, return NA
    sd = ifelse(n() > 1, sd(Bacterial_count, na.rm = TRUE), 0), # Avoid NA SD
    sem = ifelse(n > 1, sd / sqrt(n), 0), # Set SEM to 0 if only 1 observation
    q1 = quantile(Bacterial_count, 0.25), # First quartile
    q3 = quantile(Bacterial_count, 0.75), # Third quartile
    .groups = "drop"
  )

summary <- dapi_df %>%
  group_by(Age, Section) %>%
  summarise(
    n = n(),
    mean = mean(Bacterial_count),
    median = median(Bacterial_count),
    sd_na = ifelse(n > 1, sd(Bacterial_count), NA_real_), # If only 1 obs, return NA
    sd = ifelse(n() > 1, sd(Bacterial_count, na.rm = TRUE), 0), # Avoid NA SD
    q1 = quantile(Bacterial_count, 0.25), # First quartile
    q3 = quantile(Bacterial_count, 0.75), # Third quartile
    .groups = "drop"
  )

figs5a <- ggplot(dapi, aes(x = Distance_from_one_side_mn, y = median, color = Section)) +
  geom_line(linewidth = 1) + # Mean line
  geom_ribbon(aes(ymin = q1, ymax = q3, fill = Section),
    alpha = 0.2
  ) + # SEM shading
  labs(
    title = "a",
    x = "Distance from Villi (μm)",
    y = "Bacterial Counts (Median)"
  ) +
  facet_wrap(~Age, nrow = 2, scale = "free_x") +
  scale_fill_manual(values = c("#2A9D8F", "#264653", "#E76F51"), drop = FALSE) +
  scale_color_manual(values = c("#2A9D8F", "#264653", "#E76F51"), drop = FALSE) +
  custom_ggplot_theme +
  theme(legend.position = "bottom")
figs5a
```

Quantification of bacterial cells within a 5,000 μm² area using QuPath provides visual confirmation that differences in bacterial density across intestinal regions (caecum, colon, and ileum) were the primary factor influencing MSSM performance.


## Microbial Genomic Reference Data
### Summary Statistics
#### Reference MAG Catalogue {-}
```{r stat_macro,  echo = TRUE, warning = FALSE, comments = "", message = FALSE}
print_median_iqr(macromag_genomemetadata$completeness, "Completeness:")
print_median_iqr(macromag_genomemetadata$contamination, "Contamination:")
```
#### Micro-scale MAG Catalogue {-}
```{r stat_micro,  echo = TRUE, warning = FALSE, comments = "", message = FALSE}
print_median_iqr(micromag_genomemetadata$completeness, "Completeness:")
print_median_iqr(micromag_genomemetadata$contamination, "Contamination:")
```

### Macro vs Micro MAGs: Completeness And Abundance (Fig. 3a)

```{r fig_3a, echo = TRUE, warning = FALSE, comments = "", message = FALSE, fig.height=5, fig.width=7}
# Load the genome metadata for macro- and micro- MAG catalogues
genome_biplot_data <- macromag_genomemetadata %>%
  select(genome,completeness, contamination, length) %>%
  mutate(dataset = "Macro-scale")  

genome_biplot_micro_data <- micromag_genomemetadata %>%
  select(genome, completeness, contamination, length) %>%
  mutate(dataset = "Micro-scale")  

# combine the dfs
combined_data <- bind_rows(genome_biplot_data, genome_biplot_micro_data)

# dynamic limits
completeness_limits <- range(combined_data$completeness, na.rm = TRUE)
contamination_limits <- range(combined_data$contamination, na.rm = TRUE)
if (diff(completeness_limits) == 0) completeness_limits <- completeness_limits + c(-0.5, 0.5)
if (diff(contamination_limits) == 0) contamination_limits <- contamination_limits + c(-0.5, 0.5)

# Plot the combined data
combined_plot <- ggplot(combined_data, aes(x = completeness, y = contamination, size = length, fill = dataset)) +
  scale_fill_manual(values = c("#E8BD50", "#B92D65")) +  
  geom_point(shape = 21, color = "black", alpha = 0.9) +
  scale_x_continuous(limits = completeness_limits) +
  scale_y_reverse(limits = rev(contamination_limits)) +
  labs(y = "Contamination %", x = "Completeness %", color = "MAG catalogue", size = "Genome length") +
  theme_minimal() +
  custom_ggplot_theme +
  theme(legend.position = "none")

# Plot contamination boxplot
genome_contamination <- combined_data %>%
            ggplot(aes(y=contamination, x=dataset, fill=dataset)) +
                    scale_y_reverse(limits = rev(contamination_limits)) +
                    geom_boxplot() +
                    theme_void() +
                    scale_fill_manual(values = c("#E8BD50", "#B92D65")) + 
                    guides(fill = "none") +  
                    theme(
                        axis.title.x = element_blank(),
                        axis.title.y = element_blank(),
                        axis.text.y=element_blank(),
                        axis.ticks.y=element_blank(),
                        axis.text.x=element_blank(),
                        axis.ticks.x=element_blank(),
                        plot.margin = unit(c(0, 0, 0.40, 0),"inches")) 

# Plot completeness boxplot
genome_completeness <- combined_data %>%
        ggplot(aes(x=completeness, y=dataset, fill = dataset)) +
                scale_x_continuous(limits = completeness_limits) +
                geom_boxplot() +
                theme_void() +
                scale_fill_manual(values = c("#E8BD50", "#B92D65")) +
                guides(fill = "none") +  
                theme(
                    axis.title.x = element_blank(),
                    axis.title.y = element_blank(),
                    axis.text.y=element_blank(),
                    axis.ticks.y=element_blank(),
                    axis.text.x=element_blank(),
                    axis.ticks.x=element_blank(),
                    plot.margin = unit(c(0, 0, 0, 0.50),"inches")) 

# align
genome_biplot <- combined_plot + theme(plot.margin = margin(5, 5, 5, 5))
genome_completeness <- genome_completeness + theme(plot.margin = margin(0, 0, 0, 0))
genome_contamination <- genome_contamination + theme(plot.margin = margin(0, 0, 0, 0))

# compose final fig
top_row <- genome_completeness + plot_spacer() + plot_layout(widths = c(15, 1.5))
bottom_row <- genome_biplot + genome_contamination + plot_layout(widths = c(15, 1.5))
fig3a <- (top_row / bottom_row + plot_layout(heights = c(5, 15))) + plot_layout(guides = "collect") &
  theme(legend.position = "bottom", legend.box = "vertical") 

fig3a
```

### Macro MAGs Overview (Fig. 3b)

```{r fig_3b, echo = TRUE, warning = FALSE, comments = "", message = FALSE, fig.height=12, fig.width=7}
fig3b <- MAG_tree_plot(genome_metadata = macromag_genomemetadata, tree = macromag_tree)
fig3b
```

### Micro MAGs Overview (Suppl.Fig. 6)

```{r suppl_fig_1, echo = TRUE, warning = FALSE, comments = "", message = FALSE, fig.height=12, fig.width=7}
figs1 <- MAG_tree_plot(genome_metadata = micromag_genomemetadata, tree = micromag_tree)
figs1
```

### Comparative Analysis Of Genome Catalogues

```{r catalogues_genomes, echo = TRUE, warning = FALSE, comments = "", message = FALSE}
cat("Number Genomes in Reference MAG Catalogue =", length(macromag_taxonomy$genome), "\n")
cat("Number Genomes in Micro-scale MAG Catalogue =", length(micromag_taxonomy$genome), "\n")
```

```{r catalogues_phylo, echo = TRUE, warning = FALSE, comments = "", message = FALSE}
cat_sample_id <- microquant_metrics_filt_30 %>% # select microquant dataset
  filter(
    batch %in% c("MSEB0011"),
    type %in% c("positive")
  ) %>% pull(microsample)

cat_phylo_count <- full_join(
  microquant_genome_counts_filt_30 %>%
    select(genome, all_of(cat_sample_id))%>%
  filter(rowSums(across(-genome)) != 0) %>%
    rename_with(~ paste0(.x, "_micro"), .cols = -genome),
  macroquant_genome_counts_filt_30 %>%
    select(genome, all_of(cat_sample_id)) %>%
  filter(rowSums(across(-genome)) != 0) %>%
    rename_with(~ paste0(.x, "_macro"), .cols = -genome),
  by = "genome"
) %>%
  mutate(across(
    where(is.numeric),
    ~ replace_na(.x, 0)
  )) %>%
  filter(if_any(where(is.numeric), ~ . != 0)) %>%
  select(
    genome,
    where(~ !is.numeric(.x) || sum(.x, na.rm = TRUE) != 0)
  ) %>%
  column_to_rownames("genome")

cat_phylo_metadata <- bind_rows(
  microquant_metrics_filt_30 %>%
    mutate(
      sample_id = microsample,
      microsample = paste0(microsample, "_micro"),
      dataset = "Micro-scale"
    ),
  macroquant_metrics_filt_30 %>%
    mutate(
      sample_id = microsample,
      microsample = paste0(microsample, "_macro"),
      dataset = "Reference"
    )
) %>%
  filter(microsample %in% c(names(cat_phylo_count))) %>%
  column_to_rownames("microsample")

cat_phylo_taxonomy <- bind_rows(micromag_taxonomy, macromag_taxonomy) %>%
  filter(genome %in% c(rownames(cat_phylo_count))) %>%
  column_to_rownames("genome")

cat_physeq <- phyloseq(
  cat_phylo_count %>% otu_table(., taxa_are_rows = TRUE),
  cat_phylo_metadata %>% sample_data(),
  cat_phylo_taxonomy %>% as.matrix() %>% tax_table()
)

cat_physeq_species <- tax_glom(cat_physeq, taxrank = "species")
if (nrow(data.frame(cat_physeq_species@tax_table) %>%
  rownames_to_column("genome") %>%
  count(species) %>%
  filter(n > 1)) > 0) {
  cat("warning: Some species appear more than once in the taxonomy table")
} else {
  cat("Verification: The combined macro and micro phyloseq object contains only unique species.")
}
cat_physeq_G121e <- subset_samples(cat_physeq_species, animal == "G121e")
cat_physeq_G121e <- prune_taxa(taxa_sums(cat_physeq_G121e) > 0, cat_physeq_G121e)
```

#### Read Counts 
```{r catalogues_read_count, echo = TRUE, warning = FALSE, comments = "", message = FALSE}
cat_read_df <- bind_rows(
  lapply(list(
    micro = microquant_read_counts_filt_30 %>%
      select(genome, all_of(cat_sample_id)),
    macro = macroquant_read_counts_filt_30 %>%
      select(genome, all_of(cat_sample_id))
  ), function(df) {
    df %>%
      summarise(across(-1, sum)) %>%
      pivot_longer(cols = everything(), names_to = "microsample", values_to = "number_reads")
  }),
  .id = "source"
) %>% pivot_wider(names_from = source, values_from = number_reads)

cat_cor_result <- cor.test(cat_read_df$micro, cat_read_df$macro, method = "spearman")
rho <- cat_cor_result$estimate
pval <- cat_cor_result$p.value
n <- length(cat_read_df$micro)
# cat(sprintf("Correlation Spearman's ρ = %.2f, p = %.3f, n = %d", rho, pval, n), "\n")

cat_read_count <- cat_read_df %>%
  ggplot(aes(x = macro, y = micro)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  stat_cor(
    method = "spearman", label.x = Inf, label.y = -Inf,
    hjust = 1.1, vjust = -1.1, size = 4
  ) + # adds the correlation
  custom_ggplot_theme +
  labs(
    x = "Mapped Reads \nReference Catalogue",
    y = "Mapped Reads \nMicro-scale Catalogue"
  )

print_median_iqr(cat_read_df$macro / 1e6, "Reference genome catalogue: ")
print_median_iqr(cat_read_df$micro / 1e6, "Micro-scale genome catalogue: ")
cat_cor_result 
```
Read mapping strongly correlated between catalogues, with slightly higher mapped read counts observed for the micro-scale catalogue (11.8 million reads (8.54–16.0) versus 10.58 (7.72–14.21). 

#### Alpha Diversity (Suppl.Fig. 7a)

```{r suppl_fig_7a, echo = TRUE, warning = FALSE, comments = "", message = FALSE}
cat_alpha_df <- sample_data(cat_physeq_species) %>%
  data.frame() %>%
  filter(richness > 0) %>%
  select(dataset, sample_id, richness, neutral, phylogenetic, animal) %>%
  rename(
    `MAG Catalogue` = dataset,
    microsample = sample_id
  )

cat_alpha_df_lg <- cat_alpha_df %>%
  pivot_longer(
    cols = c(richness, neutral, phylogenetic),
    names_to = "metric",
    values_to = "value"
  ) %>%
  mutate(
    metric = factor(metric, levels = c("richness", "neutral", "phylogenetic"))
  )

plot_df <- cat_alpha_df_lg %>%
  filter(
    metric != "phylogenetic",
    (metric == "neutral" & value > 1) |
      (metric == "richness" & value > 0)
  ) %>%
  mutate(
    animal = recode(animal, "G121e" = "G121", "G103b" = "G103"),
    metric = fct_recode(
      metric,
      "Richness" = "richness",
      "Neutral" = "neutral"
    )
  )

figs7_a <- ggplot(plot_df, aes(x = metric, y = value, fill = `MAG Catalogue`)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(
    aes(group = `MAG Catalogue`),
    position = position_jitterdodge(
      jitter.width = 0.2,
      dodge.width = 0.8
    ),
    size = 1.5,
    alpha = 0.3
  ) +
  stat_compare_means(
    aes(group = `MAG Catalogue`),
    method = "wilcox.test",
    paired = TRUE,
    label = "p.format"
  ) +
  facet_wrap2(~ animal) +
  scale_fill_manual(values = c("#5c7fba", "#79b1a3")) +
  pnas_theme
figs7_a
```
Mapping to the micro-scale catalogue resulted in significantly higher species richness and Shannon diversity in one of the two examined animals, whereas only species richness was significantly increased in the other, suggesting differential sensitivity to rare versus abundant taxa.

##### Taxa Overlap (Suppl.Fig. 8)
**Note:** The pivot_phylo() and compare_taxa_overlap() function needs to be loaded.
```{r catalogues_taxa_table, echo = TRUE, warning = FALSE, comments = "", message = FALSE}
cat_phylo_lg <- pivot_phylo(
  phyloseq_obj = subset_samples(
    cat_physeq_species
  ),
  glom = TRUE,
  tax_transform = TRUE,
  taxon_level = "species",
  tr_method = "compositional" # remove this 
)

cat_taxa <- compare_taxa_overlap(
  data = cat_phylo_lg,
  group_var = "dataset",
  group_a = "Micro-scale",
  group_b = "Reference",
  taxon_var = "species",
  facet_var = NULL,
  abundance_var = "abundance",
  sample_var = "microsample"
)

cat_taxa$plot_df %>%
  distinct(Detection, taxon) %>%
  count(Detection, name = "Number of species") %>%
  arrange(Detection)  %>%
  knitr::kable(
    col.names = c("Detection category", "Number of species"),
    align = c("l", "r"))

```

A greater number of unique taxa were detected for the microscale catalogue.

```{r catalogues_taxa_plot, echo = TRUE, warning = FALSE, comments = "", message = FALSE}
cat_taxa_plot <- ggplot(cat_taxa$plot_df,
       aes(x = prevalence,
           y = mean_abundance,
           color = Detection)) +
  geom_point(size = 2, alpha = 0.85)  +
  scale_color_manual(
    values = c(
      "Micro-scale only" = "#5c7fba", 
      "Reference only" = "#79b1a3",
      "Shared" = "grey80"
    )
  ) +
  scale_y_log10(labels = label_scientific()) +
  scale_x_continuous(labels = percent_format()) +
  labs(
    x = "Prevalence in microsample",
    y = "Mean relative abundance (log scale)"
    ) +
  pnas_theme + theme(legend.position = "bottom")
cat_taxa_plot
```

```{r suppl_fig_8, echo = TRUE,  warning = FALSE, comments = "", message = FALSE}
cat_taxa_animal <- compare_taxa_overlap(
  data = cat_phylo_lg,
  group_var = "dataset",
  group_a = "Micro-scale",
  group_b = "Reference",
  facet_var = "animal",
  taxon_var = "species",
  abundance_var = "abundance",
  sample_var = "microsample"
)
```


#### Beta Diversity (Suppl.Fig. 7b)
```{r suppl_fig_7b_data, echo = TRUE,  warning = FALSE, comments = "", message = FALSE}
cat_betacount_df <- data.frame(otu_table(cat_physeq_species)) %>%
  rownames_to_column("genome") %>%
  melt(id = "genome") %>%
  setNames(c("genome", "microsample", "count")) %>%
  left_join(data.frame(tax_table(cat_physeq_species)) %>%
    rownames_to_column("genome"), by = "genome") %>%
  select(species, microsample, count) %>%
  filter(count > 0) %>%
  pivot_wider(names_from = species, values_from = count, values_fill = 0) %>% # change here
  select(where(~ any(. != 0))) %>%
  mutate(across(-microsample, ~ . / rowSums(across(where(is.numeric))) * 100)) %>%
  column_to_rownames(var = "microsample")
```

##### Centered Log-Ratio (CLR) Transformation And Principal Component Analysis (PCA) {-}
**Note:** The perform_pca() function needs to be loaded.

```{r suppl_fig_7b, echo = TRUE,  warning = FALSE, comments = "", message = FALSE}
# cat("Zero replacement determined", "\n")
cat_beta_obj <- perform_pca(cat_betacount_df)
cat_beta_metadata <- data.frame(sample_data(cat_physeq_species)) %>%
  rownames_to_column("microsample") %>%
  filter(microsample %in% c(rownames(cat_beta_obj$df_clr))) %>%
  rename(
    `MAG Catalogue` = dataset,
    Animal = animal
  )

cat_scores <- rownames_to_column(as.data.frame(cat_beta_obj$pca_result$x), var = "microsample")
cat_scores <- left_join(cat_scores, cat_beta_metadata, by = join_by(microsample == microsample))
variance_explained <- (cat_beta_obj$pca_result$sdev^2) / sum(cat_beta_obj$pca_result$sdev^2) * 100
figs7_b <- ggplot() +
  geom_point(
    data = cat_scores,
    aes(x = PC1, y = PC2, group = microsample, color = `MAG Catalogue`, shape = Animal),
    size = 3, alpha = 0.5
  ) +
  scale_color_manual(values = c("#5c7fba", "#79b1a3")) + 
  geom_path(
    data = cat_scores,
    aes(x = PC1, y = PC2, group = sample_id),
    linewidth = 0.7, alpha = 0.8, color = "lightgrey"
  ) +
  labs(
    # title = "b",
    x = paste0("PC1: ", round(variance_explained[1], 0), "% variance explained"),
    y = paste0("PC2: ", round(variance_explained[2], 0), "% variance explained")
  ) +
  scale_x_continuous(limits = c(-max(abs(cat_scores$PC1)), max(abs(cat_scores$PC1)))) + # Calculate symmetric limits based on data
  scale_y_continuous(limits = c(-max(abs(cat_scores$PC2)), max(abs(cat_scores$PC2)))) + # Calculate symmetric limits based on data
  geom_hline(yintercept = 0, color = "darkgrey") + # Horizontal line at y = 0
  geom_vline(xintercept = 0, color = "darkgrey") + # Vertical line at x = 0
  pnas_theme +
  theme(legend.position = "bottom")
figs7 <- figs7_a / figs7_b +
  plot_layout(heights = c(1, 1)) +
  plot_annotation(tag_levels = "a") &
  theme(plot.margin = margin(5, 5, 5, 5))

figs7_b

#ggsave( "figures/figs7_PNAS.pdf",
#  plot = figs7, width = 7.5, height = 6.5, device = cairo_pdf
#)
#ggsave("figures/figs7_PNAS.png",
#  plot = figs7, width = 7.5, height = 6.5, dpi = 300
#)
```
The 35-day-old animal (G121e) showed the greatest species-level differences between catalogues.

##### Differential Abundance (Aldex2) {-}
```{r cat_aldex_plot, echo = TRUE, warning = FALSE, comments = "", message = FALSE}
structural_zero <- data.frame(otu_table(tax_glom(cat_physeq_G121e, taxrank = "species"))) %>%
  rownames_to_column("genome") %>%
  pivot_longer(cols = -c(genome), names_to = "microsample", values_to = "abundance") %>%
  mutate(dataset = str_sub(microsample, -5, -1)) %>%
  group_by(dataset, genome) %>%
  summarise(sum_abundance = sum(abundance)) %>%
  filter(sum_abundance == 0) %>%
  pull(genome)

cat_aldex_G121e_counts <- data.frame(otu_table(tax_glom(cat_physeq_G121e, "species"))) # %>%
# rownames_to_column("genome") %>%
# filter(!genome %in%c(structural_zero)) %>%
# column_to_rownames("genome")
cat_aldex_G121e_pseudo_counts <- round(cat_aldex_G121e_counts * 1e6)
cat_aldex_G121e_conditions <- data.frame(sample_data(tax_glom(cat_physeq_G121e, "species"))) %>% pull(dataset)
cat_aldex_G121e_results <- aldex(cat_aldex_G121e_pseudo_counts,
  cat_aldex_G121e_conditions,
  test = "t",
  effect = TRUE, # include effect sizes
  denom = "all", # normalization denominator
  mc.samples = 256, # number of Monte Carlo instances
  verbose = TRUE
)
# - we.ep: expected p-value from Welch's t-test
# - wi.ep: expected p-value from Wilcoxon test
# - we.eBH: BH-adjusted p-value (q-value)
# - effect: effect size (difference in clr means / variability)
# Filter
cat_aldex_G121e_sig <- cat_aldex_G121e_results[cat_aldex_G121e_results$we.eBH < 0.05, ]
cat_aldex_G121e_effect <- cat_aldex_G121e_sig[abs(cat_aldex_G121e_sig$effect) > 0.5, ]
cat_aldex <-
  cat_aldex_G121e_effect %>%
  rownames_to_column("genome") %>%
  left_join(data.frame(tax_table(tax_glom(cat_physeq_G121e, "species"))) %>% rownames_to_column("genome"), by = "genome")

cat_aldex_plot <- cat_aldex %>%
  mutate(structural_zero = ifelse(genome %in% c(structural_zero), "Present in only 1 group", "Present in both groups")) %>%
  ggplot(., aes(x = diff.btw, y = reorder(species, effect), fill = abs(effect) > 1)) +
  geom_bar(stat = "identity", show.legend = FALSE) +
  coord_flip() +
  labs(
    x = "Median difference in \nCLR values between groups",
    y = "Species"
  ) +
  scale_fill_manual(values = c("TRUE" = "#e35d51", "FALSE" = "darkgrey")) +
  custom_ggplot_theme +
  facet_wrap(~structural_zero, scale = "free") +
  theme(axis.text.x = element_text(angle = 40, hjust = 1))
cat_aldex_plot
```
Differentially abundant species in the 35-day-old animal (G121e) are primarily structural 0.

## Environmental and host contamination 

### Controls Sequencing Performance (Suppl.Fig. 9)
```{r suppl_fig_9,  echo=TRUE, warning = FALSE, comments = "", message = FALSE, fig.width=8, fig.height=11}
ordered_metrics <- c(
  "total_before_filtering.total_reads",
  "lowquality_reads_pct",
  "adapter_contamination_pct",
  "human_pct",
  "raw_total_sequences_mag",
  "mapmag_prokreads_pct"
)

figs9 <- macroquant_metrics_filt_30 %>%
  filter(!section %in% c("Ileum")) %>%
  mutate(
    batch = gsub("MSEB00", "B", batch),
    lowquality_reads_pct = lowquality_reads_pct * 100,
    adapter_contamination_pct = adapter_contamination_pct * 100,
    human_pct = human_pct * 100,
    mapmag_prokreads_pct = mapmag_prokreads_pct * 100
  ) %>%
  pivot_longer(
    cols = -c(names(mssm_sample_metadata), "cov_filtering"),
    names_to = "metric",
    values_to = "value"
  ) %>%
  filter(metric %in% ordered_metrics) %>%
  mutate(
    metric = factor(metric, levels = ordered_metrics),
    metric = fct_recode(metric,
      "a. Sequencing Yield Pre-quality Filtering" = "total_before_filtering.total_reads",
      "b. Low Quality Reads (%)" = "lowquality_reads_pct",
      "c. Adapter Contamination (% bases)" = "adapter_contamination_pct",
      "d. Reads Mapping Human (%)" = "human_pct",
      "e. Read Counts for Bacterial Quantification " = "raw_total_sequences_mag",
      "f. Reads Mapping MAG Catalogue (%)" = "mapmag_prokreads_pct"
    ),
    type = recode(type,
      "positive" = "Microsample",
      "negative" = "Negative"
    )
  ) %>%
  ggplot(aes(x = batch, y = value, fill = type)) +
  geom_boxplot(alpha = 0.6) +
  facet_wrap(~metric, scales = "free_y", ncol = 1) +
  labs(
    fill = "",
    y = "",
    x = "Batch"
  ) +
  # geom_violin(trim = TRUE, alpha = 0.7, position = position_dodge(0.8), na.rm = TRUE) +  # Violin plot
  geom_jitter(
    position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.8),
    size = 1, alpha = 0.3
  ) +
  scale_fill_manual(values = c("#79b1a3", "#E76F51")) +
  custom_ggplot_theme + 
  pnas_theme + theme(legend.position = "bottom") 
#ggsave("figures/figs9_PNAS.png",
#  plot = figs9, width = 7.5, height = 10, dpi = 300
#)

figs9
```
Although controls yielded sequencing data, they consistently exhibited poorer quality metrics.

### Host Contamination (Suppl.Fig. 10)
```{r host_contamination, echo=TRUE, fig.height=8, fig.width=8, message=FALSE, warning=FALSE, comments=""}
#format_median_iqr(macroquant_metrics_filt_30 %>%
#  filter(type == "positive", richness > 0) %>%
#  mutate(host_pct = host_pct * 100) %>%
#  pull(host_pct))

host_df <- macroquant_metrics_filt_30 %>%
  filter(type == "positive", richness > 0) %>%
  mutate(
    host = host_pct * 100,
    host_log = log(host + 1e-6)
  )

df_bin <- host_df  %>% # binned summary (geometric mean + CI)
  mutate(bin = ntile(distance_host, 10)) %>%
  group_by(bin) %>%
  summarise(
    distance_mid = mean(distance_host),
    mean_log = mean(host_log),
    mean_host = mean(host),
    se_log = sd(host_log) / sqrt(n()),
    lower = exp(mean_log - 1.96 * se_log),
    upper = exp(mean_log + 1.96 * se_log),
    mean_host = exp(mean_log),
    .groups = "drop"
  )

figs10 <- ggplot(host_df, aes(x = distance_host, y = host)) +
  geom_point(alpha = 0.3, size = 1.5) +
  geom_ribbon(
    data = df_bin,
    aes(x = distance_mid, ymin = lower, ymax = upper),
    inherit.aes = FALSE,
    fill = "grey70",
    alpha = 0.5
  ) +
  geom_line(
    data = df_bin,
    aes(x = distance_mid, y = mean_host),
    inherit.aes = FALSE,
    color = "black",
    size = 1
  ) +
  scale_y_log10() +
  labs(
    x = "Distance to host (µm)",
    y = "Host reads (%) (log scale)"
  ) +
  pnas_theme
#ggsave("figures/host_distance_final.pdf",
#  plot = p_raw, width = 7.5, height = 3.5, device = cairo_pdf
#)
#ggsave("figures/host_distance_final.png",
#   plot = p_raw,width = 7.5, height = 3.5, dpi = 300
#)
figs10 
```

## Discriminative power and replicability of MSSM data

### Centered Log-Ratio (CLR) Transformation And Principal Component Analysis (PCA, Suppl Fig. 5)
**Note:** The perform_pca() function needs to be loaded.

```{r mssm_discriminative_object, echo = TRUE,  warning = FALSE, comments = "", message = FALSE}
# cat("Zero replacement determined", "\n")
mssm_discriminative_df <- macroquant_genome_counts_filt_30 %>%
  select(genome, macroquant_metrics_filt_30 %>% 
    filter(
      batch %in% c("MSEB0011", "MSEB0060", "MSEB0061"),
      type %in% c("positive")
    ) %>% pull(microsample)) %>%
  select(where(~ !all(. == 0))) %>%
  melt(id = "genome") %>%
  setNames(c("genome", "microsample", "count")) %>%
  pivot_wider(names_from = genome, values_from = count, values_fill = 0) %>% 
  select(where(~ any(. != 0))) %>%
  mutate(across(-microsample, ~ . / rowSums(across(where(is.numeric))) * 100)) %>% 
  column_to_rownames(var = "microsample")
mssm_discriminative_obj <- perform_pca(mssm_discriminative_df)
mssm_discriminative_metadata <- macroquant_metrics_filt_30 %>%
  filter(microsample %in% c(rownames(mssm_discriminative_obj$df_clr_dist)))
```
```{r supplementary_figure_4, echo = TRUE,  warning = FALSE, comments = "", message = FALSE, fig.height=6, fig.width=8}
figs_4 <- plot_pca(mssm_discriminative_obj$pca_result,
         samples_color_metadata = "animal", 
         samples_shape_metadata = "cryosection",
         samples_color_value = c('#ffc273','#c1558b'), 
         loadings_color_metadata = "order", 
         loadings_color_value = order_colors, 
         loadings_taxon_level = "species",
         sample_metadata = mssm_discriminative_metadata, 
         genome_metadata = macromag_genomemetadata, 
         custom_ggplot_theme = custom_ggplot_theme, 
         order_colors, loadings_number = 20)
figs_4
```


### Permutational Multivariate Analysis of Variance (PERMANOVA)

#### Tests Homogeneity Of Dispersion Between Groups {-}
##### Animal {-}
```{r mssm_discriminative_betadispersion_animal, echo = TRUE,  warning = FALSE, comments = "", message = FALSE}
set.seed(123)
betadisper_animal <- vegdist(mssm_discriminative_obj$df_clr_dist, method = "euclidean") %>%
  betadisper(., mssm_discriminative_metadata %>% pull(animal))
knitr::kable(permutest(betadisper_animal, pairwise = TRUE, permutations = 9999)$tab, digits = 4)
```
##### Cryosection {-}

```{r mssm_discriminative_betadispersion_cryosection, echo = TRUE,  warning = FALSE, comments = "", message = FALSE}
betadisper_cryo <- vegdist(mssm_discriminative_obj$df_clr_dist, method = "euclidean") %>%
  betadisper(., mssm_discriminative_metadata %>% pull(cryosection))
knitr::kable(permutest(betadisper_cryo, pairwise = TRUE, permutations = 9999)$tab, digits = 4)
```

#### PERMANOVA

```{r mssm_discriminative_permanova, echo = TRUE,  warning = FALSE, comments = "", message = FALSE}
knitr::kable(vegdist(mssm_discriminative_obj$df_clr_dist, method = "euclidean") %>%
  adonis2(
    formula = . ~ animal * cryosection,
    data = mssm_discriminative_metadata,
    permutations = 9999, by = "terms"
  ) %>%
  broom::tidy(), digits = 4)
```
MSSM data across individuals revealed highly distinct microbial communities. In contrast, differences among sequential cryosections of the same individual, were negligible.

## MSSM compared to FISH and Macro-scale Metagenomics
### MSSM vs FISH
```{r MSSM_vs_FISH_plot_settings, echo = TRUE,  warning = FALSE, comments = "", message = FALSE}
mssm_vs_fish_batch <- "MSEB0011"
mssm_vs_fish_palette <- c("G103b" = "#4f5bd5", "G121e" = "#c1558b")
mssm_vs_fish_method_shapes <- c("FISH" = 1, "MSSM" = 2)

custom_ggplot_theme_fish_mssm <- custom_ggplot_theme +
  theme(
    legend.text = element_text(size = 7),
    legend.title = element_text(size = 7),
    axis.text = element_text(size = 5),
    axis.title = element_text(size = 7, face = "bold"),
    strip.text = element_text(size = 7, face = "bold", color = "black"),
    plot.title = element_text(size = 7, face = "bold", color = "black"),
    plot.margin = unit(c(0, 0, 0, 0), "cm")
  )
```

#### PCA (Fig. 4b)
```{r PCA_MSSM_vs_FISH, echo = TRUE,  warning = FALSE, comments = "", message = FALSE, fig.height=6, fig.width=8}
# FISH genus counts to sample x genus matrix
fish_df_wide <- fish_counts %>%
  select(-Total, -animal, -cryosection, -sample_type) %>%
  select(where(~ any(. != 0)) | where(is.character) | where(is.factor)) %>%
  mutate(across(-microsample, ~ . / rowSums(across(where(is.numeric))) * 100)) %>%
  column_to_rownames(var = "microsample")

# Matching MSSM matrix at genus level
mssm_df_wide <- macroquant_genome_counts_filt_30_tidy %>%
  filter(type == "positive", batch == mssm_vs_fish_batch) %>%
  select(domain, phylum, class, order, family, genus, genome, microsample, count) %>%
  group_by(microsample, genus) %>%
  summarise(genus_count = sum(count, na.rm = TRUE), .groups = "drop") %>%
  filter(genus != "") %>%
  mutate(count = genus_count) %>%
  select(-genus_count) %>%
  pivot_wider(names_from = genus, values_from = count, values_fill = 0) %>%
  select(where(~ any(. != 0)) | where(is.character) | where(is.factor)) %>%
  column_to_rownames(var = "microsample")

# Remove sparse rows/columns
mssm_df_wide <- remove_samples_or_taxa(df = mssm_df_wide,
  min_samples_per_taxon = 30,
  min_taxa_per_sample = 26
)

# Aggregate all non-target genera as "Other" to match the FISH comparison
mssm_df_wide <- mssm_df_wide %>%
  mutate(Other = rowSums(select(., -all_of(fish_selected_genera)), na.rm = TRUE)) %>%
  select(any_of(fish_selected_genera), Other) %>%
  mutate(across(where(is.numeric), ~ . / rowSums(across(where(is.numeric))) * 100))

# Combine matrices and align matching sample metadata
fish_mssm_df_wide <- bind_rows(fish_df_wide, mssm_df_wide)
fish_mssm_sample_metadata <- bind_rows(
  fish_sample_metadata,
  macroquant_metrics_filt_30 %>%
    filter(type == "positive", batch == mssm_vs_fish_batch) %>%
    select(microsample, animal, cryosection) %>%
    mutate(sample_type = "MSSM")
)

# PCA
fish_mssm_pca <- perform_pca(fish_mssm_df_wide, z_delete = TRUE)

# tidy CLR table
fish_mssm_df_clr_tidy <- fish_mssm_pca$df_clr_dist %>%
  rownames_to_column(var = "microsample") %>%
  pivot_longer(
    cols = -microsample,
    names_to = "genome",
    values_to = "relabun"
  ) %>%
  left_join(fish_genome_metadata, by = join_by(genome == genome)) %>%
  left_join(fish_mssm_sample_metadata, by = join_by(microsample == microsample))

fig4_b <- plot_pca(fish_mssm_pca$pca_result,
  samples_color_metadata = "animal",
  samples_shape_metadata = "sample_type",
  samples_color_value = unname(mssm_vs_fish_palette),
  loadings_color_metadata = "order",
  loadings_color_value = order_colors,
  loadings_taxon_level = "genus",
  sample_metadata = fish_mssm_sample_metadata,
  genome_metadata = fish_genome_metadata,
  order_colors = order_colors,
  scaling_factor_value = 1,
  custom_ggplot_theme = custom_ggplot_theme_fish_mssm,
  group_overlay = "kde",
  overlay_group_metadata = "animal",
  overlay_legend = FALSE,
  kde_contour_type = "filled",
  kde_contour_var = "ndensity",
  kde_bins = 6,
  kde_breaks = c(0.03, 0.06, 0.10, 0.16, 0.24, 0.35),
  kde_adjust = c(2.2, 2.2),
  kde_linewidth = 0.7,
  kde_alpha = 0.32,
  plot_outliers = FALSE,
  kde_exclude_outliers_from_fit = FALSE,
  percent_outliers = 0.10
)
fig4_b
```

#### PERMANOVA
```{r PERMANOVA_MSSM_vs_FISH, echo = TRUE,  warning = FALSE, comments = "", message = FALSE}
# Calculate Euclidean distance on CLR data
fish_mssm_dist_matrix <- vegdist(fish_mssm_pca$df_clr_dist, method = "euclidean")

# Make sure that filtered df and metadata have the same number & order of rows
fish_mssm_metadata_adonis <- fish_mssm_sample_metadata %>%
  filter(microsample %in% rownames(fish_mssm_pca$df_clr_dist)) %>%
  arrange(match(microsample, rownames(fish_mssm_pca$df_clr_dist)))
all(fish_mssm_metadata_adonis$microsample == rownames(fish_mssm_pca$df_clr_dist))

# Run PERMANOVA
adonis2_result <- adonis2(
  fish_mssm_dist_matrix ~ animal + sample_type,
  data = fish_mssm_metadata_adonis,
  permutations = 9999,
  by = "terms"
)
adonis2_result %>%
  knitr::kable()
```
#### Fig. 4a
```{r barplot_MSSM_vs_FISH,echo = TRUE,  warning = FALSE, comments = "", message = FALSE, fig.height=4, fig.width=11}
# Shared genera order based on average absolute MSSM-FISH median difference
median_diff_order <- fish_mssm_df_clr_tidy %>%
  group_by(animal, genome, sample_type) %>%
  summarise(median_clr = median(relabun, na.rm = TRUE), .groups = "drop") %>%
  tidyr::pivot_wider(names_from = sample_type, values_from = median_clr) %>%
  mutate(abs_median_diff = abs(MSSM - FISH)) %>%
  group_by(genome) %>%
  summarise(mean_abs_median_diff = mean(abs_median_diff, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(mean_abs_median_diff))

plot_df_fish_mssm <- fish_mssm_df_clr_tidy %>%
  mutate(genome = factor(genome, levels = median_diff_order$genome))

axis_limits <- plot_df_fish_mssm %>%
  summarise(
    x_min = min(relabun, na.rm = TRUE),
    x_max = max(relabun, na.rm = TRUE)
  )

agree_df <- plot_df_fish_mssm %>%
  group_by(animal, genome) %>%
  summarise(
    med_fish = median(relabun[sample_type == "FISH"], na.rm = TRUE),
    med_mssm = median(relabun[sample_type == "MSSM"], na.rm = TRUE),
    delta = med_mssm - med_fish,
    .groups = "drop"
  ) %>%
  mutate(genome = factor(genome, levels = levels(plot_df_fish_mssm$genome)))

delta_xlim <- max(0.05, max(abs(agree_df$delta), na.rm = TRUE))

make_abundance_panel <- function(animal_id, show_y = FALSE) {
  p <- plot_df_fish_mssm %>%
    filter(animal == animal_id) %>%
    ggplot(aes(y = genome, x = relabun)) +
    geom_point(
      aes(color = animal, shape = sample_type, group = sample_type),
      size = 1.2,
      alpha = 0.45,
      stroke = 0.45,
      position = position_jitterdodge(jitter.width = 0.045, jitter.height = 0.10, dodge.width = 0.70)
    ) +
    geom_boxplot(
      aes(group = interaction(genome, sample_type)),
      outlier.shape = NA,
      alpha = 0.95,
      width = 0.68,
      size = 0.35,
      color = "grey15",
      fill = NA,
      position = position_dodge(width = 0.70)
    ) +
    scale_color_manual(values = mssm_vs_fish_palette) +
    scale_shape_manual(values = mssm_vs_fish_method_shapes) +
    scale_x_continuous(
      limits = c(axis_limits$x_min, axis_limits$x_max),
      expand = expansion(mult = c(0.02, 0.02))
    ) +
    labs(
      title = animal_id,
      x = "CLR abundance",
      y = if (show_y) "Genus" else NULL,
      color = "Animal",
      shape = "Sample type"
    ) +
    custom_ggplot_theme_fish_mssm +
    theme(
      axis.text.y = element_text(size = 8),
      plot.title = element_text(size = 9, hjust = 0.5),
      panel.border = element_rect(color = "grey35", fill = NA, linewidth = 0.45),
      legend.position = "none"
    )

  if (!show_y) {
    p <- p +
      theme(
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank()
      )
  }

  p
}

make_agreement_panel <- function(animal_id) {
  agree_df %>%
    filter(animal == animal_id) %>%
    ggplot(aes(x = delta, y = genome, color = animal)) +
    annotate("rect", xmin = -0.25, xmax = 0.25, ymin = -Inf, ymax = Inf, fill = "grey70", alpha = 0.16) +
    geom_vline(xintercept = 0, linetype = 2, color = "grey35", linewidth = 0.35) +
    geom_segment(aes(x = 0, xend = delta, y = genome, yend = genome), linewidth = 0.4, alpha = 0.7) +
    geom_point(size = 2.2, shape = 8) +
    scale_color_manual(values = mssm_vs_fish_palette, guide = "none") +
    scale_x_continuous(limits = c(-1.15 * delta_xlim, 1.15 * delta_xlim), expand = expansion(mult = c(0, 0))) +
    labs(
      title = animal_id,
      x = "MSSM - FISH (median CLR)",
      y = NULL
    ) +
    custom_ggplot_theme_fish_mssm +
    theme(
      axis.text.y = element_blank(),
      axis.ticks.y = element_blank(),
      plot.title = element_text(size = 9, hjust = 0.5),
      panel.border = element_rect(color = "grey35", fill = NA, linewidth = 0.45),
      legend.position = "none"
    )
}

abundance_g103 <- make_abundance_panel("G103b", show_y = TRUE)
agreement_g103 <- make_agreement_panel("G103b")
abundance_g121 <- make_abundance_panel("G121e", show_y = FALSE)
agreement_g121 <- make_agreement_panel("G121e")

fig4_a <- abundance_g103 + agreement_g103 + abundance_g121 + agreement_g121 +
  plot_layout(widths = c(1.8, 1.1, 1.8, 1.1), guides = "collect") &
  theme(legend.position = "top")

fig4_a

# cat("Grey band in agreement panels indicates practical agreement zone (\u00b10.25 CLR).\n")
```


### MSSM vs Macro-scale Metagenomics

##### Taxa Overlap (Suppl.Fig. 11)
**Note:** The pivot_phylo() and compare_taxa_overlap() function needs to be loaded.
```{r mivsma_overlap, echo = TRUE,  warning = FALSE, comments = "", message = FALSE}
mivsma_phylo_lg <- pivot_phylo(
  phyloseq_obj = subset_samples(
    physeq_all,
    type == "positive" & !section %in% c("Colon", "Ileum")
  ),
  glom = TRUE,
  tax_transform = TRUE,
  taxon_level = "species",
  tr_method = "compositional"
) %>%
  filter(abundance > 0)

mivsma_overlap <- compare_taxa_overlap(
  data = mivsma_phylo_lg,
  group_var = "group",
  group_a = "Micro-scale Spatial Metagenomics",
  group_b = "Macro-scale Metagenomics",
  facet_var = "animal",
  taxon_var = "species",
  abundance_var = "abundance",
  sample_var = "microsample"
)

mivsma_overlap$plot_df %>%
  count(animal, Detection, name = "Number of species") %>%
  arrange(animal, Detection) %>%
  knitr::kable(
    col.names = c("Animal", "Detection category", "Number of species"),
    align = c("l", "l", "r")
  )
```

```{r fig_s11, echo = TRUE,  warning = FALSE, comments = "", message = FALSE}
fig_s11 <- ggplot(
  mivsma_overlap$plot_df %>% filter(group %in% c("Micro-scale Spatial Metagenomics")),
  aes(
    x = prevalence,
    y = mean_abundance,
    color = Detection
  )
) +
  geom_point(size = 3, alpha = 0.85) +
  facet_wrap(~ animal) +
  scale_color_manual(
    values = c(
      "Micro-scale Spatial Metagenomics only" = "#5c7fba",
      "Shared" = "grey80"
    )
  ) +
  scale_y_log10(labels = scales::label_scientific()) +
  scale_x_continuous(labels = scales::percent_format()) +
  labs(
    x = "Prevalence in microsample",
    y = "Mean relative abundance (log scale)",
    color = NULL
  ) +
  pnas_theme +
  theme(legend.position = "bottom")
#ggsave(
#  "figures/fig_s11.pdf",
#  plot = fig_s11,
#  width = 7.5,
#  height = 3.5,
#  device = cairo_pdf
#)

#ggsave(
#  "figures/fig_s11.png",
#  plot = fig_s11,
#  width = 7.5,
#  height = 3.5,
#  dpi = 300
#)
fig_s11
```


## Investigate Spatial Pattern

### Spatial and functional analysis of bacterial community structure

#### Data preparation
**Note:** The prepare_spatial_section() function needs to be loaded.
```{r data_spatial, echo = TRUE,  warning = FALSE, comments = "", message = FALSE, eval = FALSE}
comm_data <- macroquant_genome_counts_filt_30
genome_id <- macroquant_genome_counts_filt_30$genome

metadata <- macroquant_metrics_filt_30 %>%
  filter(
    type == "positive",
    cryosection %in% c(
      "G121eI114B", "G121eI117A", "G121eI118B",
      "G121eO306A", "G121eO307A", "G121eO308A"
    ),
    !is.na(Xcoord)
  )

table(metadata$cryosection, metadata$batch)

comm_data <- comm_data %>%
  select(-genome) %>%
  t() %>%
  as.data.frame() %>%
  setNames(genome_id)

comm_data <- comm_data %>%
  filter(rownames(.) %in% metadata$microsample) %>%
  arrange(match(rownames(.), metadata$microsample)) %>%
  filter(rowSums(.) != 0) %>%
  select(where(~ sum(.x, na.rm = TRUE) > 0))

metadata <- metadata %>%
  filter(microsample %in% rownames(comm_data)) %>%
  arrange(match(microsample, rownames(comm_data)))

paste0(
  "comm_data and metadata in the same order: ",
  all(rownames(comm_data) == metadata$microsample)
)

macromag_genomemetadata_spatial <- macromag_genomemetadata[
  match(colnames(comm_data), macromag_genomemetadata$genome),
  ,
  drop = FALSE
]

paste0(
  "comm_data and genome_metadata in the same order: ",
  all(colnames(comm_data) == macromag_genomemetadata_spatial$genome)
)

z.war <- 0.95
spatial_sections <- c(
  "G121eO306A",
  "G121eO307A",
  "G121eO308A",
  "G121eI117A",
  "G121eI118B",
  "G121eI114B"
)

## community matrices
spatial_datasets <- lapply(
  spatial_sections,
  function(x) {
    prepare_spatial_section(
      cryosection_id = x,
      comm_data = comm_data,
      metadata = metadata,
      z.warning = z.war
    )
  }
)
names(spatial_datasets) <- spatial_sections
```

```{r functional_ordination, echo = TRUE,  warning = FALSE, comments = "", message = FALSE, eval = FALSE}
#genome_list <- Reduce(
#  union,
#  lapply(spatial_datasets, function(x) colnames(x$comm_red))
#)

genome_list <- colnames(spatial_datasets$G121eI117A$comm_red)
genome_gifts_red <- macromag_gifts %>%
  as.data.frame() %>%
  select(-starts_with(c("S0", "D04"))) %>%
  filter(rownames(.) %in% genome_list)

genome_funct_red <- genome_gifts_red %>%
  as.data.frame() %>%
  rownames_to_column(var = "genome") %>%
  pivot_longer(-genome, names_to = "column", values_to = "value") %>%
  left_join(macromag_genomemetadata_spatial, by = join_by("genome")) %>%
  select(genome, column, value, length) %>%
  mutate(
    group = substr(column, 1, 3),
    log_length = log(length)
  ) %>%
  group_by(genome, group) %>%
  summarise(mean_value = mean(value, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = group, values_from = mean_value) %>%
  column_to_rownames(var = "genome")

gift_pcoa <- ape::pcoa(dist(genome_gifts_red))
```


```{r rlqESLTP, echo = TRUE,  warning = FALSE, comments = "", message = FALSE, eval = FALSE}
spatial_sections <- c(
  "G121eO306A",
  "G121eO307A",
  "G121eO308A",
  "G121eI117A",
  "G121eI118B",
  "G121eI114B"
)

correlogram_orders <- c(
  G121eO306A = 8,
  G121eO307A = 8,
  G121eO308A = 8,
  G121eI117A = 8,
  G121eI118B = 8,
  G121eI114B = 7
)

# Exponent (not the root index) applied to the relative abundances, so 0.5 is
# the square root and makes the transform Hellinger; 0.25 would be the 4th root.
root_power <- 0.5

spatial_rlq_results <- lapply(
  spatial_sections,
  function(x) {
    run_spatial_rlq(
      cryosection_id = x,
      comm_red = spatial_datasets[[x]]$comm_red,
      metadata_section = spatial_datasets[[x]]$metadata,
      gift_pcoa = gift_pcoa,
      genome_tree = macromag_tree,
      root = root_power,
      correlogram_order = correlogram_orders[[x]]
      )
  }
)
names(spatial_rlq_results) <- spatial_sections

head(spatial_rlq_results$G121eO306A$rlq_scores) #invert
head(spatial_rlq_results$G121eO307A$rlq_scores) #invert
head(spatial_rlq_results$G121eO308A$rlq_scores)
head(spatial_rlq_results$G121eI117A$rlq_scores) #invert
head(spatial_rlq_results$G121eI118B$rlq_scores) #invert
head(spatial_rlq_results$G121eI114B$rlq_scores)
```


```{r flip_axis, echo = TRUE,  warning = FALSE, comments = "", message = FALSE, eval = FALSE}
spatial_rlq_results_raw <- spatial_rlq_results
spatial_rlq_results_flipped <- spatial_rlq_results_raw
axes_to_flip <- list(
  G121eO306A = 1,
  G121eO307A = 1,
  G121eI117A = c(1, 2),
  G121eI118B = c(1, 2)
)

flip_rlq_axis <- function(x, ax = 1) {
  slots <- c(
    "lR", "mR", "lQ", "mQ",
    "lR_givenE", "lR_givenS",
    "lQ_givenT", "lQ_givenP"
  )
  
  for (s in slots) {
    if (!is.null(x[[s]]) && ncol(x[[s]]) >= ax) {
      x[[s]][, ax] <- -x[[s]][, ax]
    }
  }
  
  x
}

for (section_id in names(axes_to_flip)) {
  for (ax in axes_to_flip[[section_id]]) {
    spatial_rlq_results_flipped[[section_id]]$rlqmix <- flip_rlq_axis(
      spatial_rlq_results_flipped[[section_id]]$rlqmix,
      ax = ax
    )
  }
}

for (section_id in names(axes_to_flip)) {
  spatial_rlq_results_flipped[[section_id]]$rlq_scores <- data.frame(
    microsample = rownames(spatial_rlq_results_flipped[[section_id]]$rlqmix$lR),
    spatial_rlq_results_flipped[[section_id]]$rlqmix$lR
  )

  spatial_rlq_results_flipped[[section_id]]$rlq_scores$microsample <-
    rownames(spatial_datasets[[section_id]]$comm_red)[
      as.numeric(spatial_rlq_results_flipped[[section_id]]$rlq_scores$microsample)
    ]

  spatial_rlq_results_flipped[[section_id]]$rlq_scores <-
    spatial_rlq_results_flipped[[section_id]]$rlq_scores %>%
    left_join(
      spatial_datasets[[section_id]]$metadata %>%
        select(microsample, Xcoord, Ycoord),
      by = "microsample"
    )
}

head(spatial_rlq_results_flipped$G121eO306A$rlq_scores) 
head(spatial_rlq_results_flipped$G121eO307A$rlq_scores) 
head(spatial_rlq_results$G121eO308A$rlq_scores)
head(spatial_rlq_results_flipped$G121eI117A$rlq_scores) 
head(spatial_rlq_results_flipped$G121eI118B$rlq_scores) 
head(spatial_rlq_results$G121eI114B$rlq_scores)
spatial_rlq_results_flipped$G121eO306A$rlqmix$lQ
spatial_rlq_results_flipped$G121eO307A$rlqmix$lQ
spatial_rlq_results$G121eO308A$rlqmix$lQ
spatial_rlq_results_flipped$G121eI117A$rlqmix$lQ
spatial_rlq_results_flipped$G121eI118B$rlqmix$lQ
spatial_rlq_results$G121eI114B$rlqmix$lQ
```




### Subspecies Segregation 
#### Pangenome: Genes Presence/Absence  (Suppl.Fig. 12a)
```{r fig_s12a, echo = TRUE,  warning = FALSE, comments = "", message = FALSE, fig.width=8, fig.height=9}
wide_data <- left_join(genepa_pg, duplication_pg, by = "gene") %>%
  left_join(partition_pg, by = "gene") %>%
  pivot_longer(!c(gene, is_single_copy_marker, `Family Partition`), names_to = "Genome", values_to = "Presence/Absence") %>%
  mutate(`Presence/Absence` = as.character(`Presence/Absence`)) %>%
  mutate(`Presence/Absence` = case_when(
    `Presence/Absence` == "1" & is_single_copy_marker == "FALSE" ~ "2",
    TRUE ~ `Presence/Absence`
  )) %>%
  mutate(`Presence/Absence` = as.numeric(`Presence/Absence`)) %>%
  select(gene, Genome, `Presence/Absence`) %>%
  pivot_wider(names_from = Genome, values_from = `Presence/Absence`, values_fill = 0)
mat <- as.matrix(wide_data[, -1])
rownames(mat) <- wide_data$gene
gene_anno <- left_join(genepa_pg, duplication_pg, by = "gene") %>%
  left_join(partition_pg, by = "gene") %>%
  pivot_longer(!c(gene, is_single_copy_marker, `Family Partition`), names_to = "Genome", values_to = "Presence/Absence") %>%
  distinct(gene, `Family Partition`) %>%
  arrange(`Family Partition`) %>%
  column_to_rownames("gene")
mat <- mat[rownames(gene_anno), ]

ann_colors <- list(
  `Family Partition` = c(
    "cloud" = "#2A9D8F",
    "shell" = "#264653",
    "persistent" = "#E76F51"
  )
)

fig_s12a <- pheatmap(mat,
  color = c("0" = "grey93", "1" = "gray32", "2" = "#7570b3"), # Named colors
  annotation_row = gene_anno,
  annotation_colors = ann_colors,
  cluster_cols = TRUE,
  cluster_rows = FALSE,
  show_rownames = FALSE,
  legend_breaks = c(0, 1, 2),
  legend_labels = c("Absence", "Presence", "Multicopy")
)
fig_s12a 
```

#### Distribution Lawsonibacter Strains (Suppl.Fig. 12b)
```{r fig_s12b, echo = TRUE,  warning = FALSE, comments = "", message = FALSE}
macroquant_lawsonibacter <- macroquant_genome_counts_filt_30_98 %>%
  filter(genome %in% c(macromag_genomemetadata_98 %>%
    filter(genus == "Lawsonibacter") %>%
    pull(genome))) %>%
  pivot_longer(-genome, names_to = "microsample", values_to = "abundance") %>%
  left_join(macroquant_metrics_filt_30 %>%
    select(microsample, type, batch, cryosection, animal, Xcoord, Ycoord), by = "microsample") %>%
  filter(
    type == "positive",
    abundance > 0,
    !is.na(Xcoord),
    !is.na(Ycoord)
  ) %>%
  left_join(macromag_genomemetadata_98, by = "genome")

lawsonibacter_pie_df <- macroquant_lawsonibacter %>%
  group_by(microsample) %>%
  mutate(
    abundance = abundance / sum(abundance) # normalize counts to proportions
  ) %>%
  arrange(microsample, genome) %>%
  mutate(
    start = cumsum(lag(abundance, default = 0)) * 2 * pi,
    end = cumsum(abundance) * 2 * pi,
    r0 = 0,
    r = 75
  )

lawsonibacter_piecharts <- split(lawsonibacter_pie_df, lawsonibacter_pie_df$cryosection) %>%
  purrr::map(~ {
    ggplot(.x) +
      geom_arc_bar(
        aes(
          x0 = Xcoord, y0 = Ycoord, r0 = r0, r = r,
          start = start, end = end, fill = genome
        ),
        color = "black", linewidth = 0.1
      ) +
      coord_fixed() +
      ggtitle(unique(.x$cryosection)) +
      theme(legend.position = "none") +
      pnas_theme +
      scale_fill_manual(values = lawsonibacter_colors, drop = FALSE)
  })

fig_s12b <- wrap_plots(
  lawsonibacter_piecharts[c(
  "G103bI301A", "G121eI103A")]
,
  ncol = 3,
  guides = "collect"
) &
theme(
  legend.position = "none",
  plot.title = element_text(hjust = 0.5)
)
fig_s12b
```


### SNP level microdiversity (Lorikeet)

### Distribution popANI by Group (Suppl.Fig. 13)
```{r fig_s13, echo = TRUE,  warning = FALSE, comments = "", message = FALSE}
lawsonibacter_popANI <- map_dfr(names(ani_list), function(bin_id) {

  ani_mat <- ani_list[[bin_id]]

  ani_mat %>%
    as.data.frame() %>%
    rownames_to_column("microsample_1") %>%
    pivot_longer(
      cols = -microsample_1,
      names_to = "microsample_2",
      values_to = "popANI"
    ) %>%
    filter(
      !is.na(popANI),
      !is.nan(popANI)
    ) %>%
    mutate(
      bin = bin_id,
      sample_pair = if_else(
        microsample_1 < microsample_2,
        paste(microsample_1, microsample_2, sep = "_"),
        paste(microsample_2, microsample_1, sep = "_")
      )
    ) %>%
    distinct(bin, sample_pair, .keep_all = TRUE) %>%
    select(-sample_pair)

})

lawsonibacter_popANI_meta <- macroquant_metrics_filt_30 %>%
  filter(
    cryosection %in% c("G121eI103A", "G103bI301A"),
    type %in% c("positive"),
    cov_filtering %in% c("Retained By Filtering"),
    !is.na(Xcoord),
    !is.na(Ycoord)
    ) %>%
  select(cryosection, microsample, animal, Xcoord, Ycoord)

stat.test <- lawsonibacter_popANI %>%
  filter(
    !is.nan(popANI),
    !is.na(popANI),
    microsample_1 %in% c(lawsonibacter_popANI_meta$microsample),
    microsample_2 %in% c(lawsonibacter_popANI_meta$microsample)#,
    # bin %in% c("GPB_bin_000044") or bin %in% c("GPB_bin_000047","GPB_bin_000063")
  ) %>%
  left_join(macroquant_metrics_filt_30 %>% select(microsample, animal), by = c("microsample_1" = "microsample")) %>%
  rename(animal_microsample_1 = animal) %>%
  left_join(macroquant_metrics_filt_30 %>% select(microsample, animal), by = c("microsample_2" = "microsample")) %>%
  rename(animal_microsample_2 = animal) %>%
  mutate(group = ifelse(animal_microsample_1 == animal_microsample_2, animal_microsample_1, "G121e-G103b")) %>%
  group_by(bin) %>%
  wilcox_test(popANI ~ group,
              comparisons = list(
                c("G121e", "G103b"),
                c("G121e", "G121e-G103b"),
                c("G103b", "G121e-G103b")
              )) %>%
  adjust_pvalue(method = "BH")


fig_s13 <- lawsonibacter_popANI %>%
  filter(
    !is.nan(popANI),
    !is.na(popANI),
    microsample_1 %in% c(lawsonibacter_popANI_meta$microsample),
    microsample_2 %in% c(lawsonibacter_popANI_meta$microsample)#,
    # bin %in% c("GPB_bin_000044") or bin %in% c("GPB_bin_000047","GPB_bin_000063")
  ) %>%
  left_join(macroquant_metrics_filt_30 %>% select(microsample, animal), by = c("microsample_1" = "microsample")) %>%
  rename(animal_microsample_1 = animal) %>%
  left_join(macroquant_metrics_filt_30 %>% select(microsample, animal), by = c("microsample_2" = "microsample")) %>%
  rename(animal_microsample_2 = animal) %>%
  mutate(group = ifelse(animal_microsample_1 == animal_microsample_2, animal_microsample_1, "G121e-G103b")) %>%
  ggplot(aes(x = group, y = popANI, fill = group)) +
  geom_boxplot(aes(group = group), outlier.shape = NA) +
  geom_jitter(
    position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.8),
    size = 0.3, alpha = 0.3
  ) +
  scale_fill_manual(values = c("#B92D65", "#E8BD50", "darkgrey")) +
  pnas_theme +
  stat_compare_means(
  method = "wilcox.test",
  comparisons = list(
    c("G121e", "G103b"),
    c("G121e", "G121e-G103b"),
    c("G103b", "G121e-G103b")
  ),
  label = "p.signif",
  size = 4,
  p.adjust.method = "BH",
  hide.ns = TRUE,
  label.y.npc = c(
    0.78,  # lowest bracket
    0.85,  # middle bracket
    0.92   # highest bracket
  )
) +
  labs(
    # title = "a",
    x = "",
    #y = "popANI GPB_bin_000044"
  ) +
  theme(legend.position = "none") +
  scale_x_discrete(labels = c(
    "G103b" = "Within \nG103b", "G121e" = "Within \nG121e",
    "G121e-G103b" = "Between \nindividuals"
  )) +
  facet_wrap(~bin) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
#png("figures/fig_s13.png", width = 7.5, height = 4.5, units = "in", res = 300)
#fig_s13 
#dev.off()
fig_s13
```

### Heatmap popANI (Suppl.Fig. 14)
```{r Fig6_c, echo = TRUE,  warning = FALSE, comments = "", message = FALSE}
plot_popani_3panel <- function(bin_ids,
                               ani_list,
                               metadata,
                               animal_order = c("G103b", "G121e"),
                               cluster = FALSE,
                               file = "popani_3panel.pdf",
                               width = 8,
                               height = 12) {


  if (length(bin_ids) != 3) {
    stop("Provide exactly 3 bins.")
  }

  dir.create(dirname(file), showWarnings = FALSE, recursive = TRUE)
  all_samples <- Reduce(union, lapply(bin_ids, function(b) {
    rownames(ani_list[[b]])
  }))

  all_samples <- intersect(all_samples, metadata$microsample)

  meta <- metadata %>%
    filter(microsample %in% all_samples) %>%
    select(microsample, animal) %>%
    column_to_rownames("microsample") %>%
    rename(Animal = animal)

  meta$Animal <- factor(meta$Animal, levels = animal_order)
  meta <- meta[order(meta$Animal), , drop = FALSE]

  animal_cols <- setNames(
    c("#E8BD50", "#B92D65")[seq_along(animal_order)],
    animal_order
  )

  top_ha <- HeatmapAnnotation(
    Animal = meta$Animal,
    show_annotation_name = FALSE,
    col = list(Animal = animal_cols)
  )

  left_ha <- rowAnnotation(
    Animal = meta$Animal,
    show_annotation_name = FALSE,
    show_legend = FALSE,
    col = list(Animal = animal_cols)  
    )


  ht_list <- lapply(seq_along(bin_ids), function(i) {

    bin_id <- bin_ids[i]

    mat <- ani_list[[bin_id]][all_samples, all_samples]
    mat <- as.matrix(mat)
    mat <- mat[rownames(meta), rownames(meta)]

    bin_vals <- as.numeric(mat)
    bin_vals <- bin_vals[is.finite(bin_vals)]

    col_fun <- colorRamp2(
      c(min(bin_vals), max(bin_vals)),
      c("black", "white")
    )

    Heatmap(
      mat,
      name = bin_id,

      col = col_fun,

      top_annotation = top_ha,
      left_annotation = left_ha,

      cluster_rows = cluster,
      cluster_columns = cluster,

      show_row_names = FALSE,
      show_column_names = FALSE,

      row_title = bin_id,

      height = unit(8, "cm")
    )
  })

  ht_stack <- Reduce(`%v%`, ht_list)
  
  draw(
    ht_stack,
    heatmap_legend_side = "left",
    annotation_legend_side = "left",
    annotation_legend_list = NULL
  )

  pdf(paste0(file, ".pdf"), width = width, height = height)

  draw(
    ht_stack,
    heatmap_legend_side = "left",
    annotation_legend_side = "left",
    annotation_legend_list = NULL
  )

  dev.off()
  
  png(paste0(file,".png"), width = width, height = height, units = "in", res = 300)

  draw(
    ht_stack,
    heatmap_legend_side = "left",
    annotation_legend_side = "left",
    annotation_legend_list = NULL
  )

  dev.off()

  invisible(ht_stack)
}

plot_popani_3panel(
  bin_ids = c(
    "GPB_bin_000044",
    "GPB_bin_000047",
    "GPB_bin_000063"
  ),
  ani_list = ani_list,
  metadata = macroquant_metrics_filt_30 %>%
    filter(
      cryosection %in% c("G121eI103A", "G103bI301A"),
      type %in% c("positive"),
      cov_filtering %in% c("Retained By Filtering"),
      !is.na(Xcoord),
      !is.na(Ycoord)
    ),
  cluster = FALSE,
  file = "figures/fig_s13"
)
```




