# Method Validation

```{r load_data_validation, echo = FALSE,  warning = FALSE, comments = "", message = FALSE}
load("resources/working_data_object.Rdata") 
```

## Discriminative power and replicability of MSSM data

```{r mssm_discriminative_df, echo = FALSE,  warning = FALSE, comments = "", message = FALSE}
mssm_discriminative_df <- macroquant_genome_counts_filt_30 %>%
  select(genome, macroquant_metrics_filt_30 %>% 
    filter(
      batch %in% c("MSEB0011"),
      type %in% c("positive")
    ) %>% pull(microsample)) %>%
  select(where(~ !all(. == 0))) %>%
  melt(id = "genome") %>%
  setNames(c("genome", "microsample", "count")) %>%
  pivot_wider(names_from = genome, values_from = count, values_fill = 0) %>% 
  select(where(~ any(. != 0))) %>%
  mutate(across(-microsample, ~ . / rowSums(across(where(is.numeric))) * 100)) %>% 
  column_to_rownames(var = "microsample")
```

### Centered Log-Ratio (CLR) Transformation And Principal Component Analysis (PCA, Suppl Fig. 5)
**Note:** The perform_pca() function needs to be loaded.

```{r mssm_discriminative_object, echo = FALSE,  warning = FALSE, comments = "", message = FALSE}
cat("Zero replacement determined", "\n")
mssm_discriminative_obj <- perform_pca(mssm_discriminative_df)
```

```{r mssm_discriminative_metadata, echo = FALSE,  warning = FALSE, comments = "", message = FALSE}
mssm_discriminative_metadata <- macroquant_metrics_filt_30 %>%
  filter(microsample %in% c(rownames(mssm_discriminative_obj$df_clr_dist)))
```

```{r supplementary_figure_5, echo = FALSE,  warning = FALSE, comments = "", message = FALSE, fig.height=6, fig.width=8}
figs_5 <- plot_pca(mssm_discriminative_obj$pca_result,
         samples_color_metadata = "animal", 
         samples_shape_metadata = "cryosection",
         samples_color_value = c('#ffc273','#c1558b'), 
         loadings_color_metadata = "order", 
         loadings_color_value = order_colors, 
         loadings_taxon_level = "species",
         sample_metadata = mssm_discriminative_metadata, 
         genome_metadata = macromag_genomemetadata, 
         custom_ggplot_theme = custom_ggplot_theme, 
         order_colors, loadings_number = 20)

figs_5
```


### Permutational Multivariate Analysis of Variance (PERMANOVA)

#### Tests Homogeneity Of Dispersion Between Groups {-}
##### Animal {-}
```{r mssm_discriminative_betadispersion_animal, echo = FALSE,  warning = FALSE, comments = "", message = FALSE}
betadisper_animal <- vegdist(mssm_discriminative_obj$df_clr_dist, method = "euclidean") %>%
  betadisper(., mssm_discriminative_metadata %>% pull(animal))
knitr::kable(permutest(betadisper_animal, pairwise = TRUE)$tab, digits = 3)
```
##### Cryosection {-}

```{r mssm_discriminative_betadispersion_cryosection, echo = FALSE,  warning = FALSE, comments = "", message = FALSE}
betadisper_cryo <- vegdist(mssm_discriminative_obj$df_clr_dist, method = "euclidean") %>%
  betadisper(., mssm_discriminative_metadata %>% pull(cryosection))
knitr::kable(permutest(betadisper_cryo, pairwise = TRUE)$tab, digits = 3)
```

#### PERMANOVA

```{r mssm_discriminative_permanova, echo = FALSE,  warning = FALSE, comments = "", message = FALSE}
knitr::kable(vegdist(mssm_discriminative_obj$df_clr_dist, method = "euclidean") %>%
  adonis2(
    formula = . ~ animal + cryosection,
    data = mssm_discriminative_metadata,
    permutations = 999, by = "terms"
  ) %>%
  broom::tidy(), digits = 3)
```
MSSM data across individuals revealed highly distinct microbial communities. In contrast, differences among sequential cryosections of the same individual, were negligible.


## MSSM compared to FISH and Macro-scale Metagenomics

### MSSM vs FISH
```{r MSSM_vs_FISH_plot_settings, echo = FALSE,  warning = FALSE, comments = "", message = FALSE}
custom_ggplot_theme_fish_mssm <- custom_ggplot_theme +
  theme( 
        legend.text = element_text(size = 7),
        legend.title = element_text(size = 7),
        axis.text = element_text(size = 5),
        axis.title = element_text(size = 7, face = "bold"),
        strip.text = element_text(size = 7, face = "bold", color = "black"),
        plot.title = element_text(size = 7, face = "bold", color = "black"),
        plot.margin = unit(c(0, 0, 0, 0), "cm"))
```

#### PCA (Fig. 4b)
```{r PCA_MSSM_vs_FISH, echo = FALSE,  warning = FALSE, comments = "", message = FALSE, fig.height=6, fig.width=8}
# Load FISH counts as wide df
fish_df_wide <- fish_counts %>%
  select(-Total, -animal, -cryosection, -sample_type, -Other) %>%
  select(where(~ any(. != 0)) | where(is.character) | where(is.factor)) %>%  
  mutate(across(-c(microsample), ~ . / rowSums(across(where(is.numeric))) * 100 )) %>%
  column_to_rownames(var = "microsample")

# Create a tidy MSSM df with the relevant data
mssm_df_wide <- macroquant_genome_counts_filt_30_tidy %>%
  filter(type == "positive") %>%
  filter(batch == "MSEB0011") %>%
  # filter(mapmag_filteredreads_pct > 0.8) %>%
  select(domain, phylum, class, order, family, genus, genome, microsample, count) %>%
  group_by(microsample, genus) %>%
  summarise(genus_count = sum(count, na.rm = TRUE))%>%
  filter(genus != "") %>%
  mutate(count=genus_count) %>%
  select(-genus_count) %>%
  pivot_wider(names_from = genus,values_from = count,values_fill = 0) %>%
  select(where(~ any(. != 0)) | where(is.character) | where(is.factor)) %>%   
  column_to_rownames(var = "microsample") #%>%

# Remove almost empty samples or taxa from entire MSSM df
mssm_df_wide <- remove_samples_or_taxa(df = mssm_df_wide,
                                       min_samples_per_taxon = 30, 
                                       min_taxa_per_sample = 26)  
  
# Sum all other genera into "Other"
mssm_df_wide <- mssm_df_wide %>%
  mutate(Other = rowSums(select(., -all_of(fish_selected_genera)), na.rm = TRUE)) %>% 
  # Keep only the selected genera
  select(any_of(fish_selected_genera)) %>%
  mutate(across(where(is.numeric), ~ . / rowSums(across(where(is.numeric))) * 100))

# Combine the dfs for MSSM and FISH
fish_mssm_df_wide <- bind_rows(fish_df_wide, mssm_df_wide) 
fish_mssm_sample_metadata <- bind_rows(
  fish_sample_metadata,
  macroquant_metrics_filt_30 %>%
    filter(type == "positive", batch == "MSEB0011") %>%
    select(microsample, animal, cryosection) %>%
    mutate(sample_type = "MSSM")
)
# Perform PCA
fish_mssm_pca <- perform_pca(fish_mssm_df_wide, z_delete = TRUE)
fig4_b <- plot_pca(fish_mssm_pca$pca_result,
                   samples_color_metadata = "sample_type", 
                   samples_shape_metadata = "animal",
                   samples_color_value = c('#4f5bd5', '#c1558b'),
                   loadings_color_metadata = "order", 
                   loadings_color_value = order_colors, 
                   loadings_taxon_level = "genus",
                   sample_metadata = fish_mssm_sample_metadata, 
                   genome_metadata = fish_genome_metadata, 
                   scaling_factor_value = 1,
                   order_colors, 
                   custom_ggplot_theme_fish_mssm)
fig4_b
```

#### PERMANOVA
```{r PERMANOVA_MSSM_vs_FISH, echo = FALSE,  warning = FALSE, comments = "", message = FALSE}
# Calculate Euclidean distance on CLR data
fish_mssm_dist_matrix <- vegdist(fish_mssm_pca$df_clr_dist, method = "euclidean")

# Make sure that filtered df and metadata have the same number & order of rows
fish_mssm_metadata_adonis <- fish_mssm_sample_metadata %>%
  filter(microsample %in% rownames(fish_mssm_pca$df_clr_dist)) %>%
  arrange(match(microsample, rownames(fish_mssm_pca$df_clr_dist)))
all(fish_mssm_metadata_adonis$microsample == rownames(fish_mssm_pca$df_clr_dist))

# Run PERMANOVA
adonis2_result <- adonis2(
  fish_mssm_dist_matrix ~ animal + sample_type,
  data = fish_mssm_metadata_adonis,
  permutations = 999,
  by = "terms"
)
adonis2_result %>%
  knitr::kable()
```
#### Fig. 4a
```{r barplot_MSSM_vs_FISH,echo = FALSE,  warning = FALSE, comments = "", message = FALSE, fig.height=7, fig.width=7}
fish_mssm_df_CLR_tidy <- fish_mssm_pca$df_clr_dist %>%
  rownames_to_column(var = "microsample") %>%
  pivot_longer(
    cols = -microsample,
    names_to = "genome",
    values_to = "relabun"
  ) %>%
  left_join(fish_genome_metadata, by = join_by(genome == genome)) %>%
  left_join(fish_mssm_sample_metadata, by = join_by(microsample == microsample)) %>%
  mutate(
    phylum = factor(phylum, levels = unique(fish_genome_metadata$phylum)),
    order = factor(order, levels = unique(fish_genome_metadata$order))
  )

# Order the genera based on abundance in animal G121 for plot
summary <- fish_mssm_df_CLR_tidy %>%
  filter(animal == "G121e") %>%
  group_by(genome) %>%
  summarise(mean = mean(relabun, na.rm = T)) %>%
  arrange(-mean)

# Plot barplot of CLR values per animal, for MSSM and FISH
fig4_a <- fish_mssm_df_CLR_tidy %>%
  mutate(genome = factor(genome, levels = rev(summary %>% pull(genome)))) %>%
  ggplot(aes(x = genome, y = relabun)) +
  geom_boxplot(aes(fill = sample_type),
    outlier.shape = NA, alpha = 0.7, width = 0.6, size = 0.3,
    position = position_dodge(width = 0.7)
  ) +
  scale_fill_manual(values = c('#4f5bd5', '#c1558b')) +
  guides(fill = "none") + 
  new_scale_fill() +
  geom_jitter(aes(fill = genome, group = sample_type),
    size = 0.8, alpha = 0.6, shape = 21,
    color = "black",
    stroke = 0.3,
    position = position_jitterdodge(jitter.width = 0.6, dodge.width = 0.7)
  ) +
  scale_fill_manual(values = fish_genus_colors) +
  facet_nested(animal ~ ., scales = "fixed", space = "fixed", switch = "y") + # switch = "y"
  labs(x = "genus", y = "CLR") +
  theme_minimal() +
  custom_ggplot_theme_fish_mssm +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + # Rotate x-axis labels
  theme(legend.position = "bottom") +
  theme(panel.spacing = unit(0.6, "lines"))
fig4_a
```


### Macro-scale Metagenomics vs MSSM

#### Compare Presence/Absence At Genus level

```{r all_overlap, echo = FALSE,  warning = FALSE, comments = "", message = FALSE}
mivsma_phylo_lg <- pivot_phylo(
  phyloseq_obj = subset_samples(
    physeq_all,
    type == "positive" & !section == "Colon" & !section == "Ileum"
  ),
  glom = TRUE,
  tax_transform = TRUE,
  taxon_level = "genus",
  tr_method = "compositional"
) %>% filter(abundance > 0)
```

##### Shared Genera Overall {-}
```{r mivsma_genus, echo = FALSE,  warning = FALSE, comments = "", message = FALSE}
mivsma_micro <- unique(mivsma_phylo_lg %>% filter(
  group %in% c("Micro-scale Spatial Metagenomics")
) %>%
  pull(genus))
mivsma_macro <- unique(mivsma_phylo_lg %>%
  filter(group %in% c("Macro-scale Metagenomics")) %>%
  pull(genus))
mivsma_micro_unique <- setdiff(mivsma_micro, mivsma_macro)
mivsma_macro_unique <- setdiff(mivsma_macro, mivsma_micro)
mivsma_df <- tibble(
  genus = c(mivsma_micro_unique, mivsma_macro_unique),
  source = c(
    rep("Micro-scale Spatial Metagenomics", length(mivsma_micro_unique)),
    rep("Macro-scale Metagenomics", length(mivsma_macro_unique))
  )
)

total <- length(unique(mivsma_phylo_lg %>% pull(genus)))

cat(
  "Shared =",
  length(intersect(mivsma_micro, mivsma_macro)),
  "(", round(length(intersect(mivsma_micro, mivsma_macro)) / total * 100, 0), "%)",
  "\n"
)
cat(
  "Unique in Micro-scale Spatial Metagenomics =", length(mivsma_micro_unique),
  "(", round(length(mivsma_micro_unique) / total * 100, 0), "%)",
  "\n"
)
cat(
  "Unique in Macro-scale Metagenomics =", length(mivsma_macro_unique),
  "(", round(length(mivsma_macro_unique) / total * 100, 0), "%)",
  "\n"
)
```

##### Shared Genera 35 Days Animal (G121e, Supplementary Table 1) {-}
```{r mivsma_35d_genus, echo = FALSE,  warning = FALSE, comments = "", message = FALSE}
mivsma_35d_micro <- unique(mivsma_phylo_lg %>% filter(
  group %in% c("Micro-scale Spatial Metagenomics"),
  animal %in% c("G121e")
) %>% pull(genus))
mivsma_35d_macro <- unique(mivsma_phylo_lg %>% filter(
  group %in% c("Macro-scale Metagenomics"),
  animal %in% c("G121e")
) %>% pull(genus))
mivsma_35d_micro_unique <- setdiff(mivsma_35d_micro, mivsma_35d_macro)
mivsma_35d_macro_unique <- setdiff(mivsma_35d_macro, mivsma_35d_micro)
mivsma_35d_df <- tibble(
  genus = c(mivsma_35d_micro_unique, mivsma_35d_macro_unique),
  source = c(
    rep("Micro-scale Spatial Metagenomics", length(mivsma_35d_micro_unique)),
    rep("Macro-scale Metagenomics", length(mivsma_35d_macro_unique))
  )
)

total <- length(unique(mivsma_phylo_lg %>% 
                         filter(animal %in% c("G121e")) %>% 
                         pull(genus)))

cat("Shared =", 
    length(intersect(mivsma_35d_micro, mivsma_35d_macro)), 
    "(", round(length(intersect(mivsma_35d_micro, mivsma_35d_macro)) / total * 100, 0),"%)",
    "\n")
cat("Unique in Micro-scale Spatial Metagenomics =", 
    length(mivsma_35d_micro_unique), 
    "(", round(length(mivsma_35d_micro_unique) / total * 100, 0),"%)",
    "\n")
cat("Unique in Macro-scale Metagenomics =", 
    length(mivsma_35d_macro_unique), 
    "(", round(length(mivsma_35d_macro_unique) / total * 100, 0),"%)",
    "\n")
```


```{r sup_table_1, echo = FALSE,  warning = FALSE, comments = "", message = FALSE}
total_microsamples <- length(unique(mivsma_phylo_lg %>% filter(
  group %in% c("Micro-scale Spatial Metagenomics"),
  animal %in% c("G121e"),
  !section %in% c("Colon", "Ileum")
) %>% pull(microsample)))

knitr::kable(mivsma_phylo_lg %>%
  filter(
    group %in% c("Micro-scale Spatial Metagenomics"),
    animal %in% c("G121e"),
    !section %in% c("Colon", "Ileum"),
    genus %in% mivsma_35d_micro_unique
  ) %>%
  group_by(genus) %>%
  summarise(
    # Prevalence = sum(abundance > 0),
    Prevalence_Total = round(sum(abundance > 0) / total_microsamples, 3),
    mean_abundance = round(mean(abundance), 3)
  ))
```

##### Shared Genera 14 Days Animal (G103b, Fig. 5a) {-}
```{r mivsma_14d_genus, echo = FALSE,  warning = FALSE, comments = "", message = FALSE}
mivsma_14d_micro <- unique(mivsma_phylo_lg %>% filter(
  group %in% c("Micro-scale Spatial Metagenomics"),
  animal %in% c("G103b")
) %>% pull(genus))
mivsma_14d_macro <- unique(mivsma_phylo_lg %>% filter(
  group %in% c("Macro-scale Metagenomics"),
  animal %in% c("G103b")
) %>% pull(genus))
mivsma_14d_micro_unique <- setdiff(mivsma_14d_micro, mivsma_14d_macro)
mivsma_14d_macro_unique <- setdiff(mivsma_14d_macro, mivsma_14d_micro)
mivsma_14d_df <- tibble(
  genus = c(mivsma_14d_micro_unique, mivsma_14d_macro_unique),
  source = c(
    rep("Micro-scale Spatial Metagenomics", length(mivsma_14d_micro_unique)),
    rep("Macro-scale Metagenomics", length(mivsma_14d_macro_unique))
  )
)

total <- length(unique(mivsma_phylo_lg %>% 
                         filter(animal %in% c("G103b")) %>% 
                         pull(genus)))


cat("Shared =", 
    length(intersect(mivsma_14d_micro, mivsma_14d_macro)), 
    "(", round(length(intersect(mivsma_14d_micro, mivsma_14d_macro)) / total * 100, 0), "%)",
    "\n")
cat("Unique in Micro-scale Spatial Metagenomics =", 
    length(mivsma_14d_micro_unique), 
    "(", round(length(mivsma_14d_micro_unique) / total * 100, 0),"%)",
    "\n")
cat("Unique in Macro-scale Metagenomics =", 
    length(mivsma_14d_macro_unique), 
    "(", round(length(mivsma_14d_macro_unique) / total * 100, 0), "%)",
    "\n")
```

```{r fig5a, echo = FALSE,  warning = FALSE, comments = "", message = FALSE, fig.width=13, fig.height=7}
mivsma_14d_phylo <- subset_samples(
  physeq_all,
  type == "positive" &
    !section == "Colon" &
    !section == "Ileum" &
    animal == "G103b"
)
mivsma_14d_phylo <- prune_taxa(taxa_sums(mivsma_14d_phylo) > 0, mivsma_14d_phylo)

cols1 <- c("#ba2760","#E9C46A")
names(cols1) <- unique(samdat_tbl(mivsma_14d_phylo)$group)
fig5a <- tax_transform(mivsma_14d_phylo, "binary", rank = "genus") %>%
  comp_heatmap(
    sample_anno = sampleAnnotation(
      "Dataset" = anno_sample_cat(
        "group",
        col = cols1,
        box_col = NA,
        border_col = "black",
        legend_title = "Dataset"
      ),
      "Sequencing Yield" = anno_sample("total_before_filtering.total_reads"),
      border = TRUE,
      show_legend = c(TRUE)
    ), # Show sample annotation legend
    sample_names_show = FALSE,
    tax_anno = taxAnnotation(
      Prev. = anno_tax_prev(),
      Abun. = anno_tax_box()
    ),
    show_heatmap_legend = FALSE,
    heatmap_legend_param = list(
      title_position = "lefttop", # Customize heatmap legend title position if needed
      legend_direction = "horizontal", # Move heatmap legend horizontally
      legend_width = unit(6, "cm") # Adjust the heatmap legend width
    )
  )
fig5a
```
