# Method Validation

```{r load_data_validation, echo = FALSE,  warning = FALSE, comments = "", message = FALSE}
load("resources/working_data_object.Rdata") 
```

## Discriminative power and replicability of MSSM data

```{r mssm_discriminative_df, echo = FALSE,  warning = FALSE, comments = "", message = FALSE}
mssm_discriminative_df <- macroquant_genome_counts_filt_30 %>%
  select(genome, macroquant_metrics_filt_30 %>% 
    filter(
      batch %in% c("MSEB0011"),
      type %in% c("positive")
    ) %>% pull(microsample)) %>%
  select(where(~ !all(. == 0))) %>%
  melt(id = "genome") %>%
  setNames(c("genome", "microsample", "count")) %>%
  pivot_wider(names_from = genome, values_from = count, values_fill = 0) %>% 
  select(where(~ any(. != 0))) %>%
  mutate(across(-microsample, ~ . / rowSums(across(where(is.numeric))) * 100)) %>% 
  column_to_rownames(var = "microsample")
```

### Centered Log-Ratio (CLR) Transformation And Principal Component Analysis (PCA, Suppl Fig. 5)
**Note:** The perform_pca() function needs to be loaded.

```{r mssm_discriminative_object, echo = FALSE,  warning = FALSE, comments = "", message = FALSE}
cat("Zero replacement determined", "\n")
mssm_discriminative_obj <- perform_pca(mssm_discriminative_df)
```

```{r mssm_discriminative_metadata, echo = FALSE,  warning = FALSE, comments = "", message = FALSE}
mssm_discriminative_metadata <- macroquant_metrics_filt_30 %>%
  filter(microsample %in% c(rownames(mssm_discriminative_obj$df_clr_dist)))
```

```{r supplementary_figure_5, echo = FALSE,  warning = FALSE, comments = "", message = FALSE}
autoplot(mssm_discriminative_obj$pca_result, data = mssm_discriminative_metadata, colour = "animal", shape = "cryosection", size = 3, alpha = 0.5, loadings = FALSE, scale = 0) + custom_ggplot_theme
```


### Permutational Multivariate Analysis of Variance (PERMANOVA)

#### Tests Homogeneity Of Dispersion Between Groups {-}
##### Animal {-}
```{r mssm_discriminative_betadispersion_animal, echo = FALSE,  warning = FALSE, comments = "", message = FALSE}
betadisper_animal <- vegdist(mssm_discriminative_obj$df_clr_dist, method = "euclidean") %>%
  betadisper(., mssm_discriminative_metadata %>% pull(animal))
knitr::kable(permutest(betadisper_animal, pairwise = TRUE)$tab, digits = 3)
```
##### Cryosection {-}

```{r mssm_discriminative_betadispersion_cryosection, echo = FALSE,  warning = FALSE, comments = "", message = FALSE}
betadisper_cryo <- vegdist(mssm_discriminative_obj$df_clr_dist, method = "euclidean") %>%
  betadisper(., mssm_discriminative_metadata %>% pull(cryosection))
knitr::kable(permutest(betadisper_cryo, pairwise = TRUE)$tab, digits = 3)
```

#### PERMANOVA

```{r mssm_discriminative_permanova, echo = FALSE,  warning = FALSE, comments = "", message = FALSE}
knitr::kable(vegdist(mssm_discriminative_obj$df_clr_dist, method = "euclidean") %>%
  adonis2(
    formula = . ~ animal + cryosection,
    data = mssm_discriminative_metadata,
    permutations = 999, by = "terms"
  ) %>%
  broom::tidy(), digits = 3)
```
MSSM data across individuals revealed highly distinct microbial communities. In contrast, differences among sequential cryosections of the same individual, were negligible.


## MSSM compared to FISH and Macro-scale Metagenomics

### MSSM vs FISH
```{r x3, echo = FALSE,  warning = FALSE, comments = "", message = FALSE}
```

### Macro-scale Metagenomics vs MSSM

#### Compare Presence/Absence At Genus level

```{r all_overlap, echo = FALSE,  warning = FALSE, comments = "", message = FALSE}
mivsma_phylo_lg <- pivot_phylo(
  phyloseq_obj = subset_samples(
    physeq_all,
    type == "positive" & !section == "Colon" & !section == "Ileum"
  ),
  glom = TRUE,
  tax_transform = TRUE,
  taxon_level = "genus",
  tr_method = "compositional"
) %>% filter(abundance > 0)
```

##### Shared Genera Overall {-}
```{r mivsma_genus, echo = FALSE,  warning = FALSE, comments = "", message = FALSE}
mivsma_micro <- unique(mivsma_phylo_lg %>% filter(
  group %in% c("Micro-scale Spatial Metagenomics")
) %>%
  pull(genus))
mivsma_macro <- unique(mivsma_phylo_lg %>%
  filter(group %in% c("Macro-scale Metagenomics")) %>%
  pull(genus))
mivsma_micro_unique <- setdiff(mivsma_micro, mivsma_macro)
mivsma_macro_unique <- setdiff(mivsma_macro, mivsma_micro)
mivsma_df <- tibble(
  genus = c(mivsma_micro_unique, mivsma_macro_unique),
  source = c(
    rep("Micro-scale Spatial Metagenomics", length(mivsma_micro_unique)),
    rep("Macro-scale Metagenomics", length(mivsma_macro_unique))
  )
)

total <- length(unique(mivsma_phylo_lg %>% pull(genus)))

cat(
  "Shared Genera =",
  length(intersect(mivsma_micro, mivsma_macro)),
  "(", round(length(intersect(mivsma_micro, mivsma_macro)) / total * 100, 0), "%)",
  "\n"
)
cat(
  "Unique Genera in Micro-scale Spatial Metagenomics =", length(mivsma_micro_unique),
  "(", round(length(mivsma_micro_unique) / total * 100, 0), "%)",
  "\n"
)
cat(
  "Unique Genera in Macro-scale Metagenomics =", length(mivsma_macro_unique),
  "(", round(length(mivsma_macro_unique) / total * 100, 0), "%)",
  "\n"
)
```

##### Shared Genera 35 Days Animal (G121e, Supplementary Table 1) {-}
```{r mivsma_35d_genus, echo = FALSE,  warning = FALSE, comments = "", message = FALSE}
mivsma_35d_micro <- unique(mivsma_phylo_lg %>% filter(
  group %in% c("Micro-scale Spatial Metagenomics"),
  animal %in% c("G121e")
) %>% pull(genus))
mivsma_35d_macro <- unique(mivsma_phylo_lg %>% filter(
  group %in% c("Macro-scale Metagenomics"),
  animal %in% c("G121e")
) %>% pull(genus))
mivsma_35d_micro_unique <- setdiff(mivsma_35d_micro, mivsma_35d_macro)
mivsma_35d_macro_unique <- setdiff(mivsma_35d_macro, mivsma_35d_micro)
mivsma_35d_df <- tibble(
  genus = c(mivsma_35d_micro_unique, mivsma_35d_macro_unique),
  source = c(
    rep("Micro-scale Spatial Metagenomics", length(mivsma_35d_micro_unique)),
    rep("Macro-scale Metagenomics", length(mivsma_35d_macro_unique))
  )
)

total <- length(unique(mivsma_phylo_lg %>% 
                         filter(animal %in% c("G121e")) %>% 
                         pull(genus)))

cat("Shared Genera =", 
    length(intersect(mivsma_35d_micro, mivsma_35d_macro)), 
    "(", round(length(intersect(mivsma_35d_micro, mivsma_35d_macro)) / total * 100, 0),"%)",
    "\n")
cat("Unique Genera in Micro-scale Spatial Metagenomics =", 
    length(mivsma_35d_micro_unique), 
    "(", round(length(mivsma_35d_micro_unique) / total * 100, 0),"%)",
    "\n")
cat("Unique Genera in Macro-scale Metagenomics =", 
    length(mivsma_35d_macro_unique), 
    "(", round(length(mivsma_35d_macro_unique) / total * 100, 0),"%)",
    "\n")
```


```{r sup_table_1, echo = FALSE,  warning = FALSE, comments = "", message = FALSE}
total_microsamples <- length(unique(mivsma_phylo_lg %>% filter(
  group %in% c("Micro-scale Spatial Metagenomics"),
  animal %in% c("G121e"),
  !section %in% c("Colon", "Ileum")
) %>% pull(microsample)))

knitr::kable(mivsma_phylo_lg %>%
  filter(
    group %in% c("Micro-scale Spatial Metagenomics"),
    animal %in% c("G121e"),
    !section %in% c("Colon", "Ileum"),
    genus %in% mivsma_35d_micro_unique
  ) %>%
  group_by(genus) %>%
  summarise(
    # Prevalence = sum(abundance > 0),
    Prevalence_Total = round(sum(abundance > 0) / total_microsamples, 3),
    mean_abundance = round(mean(abundance), 3)
  ))
```

##### Shared Genera 14 Days Animal (G103b, Fig. 5a) {-}
```{r mivsma_14d_genus, echo = FALSE,  warning = FALSE, comments = "", message = FALSE}
mivsma_14d_micro <- unique(mivsma_phylo_lg %>% filter(
  group %in% c("Micro-scale Spatial Metagenomics"),
  animal %in% c("G103b")
) %>% pull(genus))
mivsma_14d_macro <- unique(mivsma_phylo_lg %>% filter(
  group %in% c("Macro-scale Metagenomics"),
  animal %in% c("G103b")
) %>% pull(genus))
mivsma_14d_micro_unique <- setdiff(mivsma_14d_micro, mivsma_14d_macro)
mivsma_14d_macro_unique <- setdiff(mivsma_14d_macro, mivsma_14d_micro)
mivsma_14d_df <- tibble(
  genus = c(mivsma_14d_micro_unique, mivsma_14d_macro_unique),
  source = c(
    rep("Micro-scale Spatial Metagenomics", length(mivsma_14d_micro_unique)),
    rep("Macro-scale Metagenomics", length(mivsma_14d_macro_unique))
  )
)

total <- length(unique(mivsma_phylo_lg %>% 
                         filter(animal %in% c("G103b")) %>% 
                         pull(genus)))


cat("Shared Genera =", 
    length(intersect(mivsma_14d_micro, mivsma_14d_macro)), 
    "(", round(length(intersect(mivsma_14d_micro, mivsma_14d_macro)) / total * 100, 0), "%)",
    "\n")
cat("Unique Genera in Micro-scale Spatial Metagenomics =", 
    length(mivsma_14d_micro_unique), 
    "(", round(length(mivsma_14d_micro_unique) / total * 100, 0),"%)",
    "\n")
cat("Unique Genera in Macro-scale Metagenomics =", 
    length(mivsma_14d_macro_unique), 
    "(", round(length(mivsma_14d_macro_unique) / total * 100, 0), "%)",
    "\n")
```

```{r fig5a, echo = FALSE,  warning = FALSE, comments = "", message = FALSE, fig.width=13, fig.height=7}
mivsma_14d_phylo <- subset_samples(
  physeq_all,
  type == "positive" &
    !section == "Colon" &
    !section == "Ileum" &
    animal == "G103b"
)
mivsma_14d_phylo <- prune_taxa(taxa_sums(mivsma_14d_phylo) > 0, mivsma_14d_phylo)

cols1 <- c("#ba2760","#E9C46A")
names(cols1) <- unique(samdat_tbl(mivsma_14d_phylo)$group)
fig5a <- tax_transform(mivsma_14d_phylo, "binary", rank = "genus") %>%
  comp_heatmap(
    sample_anno = sampleAnnotation(
      "Dataset" = anno_sample_cat(
        "group",
        col = cols1,
        box_col = NA,
        border_col = "black",
        legend_title = "Dataset"
      ),
      "Sequencing Yield" = anno_sample("total_before_filtering.total_reads"),
      border = TRUE,
      show_legend = c(TRUE)
    ), # Show sample annotation legend
    sample_names_show = FALSE,
    tax_anno = taxAnnotation(
      Prev. = anno_tax_prev(),
      Abun. = anno_tax_box()
    ),
    show_heatmap_legend = FALSE,
    heatmap_legend_param = list(
      title_position = "lefttop", # Customize heatmap legend title position if needed
      legend_direction = "horizontal", # Move heatmap legend horizontally
      legend_width = unit(6, "cm") # Adjust the heatmap legend width
    )
  )
fig5a
```
