## MSSM Method Validation

## Load data
```{r load_data_validation, warning = FALSE, comments = "", message = FALSE}
load("data/25_07_11_selected_data.Rdata") # load the data from 1-data_preparation
```

### Discriminative power and replicability of MSSM data
#### Data selection
```{r mssm_discriminative_df, warning = FALSE, comments = "", message = FALSE}
mssm_discriminative_df <- macroquant_genome_counts_filt_30 %>%
  select(genome, macroquant_metrics_filt_30 %>% # select dataset
    filter(
      batch %in% c("MSEB0011"),
      type_binomial %in% c("positive")
    ) %>% pull(microsample)) %>%
  select(where(~ !all(. == 0))) %>%
  melt(id = "genome") %>%
  setNames(c("genome", "microsample", "count")) %>%
  pivot_wider(names_from = genome, values_from = count, values_fill = 0) %>% # samples as rows
  select(where(~ any(. != 0))) %>%
  mutate(across(-microsample, ~ . / rowSums(across(where(is.numeric))) * 100)) %>% # relative abundance
  column_to_rownames(var = "microsample")
```

#### Centered Log-Ratio (CLR) Transformation And Principal Component Analysis (PCA)
**Note:** The perform_pca() function needs to be loaded.

```{r mssm_discriminative_object, warning=FALSE, comments="", message=FALSE}
mssm_discriminative_obj <- perform_pca(mssm_discriminative_df)
```


#### Permutational Multivariate Analysis of Variance (PERMANOVA)

```{r mssm_discriminative_metadata, warning = FALSE, comments = "", message = FALSE}
mssm_discriminative_metadata <- macroquant_metrics_filt_30 %>%
  filter(microsample %in% c(rownames(mssm_discriminative_obj$df_clr_dist)))
unique(mssm_discriminative_metadata$animal)
unique(mssm_discriminative_metadata$cryosection)
```

```{r mssm_discriminative_betadispersion_animal, warning = FALSE, comments = "", message = FALSE}
# Perform betadisper analysis
betadisper_animal <- vegdist(mssm_discriminative_obj$df_clr_dist, method = "euclidean") %>%
  betadisper(., mssm_discriminative_metadata %>% pull(animal))
# Check if there are significant differences in dispersion
permutest(betadisper_animal, pairwise = TRUE)
```

```{r mssm_discriminative_betadispersion_cryosection, warning = FALSE, comments = "", message = FALSE}
# Perform betadisper analysis
betadisper_animal <- vegdist(mssm_discriminative_obj$df_clr_dist, method = "euclidean") %>%
  betadisper(., mssm_discriminative_metadata %>% pull(cryosection))
# Check if there are significant differences in dispersion
permutest(betadisper_animal, pairwise = TRUE)
```

```{r mssm_discriminative_betadispersion_animal&cryosection, warning = FALSE, comments = "", message = FALSE}
# Tests homogeneity of dispersion and pairwise comparisons between groups using permutations.
mssm_discriminative_metadata$group <- interaction(mssm_discriminative_metadata$animal, mssm_discriminative_metadata$cryosection)
betadisp_combined <- betadisper(vegdist(mssm_discriminative_obj$df_clr_dist, method = "euclidean"), mssm_discriminative_metadata$group)
permutest(betadisp_combined, permutations = 999)
```


```{r mssm_discriminative_permanova, warning = FALSE, comments = "", message = FALSE}
vegdist(mssm_discriminative_obj$df_clr_dist, method = "euclidean") %>%
  adonis2(
    formula = . ~ animal + cryosection,
    data = mssm_discriminative_metadata,
    permutations = 999, by = "terms"
  ) %>%
  broom::tidy()
```

#### Amalia: PCA Visualisation Across Animal & Cryosection (Suppl Fig. 5)
```{r supplementary_figure_5, warning = FALSE, comments = "", message = FALSE}
autoplot(mssm_discriminative_obj$pca_result, data = mssm_discriminative_metadata, colour = "animal", shape = "cryosection", size = 3, alpha = 0.5, loadings = FALSE, scale = 0) + custom_ggplot_theme
```

### Amalia: MSSM compared to FISH and Macro-scale Metagenomics

#### Amalia: MSSM vs FISH
```{r x3, warning = FALSE, comments = "", message = FALSE}
```

### Macro-scale Metagenomics vs Micro-scale Spatial Metagenomics

#### Compare Abundance At Genus level

```{r mivsma_ra, warning = FALSE, comments = "", message = FALSE}
mivsma_ra_long <- data.frame(otu_table(physeq_genus_relabund)) %>%
  rownames_to_column("genome") %>%
  melt(id = "genome", variable.name = "microsample", value.name = "ra") %>%
  left_join(data.frame(sample_data(physeq_genus_relabund)) %>%
    rownames_to_column(var = "microsample"), by = "microsample")

mivsma_ra_plot <- mivsma_ra_long %>%
  filter(ra > 0, group %in% c("Micro-scale Spatial Metagenomics")) %>%
  group_by(genome, animal) %>%
  summarise(
    mssm_mean_ra = mean(ra),
    mssm_sd_ra = sd(ra), .groups = "drop"
  ) %>%
  left_join(mivsma_ra_long %>%
    filter(group %in% c("Macro-scale Metagenomics")) %>%
    select(genome, microsample, animal, ra) %>%
    rename(macro_ra = ra), by = c("genome", "animal")) %>%
  left_join(data.frame(tax_table(physeq_genus_relabund)) %>%
    rownames_to_column(var = "genome"), by = "genome") %>%
  mutate(
    dev = abs(mssm_mean_ra - macro_ra),
    genus_label = ifelse(dev > 0.03, genus, ""),
    genus_color = ifelse(genus_label == "", "", order)
  )

ggplot(mivsma_ra_plot, aes(x = macro_ra, y = mssm_mean_ra, color = genus_color)) +
  geom_point() +
  geom_text_repel(
    aes(label = genus_label, color = genus_color),
    size = 3,
    max.overlaps = 100
  ) +
  geom_errorbar(aes(ymin = mssm_mean_ra - mssm_sd_ra, ymax = mssm_mean_ra + mssm_sd_ra), width = 0.002) + # Error bars
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") + # Diagonal reference line
  labs(x = "Relative abundance conventional", y = "Relative abundance MSSM (mean)") +
  theme(legend.position = "none") +
  facet_wrap(~animal) +
  custom_ggplot_theme +
  scale_color_manual(values = order_colors, drop = FALSE) +
  scale_x_continuous(limits = c(0, 0.3)) + #
  scale_y_continuous(limits = c(0, 0.3)) #
```


```{r mivsma_ra_nrmse, warning = FALSE, comments = "", message = FALSE}
mivsma_ra_genus_nmse <- mivsma_ra_long %>%
  filter(ra > 0, group %in% c("Micro-scale Spatial Metagenomics")) %>%
  left_join(mivsma_ra_long %>%
    filter(group %in% c("Macro-scale Metagenomics")) %>%
    select(genome, animal, ra) %>%
    rename(macro_ra = ra), by = c("genome", "animal")) %>%
  group_by(genome, animal) %>%
  summarise(
    genus_rmse = sqrt(mean((ra - macro_ra)^2, na.rm = TRUE)),
    genus_nrmse_sd = genus_rmse / sd(ra, na.rm = TRUE), # Best option for RA and CLR
    genus_nrmse_range = genus_rmse / (max(ra, na.rm = TRUE) - min(ra, na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  left_join(data.frame(tax_table(physeq_genus_relabund)) %>%
    rownames_to_column(var = "genome"), by = "genome") %>%
  mutate(
    label_sd = ifelse(genus_nrmse_sd > 6, genus, ""),
    label_rg = ifelse(genus_nrmse_range > 2.5, genus, ""),
    label_rg_ord = ifelse(label_rg == "", "", order),
    label_sd_ord = ifelse(label_sd == "", "", order)
  )

mivsma_ra_nmse <- mivsma_ra_genus_nmse %>%
  group_by(animal) %>%
  summarise(
    mean = mean(genus_rmse),
    median = median(genus_rmse),
    mean_sd = mean(genus_nrmse_sd),
    median_sd = median(genus_nrmse_sd),
    mean_range = mean(genus_nrmse_range),
    median_range = median(genus_nrmse_range)
  )

mivsma_ra_genus_nmse %>%
  ggplot(aes(x = animal, y = genus_nrmse_sd)) +
  geom_boxplot(width = 0.3, outlier.shape = NA) +
  geom_jitter(aes(color = label_rg_ord), width = 0.1, alpha = 0.9) +
  geom_text_repel(
    aes(label = label_rg, color = label_rg_ord),
    size = 3,
    max.overlaps = 100
  ) +
  scale_color_manual(values = order_colors, drop = FALSE) +
  custom_ggplot_theme +
  labs(x = "Animal", y = "Normalize RMSE \n(Root Mean Squared Error)", colour = "Order")
```


```{r mivsma_clr, warning = FALSE, comments = "", message = FALSE}
mivsma_long <- data.frame(otu_table(physeq_genus)) %>%
  #data.frame(otu_table(tax_glom(physeq_G103b, "genus"))) %>%
  rownames_to_column("genome") %>%
  melt(id = "genome", variable.name = "microsample", value.name = "count") %>%
  pivot_wider(names_from = genome, values_from = count, values_fill = 0) %>%
  # mutate(across(-microsample, ~ . / rowSums(across(where(is.numeric))) * 100)) %>%
  column_to_rownames(var = "microsample")
mivsma_clr <- perform_pca(mivsma_long)
mivsma_clr_meta <- data.frame(sample_data(physeq_genus)) %>%
  rownames_to_column("microsample") %>%
  filter(microsample %in% c(rownames(mivsma_clr$df_clr)))

mivsma_clr_long <- mivsma_clr$df_clr %>%
  rownames_to_column(var = "microsample") %>%
  pivot_longer(-microsample, names_to = "genome", values_to = "clr_value") %>%
  left_join(data.frame(tax_table(physeq_genus)) %>%
    rownames_to_column(var = "genome"), by = "genome") %>%
  left_join(data.frame(sample_data(physeq_genus)) %>%
    rownames_to_column(var = "microsample"), by = "microsample") %>%
  select(microsample, genome, clr_value, domain, phylum, class, order, family, genus, animal, cryosection, group)


```

```{r mivsma_clr_pca, warning = FALSE, comments = "", message = FALSE}
autoplot(mivsma_clr$pca_result, data = mivsma_clr_meta, colour = "group", shape = "animal", size = 3, alpha = 0.5) + custom_ggplot_theme
```

```{r mivsma_clr_nrmse, warning = FALSE, comments = "", message = FALSE}
mivsma_clr_genus_nmse <- mivsma_clr_long %>%
  filter(group %in% c("Micro-scale Spatial Metagenomics")) %>%
  left_join(mivsma_clr_long %>%
    filter(group %in% c("Macro-scale Metagenomics")) %>%
    select(genome, animal, clr_value) %>%
    rename(macro_clr = clr_value), by = c("genome", "animal")) %>%
  group_by(genome, animal) %>%
  summarise(
    genus_rmse = sqrt(mean((clr_value - macro_clr)^2, na.rm = TRUE)),
    genus_nrmse_sd = genus_rmse / sd(clr_value, na.rm = TRUE), # Best option for RA and CLR
    genus_nrmse_range = genus_rmse / (max(clr_value, na.rm = TRUE) - min(clr_value, na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  left_join(data.frame(tax_table(physeq_genus)) %>%
    rownames_to_column(var = "genome"), by = "genome") %>%
  mutate(
    label_sd = ifelse(genus_nrmse_sd > 1, genus, ""),
    label_rg = ifelse(genus_nrmse_range > 1, genus, ""),
    label_rg_ord = ifelse(label_rg == "", "", order),
    label_sd_ord = ifelse(label_sd == "", "", order)
  )

mivsma_clr_nmse <- mivsma_clr_genus_nmse %>%
  group_by(animal) %>%
  summarise(
    mean = mean(genus_rmse),
    median = median(genus_rmse),
    mean_sd = mean(genus_nrmse_sd),
    median_sd = median(genus_nrmse_sd),
    mean_range = mean(genus_nrmse_range),
    median_range = median(genus_nrmse_range)
  )

mivsma_clr_genus_nmse %>%
  ggplot(aes(x = animal, y = genus_nrmse_range)) +
  geom_boxplot(width = 0.3, outlier.shape = NA) +
  geom_jitter(aes(color = label_rg_ord), width = 0.1, alpha = 0.9) +
  geom_text_repel(
    aes(label = label_rg, color = label_rg_ord),
    size = 3,
    max.overlaps = 100
  ) +
  scale_color_manual(values = order_colors, drop = FALSE) +
  custom_ggplot_theme +
  labs(x = "Animal", y = "Normalize RMSE \n(Root Mean Squared Error)", colour = "Order")



```


```{r all_overlap_1, warning = FALSE, comments = "", message = FALSE}
mivsma_clr_genus_nmse_G121e %>%
  ggplot(aes(x = animal, y = genus_nrmse_range)) +
  geom_boxplot(width = 0.3, outlier.shape = NA) +
  geom_jitter(aes(color = label_rg_ord), width = 0.1, alpha = 0.9) +
  geom_text_repel(
    aes(label = label_rg, color = label_rg_ord),
    size = 3,
    max.overlaps = 100
  ) +
  scale_color_manual(values = order_colors, drop = FALSE) +
  custom_ggplot_theme +
  labs(x = "Animal", y = "Normalize RMSE \n(Root Mean Squared Error)", colour = "Order")


mivsma_clr_genus_nmse_G121e %>%
  group_by(animal) %>%
  summarise(
    mean = mean(genus_rmse),
    median = median(genus_rmse),
    mean_sd = mean(genus_nrmse_sd),
    median_sd = median(genus_nrmse_sd),
    mean_range = mean(genus_nrmse_range),
    median_range = median(genus_nrmse_range)
  )

```

```{r all_overlap_2, warning = FALSE, comments = "", message = FALSE}
mivsma_clr_genus_nmse_G103b %>%
  ggplot(aes(x = animal, y = genus_nrmse_range)) +
  geom_boxplot(width = 0.3, outlier.shape = NA) +
  geom_jitter(aes(color = label_rg_ord), width = 0.1, alpha = 0.9) +
  geom_text_repel(
    aes(label = label_rg, color = label_rg_ord),
    size = 3,
    max.overlaps = 100
  ) +
  scale_color_manual(values = order_colors, drop = FALSE) +
  custom_ggplot_theme +
  labs(x = "Animal", y = "Normalize RMSE \n(Root Mean Squared Error)", colour = "Order")

mivsma_clr_genus_nmse_G103b %>%
  group_by(animal) %>%
  summarise(
    mean = mean(genus_rmse),
    median = median(genus_rmse),
    mean_sd = mean(genus_nrmse_sd),
    median_sd = median(genus_nrmse_sd),
    mean_range = mean(genus_nrmse_range),
    median_range = median(genus_nrmse_range)
  )

```

#### Compare Presence/Absence At Genus level

```{r all_overlap, warning = FALSE, comments = "", message = FALSE}
mivsma_phylo_lg <- pivot_phylo(
  phyloseq_obj = subset_samples(
    physeq_all,
    type_binomial == "positive" & !section == "Colon" & !section == "Ileum"
  ),
  glom = TRUE,
  tax_transform = TRUE,
  taxon_level = "genus",
  tr_method = "compositional"
) %>% filter(abundance > 0)
```

#### Shared Genera 
```{r mivsma_genus, warning = FALSE, comments = "", message = FALSE}
mivsma_micro <- unique(mivsma_phylo_lg %>% filter(group %in% c("Micro-scale Spatial Metagenomics")) %>% pull(genus))
mivsma_macro <- unique(mivsma_phylo_lg %>% filter(group %in% c("Macro-scale Metagenomics")) %>% pull(genus))
mivsma_micro_unique <- setdiff(mivsma_micro, mivsma_macro)
mivsma_macro_unique <- setdiff(mivsma_macro, mivsma_micro)
mivsma_df <- tibble(
  genus = c(mivsma_micro_unique, mivsma_macro_unique),
  source = c(
    rep("Micro-scale Spatial Metagenomics", length(mivsma_micro_unique)),
    rep("Macro-scale Metagenomics", length(mivsma_macro_unique))
  )
)

cat("Shared Genera =", length(intersect(mivsma_micro, mivsma_macro)), "\n")
cat("Unique Genera in Micro-scale Spatial Metagenomics =", length(mivsma_micro_unique), "\n")
cat("Unique Genera in Macro-scale Metagenomics =", length(mivsma_macro_unique), "\n")

(length(intersect(mivsma_micro, mivsma_macro)) / length(unique(mivsma_phylo_lg %>% pull(genus)))) * 100
(length(mivsma_micro_unique) / length(unique(mivsma_phylo_lg %>% pull(genus)))) * 100
(length(mivsma_macro_unique) / length(unique(mivsma_phylo_lg %>% pull(genus)))) * 100
```

#### Shared Genera 35 Days Animal (G121e)  
```{r mivsma_35d_genus, warning = FALSE, comments = "", message = FALSE}
mivsma_35d_micro <- unique(mivsma_phylo_lg %>% filter(
  group %in% c("Micro-scale Spatial Metagenomics"),
  animal %in% c("G121e")
) %>% pull(genus))
mivsma_35d_macro <- unique(mivsma_phylo_lg %>% filter(
  group %in% c("Macro-scale Metagenomics"),
  animal %in% c("G121e")
) %>% pull(genus))
mivsma_35d_micro_unique <- setdiff(mivsma_35d_micro, mivsma_35d_macro)
mivsma_35d_macro_unique <- setdiff(mivsma_35d_macro, mivsma_35d_micro)
mivsma_35d_df <- tibble(
  genus = c(mivsma_35d_micro_unique, mivsma_35d_macro_unique),
  source = c(
    rep("Micro-scale Spatial Metagenomics", length(mivsma_35d_micro_unique)),
    rep("Macro-scale Metagenomics", length(mivsma_35d_macro_unique))
  )
)

cat("Shared Genera =", length(intersect(mivsma_35d_micro, mivsma_35d_macro)), "\n")
cat("Unique Genera in Micro-scale Spatial Metagenomics =", length(mivsma_35d_micro_unique), "\n")
cat("Unique Genera in Macro-scale Metagenomics =", length(mivsma_35d_macro_unique), "\n")

(length(intersect(mivsma_35d_micro, mivsma_35d_macro)) / length(unique(mivsma_phylo_lg %>% filter(
  animal %in% c("G121e")
) %>% pull(genus)))) * 100
(length(mivsma_35d_micro_unique) / length(unique(mivsma_phylo_lg %>% filter(
  animal %in% c("G121e")
) %>% pull(genus)))) * 100
(length(mivsma_35d_macro_unique) / length(unique(mivsma_phylo_lg %>% filter(
  animal %in% c("G121e")
) %>% pull(genus)))) * 100
```

#### Shared Genera 14 Days Animal (G103b)  
```{r mivsma_14d_genus, warning = FALSE, comments = "", message = FALSE}
mivsma_14d_micro <- unique(mivsma_phylo_lg %>% filter(
  group %in% c("Micro-scale Spatial Metagenomics"),
  animal %in% c("G103b")
) %>% pull(genus))
mivsma_14d_macro <- unique(mivsma_phylo_lg %>% filter(
  group %in% c("Macro-scale Metagenomics"),
  animal %in% c("G103b")
) %>% pull(genus))
mivsma_14d_micro_unique <- setdiff(mivsma_14d_micro, mivsma_14d_macro)
mivsma_14d_macro_unique <- setdiff(mivsma_14d_macro, mivsma_14d_micro)
mivsma_14d_df <- tibble(
  genus = c(mivsma_14d_micro_unique, mivsma_14d_macro_unique),
  source = c(
    rep("Micro-scale Spatial Metagenomics", length(mivsma_14d_micro_unique)),
    rep("Macro-scale Metagenomics", length(mivsma_14d_macro_unique))
  )
)

cat("Shared Genera =", length(intersect(mivsma_14d_micro, mivsma_14d_macro)), "\n")
cat("Unique Genera in Micro-scale Spatial Metagenomics =", length(mivsma_14d_micro_unique), "\n")
cat("Unique Genera in Macro-scale Metagenomics =", length(mivsma_14d_macro_unique), "\n")

(length(intersect(mivsma_14d_micro, mivsma_14d_macro)) / length(unique(mivsma_phylo_lg %>% filter(
  animal %in% c("G103b")
) %>% pull(genus)))) * 100
(length(mivsma_14d_micro_unique) / length(unique(mivsma_phylo_lg %>% filter(
  animal %in% c("G103b")
) %>% pull(genus)))) * 100
(length(mivsma_14d_macro_unique) / length(unique(mivsma_phylo_lg %>% filter(
  animal %in% c("G103b")
) %>% pull(genus)))) * 100
```

#### Supplementary Table 1
```{r sup_table_1, warning = FALSE, comments = "", message = FALSE}
total_microsamples <- length(unique(mivsma_phylo_lg %>% filter(
  group %in% c("Micro-scale Spatial Metagenomics"),
  animal %in% c("G121e"),
  !section %in% c("Colon", "Ileum")
) %>% pull(microsample)))

mivsma_phylo_lg %>%
  filter(
    group %in% c("Micro-scale Spatial Metagenomics"),
    animal %in% c("G121e"),
    !section %in% c("Colon", "Ileum"),
    genus %in% mivsma_35d_micro_unique
  ) %>%
  group_by(genus) %>%
  summarise(
    # Prevalence = sum(abundance > 0),
    Prevalence_Total = round(sum(abundance > 0) / total_microsamples, 3),
    mean_abundance = round(mean(abundance), 3)
  )
```


#### Fig. 5a

```{r fig5a, warning = FALSE, comments = "", message = FALSE}
total_microsamples <- length(unique(mivsma_phylo_lg %>% filter(
  group %in% c("Micro-scale Spatial Metagenomics"),
  animal %in% c("G103b"),
  !section %in% c("Colon", "Ileum")
) %>% pull(microsample)))

mivsma_14d_phylo <- subset_samples(
  physeq_all,
  type_binomial == "positive" &
    !section == "Colon" &
    !section == "Ileum" &
    animal == "G103b"
)

mivsma_14d_phylo <- prune_taxa(taxa_sums(mivsma_14d_phylo) > 0, mivsma_14d_phylo)

cols1 <- c("#346254", "grey")
names(cols1) <- unique(samdat_tbl(mivsma_14d_phylo)$group)
fig5a <- tax_transform(mivsma_14d_phylo, "binary", rank = "genus") %>%
  comp_heatmap(
    sample_anno = sampleAnnotation(
      "Dataset" = anno_sample_cat(
        "group",
        col = cols1,
        box_col = NA,
        border_col = "black",
        legend_title = "Dataset"
      ),
      "Sequencing Yield" = anno_sample("total_before_filtering.total_reads"),
      border = TRUE,
      show_legend = c(TRUE)
    ), # Show sample annotation legend
    sample_names_show = FALSE,
    tax_anno = taxAnnotation(
      Prev. = anno_tax_prev(),
      Abun. = anno_tax_box()
    ),
    show_heatmap_legend = FALSE,
    heatmap_legend_param = list(
      title_position = "lefttop", # Customize heatmap legend title position if needed
      legend_direction = "horizontal", # Move heatmap legend horizontally
      legend_width = unit(6, "cm") # Adjust the heatmap legend width
    )
  )

save_heatmap <- function() {
  # Save as PDF with appropriate size in inches
  pdf("figures/Fig5a.pdf", width = 10, height = 8.8) # Adjust to desired paper size
  fig5a %>% ComplexHeatmap::draw(
    annotation_legend_list = attr(fig5a, "AnnoLegends"),
    heatmap_legend_side = "bottom",
    annotation_legend_side = "bottom"
  )
  dev.off()
}

save_heatmap()

save_heatmap <- function() {
  png("figures/Fig5a.png", width = 2500, height = 2200, res = 300)
  fig5a %>% ComplexHeatmap::draw(
    annotation_legend_list = attr(fig5a, "AnnoLegends"),
    heatmap_legend_side = "bottom",
    annotation_legend_side = "bottom"
  )
  dev.off()
}
save_heatmap()
```

