# Method Implementation
```{r load_data_implementation, echo = FALSE,  warning = FALSE, comments = "", message = FALSE}
load("resources/working_data_object.Rdata") # load the data from 1-data_preparation
```

## Investigate Spatial Pattern
### Data selection
```{r data_spatial, echo = FALSE,  warning = FALSE, comments = "", message = FALSE}
metadata <- macroquant_metrics_filt_30 %>%
  filter(
    type %in% c("positive"),
    cov_filtering %in% c("Retained By Filtering"),
    !section %in% "Ileum",
    !is.na(Xcoord),
    !is.na(Ycoord),
    !animal %in% "G103b"
  )
comm_data <- macroquant_genome_counts_filt_30 %>%
  select(genome, metadata$microsample) %>%
  t() %>%
  data.frame() %>%
  row_to_names(row_number = 1) %>%
  mutate(across(everything(), as.numeric))
comm_ca <- comm_data[metadata$section == "Caecum right", ]
comm_ca <- comm_ca[, colSums(comm_ca) > 0]
comm_co <- comm_data[metadata$section == "Colon", ]
comm_co <- comm_co[, colSums(comm_co) > 0]
metadata_ca <- metadata[metadata$section == "Caecum right", ]
metadata_co <- metadata[metadata$section == "Colon", ]
cat("Number microsamples for each cryosection", "\n")
knitr::kable(table(metadata$section, metadata$cryosection))%>%
  kableExtra::kable_styling(font_size = 15)  # smaller font size
```

### Centered Log-Ratio (CLR) Transformation

```{r data_spatial_clr, echo = FALSE,  warning = FALSE, comments = "", message = FALSE}
cat("Zero replacement determined", "\n")
comm_ca_zeroRepl <- cmultRepl(comm_ca, method = "GBM", output = "prop", z.warning = 0.95)
comm_co_zeroRepl <- cmultRepl(comm_co, method = "GBM", output = "prop", z.warning = 0.95)

metadata_ca_clr <- metadata_ca[metadata_ca$microsample %in% rownames(comm_ca_zeroRepl), ]
metadata_co_clr <- metadata_co[metadata_co$microsample %in% rownames(comm_co_zeroRepl), ]

clr_transform <- function(x) {
  log(x) - mean(log(x), na.rm = TRUE)
}
comm_ca_clr <- data.frame(t(apply(comm_ca_zeroRepl, 1, clr_transform)))
comm_co_clr <- data.frame(t(apply(comm_co_zeroRepl, 1, clr_transform)))
```

### Diversity partitioning

```{r diversity_partitioning, echo = FALSE,  warning = FALSE, comments = "", message = FALSE}
## Diversity partitioning in caecum and colon
ca_adipart <- adipart(as.matrix(comm_ca > 0) * 1, as.matrix(metadata_ca[, c("microsample", "cryosection")]),
  index = "richness", weights = "unif", relative = FALSE, nsimul = 999, method = "r2dtable"
)

co_adipart <- adipart(as.matrix(comm_co > 0) * 1, as.matrix(metadata_co[, c("microsample", "cryosection")]),
  index = "richness", weights = "unif", relative = FALSE, nsimul = 999, method = "r2dtable"
)

ca <- cbind(
  c(ca_adipart$oecosimu$mean[1], ca_adipart$statistic[1]),
  c(ca_adipart$oecosimu$mean[3], ca_adipart$statistic[3]),
  c(ca_adipart$oecosimu$mean[2], ca_adipart$statistic[2])
)
rownames(ca) <- c("observed", "expected")
colnames(ca) <- c("alpha", "beta", "gamma")

co <- cbind(
  c(co_adipart$oecosimu$mean[1], co_adipart$statistic[1]),
  c(co_adipart$oecosimu$mean[3], co_adipart$statistic[3]),
  c(co_adipart$oecosimu$mean[2], co_adipart$statistic[2])
)
rownames(co) <- c("observed", "expected")
colnames(co) <- c("alpha", "beta", "gamma")

#png("figures/Fig_S4_1.png", width = 1600, height = 800, res = 200)
par(mfrow = c(1, 2))  # 1 row, 2 columns
bp <- barplot(ca,
  beside = TRUE,
  col = c("black", "white"),
  ylim = c(0, 110)
)
text(x = apply(bp, 2, mean), y = apply(ca, 2, max) + 4, labels = "***")
title(main = "Caecum", font.main = 4)
bp <- barplot(co,
  beside = TRUE,
  col = c("black", "white"),
  ylim = c(0, 110)
)
text(x = apply(bp, 2, mean), y = apply(co, 2, max) + 4, labels = "***")
title(main = "Colon", font.main = 4)
par(mfrow = c(1, 1))  # reset layout
#dev.off()

```

### Explore Spatial Pattern
**Note:** The spatial_cryosections() function needs to be loaded.

#### Ceacum Cryosection (G121eI104C)
```{r cryosection_ca, echo = FALSE,  warning = FALSE, comments = "", message = FALSE}
metadata %>% 
  filter(cryosection %in% c("G121eI104C")) %>% 
  ggplot(aes(
    x = Xcoord,
    y = Ycoord,
  )) +
  geom_point() +
  coord_fixed() +
  theme_minimal()

results_ca <- spatial_cryosections(
  cryosection_list = "G121eI104C",
  metadata_df = metadata_ca_clr,
  comm_clr = comm_ca_clr
)
```

##### Mantel Test {-}
```{r caecum_spatial_mantel, echo = FALSE,  warning = FALSE, comments = "", message = FALSE}
results_ca$mantel_results$G121eI104C
# results_ca$mantel_results$G121eI104C$statistic
# results_ca$mantel_results$G121eI104C$signif
# p < 0.05 → Statistically significant correlation between two distance matrices
# p > 0.05 → No significant correlation between two distance matrices
```


##### Mantel Correlogram {-}
```{r caecum_spatial_mantelcor, echo = FALSE,  warning = FALSE, comments = "", message = FALSE}
results_ca$mantelcor_results
```

```{r caecum_spatial_mantelcor_plot, echo = FALSE,  warning = FALSE, comments = "", message = FALSE}
mantel_ca_grob <- ggplot(
  data.frame(results_ca$mantelcor_results$G121eI104C$mantel.res) %>%
    mutate(sign_label = ifelse(`Pr.corrected.` <= 0.05, "Significant", "Not significant")) %>%
    filter(!is.na(Mantel.cor)),
  aes(x = class.index, y = Mantel.cor)
) +
  geom_point(aes(fill = sign_label), shape = 23, size = 3) +
  geom_line() +
  scale_fill_manual(values = c("Significant" = "black", "Not significant" = "white")) +
  labs(
    title = "G121eI104C",
    x = "Distance Class Index",
    y = "Mantel Correlation",
    color = "Significance"
  ) +
  geom_hline(yintercept = 0, color = "red") +
  custom_ggplot_theme +
  theme(legend.position = "none")
mantel_ca_grob
```

#####  Distance Decay {-}
```{r caecum_spatial_decay, echo = FALSE,  warning = FALSE, comments = "", message = FALSE}
aovperm(lmperm(comm_dist ~ spat_dist, data = results_ca$decay_dfs$G121eI104C), np = 10000)
# p < 0.05 → Significant correlation: Spatial distance influences microbial composition.
# p > 0.05 → No significant relationship: Microbial communities do not change with distance.
```

```{r caecum_spatial_decay_plot, echo = FALSE,  warning = FALSE, comments = "", message = FALSE}
results_ca$distance_decay_plots$G121eI104C
```

#### Colon Cryosection (G121eO301A)
```{r cryosection_c0, echo = FALSE,  warning = FALSE, comments = "", message = FALSE}
metadata %>% 
  filter(cryosection %in% c("G121eO301A")) %>% 
  ggplot(aes(
    x = Xcoord,
    y = Ycoord,
  )) +
  geom_point() +
  coord_fixed() +
  theme_minimal()

results_co <- spatial_cryosections(
  cryosection_list = "G121eO301A",
  metadata_df = metadata_co_clr,
  comm_clr = comm_co_clr
)
```

##### Mantel Test {-}
```{r colon_spatial_mantel, echo = FALSE,  warning = FALSE, comments = "", message = FALSE}
results_co$mantel_results$G121eO301A
# p < 0.05 → Statistically significant correlation between two distance matrices
# p > 0.05 → No significant correlation between two distance matrices
```

##### Mantel Correlogram {-}

```{r colon_spatial_mantelcor, echo = FALSE,  warning = FALSE, comments = "", message = FALSE}
results_co$mantelcor_results
```

```{r colon_spatial_mantelcor_plot, echo = FALSE,  warning = FALSE, comments = "", message = FALSE}
mantel_co_grob <- ggplot(
  data.frame(results_co$mantelcor_results$G121eO301A$mantel.res) %>%
    mutate(sign_label = ifelse(`Pr.corrected.` <= 0.05, "Significant", "Not significant")) %>%
    filter(!is.na(Mantel.cor)),
  aes(x = class.index, y = Mantel.cor)
) +
  geom_point(aes(fill = sign_label), shape = 23, size = 3) +
  geom_line() +
  scale_fill_manual(values = c("Significant" = "black", "Not significant" = "white")) +
  labs(
    title = "G121eO301A",
    x = "Distance Class Index",
    y = "Mantel Correlation",
    color = "Significance"
  ) +
  geom_hline(yintercept = 0, color = "red") +
  custom_ggplot_theme +
  theme(legend.position = "none")
mantel_co_grob
```

#####  Distance Decay {-}

```{r colon_spatial_decay, echo = FALSE,  warning = FALSE, comments = "", message = FALSE}
# summary(lmPerm::lmp(comm_dist ~ spat_dist, data = results_co$decay_dfs$G121eO301A))
aovperm(lmperm(comm_dist ~ spat_dist, data = results_co$decay_dfs$G121eO301A), np = 10000)
# Performs an ANOVA to check if spat_dist significantly explains variation in comm_dist.
# p < 0.05 → Significant correlation: Spatial distance influences microbial composition.
# p > 0.05 → No significant relationship: Microbial communities do not change with distance.
```

```{r colon_spatial_decay_plot, echo = FALSE,  warning = FALSE, comments = "", message = FALSE}
results_co$distance_decay_plots$G121eO301A
```


##### RLQ-rlqESLTP Analysis {-}

```{r prepare_data_rlqESLTP, echo = FALSE,  warning = FALSE, comments = "", message = FALSE}
comm_co_G121eO301A <- comm_co %>%
  as.data.frame() %>%
  rownames_to_column(var = "microsample") %>%
  bind_cols(metadata_co) %>%
  filter(cryosection == "G121eO301A") %>%
  select(contains("bin_"))
metadata_co_G121eO301A <- metadata_co %>%
  filter(cryosection == "G121eO301A")

comp <- decostand(comm_co_G121eO301A, MARGIN = 1, method = "hellinger") # Standardize composition by rows
colnames(comp) <- gsub("\\.", ":", names(comp))
env <- data.frame(
  log_seq_counts = log(metadata_co_G121eO301A$total_after_filtering.total_reads),
  div = rowSums(comp > 0)
) # Environmental matrix

genome_gifts <- macromag_gifts[rownames(macromag_gifts) %in% colnames(comp), ]
genome_funct <- genome_gifts %>%
  data.frame() %>%
  rownames_to_column(var = "id") %>% # Preserve row names as a column
  pivot_longer(-id, names_to = "column", values_to = "value") %>% # Reshape to long format
  mutate(group = substr(column, 1, 3)) %>% # Extract the first 3 characters of column names
  group_by(id, group) %>% # Group by row (id) and prefix
  summarise(mean_value = mean(value, na.rm = TRUE), .groups = "drop") %>% # Calculate rowMeans for each group
  pivot_wider(names_from = group, values_from = mean_value) %>% # Reshape back to wide format
  column_to_rownames(var = "id")
genome_funct <- genome_funct[, -c(19:21)]

phy <- macromag_tree # Phylogenetic tree constructed with Phylomatic V3 + Phylocom

phy <- drop.tip(phy, setdiff(phy$tip.label, rownames(genome_funct)))
spa <- metadata_co_G121eO301A[, c("Xcoord", "Ycoord")] # Matrix of X Y spatial coordinates
comp <- comp[, match(phy$tip.label, colnames(comp))]
genome_funct <- genome_funct[match(phy$tip.label, rownames(genome_funct)), ]

mean(phy$tip.label == colnames(comp))

# Change the phylogenetic tree (an object of the class "phylo") into an object of the class "phylog", used by ade4
phylog <- newick2phylog(write.tree(phy)) # => "phylog" class
colnames(comp) <- gsub(":", "_", colnames(comp))
rownames(genome_funct) <- gsub(":", "_", rownames(genome_funct))
```

###### Correspondence analysis of the composition matrix {-}

```{r correspondence_rlqESLTP, echo = FALSE,  warning = FALSE, comments = "", message = FALSE}
coacomp <- dudi.coa(comp, scan = FALSE, nf = 108)
```

###### Gabriel graph {-}

```{r gabriel_graph, echo = FALSE,  warning = FALSE, comments = "", message = FALSE}
# Create the Gabriel graph
nb1 <- graph2nb(gabrielneigh(as.matrix(spa)), sym = T)

# Spatial autocorrelation in diversity and sequence count distribution
sp.correlogram(nb1, log(env$div), order = 8, method = "I")
sp.correlogram(nb1, env$log_seq_counts, order = 8, method = "I")

lw1 <- nb2listw(nb1) # gives a neighbours list with spatial weights (i.e. matrix W)
```

```{r gabriel_graph_plot, echo = FALSE,  warning = FALSE, comments = "", message = FALSE}
nb1.neigh <- nb2neig(nb1)
vecspa <- scores.neig(nb1.neigh)
pcaspa <- dudi.pca(vecspa, coacomp$lw, scan = FALSE, nf = ncol(vecspa))
```

###### PCA analysis of the env matrix {-}

```{r env_pca, echo = FALSE,  warning = FALSE, comments = "", message = FALSE}
pcaenv <- dudi.pca(env, row.w = coacomp$lw, scannf = FALSE, nf = 2)
```

###### Trait-based distances {-}

```{r distances_traits, echo = FALSE,  warning = FALSE, comments = "", message = FALSE}
# Distance matrices for traits separately
listdis <- ldist.ktab(ktab.list.df(list(genome_funct)), c("Q"), scan = FALSE)
# choose 1

# Distance matrix for traits together
disT <- dist.ktab(ktab.list.df(list(genome_funct)), c("Q"), scan = FALSE)
# choose 1
pcotraits <- dudi.pco(disT, coacomp$cw, full = TRUE)
```

###### Phylogeny-based distances {-}

```{r distances_phylogeny, echo = FALSE,  warning = FALSE, comments = "", message = FALSE}
pcophy <- dudi.pco(as.dist(as.matrix(phylog$Wdist)[names(comp), names(comp)]), coacomp$cw, full = TRUE)

## 11. Tests for phylogenetic signals in traits
## *******************************************
# Function rtest.decdiv() in appendix S5, Pavoine et al., 2011. [1]
# All traits together
phystot <- rtest.decdiv(phylog, rep(1, 108), as.dist(as.matrix(disT)[names(phylog$leaves), names(phylog$leaves)]),
  nrep = 99, vranking = "droot", optiontest = "less", ties.method = "average", option = 3
)

trait_names <- c("D01","D02","D03","D05","D06","D07","D08","D09","B01","B02","B03","B06","B07","B08","B09","B10")
phy_names   <- names(phylog$leaves)

set.seed(1234)

pval_tbl <- map_dfr(trait_names, function(nm) {
  mat <- as.matrix(listdis[[nm]])

  # keep only taxa present in both; order like the tree
  idx <- phy_names[phy_names %in% rownames(mat)]
  if (length(idx) < 2L) {
    return(tibble(trait = nm, n_taxa = length(idx), pvalue = NA_real_))
  }

  mat2 <- mat[idx, idx, drop = FALSE]

  pv <- tryCatch(
    rtest.decdiv(
      phylog,
      rep(1, ncol(comp)),
      as.dist(mat2),
      nrep = 99,
      vranking = "droot",
      optiontest = "less",
      ties.method = "average",
      option = 3
    )$pvalue,
    error = function(e) NA_real_
  )
  tibble(trait = nm, pvalue = pv)
}) %>%
  mutate(pvalue = signif(pvalue, 3)) %>%
  arrange(pvalue)

considered_traits <- pval_tbl %>% 
  filter(pvalue<0.05) %>% 
  pull(trait)

## Update the PCOA of traits to remove traits without phylogenetic signal
# Distance matrix for traits together
disT <- dist.ktab(ktab.list.df(list(genome_funct[, considered_traits])), c("Q"), scan = FALSE)
# choose 1
pcotraits <- dudi.pco(disT, coacomp$cw, full = TRUE)
```

###### Extended RLQ analysis {-}

```{r extended_rlq, echo = FALSE,  warning = FALSE, comments = "", message = FALSE}
rlqmix <- rlqESLTP(pcaenv, pcaspa, coacomp, pcotraits, pcophy, scan = F, nf = 2)
```

```{r extended_rlq_barplot, echo = FALSE,  warning = FALSE, comments = "", message = FALSE}
barplot(rlqmix$eig)
```

```{r extended_rlq_table, echo = FALSE,  warning = FALSE, comments = "", message = FALSE}
rlqmix$eig %>%
  tibble(eigenvalue = .) %>%
  mutate(
    axis = paste0("RLQ", row_number()),
    prop_var = eigenvalue / sum(eigenvalue),
    cum_prop_var = cumsum(prop_var)
  ) %>%
  select(axis, eigenvalue, prop_var, cum_prop_var) %>%
  slice_head(n = 5)
```

```{r extended_rlq_axis1_1, echo = FALSE,  warning = FALSE, comments = "", message = FALSE}
plot(rlqmix, xy = spa, ax = 1, wh = "S")
plot(rlqmix, phy = phylog, ax = 1, wh = "P")
plot(rlqmix, traits = genome_funct[, considered_traits], ax = 1, type = "Q", wh = "T")
plot(rlqmix, env = pcaenv$tab, ax = 1, type = "Q", wh = "E")
```

```{r extended_rlq_axis1_positive_genomes, echo = FALSE,  warning = FALSE, comments = "", message = FALSE}
rlqmix$lQ %>% 
  rownames_to_column(var="genome") %>%
  left_join(macromag_taxonomy %>%  mutate(genome=sub(":", "_", genome)),by="genome") %>% 
  arrange(-AxcQ1) %>% 
  select(genome,AxcQ1,family,species) %>% 
  head(n=10)
```

```{r extended_rlq_axis1_negative_genomes, echo = FALSE,  warning = FALSE, comments = "", message = FALSE}
rlqmix$lQ %>% 
  rownames_to_column(var="genome") %>%
  left_join(macromag_taxonomy %>%  mutate(genome=sub(":", "_", genome)),by="genome") %>% 
  arrange(AxcQ1) %>% 
  select(genome,AxcQ1,family,species) %>% 
  head(n=10)
```

## Subspecies Segregation and Spatial Structuring

### Pangenome: Genes Presence/Absence 
```{r pangenome_lawsonibacte_Fig6_a, echo = FALSE,  warning = FALSE, comments = "", message = FALSE, fig.width=8, fig.height=9}
wide_data <- left_join(genepa_pg, duplication_pg, by = "gene") %>%
  left_join(partition_pg, by = "gene") %>%
  pivot_longer(!c(gene, is_single_copy_marker, `Family Partition`), names_to = "Genome", values_to = "Presence/Absence") %>%
  mutate(`Presence/Absence` = as.character(`Presence/Absence`)) %>%
  mutate(`Presence/Absence` = case_when(
    `Presence/Absence` == "1" & is_single_copy_marker == "FALSE" ~ "2",
    TRUE ~ `Presence/Absence`
  )) %>%
  mutate(`Presence/Absence` = as.numeric(`Presence/Absence`)) %>%
  select(gene, Genome, `Presence/Absence`) %>%
  pivot_wider(names_from = Genome, values_from = `Presence/Absence`, values_fill = 0)
mat <- as.matrix(wide_data[, -1])
rownames(mat) <- wide_data$gene
gene_anno <- left_join(genepa_pg, duplication_pg, by = "gene") %>%
  left_join(partition_pg, by = "gene") %>%
  pivot_longer(!c(gene, is_single_copy_marker, `Family Partition`), names_to = "Genome", values_to = "Presence/Absence") %>%
  distinct(gene, `Family Partition`) %>%
  arrange(`Family Partition`) %>%
  column_to_rownames("gene")
mat <- mat[rownames(gene_anno), ]

ann_colors <- list(
  `Family Partition` = c(
    "cloud" = "#2A9D8F",
    "shell" = "#264653",
    "persistent" = "#E76F51"
  )
)

fig6_a <- pheatmap(mat,
  color = c("0" = "grey93", "1" = "gray32", "2" = "#7570b3"), # Named colors
  annotation_row = gene_anno,
  annotation_colors = ann_colors,
  cluster_cols = TRUE,
  cluster_rows = FALSE,
  show_rownames = FALSE,
  legend_breaks = c(0, 1, 2),
  legend_labels = c("Absence", "Presence", "Multicopy")
)
fig6_a
```

### Distribution Lawsonibacter Strains
```{r Fig6_b, echo = FALSE,  warning = FALSE, comments = "", message = FALSE}
macroquant_lawsonibacter <- macroquant_counts_filt_30_98 %>%
  filter(genome %in% c(macromag_genomemetadata_98 %>%
    filter(genus == "Lawsonibacter") %>%
    pull(genome))) %>%
  pivot_longer(-genome, names_to = "microsample", values_to = "abundance") %>%
  left_join(macroquant_metrics_filt_30 %>%
    select(microsample, batch, cryosection, animal, Xcoord, Ycoord), by = "microsample") %>%
  filter(
    abundance > 0,
    !is.na(Xcoord),
    !is.na(Ycoord)
  ) %>%
  left_join(macromag_genomemetadata_98, by = "genome")

lawsonibacter_pie_df <- macroquant_lawsonibacter %>%
  filter(
    cryosection %in% c("G121eI103A", "G103bI301A"),
    # circul %in% c("Y")
  ) %>% 
  group_by(microsample) %>%
  mutate(
    abundance = abundance / sum(abundance) # normalize counts to proportions
  ) %>%
  arrange(microsample, genome) %>%
  mutate(
    start = cumsum(lag(abundance, default = 0)) * 2 * pi,
    end = cumsum(abundance) * 2 * pi,
    r0 = 0,
    r = 75
  )

lawsonibacter_piecharts <- split(lawsonibacter_pie_df, lawsonibacter_pie_df$cryosection) %>%
  purrr::map(~ {
    ggplot(.x) +
      geom_arc_bar(
        aes(
          x0 = Xcoord, y0 = Ycoord, r0 = r0, r = r,
          start = start, end = end, fill = genome
        ),
        color = "black", linewidth = 0.1
      ) +
      coord_fixed() +
      ggtitle(unique(.x$cryosection)) +
      theme(legend.position = "none") +
      custom_ggplot_theme +
      scale_fill_manual(values = lawsonibacter_colors, drop = FALSE)
  })

lawsonibacter_piecharts$G103bI301A
lawsonibacter_piecharts$G121eI103A
```

### Test Spatial Distribution Lawsonibacter subspecies
**Note:** The lawsonibacter_mantel_analysis() function needs to be loaded.

#### 35 Days Animal (G121e)
##### Presence/Absence {-}

```{r lawsonibacter_mantel_G121e_cryo_pa, echo = FALSE,  warning = FALSE, comments = "", message = FALSE}
lawsonibacter_G121e <- lawsonibacter_mantel_analysis(
  data = macroquant_lawsonibacter %>% filter(cryosection %in% c("G121eI103A")),
  circul_selection = "N"
)
lawsonibacter_G121e$correlogram_pa
```
##### Abundance {-}

```{r lawsonibacter_mantel_G121e_cryo, echo = FALSE,  warning = FALSE, comments = "", message = FALSE}
lawsonibacter_G121e$correlogram_clr
```


#### 14 Days Animal (G103b)
##### Presence/Absence {-}

```{r lawsonibacter_mantel_G103b_pa, echo = FALSE,  warning = FALSE, comments = "", message = FALSE}
lawsonibacter_G103b <- lawsonibacter_mantel_analysis(
  data = macroquant_lawsonibacter %>% filter(cryosection %in% c("G103bI301A")),
  circul_selection = "N"
)
lawsonibacter_G103b$correlogram_pa
```

##### Abundance {-}
```{r lawsonibacter_mantel_G103b, echo = FALSE,  warning = FALSE, comments = "", message = FALSE}
lawsonibacter_G103b$correlogram_clr
```


## SNP level microdiversity (Lorikeet)
```{r lawsonibacter_popANI, echo = FALSE,  warning = FALSE, comments = "", message = FALSE}
lawsonibacter_popANI <- map_dfr(names(ani_list), function(bin_id) {
  ani_mat <- ani_list[[bin_id]]
  ani_mat %>%
    as.data.frame() %>%
    rownames_to_column("microsample_1") %>%
    pivot_longer(-microsample_1, names_to = "microsample_2", values_to = "popANI") %>%
    filter(microsample_1 != microsample_2) %>% # Exclude self-comparisons
    mutate(
      bin = bin_id,
      # Create unordered pair identifier
      sample_pair = pmap_chr(list(microsample_1, microsample_2), ~ paste(sort(c(..1, ..2)), collapse = "_"))
    ) %>%
    distinct(bin, sample_pair, .keep_all = TRUE) %>% # Keep only one of each unordered pair
    select(-sample_pair)
})
lawsonibacter_popANI_meta <- macroquant_metrics_filt_30 %>%
  filter(
    cryosection %in% c("G121eI103A", "G103bI301A"),
    type %in% c("positive"),
    cov_filtering %in% c("Retained By Filtering"),
    !is.na(Xcoord),
    !is.na(Ycoord)
    ) %>%
  select(cryosection, microsample, animal, Xcoord, Ycoord)
```

### Distribution popANI by Group (Fig6_d and SupFig7_a) 
```{r Fig6_d, echo = FALSE,  warning = FALSE, comments = "", message = FALSE}
boxplot_popANI <- lawsonibacter_popANI %>%
  filter(
    !is.nan(popANI),
    !is.na(popANI),
    microsample_1 %in% c(lawsonibacter_popANI_meta$microsample),
    microsample_2 %in% c(lawsonibacter_popANI_meta$microsample)#,
    # bin %in% c("GPB_bin_000044") or bin %in% c("GPB_bin_000047","GPB_bin_000063")
  ) %>%
  left_join(macroquant_metrics_filt_30 %>% select(microsample, animal), by = c("microsample_1" = "microsample")) %>%
  rename(animal_microsample_1 = animal) %>%
  left_join(macroquant_metrics_filt_30 %>% select(microsample, animal), by = c("microsample_2" = "microsample")) %>%
  rename(animal_microsample_2 = animal) %>%
  mutate(group = ifelse(animal_microsample_1 == animal_microsample_2, animal_microsample_1, "G121e-G103b")) %>%
  ggplot(aes(x = group, y = popANI, fill = group)) +
  geom_boxplot(aes(group = group), outlier.shape = NA) +
  geom_jitter(
    position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.8),
    size = 0.3, alpha = 0.3
  ) +
  scale_fill_manual(values = c("#79b1a3", "#e35d51", "#ffcc62")) +
  custom_ggplot_theme +
  stat_compare_means(
    method = "wilcox.test",
    comparisons = list(
      c("G121e", "G103b"),
      c("G121e", "G121e-G103b"),
      c("G103b", "G121e-G103b")
    ),
    label = "p.signif",
    p.adjust.method = "BH",
    hide.ns = TRUE
  ) +
  labs(
    # title = "a",
    x = "",
    #y = "popANI GPB_bin_000044"
  ) +
  theme(legend.position = "none") +
  scale_x_discrete(labels = c(
    "G103b" = "Within Animal \nG103b", "G121e" = "Within Animal \nG121e",
    "G121e-G103b" = "Between Animals"
  )) +
  # Hide last tick label by replacing it with blank
  scale_y_continuous(labels = function(x) {
    lbls <- as.character(x)
    lbls[length(lbls)] <- "" # blank out last label
    lbls
  })+ facet_wrap(~bin) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
boxplot_popANI
```

### Heatmap popANI GPB_bin_000044 (Fig6_c) 
```{r Fig6_c, echo = FALSE,  warning = FALSE, comments = "", message = FALSE}
popani_044 <- ani_list$GPB_bin_000044[rownames(ani_list$GPB_bin_000044) %in% lawsonibacter_popANI_meta$microsample, ] %>%
  select(lawsonibacter_popANI_meta$microsample)

metadata_044 <- macroquant_metrics_filt_30 %>%
  filter(microsample %in% colnames(popani_044)) %>%
  select(microsample, animal) %>%
  column_to_rownames(var = "microsample") %>%
  rename(    Animal = animal
  )

annotation_colors <- list(
  Animal = c(
    "G121e" = "darkgrey",
    "G103b" = "#E9C46A"
  ) 
)

fig6_c <- pheatmap(
  popani_044,
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  clustering_method = "ward",
  display_numbers = FALSE,
  legend = TRUE,
  show_rownames = FALSE,
  show_colnames = FALSE,
  color = colorRampPalette(c("black", "#fbffe0"))(100),
  annotation_col = metadata_044,
  annotation_row = metadata_044,
  annotation_colors = annotation_colors,
)
fig6_c
```

### Spatial Pattern Microdiversity

```{r mantel_cor_bins, echo = FALSE,  warning = FALSE, comments = "", message = FALSE}
bins_id <- c("GPB_bin_000044","GPB_bin_000047","GPB_bin_000063")  
ids_35day <- lawsonibacter_popANI_meta %>%
  filter(cryosection %in% c("G121eI103A")) %>%
  pull(microsample)

results_list <- list()
for (bin in bins_id) {
  cat("Running analysis for", bin, "\n")
  res <- ani_spatial_analysis(bin, ani_list, ids_35day, lawsonibacter_popANI_meta)
  results_list[[bin]] <- res
}
results_list$GPB_bin_000063$mantel
```


