## MSSM Method Implementation
## Load data
```{r load_data_implementation, warning = FALSE, comments = "", message = FALSE}
load("data/25_07_11_selected_data.Rdata") # load the data from 1-data_preparation
```

### Investigate Spatial Pattern
#### Data selection
```{r data_spatial, warning = FALSE, comments = "", message = FALSE}
metadata <- macroquant_metrics_filt_30 %>%
  filter(
    type_binomial %in% c("positive"),
    cov_filtering %in% c("Retained By Filtering"),
    !section %in% "Ileum",
    !is.na(Xcoord),
    !animal %in% "G103b"
  )

comm_data <- macroquant_genome_counts_filt_30 %>%
  select(genome, metadata$microsample) %>%
  t() %>%
  data.frame() %>%
  row_to_names(row_number = 1) %>%
  mutate(across(everything(), as.numeric))

# table(rowSums(comm_data) > 0)
# all(metadata$microsample == rownames(comm_data))
# table(is.na(comm_data))
# table(metadata$section, metadata$cryosection)

comm_ca <- comm_data[metadata$section == "Caecum right", ]
comm_ca <- comm_ca[, colSums(comm_ca) > 0]
comm_co <- comm_data[metadata$section == "Colon", ]
comm_co <- comm_co[, colSums(comm_co) > 0]
metadata_ca <- metadata[metadata$section == "Caecum right", ]
metadata_co <- metadata[metadata$section == "Colon", ]
table(metadata$section, metadata$cryosection)
table(metadata$section, metadata$animal)

metadata %>% 
  filter(cryosection=="G121eO301A") %>% 
  ggplot() +
  geom_point(aes(
    x = Xcoord,
    y = Ycoord,
    color = cycles
  )) + coord_fixed()
```

#### Transform the data for posterior multivariate analyses

```{r data_spatial_clr, comment="", message=FALSE, warning=FALSE}
comm_ca_zeroRepl <- cmultRepl(comm_ca, method = "GBM", output = "prop", z.warning = 0.95)
comm_co_zeroRepl <- cmultRepl(comm_co, method = "GBM", output = "prop", z.warning = 0.95)

metadata_ca_clr <- metadata_ca[metadata_ca$microsample %in% rownames(comm_ca_zeroRepl), ]
metadata_co_clr <- metadata_co[metadata_co$microsample %in% rownames(comm_co_zeroRepl), ]

clr_transform <- function(x) {
  log(x) - mean(log(x), na.rm = TRUE)
}
comm_ca_clr <- data.frame(t(apply(comm_ca_zeroRepl, 1, clr_transform)))
comm_co_clr <- data.frame(t(apply(comm_co_zeroRepl, 1, clr_transform)))
```

#### Diversity partitioning

```{r diversity_partitioning, comment="", message=FALSE, warning=FALSE}
## Diversity partitioning in caecum and colon
ca_adipart <- adipart(as.matrix(comm_ca > 0) * 1, as.matrix(metadata_ca[, c("microsample", "cryosection")]),
  index = "richness", weights = "unif", relative = FALSE, nsimul = 999, method = "r2dtable"
)

co_adipart <- adipart(as.matrix(comm_co > 0) * 1, as.matrix(metadata_co[, c("microsample", "cryosection")]),
  index = "richness", weights = "unif", relative = FALSE, nsimul = 999, method = "r2dtable"
)

ca <- cbind(
  c(ca_adipart$oecosimu$mean[1], ca_adipart$statistic[1]),
  c(ca_adipart$oecosimu$mean[3], ca_adipart$statistic[3]),
  c(ca_adipart$oecosimu$mean[2], ca_adipart$statistic[2])
)
rownames(ca) <- c("observed", "expected")
colnames(ca) <- c("alpha", "beta", "gamma")

co <- cbind(
  c(co_adipart$oecosimu$mean[1], co_adipart$statistic[1]),
  c(co_adipart$oecosimu$mean[3], co_adipart$statistic[3]),
  c(co_adipart$oecosimu$mean[2], co_adipart$statistic[2])
)
rownames(co) <- c("observed", "expected")
colnames(co) <- c("alpha", "beta", "gamma")


par(mfrow = c(1, 2))
bp <- barplot(ca,
  beside = TRUE,
  col = c("black", "white"),
  ylim = c(0, 110)
)
text(x = apply(bp, 2, mean), y = apply(ca, 2, max) + 7, labels = "***")
title(main = "Caecum", font.main = 4)
bp <- barplot(co,
  beside = TRUE,
  col = c("black", "white"),
  ylim = c(0, 110)
)
text(x = apply(bp, 2, mean), y = apply(co, 2, max) + 7, labels = "***")
title(main = "Colon", font.main = 4)
par(mfrow = c(1, 1))
```

#### Explore Spatial Pattern
**Note:** The spatial_cryosections() function needs to be loaded.
##### Ceacum Cryosection (G121eI104C)
```{r cryosection_ca, warning = FALSE, comments = "", message = FALSE}
results_ca <- spatial_cryosections(
  cryosection_list = "G121eI104C",
  metadata_df = metadata_ca_clr,
  comm_clr = comm_ca_clr
)
```

###### Mantel test
```{r caecum_spatial_mantel, comment="", message=FALSE, warning=FALSE}
results_ca$mantel_results$G121eI104C
# results_ca$mantel_results$G121eI104C$statistic
# results_ca$mantel_results$G121eI104C$signif
# p < 0.05 → Statistically significant correlation between two distance matrices
# p > 0.05 → No significant correlation between two distance matrices
```


###### Mantel Correlogram Stat
```{r caecum_spatial_mantelcor, comment="", message=FALSE, warning=FALSE}
results_ca$mantelcor_results
```

###### Mantel Correlogram Plot
```{r caecum_spatial_mantelcor_plot, comment="", message=FALSE, warning=FALSE}
mantel_ca_grob <- ggplot(
  data.frame(results_ca$mantelcor_results$G121eI104C$mantel.res) %>%
    mutate(sign_label = ifelse(`Pr.corrected.` <= 0.05, "Significant", "Not significant")) %>%
    filter(!is.na(Mantel.cor)),
  aes(x = class.index, y = Mantel.cor)
) +
  geom_point(aes(fill = sign_label), shape = 23, size = 3) +
  geom_line() +
  scale_fill_manual(values = c("Significant" = "black", "Not significant" = "white")) +
  labs(
    title = "G121eI104C",
    x = "Distance Class Index",
    y = "Mantel Correlation",
    color = "Significance"
  ) +
  geom_hline(yintercept = 0, color = "red") +
  custom_ggplot_theme +
  theme(legend.position = "none")

mantel_ca_grob
```

######  Distance Decay Stat
```{r caecum_spatial_decay, comment="", message=FALSE, warning=FALSE}
aovperm(lmperm(comm_dist ~ spat_dist, data = results_ca$decay_dfs$G121eI104C), np = 10000)
# p < 0.05 → Significant correlation: Spatial distance influences microbial composition.
# p > 0.05 → No significant relationship: Microbial communities do not change with distance.
```

######  Distance Decay Plot

```{r caecum_spatial_decay_plot, comment="", message=FALSE, warning=FALSE}
results_ca$distance_decay_plots$G121eI104C
```

##### Colon Cryosection (G121eO301A)
```{r cryosection_c0, warning = FALSE, comments = "", message = FALSE}
results_co <- spatial_cryosections(
  cryosection_list = "G121eO301A",
  metadata_df = metadata_co_clr,
  comm_clr = comm_co_clr
)
```

###### Mantel test
```{r colon_spatial_mantel, comment="", message=FALSE, warning=FALSE}
results_co$mantel_results$G121eO301A
# p < 0.05 → Statistically significant correlation between two distance matrices
# p > 0.05 → No significant correlation between two distance matrices
```

###### Mantel Correlogram Stat
```{r colon_spatial_mantelcor, comment="", message=FALSE, warning=FALSE}
results_co$mantelcor_results
```

###### Mantel Correlogram Plot
```{r colon_spatial_mantelcor_plot, comment="", message=FALSE, warning=FALSE}
mantel_co_grob <- ggplot(
  data.frame(results_co$mantelcor_results$G121eO301A$mantel.res) %>%
    mutate(sign_label = ifelse(`Pr.corrected.` <= 0.05, "Significant", "Not significant")) %>%
    filter(!is.na(Mantel.cor)),
  aes(x = class.index, y = Mantel.cor)
) +
  geom_point(aes(fill = sign_label), shape = 23, size = 3) +
  geom_line() +
  scale_fill_manual(values = c("Significant" = "black", "Not significant" = "white")) +
  labs(
    title = "G121eO301A",
    x = "Distance Class Index",
    y = "Mantel Correlation",
    color = "Significance"
  ) +
  geom_hline(yintercept = 0, color = "red") +
  custom_ggplot_theme +
  theme(legend.position = "none")
mantel_co_grob
```

######  Distance Decay Stat
```{r colon_spatial_decay, comment="", message=FALSE, warning=FALSE}
# summary(lmPerm::lmp(comm_dist ~ spat_dist, data = results_co$decay_dfs$G121eO301A))
aovperm(lmperm(comm_dist ~ spat_dist, data = results_co$decay_dfs$G121eO301A), np = 10000)
# Performs an ANOVA to check if spat_dist significantly explains variation in comm_dist.
# p < 0.05 → Significant correlation: Spatial distance influences microbial composition.
# p > 0.05 → No significant relationship: Microbial communities do not change with distance.
```

######  Distance Decay Plot

```{r colon_spatial_decay_plot, comment="", message=FALSE, warning=FALSE}
results_co$distance_decay_plots$G121eO301A
```

```{r figs4_2, warning = FALSE, comments = "", message = FALSE}
figs4_2 <- cowplot::plot_grid(
  results_co$distance_decay_plots$G121eO301A,
  mantel_co_grob,
  results_ca$distance_decay_plots$G121eI104C,
  mantel_ca_grob,
  ncol = 2,
  labels = c("a", "b", "c", "d"),
  label_size = 16
)
figs4_2
# ggsave("figures/FigS4_2.png", figs4_2, width = 10, height = 9, dpi = 300)
```

###### RLQ-rlqESLTP Analysis (G121eO301A)

```{r load_data_rlqESLTP, comment="", message=FALSE, warning=FALSE}
source("data/JEC_1743_sm_apps5.txt")
```

```{r prepare_data_rlqESLTP, comment="", message=FALSE, warning=FALSE}
# comm_co_G121eO301A <- results_co$cryosection_dfs[[cryosection]]$comm_clr
# metadata_co_G121eO301A <- results_co$cryosection_dfs[[cryosection]]$metadata
comm_co_G121eO301A <- comm_co %>%
  as.data.frame() %>%
  rownames_to_column(var = "microsample") %>%
  bind_cols(metadata_co) %>%
  filter(cryosection == "G121eO301A") %>%
  select(contains("bin_"))
metadata_co_G121eO301A <- metadata_co %>%
  filter(cryosection == "G121eO301A")

comp <- decostand(comm_co_G121eO301A, MARGIN = 1, method = "hellinger") # Standardize composition by rows
colnames(comp) <- gsub("\\.", ":", names(comp))
env <- data.frame(
  log_seq_counts = log(metadata_co_G121eO301A$total_after_filtering.total_reads),
  div = rowSums(comp > 0)
) # Environmental matrix

genome_gifts <- macromag_gifts[rownames(macromag_gifts) %in% colnames(comp), ]
genome_funct <- genome_gifts %>%
  data.frame() %>%
  rownames_to_column(var = "id") %>% # Preserve row names as a column
  pivot_longer(-id, names_to = "column", values_to = "value") %>% # Reshape to long format
  mutate(group = substr(column, 1, 3)) %>% # Extract the first 3 characters of column names
  group_by(id, group) %>% # Group by row (id) and prefix
  summarise(mean_value = mean(value, na.rm = TRUE), .groups = "drop") %>% # Calculate rowMeans for each group
  pivot_wider(names_from = group, values_from = mean_value) %>% # Reshape back to wide format
  column_to_rownames(var = "id")
genome_funct <- genome_funct[, -c(19:21)]

phy <- macromag_tree # Phylogenetic tree constructed with Phylomatic V3 + Phylocom

phy <- drop.tip(phy, setdiff(phy$tip.label, rownames(genome_funct)))
spa <- metadata_co_G121eO301A[, c("Xcoord", "Ycoord")] # Matrix of X Y spatial coordinates
comp <- comp[, match(phy$tip.label, colnames(comp))]
genome_funct <- genome_funct[match(phy$tip.label, rownames(genome_funct)), ]

mean(phy$tip.label == colnames(comp))

# Change the phylogenetic tree (an object of the class "phylo") into an object of the class "phylog", used by ade4
phylog <- newick2phylog(write.tree(phy)) # => "phylog" class
colnames(comp) <- gsub(":", "_", colnames(comp))
rownames(genome_funct) <- gsub(":", "_", rownames(genome_funct))
```

####### Antton: Correspondence analysis of the composition matrix 
```{r correspondence_rlqESLTP, comment="", message=FALSE, warning=FALSE}
coacomp <- dudi.coa(comp, scan = FALSE, nf = 108)
summary(coacomp)
```

###### Spatial analysis 

###### Gabriel graph

```{r gabriel_graph, comment="", message=FALSE, warning=FALSE}
# Create the Gabriel graph
nb1 <- graph2nb(gabrielneigh(as.matrix(spa)), sym = T)
nb1

# Spatial autocorrelation in diversity and sequence count distribution
sp.correlogram(nb1, log(env$div), order = 8, method = "I")
sp.correlogram(nb1, env$log_seq_counts, order = 8, method = "I")

lw1 <- nb2listw(nb1) # gives a neighbours list with spatial weights (i.e. matrix W)
lw1
```

```{r gabriel_graph_plot, comment="", message=FALSE, warning=FALSE}
par(mfrow = c(1, 1))
plot(nb1, spa, pch = 21, bg = "red")
title(main = "Gabriel Graph")
class(nb1)

# [1] "nb"
# This is the Gabriel graph -the graph that in this case defines which points
# are connected

# The matrix of spatial variables is obtained as the eigenvectors of a
# neighbour matrix. This matrix is analysed by principal component analysis.
nb1.neigh <- nb2neig(nb1)
vecspa <- scores.neig(nb1.neigh)
pcaspa <- dudi.pca(vecspa, coacomp$lw, scan = FALSE, nf = ncol(vecspa))
summary(pcaspa)
```

###### PCA analysis of the env matrix
```{r env_pca, comment="", message=FALSE, warning=FALSE}
pcaenv <- dudi.pca(env, row.w = coacomp$lw, scannf = FALSE, nf = 2)
summary(pcaenv)
```

###### The distances between species based on their biological traits, analyzed by PCoA.

```{r distances_traits, comment="", message=FALSE, warning=FALSE}
# Distance matrices for traits separately
listdis <- ldist.ktab(ktab.list.df(list(genome_funct)), c("Q"), scan = FALSE)
# choose 1

# Distance matrix for traits together
disT <- dist.ktab(ktab.list.df(list(genome_funct)), c("Q"), scan = FALSE)
# choose 1
pcotraits <- dudi.pco(disT, coacomp$cw, full = TRUE)
summary(pcotraits)
```

###### The distances between species based on their phylogenetic relatedness, analysed by PCoA.

```{r distances_phylogeny, comment="", message=FALSE, warning=FALSE}
pcophy <- dudi.pco(as.dist(as.matrix(phylog$Wdist)[names(comp), names(comp)]), coacomp$cw, full = TRUE)
summary(pcophy)

## 11. Tests for phylogenetic signals in traits
## *******************************************
# Function rtest.decdiv() in appendix S5, Pavoine et al., 2011. [1]
# All traits together
phystot <- rtest.decdiv(phylog, rep(1, 108), as.dist(as.matrix(disT)[names(phylog$leaves), names(phylog$leaves)]),
  nrep = 99, vranking = "droot", optiontest = "less", ties.method = "average", option = 3
)
phystot

# All traits independently
rtest.decdiv(phylog, rep(1, ncol(comp)), as.dist(as.matrix(listdis$D01)[
  names(phylog$leaves),
  names(phylog$leaves)
]),
nrep = 99, vranking = "droot", optiontest = "less",
ties.method = "average", option = 3
)
rtest.decdiv(phylog, rep(1, ncol(comp)), as.dist(as.matrix(listdis$D02)[
  names(phylog$leaves),
  names(phylog$leaves)
]),
nrep = 99, vranking = "droot", optiontest = "less",
ties.method = "average", option = 3
)
rtest.decdiv(phylog, rep(1, ncol(comp)), as.dist(as.matrix(listdis$D03)[
  names(phylog$leaves),
  names(phylog$leaves)
]),
nrep = 99, vranking = "droot", optiontest = "less",
ties.method = "average", option = 3
)
rtest.decdiv(phylog, rep(1, ncol(comp)), as.dist(as.matrix(listdis$D05)[
  names(phylog$leaves),
  names(phylog$leaves)
]),
nrep = 99, vranking = "droot", optiontest = "less",
ties.method = "average", option = 3
)
rtest.decdiv(phylog, rep(1, ncol(comp)), as.dist(as.matrix(listdis$D06)[
  names(phylog$leaves),
  names(phylog$leaves)
]),
nrep = 99, vranking = "droot", optiontest = "less",
ties.method = "average", option = 3
)
rtest.decdiv(phylog, rep(1, ncol(comp)), as.dist(as.matrix(listdis$D07)[
  names(phylog$leaves),
  names(phylog$leaves)
]),
nrep = 99, vranking = "droot", optiontest = "less",
ties.method = "average", option = 3
)
rtest.decdiv(phylog, rep(1, ncol(comp)), as.dist(as.matrix(listdis$D08)[
  names(phylog$leaves),
  names(phylog$leaves)
]),
nrep = 99, vranking = "droot", optiontest = "less",
ties.method = "average", option = 3
)
rtest.decdiv(phylog, rep(1, ncol(comp)), as.dist(as.matrix(listdis$D09)[
  names(phylog$leaves),
  names(phylog$leaves)
]),
nrep = 99, vranking = "droot", optiontest = "less",
ties.method = "average", option = 3
)
rtest.decdiv(phylog, rep(1, ncol(comp)), as.dist(as.matrix(listdis$B01)[
  names(phylog$leaves),
  names(phylog$leaves)
]),
nrep = 99, vranking = "droot", optiontest = "less",
ties.method = "average", option = 3
)
rtest.decdiv(phylog, rep(1, ncol(comp)), as.dist(as.matrix(listdis$B02)[
  names(phylog$leaves),
  names(phylog$leaves)
]),
nrep = 99, vranking = "droot", optiontest = "less",
ties.method = "average", option = 3
)
rtest.decdiv(phylog, rep(1, ncol(comp)), as.dist(as.matrix(listdis$B03)[
  names(phylog$leaves),
  names(phylog$leaves)
]),
nrep = 99, vranking = "droot", optiontest = "less",
ties.method = "average", option = 3
)
rtest.decdiv(phylog, rep(1, ncol(comp)), as.dist(as.matrix(listdis$B06)[
  names(phylog$leaves),
  names(phylog$leaves)
]),
nrep = 99, vranking = "droot", optiontest = "less",
ties.method = "average", option = 3
)
rtest.decdiv(phylog, rep(1, ncol(comp)), as.dist(as.matrix(listdis$B07)[
  names(phylog$leaves),
  names(phylog$leaves)
]),
nrep = 99, vranking = "droot", optiontest = "less",
ties.method = "average", option = 3
)
rtest.decdiv(phylog, rep(1, ncol(comp)), as.dist(as.matrix(listdis$B08)[
  names(phylog$leaves),
  names(phylog$leaves)
]),
nrep = 99, vranking = "droot", optiontest = "less",
ties.method = "average", option = 3
)
rtest.decdiv(phylog, rep(1, ncol(comp)), as.dist(as.matrix(listdis$B09)[
  names(phylog$leaves),
  names(phylog$leaves)
]),
nrep = 99, vranking = "droot", optiontest = "less",
ties.method = "average", option = 3
)
rtest.decdiv(phylog, rep(1, ncol(comp)), as.dist(as.matrix(listdis$B10)[
  names(phylog$leaves),
  names(phylog$leaves)
]),
nrep = 99, vranking = "droot", optiontest = "less",
ties.method = "average", option = 3
)

## Update the PCOA of traits to remove traits without phylogenetic signal
# Distance matrix for traits together
disT <- dist.ktab(ktab.list.df(list(genome_funct[, -c(4, 8, 13)])), c("Q"), scan = FALSE)
# choose 1
pcotraits <- dudi.pco(disT, coacomp$cw, full = TRUE)
summary(pcotraits)
```

###### Extended RLQ analysis

```{r extended_rlq, comment="", message=FALSE, warning=FALSE}
rlqmix <- rlqESLTP(pcaenv, pcaspa, coacomp, pcotraits, pcophy, scan = F, nf = 2)
barplot(rlqmix$eig)
rlqmix$eig[1] / sum(rlqmix$eig)
# [1] 0.73
rlqmix$eig[2] / sum(rlqmix$eig)
# [1] 0.12
```

**1st axis**
```{r extended_rlq_axis1_1, comment="", message=FALSE, warning=FALSE}
plot(rlqmix, xy = spa, ax = 1, wh = "S")
plot(rlqmix, phy = phylog, ax = 1, wh = "P")
plot(rlqmix, traits = genome_funct[, -c(4, 8, 13)], ax = 1, type = "Q", wh = "T")
plot(rlqmix, env = pcaenv$tab, ax = 1, type = "Q", wh = "E")

s.value(metadata_co_G121eO301A[, c("Xcoord", "Ycoord")],
  rlqmix$lR[, 1],
  sub = "Patterns in S and E",
  csub = 1.5,
  include.origin = F,
  csize = 0.8
)

plot(rlqmix, env = pcaenv$tab, ax = 1, type = "Q", wh = "E")
dotchart.phylog(phylog, rlqmix$lQ[names(phylog$leaves), 1],
  cleav = 0, cdot = 1,
  scaling = F, yjoi = 0, cex.axis = 1.5, sub = "patterns in T and P", csub = 0
)
```

**1st axis**
```{r extended_rlq_axis1_2, comment="", message=FALSE, warning=FALSE}
png("figures/plot_a.png", width = 800, height = 400, res = 150)
s.value(
  metadata_co_G121eO301A[, c("Xcoord", "Ycoord")],
  rlqmix$lR[, 1],
  sub = "Patterns in S and E",
  csub = 1.5,
  include.origin = FALSE,
  csize = 0.8
)
mtext("a", side = 3, line = 0.5, at = par("usr")[1], font = 2, cex = 1.5)
dev.off()

png("figures/plot_b.png", width = 800, height = 400, res = 150)
plot(rlqmix, env = pcaenv$tab, ax = 1, type = "Q", wh = "E")
# mtext("b", side = 3, line = 0.5, at = par("usr")[1], font = 2, cex = 1.5)
dev.off()

png("figures/plot_c.png", width = 800, height = 600, res = 150)
dotchart.phylog(
  phylog,
  rlqmix$lQ[names(phylog$leaves), 1],
  cleav = 0,
  cdot = 1,
  scaling = FALSE,
  yjoi = 0,
  cex.axis = 1.5,
  sub = "patterns in T and P",
  csub = 0
)
# mtext("c", side = 3, line = 0.5, at = par("usr")[1], font = 2, cex = 1.5)
dev.off()

img_a <- image_read("figures/plot_a.png")
img_b <- image_read("figures/plot_b.png")
img_c <- image_read("figures/plot_c.png")

combined <- image_append(c(img_b, img_c), stack = TRUE)
image_write(combined, "figures/combined_plots.png")
```

**2nd axis**
```{r extended_rlq_axis2, comment="", message=FALSE, warning=FALSE}
plot(rlqmix, xy = spa, ax = 2, wh = "S")
plot(rlqmix, phy = phylog, ax = 2, wh = "P")
plot(rlqmix, traits = genome_funct, ax = 2, type = "Q", wh = "T")
plot(rlqmix, env = pcaenv$tab, ax = 2, type = "Q", wh = "E")
```


### Subspecies Segregation and Spatial Structuring

#### Visual Gene based on Pangenome
```{r pangenome_lawsonibacte_Fig6_b, warning = FALSE, comments = "", message = FALSE}
# duplication 2025_07_mean_persistent_duplication.tsv
# gene category 2025_07_22_matrix.csv
# presence/absence 2025_07_22_gene_presence_absence.Rtab
genepa_pg <- read_tsv("data/data_preparation/2025_07_22_gene_presence_absence.Rtab", show_col_types = FALSE) %>%
  rename(gene = Gene)
duplication_pg <- read_tsv("data/data_preparation/2025_07_22_mean_persistent_duplication.tsv", show_col_types = FALSE) %>%
  rename(gene = persistent_family) %>% 
  select(gene, is_single_copy_marker)
partition_pg <- read_csv("data/data_preparation/2025_07_22_matrix.csv", show_col_types = FALSE) %>%
  rename(gene = Gene, `Family Partition`=`Non-unique Gene name`) %>% select(gene,`Family Partition`)



# Convert to numeric
wide_data <- left_join(genepa_pg, duplication_pg, by = "gene") %>% left_join(partition_pg, by = "gene") %>% 
  pivot_longer(!c(gene, is_single_copy_marker, `Family Partition`), names_to = "Genome",  values_to = "Presence/Absence") %>%
  mutate(`Presence/Absence` = as.character(`Presence/Absence`)) %>%
  mutate(`Presence/Absence` = case_when(
    `Presence/Absence` == "1" & is_single_copy_marker == "FALSE" ~ "2",
    TRUE ~ `Presence/Absence`
  )) %>%
  mutate(`Presence/Absence` = as.numeric(`Presence/Absence`)) %>% 
  select(gene, Genome, `Presence/Absence`) %>%
  pivot_wider(names_from = Genome, values_from = `Presence/Absence`, values_fill = 0)

# Matrix for heatmap
mat <- as.matrix(wide_data[,-1])
rownames(mat) <- wide_data$gene

# Annotation for rows (genes)
gene_anno <- left_join(genepa_pg, duplication_pg, by = "gene") %>% left_join(partition_pg, by = "gene") %>% 
  pivot_longer(!c(gene, is_single_copy_marker, `Family Partition`), names_to = "Genome",  values_to = "Presence/Absence") %>%
  distinct(gene, `Family Partition`) %>%
  arrange(`Family Partition`) %>% 
  column_to_rownames("gene")

# Reorder mat rows based on gene_anno rownames
mat <- mat[rownames(gene_anno), ]


# Plot
fig6_b <- pheatmap(mat,
         color = c("0" = "white", "1" = "#1b9e77", "2" = "#7570b3"),  # Named colors
         annotation_row = gene_anno,
         cluster_cols = TRUE,
         cluster_rows = FALSE,
         show_rownames = FALSE,
         legend_breaks = c(0, 1, 2),
         legend_labels = c("Absence", "Presence", "Multicopy"))

ggsave(filename = "figures/fig6_b.pdf", fig6_b, width = 8, height = 9)

pdf("figures/fig6_b.pdf", width = 8, height = 9)
pheatmap(mat,
         color = c("0" = "lightgrey", "1" = "#1b9e77", "2" = "#7570b3"),
         annotation_row = gene_anno,
         cluster_cols = TRUE,
         cluster_rows = FALSE,
         show_rownames = FALSE,
         legend_breaks = c(0, 1, 2),
         legend_labels = c("Absence", "Presence", "Multicopy"))
dev.off()

```

#### Visual Distribution Lawsonibacter subspecies
```{r plot_98ANI, warning = FALSE, comments = "", message = FALSE}
macromag_genomemetadata_98 %>% 
  filter(genus %in% c("Lawsonibacter"))

macroquant_lawsonibacter <- macroquant_counts_filt_30_98 %>%
  filter(genome %in% c(macromag_genomemetadata_98 %>%
    filter(genus == "Lawsonibacter") %>%
    pull(genome))) %>%
  pivot_longer(-genome, names_to = "microsample", values_to = "abundance") %>%
  left_join(macroquant_metrics_filt_30 %>%
    select(microsample, batch, cryosection, animal, Xcoord, Ycoord), by = "microsample") %>%
  filter(
    abundance > 0,
    !is.na(Xcoord),
    !is.na(Ycoord)
  ) %>%
  left_join(macromag_genomemetadata_98, by = "genome")

macroquant_lawsonibacter %>% 
  filter(abundance >0,
         cryosection %in% c("G121eI103A", "G103bI301A", "G121eI104C")) %>% 
  group_by(genome) %>% distinct(genome)

macroquant_lawsonibacter_bar <- macroquant_lawsonibacter %>%
  filter(cryosection %in%  c("G121eI104C")) %>% 
  arrange(microsample) %>%
  mutate(mag_species = paste0(genome, "_", species, "_", circul)) %>%
  ggplot(aes(x = abundance, y = microsample, fill = mag_species)) +
  geom_bar(position = "fill", stat = "identity", width = 0.8) +
  facet_wrap(~animal, scales = "free_y", ncol = 1) +
  scale_fill_viridis_d(option = "turbo", name = "MAG (98% ANI)") +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank())
# ggsave(filename = "figures/macroquant_lawsonibacter_bar.png", macroquant_lawsonibacter_bar, width = 10, height = 7)
macroquant_lawsonibacter_bar
```

```{r Fig6_a, warning = FALSE, comments = "", message = FALSE}
# Order and group data for cumulative pie angles
lawsonibacter_pie_df <- macroquant_lawsonibacter %>%
  filter(
    cryosection %in% c("G121eI103A", "G103bI301A", "G121eI104C"),
    !microsample %in% c("M300840", "M301068", "M301085", "M301084", "M301055", # exclude based on spatial coordinate OUT
      "M300833", "M300886") # exclude based on spatial coordinate SAME
    # circul %in% c("Y")
  ) %>% 
  group_by(microsample) %>%
  mutate(
    abundance = abundance / sum(abundance) # normalize counts to proportions
  ) %>%
  arrange(microsample, genome) %>%
  mutate(
    start = cumsum(lag(abundance, default = 0)) * 2 * pi,
    end = cumsum(abundance) * 2 * pi,
    r0 = 0,
    r = 75
  )

lawsonibacter_piecharts <- split(lawsonibacter_pie_df, lawsonibacter_pie_df$cryosection) %>%
  map(~ {
    ggplot(.x) +
      geom_arc_bar(
        aes(
          x0 = Xcoord, y0 = Ycoord, r0 = r0, r = r,
          start = start, end = end, fill = genome
        ),
        color = "black", linewidth = 0.1
      ) +
      coord_fixed() +
      ggtitle(unique(.x$cryosection)) +
      theme(legend.position = "none") +
      custom_ggplot_theme +
      scale_fill_manual(values = lawsonibacter_colors, drop = FALSE)
  })
fig6_a <- cowplot::plot_grid(plotlist = lawsonibacter_piecharts, ncol = 3)
fig6_a
# ggsave(filename = "figures/Fig6_a.pdf", fig6_a, width = 10, height = 5)
# ggsave(filename = "figures/Fig6_a.png", fig6_a, width = 10, height = 5)

# ggsave(filename = "figures/Fig6_a_legend.png",lawsonibacter_piecharts$G121eI103A + theme(legend.position = "right"), width = 10, height = 5)
```

#### Test Spatial Distribution Lawsonibacter subspecies
**Note:** The lawsonibacter_mantel_analysis() function needs to be loaded.
##### 35 Days Animal (G121e)
```{r lawsonibacter_mantel_G121e, warning = FALSE, comments = "", message = FALSE}
lawsonibacter_G121e <- lawsonibacter_mantel_analysis(
  data = macroquant_lawsonibacter %>% filter(cryosection %in% c("G121eI103A")),
  circul_selection = "N"
)
lawsonibacter_G121e
```

```{r lawsonibacter_mantel_G121e, warning = FALSE, comments = "", message = FALSE}
lawsonibacter_G121e <- lawsonibacter_mantel_analysis(
  data = macroquant_lawsonibacter %>% filter(cryosection %in% c("G121eI104C")),
  circul_selection = "N"
)
lawsonibacter_G121e
```


##### 14 Days Animal (G103b)

```{r lawsonibacter_mantel_G103b, warning = FALSE, comments = "", message = FALSE}
lawsonibacter_G103b <- lawsonibacter_mantel_analysis(
  data = macroquant_lawsonibacter %>% filter(cryosection %in% c("G103bI301A")),
  circul_selection = "N"
)
lawsonibacter_G103b
```


### SNP level microdiversity (Lorikeet)
#### Visual PopANI for GPB_bin_000044
```{r lawsonibacter_popANI, warning = FALSE, comments = "", message = FALSE}
lawsonibacter_popANI <- map_dfr(names(ani_list), function(bin_id) {
  ani_mat <- ani_list[[bin_id]]
  ani_mat %>%
    as.data.frame() %>%
    rownames_to_column("microsample_1") %>%
    pivot_longer(-microsample_1, names_to = "microsample_2", values_to = "popANI") %>%
    filter(microsample_1 != microsample_2) %>% # Exclude self-comparisons
    mutate(
      bin = bin_id,
      # Create unordered pair identifier
      sample_pair = pmap_chr(list(microsample_1, microsample_2), ~ paste(sort(c(..1, ..2)), collapse = "_"))
    ) %>%
    distinct(bin, sample_pair, .keep_all = TRUE) %>% # Keep only one of each unordered pair
    select(-sample_pair)
})

lawsonibacter_popANI_meta <- macroquant_metrics_filt_30 %>%
  filter(
    cryosection %in% c("G121eI103A", "G103bI301A","G121eI104C"),
    type_binomial %in% c("positive"),
    cov_filtering %in% c("Retained By Filtering"),
    !is.na(Xcoord),
    !is.na(Ycoord),
    !microsample %in% c(
      "M300840", "M301068", "M301085", "M301084", "M301055", # exclude based on spatial coordinate OUT
      "M300833", "M300886" # exclude based on spatial coordinate SAME
    ) 
  ) %>%
  select(cryosection, microsample, animal, Xcoord, Ycoord)

```

##### Distribution popANI by Group (Fig6_c and SupFig7_a)
```{r Fig6_c, warning = FALSE, comments = "", message = FALSE}
fig6_c <- lawsonibacter_popANI %>%
  filter(
    !is.nan(popANI),
    !is.na(popANI),
    microsample_1 %in% c(lawsonibacter_popANI_meta$microsample),
    microsample_2 %in% c(lawsonibacter_popANI_meta$microsample),
    bin %in% c("GPB_bin_000044")
  ) %>%
  left_join(macroquant_metrics_filt_30 %>% select(microsample, animal), by = c("microsample_1" = "microsample")) %>%
  rename(animal_microsample_1 = animal) %>%
  left_join(macroquant_metrics_filt_30 %>% select(microsample, animal), by = c("microsample_2" = "microsample")) %>%
  rename(animal_microsample_2 = animal) %>%
  mutate(group = ifelse(animal_microsample_1 == animal_microsample_2, animal_microsample_1, "G121e-G103b")) %>%
  ggplot(aes(x = group, y = popANI, fill = group)) +
  geom_boxplot(aes(group = group), outlier.shape = NA) +
  geom_jitter(
    position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.8),
    size = 0.3, alpha = 0.3
  ) +
  scale_fill_manual(values = c("#79b1a3", "#e35d51", "#ffcc62")) +
  custom_ggplot_theme +
  stat_compare_means(
    method = "wilcox.test",
    comparisons = list(
      c("G121e", "G103b"),
      c("G121e", "G121e-G103b"),
      c("G103b", "G121e-G103b")
    ),
    label = "p.signif",
    p.adjust.method = "BH",
    hide.ns = TRUE
  ) +
  labs(
    # title = "a",
    x = "",
    y = "popANI GPB_bin_000044"
  ) +
  theme(legend.position = "none") +
  scale_x_discrete(labels = c(
    "G103b" = "Within Animal \nG103b", "G121e" = "Within Animal \nG121e",
    "G121e-G103b" = "Between Animals"
  )) +
  # Hide last tick label by replacing it with blank
  scale_y_continuous(labels = function(x) {
    lbls <- as.character(x)
    lbls[length(lbls)] <- "" # blank out last label
    lbls
  })
# facet_wrap(~bin)
fig6_c
# ggsave(filename = "figures/Fig6_c.pdf", fig6_c, width = 10, height = 5)
# ggsave(filename = "figures/Fig6_c.png", fig6_c, width = 10, height = 5)
```

```{r supfig7_a, warning = FALSE, comments = "", message = FALSE}
supfig7_a <- lawsonibacter_popANI %>%
  filter(
    !is.nan(popANI),
    !is.na(popANI),
    microsample_1 %in% c(lawsonibacter_popANI_meta$microsample),
    microsample_2 %in% c(lawsonibacter_popANI_meta$microsample),
    bin %in% c("GPB_bin_000047","GPB_bin_000063")
  ) %>%
  left_join(macroquant_metrics_filt_30 %>% select(microsample, animal), by = c("microsample_1" = "microsample")) %>%
  rename(animal_microsample_1 = animal) %>%
  left_join(macroquant_metrics_filt_30 %>% select(microsample, animal), by = c("microsample_2" = "microsample")) %>%
  rename(animal_microsample_2 = animal) %>%
  mutate(group = ifelse(animal_microsample_1 == animal_microsample_2, animal_microsample_1, "G121e-G103b")) %>%
  ggplot(aes(x = group, y = popANI, fill = group)) +
  geom_boxplot(aes(group = group), outlier.shape = NA) +
  geom_jitter(
    position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.8),
    size = 0.3, alpha = 0.3
  ) +
  scale_fill_manual(values = c("#79b1a3", "#e35d51", "#ffcc62")) +
  custom_ggplot_theme +
  stat_compare_means(
    method = "wilcox.test",
    comparisons = list(
      c("G121e", "G103b"),
      c("G121e", "G121e-G103b"),
      c("G103b", "G121e-G103b")
    ),
    label = "p.signif",
    p.adjust.method = "BH",
    hide.ns = TRUE
  ) +
  labs(
    # title = "a",
    x = "",
    y = "popANI"
  ) +
  theme(legend.position = "none") +
  scale_x_discrete(labels = c(
    "G103b" = "Within Animal \nG103b", "G121e" = "Within Animal \nG121e",
    "G121e-G103b" = "Between Animals"
  )) +
  # Hide last tick label by replacing it with blank
  scale_y_continuous(labels = function(x) {
    lbls <- as.character(x)
    lbls[length(lbls)] <- "" # blank out last label
    lbls
  }) + facet_wrap(~bin)
supfig7_a 
# ggsave(filename = "figures/SupFig7_a.pdf", supfig7_a , width = 10, height = 5)
# ggsave(filename = "figures/SupFig7_a.png", supfig7_a , width = 10, height = 5)
```

##### Heatmap popANI GPB_bin_000044 (Fig6_b)
```{r Fig6_b, warning = FALSE, comments = "", message = FALSE}
# Remove pairs with NA for visualisation
popani_044 <- ani_list$GPB_bin_000044[rownames(ani_list$GPB_bin_000044) %in% lawsonibacter_popANI_meta$microsample, ] %>%
  select(lawsonibacter_popANI_meta$microsample)

sum(is.na(popani_044)) # Count of NA

metadata_044 <- macroquant_metrics_filt_30 %>%
  filter(microsample %in% colnames(popani_044)) %>%
  select(microsample, animal, cryosection) %>%
  column_to_rownames(var = "microsample") %>%
  rename(
    Cryosection = cryosection,
    Animal = animal
  )

annotation_colors <- list(
  Animal = c(
    "G121e" = "#e35d51",
    "G103b" = "#79b1a3"
  ), # light blue
  Cryosection = c(
    "G121eI103A" = "darkgrey",
    "G103bI301A" = "#ffcc62",
    "G121eI104C" = "blue"
  ) # soft orange
)

fig6_b <- pheatmap(
  popani_044,
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  clustering_method = "ward",
  display_numbers = FALSE,
  legend = TRUE,
  show_rownames = TRUE,
  show_colnames = TRUE,
  color = colorRampPalette(c("black", "#fbffe0"))(100),
  annotation_col = metadata_044,
  annotation_row = metadata_044,
  annotation_colors = annotation_colors,
  main = "Clustered ANI Heatmap"
)
fig6_b

# ggsave(filename = "figures/Fig6_b.pdf", fig6_b, width = 10, height = 5)
# ggsave(filename = "figures/Fig6_b.png", fig6_b, width = 10, height = 5)
```




#### Spatial Pattern Microdiversity
##### Stat
```{r mantel_cor_bins, warning = FALSE, comments = "", message = FALSE}
bins_id <- c("GPB_bin_000044","GPB_bin_000047","GPB_bin_000063")  

ids_35day <- lawsonibacter_popANI_meta %>%
  filter(animal == "G121e",
         cryosection %in% c("G121eI103A")) %>%
  pull(microsample)

results_list <- list()
for (bin in bins_id) {
  cat("Running analysis for", bin, "\n")
  res <- run_ani_spatial_analysis(bin, ani_list, ids_35day, lawsonibacter_popANI_meta)
  results_list[[bin]] <- res
}

results_list$GPB_bin_000044$mantel
results_list$GPB_bin_000047$mantel
results_list$GPB_bin_000063$mantel

222.67176-43.25015
402.09338-222.67176
222.67176+
  

  bins_id <- c("GPB_bin_000044","GPB_bin_000047","GPB_bin_000063")  

ids_35day <- lawsonibacter_popANI_meta %>%
  filter(animal == "G121e",
         cryosection %in% c("G121eI104C")) %>%
  pull(microsample)

results_list <- list()
for (bin in bins_id) {
  cat("Running analysis for", bin, "\n")
  res <- run_ani_spatial_analysis(bin, ani_list, ids_35day, lawsonibacter_popANI_meta)
  results_list[[bin]] <- res
}

results_list$GPB_bin_000044$mantel
results_list$GPB_bin_000047$mantel
results_list$GPB_bin_000063$mantel


results_list$GPB_bin_000063$mantel$mantel.res[ , 2]
```

##### Visual (SupFig7_b)
```{r mantel_bins, warning = FALSE, comments = "", message = FALSE}
mantel_GPB_bin_000044 <- ggplot(
  data.frame(results_list$GPB_bin_000044$mantel$mantel.res) %>%
    mutate(sign_label = ifelse(`Pr.corrected.` <= 0.05, "Significant", "Not significant")) %>%
    filter(!is.na(Mantel.cor)),
  aes(x = class.index, y = Mantel.cor)
) +
  geom_point(aes(fill = sign_label), shape = 23, size = 3) +
  geom_line() +
  scale_fill_manual(values = c("Significant" = "black", "Not significant" = "white")) +
  labs(
    title = "GPB_bin_000044",
    x = "Distance Class Index",
    y = "Mantel Correlation",
    color = "Significance"
  ) +
  geom_hline(yintercept = 0, color = "red") +
  custom_ggplot_theme +
  theme(legend.position = "none")

mantel_GPB_bin_000047 <- ggplot(
  data.frame(results_list$GPB_bin_000047$mantel$mantel.res) %>%
    mutate(sign_label = ifelse(`Pr.corrected.` <= 0.05, "Significant", "Not significant")) %>%
    filter(!is.na(Mantel.cor)),
  aes(x = class.index, y = Mantel.cor)
) +
  geom_point(aes(fill = sign_label), shape = 23, size = 3) +
  geom_line() +
  scale_fill_manual(values = c("Significant" = "black", "Not significant" = "white")) +
  labs(
    title = "GPB_bin_000047",
    x = "Distance Class Index",
    y = "Mantel Correlation",
    color = "Significance"
  ) +
  geom_hline(yintercept = 0, color = "red") +
  custom_ggplot_theme +
  theme(legend.position = "none")

mantel_GPB_bin_000063 <- ggplot(
  data.frame(results_list$GPB_bin_000063$mantel$mantel.res) %>%
    mutate(sign_label = ifelse(`Pr.corrected.` <= 0.05, "Significant", "Not significant")) %>%
    filter(!is.na(Mantel.cor)),
  aes(x = class.index, y = Mantel.cor)
) +
  geom_point(aes(fill = sign_label), shape = 23, size = 3) +
  geom_line() +
  scale_fill_manual(values = c("Significant" = "black", "Not significant" = "white")) +
  labs(
    title = "GPB_bin_000063",
    x = "Distance Class Index",
    y = "Mantel Correlation",
    color = "Significance"
  ) +
  geom_hline(yintercept = 0, color = "red") +
  custom_ggplot_theme +
  theme(legend.position = "none")


supfigs7_b <- cowplot::plot_grid(mantel_GPB_bin_000044, 
                            mantel_GPB_bin_000047, 
                            mantel_GPB_bin_000063,
  ncol = 2, label_size = 16
)
supfigs7_b 

supfigs7 <- cowplot::plot_grid(supfig7_a, supfigs7_b, 
  ncol = 1, labels = c("a", "b"), label_size = 16
)

ggsave(filename = "figures/supfig7.png", supfigs7, width = 10, height = 12)



supfigs7c <- cowplot::plot_grid(results_list$GPB_bin_000044$plot_cor, 
                            results_list$GPB_bin_000047$plot_cor, 
                            results_list$GPB_bin_000063$plot_cor,
  ncol = 1, label_size = 16
)
```

```{r test_distance, warning = FALSE, comments = "", message = FALSE}
# Subset coordinate
coords <- lawsonibacter_popANI_meta %>%
  filter(cryosection %in% c("G121eI103A")) %>%
  select(microsample, Xcoord, Ycoord)
# Compute pairwise distances
dist_matrix <- dist(coords %>% select(Xcoord, Ycoord))
# Convert to matrix and name the rows/cols with microsample
dist_matrix_named <- as.matrix(dist_matrix)
rownames(dist_matrix_named) <- coords$microsample
colnames(dist_matrix_named) <- coords$microsample
# Convert to long format and filter for distances > 1000 µm
dist_df <- as.data.frame(as.table(dist_matrix_named)) %>%
  rename(sample1 = Var1, sample2 = Var2, distance = Freq) %>%
  filter(sample1 != sample2) 

# Group sizes based on n.dist
group_sizes <- c(18, 44, 48, 48, 40, 60, 60, 86, 88, 112, 78)  # 
# Create group labels
group_labels <- rep(sprintf("br_%02d", seq_along(group_sizes)), times = group_sizes)

# Slice comparison based on total  n.dist
dist_df_unique <- dist_df %>%
  #rowwise() %>%
  #mutate(pair_id = paste(sort(c(sample1, sample2)), collapse = "_")) %>%
  #ungroup() %>%
  #distinct(pair_id, .keep_all = TRUE) %>%
  #select(-pair_id) %>%
  arrange(distance) %>% 
  slice(1:682)


# Apply to your data frame
dist_df_unique$distance_bin <- group_labels


#dist_df_unique <- dist_df_unique %>%
#  rowwise() %>%
#  mutate(pair_id = paste(sort(c(sample1, sample2)), collapse = "_")) %>%
#  ungroup() %>%
#  distinct(pair_id, .keep_all = TRUE) %>%
#  select(-pair_id) 

# View the result
distance_visual <- dist_df_unique %>%
  left_join(macroquant_metrics_filt_30 %>% select(microsample, Xcoord, Ycoord), by = c("sample1" = "microsample")) %>%
  rename(Xcoord_microsample_1 = Xcoord,
         Ycoord_microsample_1 = Ycoord) %>%
  left_join(macroquant_metrics_filt_30 %>% select(microsample, Xcoord, Ycoord), by = c("sample2" = "microsample")) %>%
  rename(Xcoord_microsample_2 = Xcoord,
         Ycoord_microsample_2 = Ycoord) %>% 
  #filter(distance_bin %in% c("br_01", "br_06", "br_11")) %>% 
  ggplot() +
  geom_segment(aes(
    x = Xcoord_microsample_1,
    y = Ycoord_microsample_1,
    xend = Xcoord_microsample_2,
    yend = Ycoord_microsample_2,
    color = distance_bin
  ),
  size = 1, alpha = 0.6
  ) +
  # Add dots for sample1
  geom_point(aes(x = Xcoord_microsample_1, y = Ycoord_microsample_1, color = distance_bin), size = 2) +
  # Add dots for sample2
  geom_point(aes(x = Xcoord_microsample_2, y = Ycoord_microsample_2, color = distance_bin), size = 2) +
  coord_fixed() +  # Preserves aspect ratio
  #scale_color_brewer(palette = "Set2") +  # Optional: nicer color palette
  custom_ggplot_theme +
  labs(
    x = "X coordinate",
    y = "Y coordinate",
    color = "Distance bin",
    title = "Microsample Pairs Colored by Distance Bin"
  ) + facet_wrap(~distance_bin, ncol=4)

ggsave(filename = "figures/distance_visual.png", distance_visual, width = 20, height = 10)
```

```{r test_distance, warning = FALSE, comments = "", message = FALSE}
# Subset coordinate
coords <- lawsonibacter_popANI_meta %>%
  filter(cryosection %in% c("G121eI104C")) %>%
  select(microsample, Xcoord, Ycoord)
# Compute pairwise distances
dist_matrix <- dist(coords %>% select(Xcoord, Ycoord))
# Convert to matrix and name the rows/cols with microsample
dist_matrix_named <- as.matrix(dist_matrix)
rownames(dist_matrix_named) <- coords$microsample
colnames(dist_matrix_named) <- coords$microsample
# Convert to long format and filter for distances > 1000 µm
dist_df <- as.data.frame(as.table(dist_matrix_named)) %>%
  rename(sample1 = Var1, sample2 = Var2, distance = Freq) %>%
  filter(sample1 != sample2) 

# Group sizes based on n.dist
group_sizes <- results_list$GPB_bin_000063$mantel$mantel.res[ , 2]  # 
# Create group labels
group_labels <- rep(sprintf("br_%02d", seq_along(group_sizes)), times = group_sizes)

# Slice comparison based on total  n.dist
dist_df_unique <- dist_df %>%
  #rowwise() %>%
  #mutate(pair_id = paste(sort(c(sample1, sample2)), collapse = "_")) %>%
  #ungroup() %>%
  #distinct(pair_id, .keep_all = TRUE) %>%
  #select(-pair_id) %>%
  arrange(distance) %>% 
  slice(1:length(group_labels))


# Apply to your data frame
dist_df_unique$distance_bin <- group_labels


#dist_df_unique <- dist_df_unique %>%
#  rowwise() %>%
#  mutate(pair_id = paste(sort(c(sample1, sample2)), collapse = "_")) %>%
#  ungroup() %>%
#  distinct(pair_id, .keep_all = TRUE) %>%
#  select(-pair_id) 

# View the result
distance_visual <- dist_df_unique %>%
  left_join(macroquant_metrics_filt_30 %>% select(microsample, Xcoord, Ycoord), by = c("sample1" = "microsample")) %>%
  rename(Xcoord_microsample_1 = Xcoord,
         Ycoord_microsample_1 = Ycoord) %>%
  left_join(macroquant_metrics_filt_30 %>% select(microsample, Xcoord, Ycoord), by = c("sample2" = "microsample")) %>%
  rename(Xcoord_microsample_2 = Xcoord,
         Ycoord_microsample_2 = Ycoord) %>% 
  #filter(distance_bin %in% c("br_01", "br_06", "br_11")) %>% 
  ggplot() +
  geom_segment(aes(
    x = Xcoord_microsample_1,
    y = Ycoord_microsample_1,
    xend = Xcoord_microsample_2,
    yend = Ycoord_microsample_2,
    color = distance_bin
  ),
  size = 1, alpha = 0.6
  ) +
  # Add dots for sample1
  geom_point(aes(x = Xcoord_microsample_1, y = Ycoord_microsample_1, color = distance_bin), size = 2) +
  # Add dots for sample2
  geom_point(aes(x = Xcoord_microsample_2, y = Ycoord_microsample_2, color = distance_bin), size = 2) +
  coord_fixed() +  # Preserves aspect ratio
  #scale_color_brewer(palette = "Set2") +  # Optional: nicer color palette
  custom_ggplot_theme +
  labs(
    x = "X coordinate",
    y = "Y coordinate",
    color = "Distance bin",
    title = "Microsample Pairs Colored by Distance Bin"
  ) + facet_wrap(~distance_bin, ncol=4)

distance_visual 
ggsave(filename = "figures/distance_visual.png", distance_visual, width = 20, height = 10)
```