---
title: "3D'omics | Micro-scale spatial metagenomics (MSSM)"
author:
- Carlotta Pietroni^[University of Copenhagen, carlotta.pietroni@sund.ku.dk]
- Antton Alberdi^[University of Copenhagen, antton.alberdi@sund.ku.dk]
- Amalia Bogri^[University of Copenhagen, amalia.bogri@sund.ku.dk]
date: "`r Sys.Date()`"
subtitle: Work in progress
site: bookdown::bookdown_site
documentclass: book
bibliography:
- book.bib
- packages.bib
url: "https://3d-omics.github.io/XXXX"
description: |
  Data analysis code for 3D'omics microdissection methods manuscript (MSSM)
link-citations: true
github-repo: "3d-omics/XXX"
---

```{r knitr_opts, echo=FALSE}
knitr::opts_chunk$set(
    class.source = "script-source",
    class.output = "script-output",
    comment = NA)
```

# Introduction

This webbook contains all the code used for the analyses of in "Micro-scale spatial metagenomics: revealing high-resolution spatial biogeography of gut microbiomes".

## Prepare the R environment

### Environment

To reproduce all the analyses locally, clone this repository in your computer using:

```
RStudio > New Project > Version Control > Git
```

And indicating the following git repository:

> https://github.com/3d-omics/XXX

Once the R project has been created, follow the instructions and code chunks shown in this webbook.

### Libraries

The following R packages are required for the data analysis.

```{r load_libraries, warning=FALSE, comments="", message=FALSE}
# install.packages('BiocManager', dependencies = TRUE)
# BiocManager::install(c("phyloseq", "ggtreeExtra","ggtree","ANCOMBC", "ALDEx2"))'
# install.packages('remotes', dependencies = TRUE)
# remotes::install_github('anttonalberdi/hilldiv2')
# remotes::install_github('anttonalberdi/distillR')'
# remotes::install_github('wilkelab/cowplot')

if(!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, rairtable, jsonlite, distillR, ape,
               hilldiv2, ggpubr, broom, reshape2, zCompositions, ANCOMBC,
               cowplot, ALDEx2, car, DHARMa, sjPlot, ggeffects, microViz,
               ggh4x, vegan, ggfortify, ggrepel, microbiome, ggforce, patchwork,
               grid, pheatmap, janitor, spdep, adespatial, adegraphics, # lmPerm
               ade4, magick, permuco, airtabler2, rmarkdown, purrr)
select <- dplyr::select
recode <- dplyr::recode
```
### Notes:
#### Installation of modules
Most of the modules can be downloaded from CRAN with 'install.packages('XXX', dependencies = TRUE)'.
Four modules are from Bioconductor. 
Install with: 'install.packages('BiocManager', dependencies = TRUE)
BiocManager::install(c("phyloseq", "ggtreeExtra","ggtree","ANCOMBC", "ALDEx2"))'
Three modules are github. 
Install with: 'install.packages('remotes', dependencies = TRUE)
remotes::install_github('anttonalberdi/hilldiv2')
remotes::install_github('anttonalberdi/distillR')'
remotes::install_github('wilkelab/cowplot')'

#### Airtable access:
Currently the metadata is loaded from the 3D'omics airtable. In order to access it you need your own API key (from Antton). set_airtable_api_key('XXXXX', install = TRUE)

## Define Plotting Settings
### Color Schemes
```{r get_colors, warning=FALSE, comments="", message=FALSE, eval=FALSE}
phylum_colors <- c(Actinomycetota = "#346254",
                   Bacillota = "#4a6ab7",
                   Bacillota_A = "#8c1c47",
                   Bacteroidota = "#9c8464",
                   Pseudomonadota = "#c49d4b",
                   Verrucomicrobiota = "#462410")

order_colors <- c(Actinomycetales	= "#79b1a3",
                  Coriobacteriales =	"#498a77",
                  CAJFEE01 =	"#a0c5e8",
                  Erysipelotrichales =	"#7acef4",
                  Lactobacillales =	"#6390fb",
                  ML615J_28 =	"#70c9de",
                  Paenibacillales =	"#70a8dc",
                  RF39 =	"#5c7fba",
                  Christensenellales =	"#f58262",
                  Clostridiales =	"#cf4a82",
                  Lachnospirales =	"#e15f7d",
                  Monoglobales = "#e18299",
                  Oscillospirales =	"#ba2760",
                  Peptostreptococcales =	"#e35d51",
                  TANB77 =	"#ef3d26",
                  UBA1381 =	"#be5643",
                  Bacteroidales =	"#e3c7a0",
                  Enterobacterales =	"#ffcc62",
                  Verrucomicrobiales = "#823f1d")

lawsonibacter_colors <- c(`GPB:bin_000217`	= "#79b1a3",
                  `GPB:bin_000137` =	"darkgrey",
                  `GPB:bin_000118` =	"#7acef4", 
                  `GPB:bin_000080` =	"#6390fb", 
                  `GPB:bin_000077` =	"#5c7fba",
                  `GPB:bin_000072` =	"lightgreen", 
                  `GPB:bin_000015` =	"#cf4a82", 
                  `GPB:bin_000047` =	"#e35d51", 
                  `GPB:bin_000044` =	"black",
                  `GPB:bin_000034` =	"#e3c7a0", 
                  `GPB:bin_000063` =	"#ffcc62", 
                  `D300470:bin_000020` = "darkseagreen3") 

```

### Plotting Theme
```{r custom_ggplot_theme, warning=FALSE, comments="", message=FALSE}
custom_ggplot_theme <- theme(
  strip.text.y.left = element_text(angle = 0),
  strip.text.y.right = element_text(angle = 0),
  axis.text = element_text(size = 10),
  axis.title = element_text(size = 12, face = "bold"),
  strip.background = element_rect(fill = "#dde3e9", color = "white", size = 0.8), # Custom facet strip background
  strip.text = element_text(size = 11, face = "bold", color = "black"), # Custom facet text 8
  strip.placement = "outside", # Place strip outside the panel grid
  panel.spacing = unit(0.1, "lines"), # Adjust space between panels
  panel.grid.major = element_line(color = "#dde3e9"), # Customize major grid lines
  panel.grid.minor = element_blank(), # Remove minor grid lines
  panel.background = element_rect(fill = "white"), # Change panel background color
  plot.margin = unit(c(1, 1, 1, 1), "cm") # Adjust plot margins to ensure content fits
)
```

## Load Functions
### parse_fastp_function
```{r parse_fastp_function, warning=FALSE, message=FALSE, comments=""}
parse_fastp <- function(multiqc_json) {
  fastp_raw <- multiqc_json$report_saved_raw_data$multiqc_fastp

  fastp_tibble <-
    tibble::tibble(
      sample_id = names(fastp_raw),
      values = fastp_raw
    ) |>
    dplyr::mutate(
      sample_id = sample_id |>
        stringr::str_remove_all("fastp \\| ") |>
        stringr::str_remove("_[12].fq.gz") |>
        stringr::str_split("_") |>
        purrr::map(1) |>
        unlist()
    ) |>
    tidyr::unnest_wider(values) |>
    tidyr::unnest_wider(-sample_id, names_sep = ".") |>
    tidyr::unnest_wider(
      c(
        summary.before_filtering,
        summary.after_filtering,
        polyx_trimming.polyx_trimmed_reads,
        polyx_trimming.polyx_trimmed_bases
      ),
      names_sep = "."
    )
  fastp_tibble
}
```

### calculate_alpha_diversity
```{r alpha_diversity_function, warning=FALSE, comments="", message=FALSE}
calculate_alpha_diversity <- function(input_data, metadata_name, dataset_name, tree_name) {
  input_data_matrix <- input_data %>% column_to_rownames(var = "genome") 
  colsum <- input_data_matrix  %>%
    summarise(across(where(is.numeric), sum)) %>%
    pivot_longer(cols = everything(), names_to = "microsample", values_to = "cov_filtering") %>% 
    mutate(cov_filtering=ifelse(cov_filtering > 0 , "Retained By Filtering", "Excluded By Filtering"))
  # Define diversity metrics and calculate
  diversity_metrics <- list(
    richness = hilldiv(input_data_matrix, q = 0),
    neutral = hilldiv(input_data_matrix, q = 1),
    phylogenetic = hilldiv(input_data_matrix, q = 1, tree = tree_name)
  )
  # Process metrics into a single data frame
  alpha_diversity <- lapply(names(diversity_metrics), function(metric) {
    diversity_metrics[[metric]] %>%
      t() %>%
      as.data.frame() %>%
      rownames_to_column(var = "microsample") %>%
      rename(!!sym(metric) := 2) # rename metric column
  }) %>%
    reduce(full_join, by = "microsample") %>% 
    right_join(metadata_name, by = "microsample") %>% # Merge with final stats
    left_join(colsum, by = "microsample")
  #output_filename <- paste0("results/alpha_div_", dataset_name, ".tsv")
  #write_tsv(alpha_diversity, output_filename)
  return(alpha_diversity)
}
```

### Amalia: perform_pca
```{r pca_function, warning=FALSE, comments="", message=FALSE, eval=FALSE}
perform_pca <- function(df, zero_method = "GBM", z_delete = TRUE) {
  
  # Store original dimensions
  original_rows <- nrow(df)
  original_cols <- ncol(df)
  
  # 1. Zero replacement
  if (any(df == 0)) { # I think cmultRepl already does that
    print("Zeros found")
    df <- cmultRepl(df, method = zero_method, output = "prop",
                        z.warning = 0.8, z.delete = z_delete)
    df <- df * 100
  }

  # Print removed rows and columns
  removed_rows <- original_rows - nrow(df)
  removed_cols <- original_cols - ncol(df)
  cat("Rows (samples) removed:", removed_rows, "\n")
  cat("Columns (taxa) removed:", removed_cols, "\n")
  
  #plot_abundance_heatmap(df)

  # Geometric mean function
  geometric_mean <- function(x) {
    # Use log to avoid underflow
    exp(mean(log(x), na.rm = TRUE))  
    }
    
  # 2. Calculate geometric mean of the parts (taxa) of the data set.
  taxa_geometric_means <- apply(df, 2, geometric_mean)
  
  # 3. Center data
  df_centered <- sweep(df, 2, taxa_geometric_means, FUN = "/")
  
  df_centered <- as.matrix(df_centered)  
  
  # Compute the Variation Matrix
  variation_matrix <- outer(
    1:ncol(df_centered), 1:ncol(df_centered), 
    Vectorize(function(i, j) var(log(df_centered[, i] / df_centered[, j]), na.rm = TRUE))
  )
 
   # Calculate Total Variance
  D <- ncol(df_centered)  # Number of taxa (columns)
  totvar <- (1 / (2 * D)) * sum(variation_matrix, na.rm = TRUE)
  
  # 4. Scale data
  power_exponent <- 1 / sqrt(totvar)
  df_scaled <- df_centered^power_exponent
  # print(df_scaled)

  # CLR transform data
  clr_transform <- function(x) {
    log(x) - mean(log(x), na.rm = TRUE)
  }
  df_clr <- t(apply(df_scaled, 1, clr_transform)) 
  df_clr <- as.data.frame(df_clr)
  
  df_clr_dist <- t(apply(df, 1, clr_transform)) 
  df_clr_dist <- as.data.frame(df_clr_dist)
  
  # Perform PCA on zero replaced, centered, scaled, and CLR transformed df
  pca_result <- prcomp(df_clr, center = FALSE, scale. = FALSE)
  pca_result_dist <- prcomp(df_clr_dist, center = TRUE, scale. = TRUE)
  
  return(list(
    df_clr = df_clr,
    df_clr_dist = df_clr_dist,
    pca_result=pca_result,
    pca_result_dist=pca_result_dist))
}
```

### fit_and_analyze_model
```{r fit_and_analyze_model(), message=FALSE, eval=FALSE}
fit_and_analyze_model <- function(model = c("lm", "glm"),
                                  distribution = NULL,
                                  response_var,
                                  explanatory_var,
                                  data) {
  model <- match.arg(model) # Restrict to "lm" or "glm"
  # Construct the model formula
  model_formula <- as.formula(paste(response_var, "~", explanatory_var))

  # Initialize all possible return objects
  anova_result <- NULL
  md <- NULL
  simResids <- NULL

  if (model == "lm") {
    # Continuous floats (any real number)
    md <- lm(model_formula, data = data)
    anova_result <- broom::tidy(Anova(md, test.statistic = "F"))
  } else if (model == "glm") {
    # Counts (integers ≥0): poisson, quasipoisson (accounts overdispersion)
    # Proportions (0–1 continuous):	quasibinomial GLM (accounts overdispersion) but beta regression (not GLM)	Preferred over quasibinomial GLM
    # Continuous floats (any real number): gaussian, Gamma (continuous positive data, skewed to the right)
    if (is.null(distribution)) {
      stop("You must specify a distribution family for glm.")
    }

    md <- glm(model_formula, family = distribution, data = data)

    if (!grepl("^quasi", distribution)) { # Use DHARMa only on supported distributions
      simResids <- simulateResiduals(md)
      anova_result <- broom::tidy(Anova(md, test.statistic = "F")) # ANOVA with F test
    } else {
      anova_result <- broom::tidy(Anova(md, test.statistic = "Wald")) # ANOVA with Chi-squared test
    }
  }
  result <- list(
    model_fit = md,
    anova = anova_result
  )

  if (!is.null(simResids)) {
    result$simResidual <- simResids
  }

  return(result)
}
```

### pivot_phylo
```{r pivot_phylo, message=FALSE, eval=FALSE}
pivot_phylo <- function(phyloseq_obj, glom = TRUE, tax_transform = TRUE, taxon_level, tr_method) {
  if (glom == TRUE && !is.null(taxon_level)) {
    phyloseq_obj <- prune_taxa(taxa_sums(phyloseq_obj) > 0, phyloseq_obj)
    phyloseq_obj <- tax_glom(phyloseq_obj, taxon_level)
  } else {
    .
  }
  if (tax_transform == TRUE && !is.null(tr_method)) {
    phyloseq_obj <- tax_transform(phyloseq_obj, tr_method)
  } else {
    .
  }

  pivot_dataframe <- data.frame(otu_table(phyloseq_obj)) %>%
    rownames_to_column(var = "genome") %>%
    pivot_longer(-genome, names_to = "microsample", values_to = "abundance") %>%
    filter(abundance > 0) %>%
    left_join(data.frame(tax_table(phyloseq_obj)) %>%
      rownames_to_column(var = "genome"), by = "genome") %>%
    left_join(data.frame(sample_data(phyloseq_obj)) %>%
      rownames_to_column(var = "microsample"), by = "microsample")

  # Re-order levels
  taxa_levels <- c("domain", "phylum", "class", "order", "family", "genus", "species")
  # Iterate over taxonomic levels
  for (taxa in taxa_levels) {
    # Check if the column has more than one unique value
    if (taxa %in% colnames(pivot_dataframe) && length(unique(pivot_dataframe[[taxa]])) > 1) {
      # Convert each taxonomic level to a factor with levels ordered by abundance
      pivot_dataframe <- pivot_dataframe %>%
        mutate(
          !!taxa := factor(
            !!sym(taxa),
            levels = pivot_dataframe %>%
              group_by(!!sym(taxa)) %>%
              summarise(total_abundance = sum(abundance, na.rm = TRUE), .groups = "drop") %>%
              arrange(desc(total_abundance)) %>%
              pull(!!sym(taxa))
          ) # Extract ordered levels
        )
    }
  }
  return(pivot_dataframe)
}
```

### spatial_cryosections
```{r spatial_cryosections_fun, message=FALSE, eval=FALSE}
spatial_cryosections <- function(cryosection_list, metadata_df, comm_clr) {
  cryosection_dfs <- list()
  mantel_results <- list()
  mantelcor_results <- list()
  decay_dfs <- list()
  distance_decay_plots <- list()
  structure_results <- list()

  for (cryosection in cryosection_list) {
    # Filter metadata for this section
    metadata_data <- metadata_df %>%
      filter(cryosection == !!cryosection, !is.na(.data$Xcoord), !is.na(.data$Ycoord))

    # Filter community data
    comm_data <- comm_clr %>%
      data.frame() %>%
      rownames_to_column(var = "microsample") %>%
      filter(microsample %in% metadata_data$microsample) %>%
      column_to_rownames(var = "microsample")

    cryosection_dfs[[cryosection]] <- list(
      comm_clr = comm_data,
      metadata = metadata_data
    )

    # Mantel correlogram
    mantel <- vegan::mantel(
      dist(comm_data),
      dist(metadata_data[, c("Xcoord", "Ycoord")]),
      permutations = 999
    )
    mantel_results[[cryosection]] <- mantel

    # Mantel correlogram
    correlog <- vegan::mantel.correlog(
      D.eco = dist(comm_data),
      D.geo = dist(metadata_data[, c("Xcoord", "Ycoord")]),
      nperm = 999
    )
    mantelcor_results[[cryosection]] <- correlog

    # Distance decay
    toplot <- data.frame(
      spat_dist = as.numeric(dist(metadata_data[, c("Xcoord", "Ycoord")])),
      comm_dist = as.numeric(dist(comm_data))
    )
    decay_dfs[[cryosection]] <- toplot

    # Plot
    p <- ggplot(toplot, aes(x = spat_dist, y = comm_dist)) +
      # geom_point() +
      geom_smooth() +
      xlab("Spatial distance (μm)") +
      ylab("Aitchison \ndistance") +
      custom_ggplot_theme +
      # ggtitle(paste("Distance Decay -", cryosection))
      ggtitle(paste(cryosection))
    distance_decay_plots[[cryosection]] <- p

    # Print summary
    # print(paste("Summary for", cryosection))
    # print(summary(lm(comm_dist ~ spat_dist, data = toplot)))

    #### Complex spatial structures
    # set.seed(111)
    # spat_SWNs<-listw.candidates(data.frame(metadata_data[,c("Xcoord","Ycoord")]),
    # nb = c("gab"),
    # weights = c("fdown", "fup"),
    # y_fdown = c(2,5), y_fup = c(0.1,0.5))
    # W_sel <- listw.select(comm_data, spat_SWNs, MEM.autocor = "positive",
    # p.adjust = T, method="FWD")
    # structure_results[[cryosection]] <- W_sel
  }
  return(list(
    cryosection_dfs = cryosection_dfs,
    mantel_results = mantel_results,
    mantelcor_results = mantelcor_results,
    decay_dfs = decay_dfs,
    distance_decay_plots = distance_decay_plots,
    structure_results = structure_results
  ))
}
```

### lawsonibacter_mantel_analysis 
```{r lawsonibacter_mantel_analysis, eval=FALSE}
lawsonibacter_mantel_analysis <- function(data, animal_id, circul_selection) {

  # 1. Filter and prepare presence-absence matrix
  if (circul_selection == "Y") {
    filtered_data <- data %>%
      filter(
        cryosection %in% c("G121eI103A", "G103bI301A"),
        !microsample %in% c("M300840", "M301068", "M301085", "M301084"),
        circul == circul_selection,
        animal == animal_id
      )
  } else {
    message("circul_selection is not 'Y'; skipping circul filter.")
    filtered_data <- data %>%
      filter(
        cryosection %in% c("G121eI103A", "G103bI301A"),
        !microsample %in% c("M300840", "M301068", "M301085", "M301084"),
        animal == animal_id
      )
  }

  # Presence Absence
  # 1. Build presence-absence matrix
  lawsonibacter_pa <- filtered_data %>%
    mutate(presence = ifelse(abundance > 0, 1, 0)) %>%
    select(microsample, genome, presence) %>%
    pivot_wider(names_from = genome, values_from = presence, values_fill = 0) %>%
    column_to_rownames("microsample")
  # 2. Community distance
  comm_dist_pa <- vegan::vegdist(lawsonibacter_pa, method = "jaccard")
  # 3. Spatial coordinates
  coords_pa <- filtered_data %>%
    distinct(microsample, Xcoord, Ycoord) %>%
    arrange(microsample)
  coords_pa <- coords_pa[match(rownames(lawsonibacter_pa), coords_pa$microsample), ]
  # 4. Spatial distance
  spatial_dist_pa <- dist(coords_pa[, c("Xcoord", "Ycoord")])
  # 5. Mantel test
  mantel_result_pa <- vegan::mantel(comm_dist_pa, spatial_dist_pa, permutations = 999)
  # 6. Mantel correlogram
  mantelcor_result_pa <- vegan::mantel.correlog(
    D.eco = comm_dist_pa,
    D.geo = spatial_dist_pa,
    nperm = 999
  )
  
  # CLR
  # 1. Build count matrix and CLR transformation
  lawsonibacter_clr <- filtered_data %>%
    select(microsample, genome, abundance) %>%
    pivot_wider(names_from = genome, values_from = abundance, values_fill = 0) %>%
    column_to_rownames("microsample")
  lawsonibacter_clr <- cmultRepl(lawsonibacter_clr, method = "GBM", output = "prop", z.warning = 0.95)
  clr_transform <- function(x) {
    log(x) - mean(log(x), na.rm = TRUE)
    }
  lawsonibacter_clr <- data.frame(t(apply(lawsonibacter_clr, 1, clr_transform)))
  # 2. Community distance
  comm_dist_clr <- vegan::vegdist(lawsonibacter_clr, method = "euclidean")
  # 2. Spatial coordinates
  coords_clr <- filtered_data %>%
    distinct(microsample, Xcoord, Ycoord) %>%
    arrange(microsample) %>% 
    filter(microsample %in% c(rownames(lawsonibacter_clr)))
  coords_clr <- coords_clr[match(rownames(lawsonibacter_clr), coords_clr$microsample), ]
  # 4. Spatial distance
  spatial_dist_clr <- dist(coords_clr[, c("Xcoord", "Ycoord")])
  # 5. Mantel test
  mantel_result_clr <- vegan::mantel(comm_dist_clr, spatial_dist_clr, permutations = 999)
  # 6. Mantel correlogram
  mantelcor_result_clr <- vegan::mantel.correlog(
    D.eco = comm_dist_clr,
    D.geo = spatial_dist_clr,
    nperm = 999
    )
  
  distance_clr_df <- data.frame(
      spat_dist = as.numeric(spatial_dist_clr),
      comm_dist = as.numeric(comm_dist_clr)
    )
  
  clr_lm <- aovperm(lmperm(comm_dist ~ spat_dist, data = distance_clr_df, np=10000))
  
                    return(list(
    animal = animal_id,
    mantel_pa = mantel_result_pa,
    correlogram_pa = mantelcor_result_pa,
    mantel_clr = mantel_result_clr,
    correlogram_clr = mantelcor_result_clr,
    clr_lm = clr_lm
  ))
}
```

<!--chapter:end:index.Rmd-->

# Loading and Tidying Data
**Note:** This step requires access to the raw data files located at xxxx. If you prefer to skip this step, you can proceed to the next section, making sure to load the R object that contains the necessary files for the subsequent analysis in location.

## Import Sample metadata
```{r sample_metadata_mssm, message=FALSE, eval=FALSE}
mssm_sample_metadata <- airtable("4-MSE-Info", "appKakM1bnKSekwuW") %>%  get base ID from Airtable browser URL
  read_airtable(., fields = c(
    "ID", "LabBatch_text", "IntestinalSection", "SampleType",
    "Xcoord", "Ycoord", "SizeApprox", "cryosection_text", "buffer_text", "Collection_method_text",
    "Collection_Success", "Collection_attempts", "UsedCycles", "animal_temp", "Protocol_text",
    "adapter_type_text"
  ), id_to_col = TRUE) %>%
  filter(
    LabBatch_text %in% c(
       "MSEB0001", "MSEB0002", "MSEB0003", "MSEB0004", "MSEB0005",
      "MSEB0006", "MSEB0009", "MSEB0010", "MSEB0011", "MSEB0012",
      "MSEB0014", "MSEB0015"
    ),
    !SampleType %in% c("ControlHuman")
  ) %>%
  rename(
    batch = LabBatch_text, microsample = ID, section = IntestinalSection, 
    type = SampleType, cryosection = cryosection_text, 
    buffer = buffer_text, collection_method = Collection_method_text,
    collection = Collection_Success, collection_attempts = Collection_attempts,
    cycles = UsedCycles, animal = animal_temp, size = SizeApprox, 
    protocol = Protocol_text, adapter = adapter_type_text
  ) %>%
  select(
    microsample, section, type, batch, cryosection, buffer, Xcoord, Ycoord, 
    size, collection_method, collection, collection_attempts, cycles, animal, 
    protocol, adapter
  ) %>%
  mutate(
    type_binomial = ifelse(type == "Positive", "positive", "negative"),
    buffer = case_when(
      buffer == "B01" ~ "L01",
      buffer == "B02" ~ "L02",
      buffer == "B03" ~ "L03",
      buffer == "B04" ~ "L04",
      buffer == "B09" ~ "L05",
      buffer == "B11" ~ "L06",
      buffer == "B12" ~ "L07",
      buffer == "B13" ~ "L08",
      TRUE ~ buffer
    )
  ) %>%
  unnest(c(section, Xcoord, Ycoord, size, collection, cycles)) %>%
  arrange(microsample)
#write.table(mssm_sample_metadata, "data/mssm_sample_metadata.tsv")
#mssm_sample_metadata <- read.table("data/mssm_sample_metadata.tsv")
```

## Sequencing Data Quality 

### Read Quality: Pre- vs Post-Filtering (fastp)
**Note:** The parse_fastp() function in index needs to be loaded.
```{r read_quality, warning=FALSE, message=FALSE, comments=""}
mssm_json_data <- fromJSON("data/multiqc_data.json") # Download file from ERDA
mssm_fastp <- parse_fastp(mssm_json_data) %>%
  rename_with(~ gsub("^(summary|filtering_result|adapter_cutting|polyx_trimming)\\.", "", .)) %>%
  select(
    sample_id,
    before_filtering.total_reads,
    after_filtering.total_reads,
    before_filtering.total_bases,
    after_filtering.total_bases,
    before_filtering.gc_content,
    after_filtering.gc_content,
    low_quality_reads,
    adapter_trimmed_reads,
    adapter_trimmed_bases,
    duplication.rate
  ) %>%
  mutate(microsample = str_remove(sample_id, "\\.lib\\d+$")) %>%
  filter(grepl("^M", microsample)) %>%
  group_by(microsample) %>%
  summarise(
    across(contains("reads"), sum, .names = "total_{.col}"),
    across(contains("bases"), sum, .names = "total_{.col}"),
    across(ends_with("content"), mean, .names = "percent_{.col}"),
    across(ends_with("rate"), mean, .names = "percent_{.col}")
  )
```


### Contamination Assessment: Host, Human, and Environmental Reads
```{r mapping_eukaryots_references, warning=FALSE, comments="", message=FALSE}
mssm_euk_mapping <- read_tsv("https://sid.erda.dk/share_redirect/aTWbNfAf1o/results/preprocess_bowtie2_multiqc_report_data/multiqc_samtools_stats.txt") %>%
  mutate(
    reference = case_when(
      grepl("GRCh38", Sample, ignore.case = TRUE) ~ "human",
      grepl("GRCg7b", Sample, ignore.case = TRUE) ~ "chicken",
      grepl("Sscrofa11.1", Sample, ignore.case = TRUE) ~ "pig",
      TRUE ~ NA_character_
    ),
    microsample = str_extract(Sample, "M\\d+")
  ) %>%
  filter(
    !is.na(reference), # Remove rows where reference is NA
    grepl("^M", microsample)
  ) %>%
  select(Sample, microsample, reference, raw_total_sequences, reads_mapped, reads_unmapped) %>%
  group_by(microsample, reference) %>%
  summarise(
    raw_total_sequences = sum(raw_total_sequences, na.rm = TRUE),
    reads_mapped = sum(reads_mapped, na.rm = TRUE),
    reads_unmapped = sum(reads_unmapped, na.rm = TRUE)
  ) %>%
  pivot_wider(names_from = reference, values_from = c(raw_total_sequences, reads_mapped, reads_unmapped))
```

### Genome-Centric Mapping Statistics: Macroscale MAG Catalogue
```{r mapping_macro_catalogue, warning=FALSE, comments="", message=FALSE}
# download.file("https://sid.erda.dk/share_redirect/aTWbNfAf1o/results/quantify.zip", "data/quantify.zip", mode = "wb")
mssm_mag_mapping <- read.table(unz("data/quantify.zip", "multiqc_samtools_stats.txt"), header = T, sep = "\t", quote = "\"") %>%
  mutate(microsample = str_extract(Sample, "M\\d+")) %>%
  filter(grepl("^M", microsample)) %>%
  select(Sample, microsample, raw_total_sequences, reads_mapped, reads_unmapped) %>%
  group_by(microsample) %>%
  summarise(
    raw_total_sequences_mag = sum(raw_total_sequences, na.rm = TRUE),
    reads_mapped_mag = sum(reads_mapped, na.rm = TRUE),
    reads_unmapped_mag = sum(reads_unmapped, na.rm = TRUE)
  )
```

### Read-Centric Mapping Statistics: Kraken (genomes) and SingleM (single-copy marker genes)
```{r mapping_singlem, warning=FALSE, comments="", message=FALSE}
mssm_singlem <- read_tsv("https://sid.erda.dk/share_redirect/aTWbNfAf1o/results/preprocess/singlem/microbial_fraction.tsv") %>%
  mutate(microsample = str_extract(sample, "M\\d+")) %>%
  filter(grepl("^M", microsample)) %>%
  select(-c(warning, sample)) %>%
  group_by(microsample) %>%
  summarise(
    read_fraction_singlem = (mean(read_fraction, na.rm = TRUE)) / 100 #,
    # bacterial_archaeal_bases_singlem = sum(bacterial_archaeal_bases, na.rm = TRUE),
    # average_bacterial_archaeal_genome_size_singlem = mean(average_bacterial_archaeal_genome_size, na.rm = TRUE)
  )
```

```{r mapping_kraken, warning=FALSE, comments="", message=FALSE}
mssm_kraken <- read_tsv("https://sid.erda.dk/share_redirect/aTWbNfAf1o/results/preprocess_kraken2_multiqc_report_data/multiqc_kraken.txt") %>%
  mutate(microsample = str_extract(Sample, "M\\d+")) %>%
  filter(grepl("^M", microsample)) %>%
  select(Sample, microsample, D) %>%
  separate(D, into = c("Eukaryota", "Bacteria", "Archaea", "Viruses"), sep = ", ", remove = TRUE) %>%
  mutate(across(!all_of("microsample"), ~ parse_number(.))) %>%
  group_by(microsample) %>%
  summarise(
    kr_Eukaryota = sum(Eukaryota, na.rm = TRUE),
    kr_Bacteria = sum(Bacteria, na.rm = TRUE),
    kr_Archaea = sum(Archaea, na.rm = TRUE),
    kr_Viruses = sum(Viruses, na.rm = TRUE),
  ) %>%
  mutate(
    total_kraken = kr_Eukaryota + kr_Bacteria + kr_Archaea + kr_Viruses,
    bacteria_pct_kraken = kr_Bacteria / total_kraken,
    eukaryota_pct_kraken = kr_Eukaryota / total_kraken
  )
```

### Redundancy & Coverage Estimation: Nonpareil
```{r estimates_nonpareil, warning=FALSE, comments="", message=FALSE}
mssm_nonpareil <- read_tsv("https://sid.erda.dk/share_redirect/aTWbNfAf1o/results/preprocess_nonpareil_multiqc_report_data/nonpareil-table.txt") %>%
  mutate(microsample = str_extract(Sample, "M\\d+")) %>%
  filter(grepl("^M", microsample)) %>%
  select(microsample, Redundancy, Coverage, Diversity) %>%
  group_by(microsample) %>%
  summarise(
    redundancy_nonpareil = (mean(Redundancy, na.rm = TRUE)) / 100,
    coverage_nonpareil = (mean(Coverage, na.rm = TRUE)) / 100,
    diversity_nonpareil = mean(Diversity, na.rm = TRUE)
  )
```

### Combined Quality Statistics
```{r data_quality_df, warning=FALSE, comments="", message=FALSE, eval=exists("mssm_sample_metadata")}
ls()
mssm_quality <- mssm_sample_metadata %>%
  full_join(mssm_fastp, by = "microsample") %>%
  full_join(mssm_euk_mapping, by = "microsample") %>%
  full_join(mssm_mag_mapping, by = "microsample") %>%
  full_join(mssm_singlem, by = "microsample") %>%
  full_join(mssm_kraken, by = "microsample") %>%
  full_join(mssm_nonpareil, by = "microsample") %>%
  mutate(
    lowquality_reads_pct = ((total_before_filtering.total_reads - total_after_filtering.total_reads) / total_before_filtering.total_reads),
    adapter_contamination_pct = (total_adapter_trimmed_bases / total_before_filtering.total_bases),
    host_pct = (reads_mapped_chicken / total_before_filtering.total_reads),
    human_pct = (reads_mapped_human / total_before_filtering.total_reads),
    pig_pct = (reads_mapped_pig / total_before_filtering.total_reads),
    prokreads_pct = (raw_total_sequences_mag / total_before_filtering.total_reads),
    mapmag_rawreads_pct = (reads_mapped_mag / total_before_filtering.total_reads),
    mapmag_filteredreads_pct = (reads_mapped_mag / total_after_filtering.total_reads),
    mapmag_prokreads_pct = (reads_mapped_mag / raw_total_sequences_mag)
  ) %>%
  filter(microsample %in% c(mssm_sample_metadata$microsample))
```

## Input Preparation For Metagenomic Analysis

### Abundance Estimation Using Macro-scale MAGs Catalogue 95 ANI

#### Import Genome Counts

```{r mag_readcount_macroquant, message=FALSE, eval=FALSE, warning=FALSE, comments=""}
macroquant_read_counts <- read_tsv("https://sid.erda.dk/share_redirect/aTWbNfAf1o/results/quantify/coverm/genome.count.REF0015-g_mg_hybrid-0.95.tsv.gz", show_col_types = FALSE) %>%
  rename(genome = 1) %>%
  pivot_longer(!genome, names_to = "data", values_to = "counts") %>%
  mutate(sample = substr(data, 1, 7)) %>%
  group_by(genome, sample) %>%
  summarise(counts = sum(counts), .groups = "drop") %>%
  pivot_wider(names_from = "sample", values_from = "counts") 
```

```{r mag_coveredbases_macroquant, message=FALSE, eval=FALSE}
macroquant_covered_bases <- read_tsv("https://sid.erda.dk/share_redirect/aTWbNfAf1o/results/quantify/coverm/genome.covered_bases.REF0015-g_mg_hybrid-0.95.tsv.gz", show_col_types = FALSE) %>%
  rename(genome = 1) %>%
  pivot_longer(!genome, names_to = "data", values_to = "counts") %>%
  mutate(sample = substr(data, 1, 7)) %>%
  group_by(genome, sample) %>%
  summarise(counts = sum(counts), .groups = "drop") %>%
  pivot_wider(names_from = "sample", values_from = "counts")
```

#### Macro-scale MAGs Catalogue
##### Taxonomy
This is the raw taxonomy table generated by GTDBtk, which is simplified for downstream analyses.
```{r mag_taxonomy_macro, message=FALSE, eval=FALSE}
macromag_taxonomy <- read_tsv("https://sid.erda.dk/share_redirect/BTwCSjaWz2/REF0015-g_mg_hybrid/gtdbtk.summary.tsv.gz", show_col_types = FALSE) %>%
  rename(genome = user_genome) %>%
  mutate(genome = str_replace_all(genome, "\\.fa", "")) %>%
  separate(classification, c("domain", "phylum", "class", "order", "family", "genus", "species"), sep = ";") %>%
  select(genome, domain, phylum, class, order, family, genus, species) %>%
  mutate(
    domain = sub("^d__", "", domain), # Remove prefix
    phylum = sub("^p__", "", phylum),
    class = sub("^c__", "", class),
    order = sub("^o__", "", order),
    family = sub("^f__", "", family),
    genus = sub("^g__", "", genus),
    species = sub("^s__", "", species)
  ) %>%
  arrange(match(genome, macroquant_read_counts$genome)) %>%
  filter(genome %in% macroquant_read_counts$genome) %>%
  mutate(
    species = if_else(
      is.na(species) | species == "",
      paste0("unclassified_", genus),
      species
    )
  ) %>%
  mutate(
    species = if_else(
      is.na(species) | species == "unclassified_",
      paste0("unclassified_", family),
      species
    )
  ) %>%
  mutate(
    genus = if_else(
      is.na(genus) | genus == "",
      paste0("unclassified_", family),
      genus
    )
  )
```

##### Quality
```{r mag_quality_macro, message=FALSE, eval=FALSE}
macromag_quality <- read_tsv("https://sid.erda.dk/share_redirect/BTwCSjaWz2/REF0015-g_mg_hybrid/checkm2.quality_report.tsv.gz", show_col_types = FALSE) %>%
  # rename first column
  rename(genome = 1) %>%
  # remove .fa from the end of each genome name
  mutate(genome = str_replace_all(genome, "\\.fa", "")) %>%
  # select relevant columns and rename
  select(genome, Completeness, Contamination, Coding_Density, Genome_Size, Contig_N50) %>%
  rename(
    completeness = Completeness,
    contamination = Contamination,
    coding_density = Coding_Density,
    length = Genome_Size,
    N50_length = Contig_N50
  ) %>%
  # estimate genome circularity (N50 should be equal to genome length)
  mutate(circularity = N50_length / length) %>%
  mutate(circularity = ifelse(circularity >= 1, "1", "0")) %>%
  # update completeness and contamination for genomes with circularity of 1
  mutate(
    completeness = ifelse(circularity == "1", 100, completeness),
    contamination = ifelse(circularity == "1", 0, contamination)
  ) %>%
  # estimate ratio of: genome length / median bacterial genome length
  mutate(length_ratio = length / median(length, na.rm = TRUE))

macromag_genomemetadata <- macromag_taxonomy %>%
  left_join(macromag_quality, by = join_by(genome == genome)) %>% # join quality
  arrange(match(genome, macroquant_read_counts$genome)) %>%
  filter(genome %in% macroquant_read_counts$genome)

```

##### Functional Annotation

```{r mag_annotation_macro, message=FALSE, eval=FALSE}
# macromag_annotations <- read_tsv("https://sid.erda.dk/share_redirect/BTwCSjaWz2/REF0015-g_mg_hybrid/dram.annotations.tsv.gz", show_col_types = FALSE) %>%
macromag_annotations <- read_tsv("data/annotations.tsv", show_col_types = FALSE) %>%
  rename(gene = 1, genome = 2) %>%
  filter(genome %in% macroquant_read_counts$genome)
macromag_gifts <- distill(macromag_annotations, GIFT_db, genomecol = 2, annotcol = c(9, 10, 19))
```

##### Phylogenetic Tree

```{r mag_tree_macro, message=FALSE, warning=FALSE, eval=FALSE}
macromag_tree <- read.tree("https://sid.erda.dk/share_redirect/BTwCSjaWz2/REF0015-g_mg_hybrid/gtdbtk.backbone.bac120.classify.tree")
macromag_tree$tip.label <- str_replace_all(macromag_tree$tip.label, "'", "") # remove single quotes in MAG names
macromag_tree <- keep.tip(macromag_tree, tip = macroquant_read_counts$genome) # keep only MAG tips
```

#### Coverage Filtering & Genome Count Normalization

##### Coverage filtering
**Note:** Genomes that have less than 30% of their length covered by reads are turned into zeros to account for the random allocation of reads across genomes due to mapping heuristics. 
```{r coverage_fitering_macroquant, eval=FALSE}
macroquant_coverage <- macroquant_covered_bases %>% # calculate coverage values by dividing the number of covered bases by the length of each genome
  mutate(across(where(is.numeric), ~ . / macromag_genomemetadata$length))

macroquant_read_counts_filtered <- list()
macroquant_summary_filtered <- list()
# Define thresholds
coverage_thresholds <- c(0.4, 0.35, 0.3, 0.25, 0.2, 0.15, 0.1, 0.05)
# Loop over thresholds
for (threshold in coverage_thresholds) {
  df <- macroquant_coverage %>%
    mutate(across(where(is.numeric), ~ ifelse(. > threshold, 1, 0))) %>%
    mutate(across(-1, ~ . * macroquant_read_counts[[cur_column()]]))
  summary <- df %>%
    summarise(across(where(is.numeric), \(x) sum(x, na.rm = TRUE))) %>%
    pivot_longer(
      cols = where(is.numeric), # columns to pivot (all numeric here)
      names_to = "microsample", # new column for former column names
      values_to = paste0("read_count_", threshold * 100) # new column for values
    )
  name <- paste0("filt_", threshold * 100)
  macroquant_read_counts_filtered[[name]] <- df
  macroquant_summary_filtered[[name]] <- summary
}
macroquant_merged_summary <- reduce(macroquant_summary_filtered, left_join, by = "microsample")

min_coverage <- 0.3
macroquant_read_counts_filt_30 <- macroquant_coverage %>%
  mutate(across(where(is.numeric), ~ ifelse(. > min_coverage, 1, 0))) %>%
  mutate(across(-1, ~ . * macroquant_read_counts[[cur_column()]]))

all(macroquant_read_counts_filtered$filt_30 == macroquant_read_counts_filt_30)

```

##### Generate genome count table
After filtering the low-coverage reads, read counts are transformed into genome counts using genome-length and read-length information.
```{r genome_counts_macroquant, eval=FALSE}
readlength <- 150 # change if sequencing read length is different
macroquant_genome_counts_filt_30 <- macroquant_read_counts_filt_30 %>%
  mutate(across(where(is.numeric), ~ . / (macromag_genomemetadata$length / readlength))) 
macroquant_genome_counts  <- macroquant_read_counts %>%
  mutate(across(where(is.numeric), ~ . / (macromag_genomemetadata$length / readlength))) 
```

#### Alpha Diversity
**Note:** The calculate_alpha_diversity() function needs to be loaded.
```{r alpha_diversities_macroquant, warning=TRUE, comments="", message=FALSE}
# Calculate alpha diversity for filtered genome counts data
macroquant_metrics_filt_30 <- calculate_alpha_diversity(
  input_data = macroquant_genome_counts_filt_30,
  metadata_name = mssm_quality,
  dataset_name = "filt_30",
  tree_name = macromag_tree
) %>% filter(!microsample %in% c("Undeter"))
``` 



<!--chapter:end:01-data_preparation.Rmd-->

