---
title: "3D'omics | Micro-scale spatial metagenomics (MSSM)"
author:
- Carlotta Pietroni^[University of Copenhagen, carlotta.pietroni@sund.ku.dk]
- Jorge Langa^[University of the Basque Country, jorgeeliseo.langa@ehu.eus]
- Iñaki Odriozola Larrañaga^[University of Copenhagen, inaki.larranaga@sund.ku.dk]
- Amalia Bogri^[University of Copenhagen, amalia.bogri@sund.ku.dk]
- Antton Alberdi^[University of Copenhagen, antton.alberdi@sund.ku.dk]
date: "`r Sys.Date()`"
subtitle: Insert manuscript link
site: bookdown::bookdown_site
documentclass: book
bibliography:
- book.bib
- packages.bib
url: "https://github.com/3d-omics/MSSM.git"
description: |
  Data analysis code for 3D'omics microdissection methods manuscript (MSSM)
link-citations: true
github-repo: "3d-omics/MSSM"
---

```{r knitr_opts, echo=FALSE}
knitr::opts_chunk$set(
  class.source = "script-source",
  class.output = "script-output",
  comment = NA
)
```

# Introduction

This webbook provides all the code used in the analyses for the study "Micro-scale spatial metagenomics: revealing high-resolution spatial biogeography of gut microbiomes". The project 3D'omics (http://www.3domics.eu) was Funded by European Union’s Horizon 2020 Research and Innovation programme. 

## Prepare the R environment

### Environment

To reproduce all the analyses locally, clone this repository in your computer using:

```
RStudio > New Project > Version Control > Git
```

And indicating the following git repository:

> https://github.com/3d-omics/MSSM.git

Once the R project has been created, follow the instructions and code chunks shown in this webbook.

### Libraries
The following R packages are required for the data analysis. Most modules can be installed directly from CRAN or GitHub.
However, four modules are provided through Bioconductor and should be installed if needed with the command below before loading:
```{r load_libraries, warning = FALSE, comments = "", message = FALSE}
# install.packages('BiocManager', dependencies = TRUE)
# BiocManager::install(c("phyloseq", "ggtreeExtra","ggtree", "ALDEx2"))'

if (!require("pacman")) install.packages("pacman")
# If the package is not installed, installs from CRAN
pacman::p_load(
  tidyverse, rairtable, jsonlite, ape,
  ggpubr, broom, reshape2, zCompositions, ANCOMBC,
  ALDEx2, car, DHARMa, sjPlot, ggeffects,
  ggh4x, vegan, ggfortify, ggrepel, microbiome, ggforce, patchwork,
  grid, pheatmap, janitor, spdep, adespatial, adegraphics,
  ade4, magick, permuco, airtabler2, rmarkdown, purrr, PerformanceAnalytics
)
# If the package is not installed, installs from GitHub
pacman::p_load_gh(
  "anttonalberdi/hilldiv2",
  "anttonalberdi/distillR",
  "wilkelab/cowplot",
  "david-barnett/microViz"
)
select <- dplyr::select
recode <- dplyr::recode
```

## Define Plotting Settings
### Color Schemes
```{r phylum_colors, echo = FALSE, warning = FALSE, message = FALSE}
phylum_colors <- c(
  Actinomycetota = "#346254",
  Bacillota = "#4a6ab7",
  Bacillota_A = "#8c1c47",
  Bacteroidota = "#9c8464",
  Pseudomonadota = "#c49d4b",
  Verrucomicrobiota = "#462410"
)

barplot(
  1:length(phylum_colors),
  col = phylum_colors,
  main = "Phylum Colors",
  yaxt = "n", 
  names.arg = names(phylum_colors), 
  las = 2
)
```

```{r order_colors, echo = FALSE, warning = FALSE, message = FALSE}
order_colors <- c(
  Actinomycetales = "#79b1a3",
  Coriobacteriales = "#498a77",
  CAJFEE01 = "#a0c5e8",
  Erysipelotrichales = "#7acef4",
  Lactobacillales = "#6390fb",
  ML615J_28 = "#70c9de",
  Paenibacillales = "#70a8dc",
  RF39 = "#5c7fba",
  Christensenellales = "#f58262",
  Clostridiales = "#cf4a82",
  Lachnospirales = "#e15f7d",
  Monoglobales = "#e18299",
  Oscillospirales = "#ba2760",
  Peptostreptococcales = "#e35d51",
  TANB77 = "#ef3d26",
  UBA1381 = "#be5643",
  Bacteroidales = "#e3c7a0",
  Enterobacterales = "#ffcc62",
  Verrucomicrobiales = "#823f1d"
)

barplot(
  1:length(order_colors),
  col = order_colors,
  main = "Order Colors",
  yaxt = "n",
  names.arg = names(order_colors),
  las = 2
)
palette_color <- order_colors
```

```{r lawsonibacter_colors, echo = FALSE, warning = FALSE, message = FALSE}
lawsonibacter_colors <- c(
  `GPB:bin_000217` = "#79b1a3",
  `GPB:bin_000137` = "darkgrey",
  `GPB:bin_000118` = "#7acef4",
  `GPB:bin_000080` = "#6390fb",
  `GPB:bin_000077` = "#5c7fba",
  `GPB:bin_000072` = "lightgreen",
  `GPB:bin_000015` = "#cf4a82",
  `GPB:bin_000047` = "#e35d51",
  `GPB:bin_000044` = "black",
  `GPB:bin_000034` = "#e3c7a0",
  `GPB:bin_000063` = "#ffcc62",
  `D300470:bin_000020` = "darkseagreen3"
)

barplot(
  1:length(lawsonibacter_colors),
  col = lawsonibacter_colors,
  main = "Lawsonibacter MAGs colors",
  yaxt = "n",
  names.arg = names(lawsonibacter_colors),
  las = 2
)
```
```{r supplementary_color, echo = FALSE, warning = FALSE, message = FALSE}
supplementary_color <- c("#2A9D8F","#264653","#E76F51", "#E9C46A")
barplot(
  1:length(supplementary_color),
  col = supplementary_color,
  main = "Palette colors",
  yaxt = "n",
  names.arg = names(supplementary_color),
  las = 2
)
```

### Plotting Theme
```{r custom_ggplot_theme, echo = FALSE, warning = FALSE, message = FALSE}
custom_ggplot_theme <- theme(
  strip.text.y.left = element_text(angle = 0),
  strip.text.y.right = element_text(angle = 0),
  axis.text = element_text(size = 10),
  axis.title = element_text(size = 12, face = "bold"),
  strip.background = element_rect(fill = "#dde3e9", color = "white", size = 0.8), # Custom facet strip background
  strip.text = element_text(size = 11, face = "bold", color = "black"), # Custom facet text 8
  strip.placement = "outside", # Place strip outside the panel grid
  panel.spacing = unit(0.1, "lines"), # Adjust space between panels
  panel.grid.major = element_line(color = "#dde3e9"), # Customize major grid lines
  panel.grid.minor = element_blank(), # Remove minor grid lines
  panel.background = element_rect(fill = "white"), # Change panel background color
  plot.margin = unit(c(1, 1, 1, 1), "cm") # Adjust plot margins to ensure content fits
)
ggplot(data.frame(x = 0, y = 0), aes(x, y)) +
  geom_blank() +
  annotate("text", x = 0, y = 0, label = "MSSM", size = 8, fontface = "bold") +
  custom_ggplot_theme
```

## Load Functions
The functions defined below are used throughout other sections of this analysis and must be loaded before running those parts.
```{r functions, warning = FALSE, comments = "", message = FALSE}
source("resources/scripts/MSSM_functions_pre.R")
source("resources/scripts/JEC_1743_sm_apps5.R")
```

